### 2016-09-28 Paper_v5.tex
First commit

\documentclass[11pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

% Use adjustwidth environment to exceed column width (see example table in text)
\usepackage{changepage}

% Use Unicode characters when possible
\usepackage[utf8]{inputenc}

% textcomp package and marvosym package for additional characters
\usepackage{textcomp,marvosym}

% fixltx2e package for \textsubscript
\usepackage{fixltx2e}

% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath,amssymb}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref,hyperref}

% line numbers
\usepackage[right]{lineno}

% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% rotating package for sideways tables
\usepackage{rotating}

% Remove comment for double spacing
%\usepackage{setspace}
%\doublespacing

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in
\textheight 8.75in

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}

% Use the PLoS provided BiBTeX style
\bibliographystyle{plos2009}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Leave date blank
\date{}

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
\lhead{\includegraphics[natwidth=1.3in,natheight=0.4in]{PLOSlogo.png}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\sf PLOS}

%% Include all macros below

\newcommand{\lorem}{{\bf LOREM}}
\newcommand{\ipsum}{{\bf IPSUM}}

%% END MACROS SECTION

\begin{document}
\vspace*{0.35in}
\begin{flushleft}
{\Large
\textbf\newline{Population heterogeneity in the epithelial to mesenchymal transition is controlled by NFAT and phosphorylated Sp1}}
\newline
% Insert Author names, affiliations and corresponding author email.
\\
Russell Gould\textsuperscript{1},
Anirikh Chakrabarti\textsuperscript{2},
Jeffrey D. Varner\textsuperscript{2},
Jonathan Butcher\textsuperscript{1,*},
\\
\bf{1} Department of Biomedical Engineering, Cornell University, Ithaca, NY, USA
\\
\bf{2} School of Chemical and Biomolecular Engineering, Cornell University, Ithaca, NY, USA

% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
%
% Remove or comment out the author notes below if they aren't used.
%
% Primary Equal Contribution Note
%\Yinyang These authors contributed equally to this work.

% Additional Equal Contribution Note
%\ddag These authors also contributed equally to this work.

% Current address notes
%\textcurrency a Insert current address of first author with an address update
% \textcurrency b Insert current address of second author with an address update
% \textcurrency c Insert current address of third author with an address update

% Deceased author note
%\dag Deceased

% Group/Consortium Author Note
%\textpilcrow Insert Collaborative Author line here

* E-mail: jtb47@cornell.edu
\end{flushleft}

%\linenumbers

%{\noindent Keywords:~ | Morphogenesis | Plasticity | Cancer | Systems Biology| Tubulogenesis}

\section*{Abstract}
Epithelial to mesenchymal transition (EMT) is an essential differentiation program during tissue morphogenesis and remodeling.
EMT is induced by soluble transforming growth factor $\beta$ (TGF-$\beta$) family members, and restricted by vascular endothelial growth factor family members.
While many downstream molecular regulators of EMT have been identified, these have been largely evaluated individually without considering potential crosstalk.
In this study, we created an ensemble of dynamic mathematical models describing TGF-$\beta$ induced EMT to better understand the operational hierarchy of this complex molecular program.
These models incorporate mass action kinetics within an ordinary differential equation (ODE) framework to describe the transcriptional and post-translational regulatory events driving EMT.
Model parameters were estimated from multiple data sets using multiobjective optimization, in combination with cross-validation.
TGF-$\beta$ exposure drove the model population toward a mesenchymal phenotype, while an epithelial phenotype was maintained following vascular endothelial growth factor A (VEGF-A) exposure.
Simulations predicted that the transcription factors phosphorylated SP1 and NFAT were master regulators promoting or inhibiting EMT, respectively.
Surprisingly, simulations also predicted that a cellular population could exhibit phenotypic heterogeneity (characterized by a significant fraction of the population with both high epithelial and mesenchymal marker expression)
if treated simultaneously with TGF-$\beta$ and VEGF-A.  We tested this prediction experimentally in both MCF10A and DLD1 cells and found that upwards of 45\% of the cellular population acquired this hybrid state in the presence of both TGF-$\beta$ and VEGF-A.
We experimentally validated the predicted NFAT/Sp1 signaling axis for each phenotype response. Lastly, we found that cells in the hybrid state had significantly different functional behavior when compared to VEGF-A or TGF-$\beta$ treatment alone.
Together, these results establish a predictive mechanistic model of EMT susceptibility, and potentially reveal a novel signaling axis which regulates carcinoma progression through an EMT versus tubulogenesis response.

\section*{Author Summary}
Tissue formation and remodeling requires a complex and dynamic balance of interactions between epithelial cells, which reside on the surface, and mesenchymal cells that reside in the tissue interior.
During embryonic development, wound healing, and cancer, epithelial cells transform into a mesenchymal cell to form new types of tissues.
It is important to understand this process so that it can be controlled to generate beneficial effects and limit pathological differentiation.
Much research over the past 20 years has identified many different molecular species that are relevant, but these have mainly been studied one at a time.
In this study, we developed and implemented a novel computational strategy to interrogate all of the known players in this transformation process to identify which are the major bottlenecks.
We determined that NFATc1 and pSP1 are essential for promoting epithelial or mesenchymal differentiation, respectively.
We then predicted the existence of a partially transformed cell that exhibits both epithelial and mesenchymal characteristics.
We found this partial cell type develops a network of invasive but stunted vascular structures that may be a unique cell target for understanding cancer progression and angiogenesis.

%as seen by the initiating the formation of endocardial cushions in the atrioventricular valves and mediating palate fusion of medial epithelial cells \cite{Zavadil:2005fu}.  In this context, EMT terminates and has been suggested as a highly controlled process \cite{Lee:2006tw}.  EMT is also important under pathological conditions, prominently in fibrosis and in invasion and metastasis of carcinomas \cite{Thiery:2003il}. This culminates decades of work within the EMT field appears shared in developmental and pathological conditions \cite{Xu:2009oq}.
%In this study, we determined the extent of EMT in normal epithelial cells exposed to pathological conditions by (1) implementing and subsequently interrogating a population based model containing canonical EMT regulatory networks, and (2) verify our mechanistic targets using established cell lines with differing induction capacities for EMT. Toward this objective, we modeled the core molecular interactions of EMT using mass-action kinetics within an ordinary differential equation (ODE) framework.  A population approach was used to estimate a family or ensemble of EMT models that was consistent with the previous literature \cite{Medici:2006qa,Medici:2008fk}.  Using the newly predicted operational paradigms, we sought to capture the role of phospho-Sp1 and NFAT on phenotype heterogeneity during TGF$\beta$ induced EMT \emph{in vitro} via key marker expression profiles and fate processes.
%Infrastructure components include TGF-$\beta$2 ligand binding, TGF-$\beta$RI,II,III transmembrane growth factor receptor phosphorylation, activation of the Ras/MAPK cascade, cytosolic release of $\beta$-catenin from E-cadherin, autocrine response of the TGF-$\beta$3 ligand, phosphorylation of the Smad proteins, and cross-talk between the MAPK/Smad signaling pathways.
%Furthermore, because these reactions occur with varying initial conditions and unknown parameter values, significant  differences may exist in the overall response of the model.
%To unravel the operation of this complex architecture, we have used mathematical modeling.
%In this program, extracellular TGF-$\beta$1/2 induces a series of  that ultimately repress epithelial gene expression
%while simultaneously activating mesenchymal gene expression.
%However, this program is not a simple switch. It involves
%Thus, the molecular architecture responsible for EMT induction is highly complex, involving multiple inputs and overlapping transcriptional networks.
%However, including these additional mechanisms increases model complexity.
%Analysis of the model predicted that the NFATc and phosphorylated Sp1 transcription factors were likely the key elements driving this heterogeneity.

\linenumbers

\section*{Introduction}
The epithelial to mesenchymal transition (EMT) is a broadly participating, evolutionarily conserved differentiation program essential for tissue morphogenesis, remodeling and pathological processes such as cancer \cite{Thiery:2003il}. During EMT polarized, tightly adhered epithelial cell monolayers are transformed into non-interacting motile mesenchymal cells that simultaneously degrade and synthesize extracellular matrix (ECM) components and invade into the underlying tissue space \cite{Stahl:2001pd}. EMT is the fundamental initiator of developmental processes such as embryonic gastrulation and valvulogenesis \cite{Eisenberg:1995ai} (also Kalluri J Clin Invest 2009, Thiery Cell 2009). Transforming growth factor $\beta$ (TGF-$\beta$) family members are important inducers of both developmental and pathological EMT \cite{Zavadil:2005fu,Xu:2009oq}. Decades of research has focused on identifying molecular regulators of EMT, but almost all on a single gene and in a nearly binary yes/no level of qualitative understanding. Medici and coworkers recently identified a core signaling program by which TGF-$\beta$ isoforms induce EMT across a variety of cell lines \cite{Medici:2006qa,Medici:2008fk}.
This program involves carefully orchestrated rounds of gene expression driven by the Smad and Snail families of transcription factors as well as
other key factors such as lymphoid enhancer-binding factor 1 (LEF-1), nuclear factor of activated T-cells, cytoplasmic 1 (NFATc1), and specificity protein 1 (Sp1).
Coregulators such as $\beta$-catenin, NF-$\kappa$B, and the ErbB family of receptor tyrosine kinases however also participate in EMT regulation, but the degree of each's influence is difficult to ascertain in isolation \cite{Kim:2002lh,Jiang:2007aa,Huber:2004aa,Hardy:2010aa}.
EMT also exhibits complex temporal dynamics that are often intractable in gain/loss of function studies.
Elucidating the master regulatory architecture controlling EMT therefore requires inclusion of these complex overlapping and non-binary behaviors.

Systems biology and mathematical modeling are essential tools for understanding complex developmental programs like EMT \cite{Ahmed:2007mi}.
Previous computational models of TGF-$\beta$ induced differentiation focused on single biological factors or EMT in single cells.
For example, Chung \emph{et al.}, constructed a model of TGF-$\beta$ receptor activation and Smad signaling using ordinary differential equations and mass-action kinetics.
Their model suggested that a reduction of functional TGF-$\beta$ receptors in cancer cells may lead to an attenuated Smad2 signal \cite{Chung:2009jl}.
Similarly, Vilar \emph{et al.} suggested that specific changes in receptor trafficking patterns could lead to phenotypes that favor tumorigenesis \cite{Vilar:2006gb}.
Although these models provided insight into the role of receptor dynamics, EMT induction involves many other components, including competing second messengers and interconnected transcriptional regulatory loops.
Integrating these additional scales of molecular signaling while maintaining the capacity for robust prediction requires a new and expanded computational and experimental strategy.
Data-driven systems approaches \cite{Cirit:2012kx} or logical model formulations \cite{Morris:2011ys} are emerging paradigms that constrain model complexity through the incorporation of training and validation data.
These are interesting techniques because the data informs model structure (which can be expanded as more data becomes available).
Alternatively, Bailey proposed more than a decade ago that a qualitative understanding of a complex biological system should not require complete definition of its structural and parametric content \cite{2001_bailey_NatBiotech}.
Shortly thereafter, Sethna and coworkers showed that complex model behavior is often controlled by only a few parameter combinations, a characteristic seemingly universal to multi-parameter models referred to as ``sloppiness'' \cite{Machta:2013by}.
Thus, reasonable model predictions are often possible with only limited parameter information.
Taking advantage of this property, we developed sloppy techniques for parameter identification using ensembles of deterministic models \cite{Song:2010fk}.
Furthermore, we proposed that the sloppy behavior of biological networks may also be seen as a source of cell-to-cell \cite{Lequieu:2011fj} or even patient-to-patient heterogeneity \cite{2010_luan_varner_MolBioSys}.
Recently, Bayesian parameter identification techniques have also been used to explore cell-to-cell heterogeneity  \cite{2011_kalita_brasier_JBC,2011_hasenauer_allgower_BMCBioinfo},
where a population of cells could be viewed as a dynamic ensemble of context-specific biochemical networks \cite{2012_creixell_linding_NatBiotech}.

In this study, we developed a family of mechanistic models describing the induction of EMT by TGF-$\beta$ isoforms in the presence and absence of vascular endothelial growth factor A (VEGF-A).
We incorporated mass action kinetics within an ordinary differential equation (ODE) framework to describe the EMT interaction network containing 995 gene, protein or mRNA components interconnected through 1700 interactions.
A family of model parameters was estimated using 41 molecular data sets generated in DLD1 colon carcinoma, MDCKII and A375 melanoma cells using the Pareto optimal ensemble technique (POETs) multiobjective optimization algorithm.
POETs identified more than 15,000 likely TGF-$\beta$ induced EMT models, from which we selected approximately 1100 models for further analysis.
Analysis of the model population suggested that both MCF10A and DLD1 cells could exhibit phenotypic heterogeneity if treated simultaneously with TGF-$\beta$1/2 and VEGF-A.
This heterogeneity was characterized by a significant fraction of the population being in a “hybrid state” having both high E-cadherin and high Vimentin expression.
We tested these predictions using qRT-PCR and flow-cytometry studies in a variety of experimental conditions.
Validation studies confirmed that upwards of 45\% of the cellular population could be put into the hybrid state in the presence of both TGF-$\beta$1/2 and VEGF-A.
Moreover, this response depended upon both activation of Sp1 by MAPK and NFATc1 transcriptional activity consistent with the predicted molecular signaling.
Lastly, the hybrid populations of both DLD1 and MCF10A cells exhibited different functional behavior than those from either TGF-$\beta$ or VEGF-A treatment.
The extent of ductal branch formation significantly increased with MCF10A cells in the hybrid phenotype, compared with cells treated with VEGF-A alone.
Together, these results establish a predictive mechanistic model of EMT susceptibility, and reveal a novel signaling axis, which possibly regulates carcinoma progression through an EMT versus tubulogenesis response.

\clearpage

\section*{Results}

\subsection*{The model population captured key features of TGF-$\beta$ induced EMT}
The EMT model architecture, based upon curated molecular connectivity,
described the expression of 80 genes following exposure to TGF-$\beta$ isoforms and VEGF-A (Fig. \ref{fg:F1}).
The EMT model contained 995 molecular species interconnected by 1700 interactions.
Model equations were formulated using mass-action kinetics within an ordinary differential equation (ODE) framework.
ODEs and mass action kinetics are common tools to model biochemical pathways \cite{Schoeberl:2002rv,Chen:2009uq,Tasseff:2011ve}.
However, while ODE models can simulate complex intracellular behavior,
they require estimates for model parameters which are often difficult to obtain.
The EMT model had 1756 unknown model parameters, 1700 kinetic constants and 56 non-zero initial conditions.
As expected, these parameters were not uniquely identifiable given the training data \cite{Gadkar:2005vn}.
Thus, instead of identifying a single best fit (but uncertain) model,
we estimated a sloppy population of models (each consistent with the training data) by simultaneously minimizing the difference between model simulations and 41 molecular
data sets using the Pareto Optimal Ensemble Technique (POETs).
The training data were generated in DLD1 colon carcinoma, MDCKII, and A375 melanoma cells following
exposure to TGF-$\beta$ isoforms \cite{Medici:2008fk}. We organized these data sets into 11 objective functions which were simultaneously minimized by POETs.
Additionally, we used 12 molecular data sets generated in HK-2 cells following VEGF-A exposure to train VEGF-A responsive model processes \cite{Lian:2011bs}.
To guard against overfitting, we augmented the multiobjective optimization with leave-one-out cross validation to independently estimate both the training and
prediction error for each objective. Thus, we generated 11 different model ensembles.
Lastly, we compared model predictions with independent data sets not used during training (both at the molecular and model population
levels) to evaluate the predictive power of the parameter ensemble.
Additional details of the signaling architecture included in the model are presented in the materials and methods and the supplement.

POETs generated a population of probable signaling models which captured the multiple phases of EMT induction (Fig. \ref{fg:F2}).
POETs sampled well over 10$^6$ probable models during each stage of the cross-validation, using a combination of both local and global random sampling.
This sampling generated approximately 15,000 highly probable models from which we selected N $\simeq$ 1100 models for further analysis.
The selected models all had the same possible molecular connectivity, but different values for model parameters and extrinsic factors such as RNA polymerase or ribosome abundance.
Model selection was based upon Pareto rank, the prediction and training error across all objectives and model to model correlation (supplemental materials).
The model population recapitulated key signaling events following TGF-$\beta$ exposure.
We subdivided the response to TGF-$\beta$ exposure into two phases.
First, TGF-$\beta$1/2 signaling initiated a program which downregulated E-cadeherin expression in a MAPK dependent manner while
simultaneously upregulating TGF-$\beta$3 expression. Second, TGF$-\beta$3 secretion initiated an autocrine feedback which upregulated the expression of
mesenchymal markers such as Vimentin and key upstream transcription factors such as LEF-1 in a SMAD dependent manner.
Each phase involved the hierarchal expression and/or post-translational modification of several key transcription factors.
During the first phase, stimulation with TGF-$\beta$1/2 (10 a.u.) activated both the SMAD and MAPK pathways.
MAPK activation resulted in the phosphorylation of the transcription factor activator protein 1 (AP-1), which in-turn upregulated the expression of Snail, a well established transcriptional repressor (Fig. \ref{fg:F2}A).
Snail expression was MAPK-dependent; the MEK inhibitor U0126 blocked AP-1 activation and Snail expression following TGF-$\beta$1/2 exposure (Fig. \ref{fg:F2}A, Lane 3).
Similar results were obtained for Slug expression, confirming initial activation through the MAPK pathway (data not shown).
Overexpression of either Snail or Slug upregulated TGF-$\beta$3 expression (Fig. \ref{fg:F2}C) while simultaneously downregulating E-cadeherin expression (Fig. \ref{fg:F2}F).
During the second phase, TGF$-\beta$3 secretion and the subsequent autocrine signaling resulted in the upregulation of mesenchymal marker expression.
The TGF$-\beta$3 induced gene expression program involves a complex hierarchy of transcriptional and post-translational regulatory events.
Absence of E-cadherin indirectly promoted TGF$-\beta$3 expression through the $\beta$-catenin/TCF4 complex following Snail or Slug expression (Fig. \ref{fg:F2}C, Lane 2 or 3).
Conversely, over-expression of E-cadherin inhibited the TGF$-\beta$3 autocrine production by sequestering cytosolic $\beta$-catenin, thereby blocking EMT (Fig. \ref{fg:F2}C, Lane 4 or 5).
TGF$-\beta$3 signaled through through the Smad pathway to regulate LEF-1 expression and downstream target EMT genes (Fig. \ref{fg:F2}G).
TGF$-\beta$3 (10 a.u.) in combination with downstream inhibitors (DN-Smad4 and DN-LEF-1) completely inhibited Vimentin expression, while elevating E-cadherin expression (Fig. \ref{fg:F2}H,I).

The predictive power of the ensemble was tested using both cross validation and by comparing simulations with data sets not used for model training.
In whole, 78\% of our training objectives were statistically significant (at a 95\% confidence interval) compared to a randomized parameter family (N = 100)
generated from the best-fit nominal set (starting point for the optimization).
Conversely, we \emph{predicted} approximately 60\% of the training objectives, at a 95\% confidence interval compared to randomized parameters.
The model also captured the temporal gene expression responses of E-cadherin, pSmad2, and LEF-1 to within one-standard deviation (up to the 48 hr time-point) (Fig. \ref{fg:F2}J-L).
This data was not used for model training.
The high predictability can be attributed to the combination of the leave-one-out cross validation scheme, diverse objective functions, and robustness of the POETs algorithm.
Taken together, the model captured the key signaling events revealed by Medici \emph{et al.} \cite{Medici:2008fk} that drive the phenotypic conversion.
A listing of data used for training is included in the supplement (Fig. \ref{fig:S5} and Fig. \ref{fig:S6}).

\subsection*{Identification of a novel LEF-1 regulator}
During model identification, we found that consistent TGF-$\beta$ induced EMT required an additional regulatory protein.
This protein, which we called hypothetical regulator 1 (YREG1), was required to mediate between SNAIL/SLUG transcriptional activity
and the upregulation of LEF-1 expression following TGF-$\beta$1/2 exposure.
SNAIL/SLUG are well known transcriptional repressors \cite{Hemavathy:2000aa,Hemavathy:2000ab,Dhasarathy:2011aa}, although there are a few studies
which suggest that at least SNAIL can also act as a transcriptional activator \cite{Guaita:2002aa}.
In the model, we assumed the expression of SNAIL/SLUG was likely regulated by AP1/SP1 \cite{Jackstadt:2013aa}.
Thus, upon receiving a TGF-$\beta$1/2 signal, the model predicted enhanced SNAIL/SLUG expression, consistent with experimental observations.
TGF-$\beta$1/2 stimulation also induces LEF-1 expression. However, literature evidence suggested that LEF-1 expression was not strongly dependent upon AP1/SP1 activity \cite{Eastman:1999aa}.
Thus, either SNAIL/SLUG are acting as inducers (contrary to substantial biochemical evidence) or, they are repressing the expression of an intermediate repressor.
Given the biochemical evidence supporting SNAIL/SLUG as repressors, we created YREG1 a hypothetical intermediate repressor whose
expression is downregulated by SNAIL/SLUG. The literature data therefore suggested that YREG1 had two transcriptional targets, LEF-1 and TGF-$\beta$3.
By adding this regulator, our simulations became consistent with training and literature data.
Medici et al. suggested a similar idea where feedback between $\beta$-catenin and LEF-1 was likely, although this feedback had yet to be identified \cite{Medici:2008fk}.
Low levels of YREG1 expression were used in all simulations to regulate the formation of the $\beta$-catenin-LEF-1 complex.
To test the potency of YREG1, we conducted knockdown and over-expression simulations following the addition of TGF-$\beta$1/2 (Fig. \ref{fig:S8}).
In the absence of YREG1, most of the population failed to consistently respond to TGF-$\beta$1/2 exposure compared to the wild-type (Fig. \ref{fig:S8}A).
Conversely, YREG1 overexpression revealed an exclusively epithelial phenotype following TGF-$\beta$1/2 stimulation (Fig. \ref{fig:S8}B).
Overexpression of YREG1 repressed LEF-1 and TGF-$\beta$3 expression,
thereby not allowing free $\beta$-catenin to form the $\beta$-catenin-LEF-1 complex which promotes mesenchymal gene expression, or SMAD activity following from autocrine TGF-$\beta$3 signaling.
Likewise, the abundance of the pSmad2/4-LEF-1 complex was also reduced in cells overexpression YREG1, which blocked the repression of E-cadherin.
Taken together, we found that low YREG1 expression was necessary for stabilizing EMT,
while elevated YREG1 levels limited the extent of EMT induction.

\clearpage

\subsection*{TGF$-\beta$1/2 and VEGF-A exposure promotes phenotype heterogeneity through NFATc and phosphorylated Sp1}
While we captured the central tendency of many of the molecular features of EMT induction following TGF$-\beta$1/2 exposure, an often neglected but important emergent
feature of developmental and pathological programs is population heterogeneity \cite{Park:2010fu}.
We (and others) previously hypothesized that deterministic model ensembles can interrogate population behavior, at least at a course grained level \cite{Lequieu:2011fj}.
We tested this hypothesis by analyzing the response of the population of EMT models to extracellular cues and then comparing this response to flow cytometry studies.
We used robustness coefficients to quantify the response of the individual members of the ensemble to TGF$-\beta$1/2 stimulation.
We have previously used robustness coefficients to systematically quantify response of a system to structural or operational perturbations,
for example gene deletions or the addition of a growth factor or hormone  \cite{Song:2010fk,Tasseff:2010pi,Lequieu:2011fj,Tasseff:2011ve}.
Robustness coefficients quantify shifts in molecular marker abundance resulting from molecular of environmental perturbations relative to an unperturbed control state.
Robustness coefficients $\gg$ 1 indicate that marker abundance increased, while robustness coefficients $\ll$ 1 indicates marker abundance decreased relative to an unperturbed control.
A value of $\sim$ 1 indicates approximately no change in marker abundance following the perturbation.
We calculated robustness coefficients for each member of the ensemble (N $\simeq$ 1100) for two downstream phenotypic markers, Vimentin (mesenchymal) and E-cadherin (epithelial)
following the addition of TGF$-\beta$1/2 alone (Fig. \ref{fg:F3}), and VEGF-A in combination with NFATc inhibitors (Fig. \ref{fg:F4}).
The absence of TGF$-\beta$1/2 or VEGF-A stimulation was used as the baseline for the robustness calculations.

We identified model subpopulations that exhibited different behaviors following exposure to TGF$-\beta$1/2 (Fig. \ref{fg:F3}A, labeled P1-P4).
Analysis of the molecular signatures of these subpopulations suggested the abundance, localization and state of the Sp1, AP-1 and NFATc transcription factors controlled population heterogeneity.
The behavior of the majority of models ($>$70\%) was similar to subpopulation one (P1) or subpopulation two (P2) in Fig \ref{fg:F3}.
These models showed the classically expected behavior, a switch from an epithelial to mesenchymal phenotype following TGF$-\beta$1/2 exposure.
Models near P1 had evaluated nuclear localized phosphorylated Sp1, relative to non-induced cells (and models near P2).
Elevated Sp1 activity decreased E-cadherin expression through Slug-mediated inhibition, which in turn increased Vimentin expression through TGF$-\beta$3 autocrine signaling and the liberation of $\beta$-catenin.
Near P2, Sp1 transcriptional activity was lower than P1, leading to only modestly increased Vimentin expression and E-cadherin repression following TGF$-\beta$1/2 stimulation.
Near subpopulation three (P3), reduced levels of nuclear phosphorylated AP-1, Sp1, and NFAT (resulting from the loss of ERK kinase activity)
were responsible for Vimentin \emph{repression} relative to the control.
However, the most biologically interesting behavior was exhibited by subpopulation four (P4).
Models near P4 had elevated Sp1 and NFAT transcriptional activity, which increased \emph{both} Vimentin and E-cadherin expression.
Analysis of these hypothetical cells suggested they had \emph{abnormal} signaling; deregulated NFAT expression and nuclear localization promoted E-cadherin expression while TGF$-\beta$1/2 induced Sp1 action promoted Vimentin expression. Analysis of the connectivity and information flow through the signaling architecture suggested that Sp1 and NFAT action could be manipulated \emph{independently} by simultaneous TGF$-\beta$1/2 and VEGF-A stimulation (Fig. \ref{fg:S1}).

To test this hypothesis, we simulated the response of the network to TGF$-\beta$1/2 and VEGF-A treatment with and without NFATc inhibitors (Fig. \ref{fg:F4}).
As expected, stimulation with VEGF-A (50 a.u.) maintained an epithelial population (Q4-43.6\%), while TGF$-\beta$1/2 (10 a.u.) exposure shifted
the population from an epithelial (Q4-5.5\%) to a mesenchymal (Q1-45.6\%) phenotype (Fig. \ref{fg:F4}A and Fig. \ref{fg:F4}B).
On the other hand, combined stimulation with TGF$-\beta$1/2 (10 a.u.) and VEGF-A (50 a.u.) increased both E-cadherin and Vimentin expression (Q2-45.3\%), resulting in a hybrid phenotype with both epithelial and mesenchymal characteristics (Fig. \ref{fg:F4}C).
To better understand this hybrid response, we quantified the simulated protein levels for E-cadherin, Vimentin, phosphorylated nuclear Sp1, nuclear NFATc1, $\alpha$-smooth muscle actin ($\alpha$-SMA) and Slug as a function of condition (Fig. \ref{fg:S2}A-C).
Vimentin expression was correlated with high levels of nuclear phosphorylated Sp1, following TGF$-\beta$1/2 exposure.
Conversely, elevated E-cadherin expression depended upon the activity of NFAT transcription factors downstream of VEGF-A stimulation.
To further isolate the role of NFAT on this hybrid state, we simulated the inhibition of NFAT transcriptional activity across all conditions (all else being equal).
NFAT inhibition in combination with VEGF-A treatment blocked all E-cadherin positive sets (Fig. \ref{fg:F4}D).
Likewise, TGF$-\beta$1/2 treatment in combination with NFATc inhibition also resulted in the loss of E-cadherin expression (Fig. \ref{fg:F4}E).
Lastly, NFATc inhibition in combination with simultaneous TGF$-\beta$1/2 and VEGF-A exposure repressed nearly all E-cadherin expression, shifting nearly the entire population towards a mesenchymal phenotype (Fig. \ref{fg:F4}F). Taken together, high levels of nuclear localized phosphorylated Sp1 correlated with Vimentin expression, while NFATc transcriptional activity was predicted to be critical for maintaining E-cadherin expression.

\subsection*{Combined TGF$-\beta$2 and VEGF-A exposure drives heterogeneity in MCF10A and DLD1 cells}
The EMT model simulations suggested the transcriptional activity of NFATc and Sp1 could be independently tuned to generate a hybrid cell population with both epithelial and mesenchymal characteristics.
To test this hypothesis, we exposed either quiescent epithelial (MCFA10, (Fig. \ref{fg:F5})) or transformed epithelial cells (DLD1, (Fig. \ref{fg:S3})) to combinations of TGF$-\beta$1/2 and/or VEGF-A.
As expected, treatment with TGF$-\beta$1/2 (10ng/ml) increased Slug and Vimentin expression, while repressing E-cadherin expression both at the transcript and protein levels in MCF10A (Fig. \ref{fg:F5}A-B) and DLD1 cells (Fig. \ref{fg:S4}C, Fig S3 D,E).
Both MCF10A (Fig. \ref{fg:F5}C) and DLD1 cells (Fig. \ref{fg:S3}E,G) transitioned from quiescent cobblestone morphology to spread spindle shapes, consistent with EMT.
As predicted, we found increased nuclear localization of phosphorylated Sp1 following TGF$-\beta$1/2 stimulation in both MCF10A (Fig. \ref{fg:F5}B,C) and DLD1 cells (Fig. \ref{fg:S3}E,F).
Consistent with model predictions, VEGF-A (50ng/ml) treatment increased the abundance of NFATc1 and E-cadherin at both the transcript and protein level in both MCF10A (Fig. \ref{fg:F5}A) and DLD1 (Fig. \ref{fg:S3}A) cells.
We also found that NFATc1 nuclear localization significantly increased in both MCF10 and DLD1 treated with VEGF-A independently of the abundance of nuclear localized phosphorylated Sp1 levels (Fig. \ref{fg:F5}B,C Fig.S3C,E ).
Interestingly, combining VEGF-A (50ng/ml) with TGF$-\beta$1/2 (10ng/ml) resulted in significantly elevated expression of both E-cadherin and Vimentin at the transcript and protein levels in both MCF10A and DLD1 cells (Fig 5A,B; Fig S3D,E; Fig S4C).
NFATc1 expression increased, while Sp1 expression was similar to the TGF$-\beta$1/2 case alone (Fig. \ref{fg:F5}A-B,  Fig S3D,E; Fig S4C)), supporting their independent regulation.
The expression of Slug, and Vimentin significantly increased, while E-cadherin levels were increased in MCF10A cells (Fig 5A) and maintained at control levels in DLD1 cells (Fig. S3D).
As further predicted, nuclear co-localization of both NFATc1 and phosphorylated Sp1 were apparent in MCF10A and DLD1 cells treated with both ligands (Fig. \ref{fg:F5}B,C Fig S3E,F).
Taken together, combined VEGF-A and TGF$-\beta$1/2 treatment elicited a hybrid phenotype expressing both mesenchymal and epithelial characteristics in both MCF10A and DLD1 cells.
This phenotype was driven by the transcriptional activity of two key transcription factors, Sp1 and NFATc, which could be modulated independently by TGF$-\beta$1/2 and VEGF-A exposure.

Our robustness analysis predicted that NFATc transcriptional activity was critical to maintaining E-cadherin expression in the presence of both VEGF-A and TGF$-\beta$1/2.
We experimentally tested this hypothesis by exposing both MCF10A (Fig. \ref{fg:F5}E,F) and DLD1 cells (Fig. \ref{fg:S4}) to combinations of VEGF-A and TGF$-\beta$1/2 in the presence or absence of VIVIT,
a soluble peptide inhibitor of NFATc transcriptional activity \cite{Aramburu:1999if}.
Treatment with VEGF-A (50ng/ml) and VIVIT (10$\mu$M) in MCF10A cells significantly reduced E-cadherin expression compared to VEGF-A alone (Fig 5D,E).
Co-treatment with VIVIT and TGF$-\beta$1/2 did not enhance EMT capacity of MCF10A cells above that of TGF$-\beta$1/2 alone (Fig 5A,B,E).
Likewise, VIVIT in combination with both TGF$-\beta$1/2 and VEGF-A resulted in a loss of E-cadherin gene and protein expression, while Slug and Vimentin levels remained increased (Fig. \ref{fg:F5}D,E ).
Quantitative flow cytometry confirmed these results in both MCF10A (Fig. \ref{fg:F5}F) and DLD1 cells (Fig. S4C).
Both epithelial cell lines initially had high levels of E-cadherin expression, and low vimentin abundance (Q1-99.5\%), but both MCF10A and DLD1 cells shifted from an epithelial to mesenchymal phenotype (Q1-33.4\%, Q4-42.8\%) following TGF$-\beta$1/2 exposure.
As expected, NFATc nuclear localization was repressed with VIVIT treatment regardless of ligand stimulation, while the abundance of nuclear phosphorylated Sp1 increased for both TGF$-\beta$1/2 and TGF$-\beta$1/2 + VIVIT conditions (Fig. 5D,E).
Combined TGF$-\beta$1/2 and VEGF-A increased both Vimentin and E-cadherin expression (Q1-42.1\%, Q2-52.3\%) compared to TGF$-\beta$1/2 alone.
Together, these results demonstrate that NFATc and phosphorylated Sp1 are critical for regulating E-cadherin and Vimentin expression during phenotype heterogeneity in MCF10A and DLD1.

\subsection*{Ductal branching during acini formation is dependent upon phenotype heterogeneity in MCF10A and DLD1 cells}
We finally employed established three-dimensional (3D) \emph{in vitro} models of invasion, migration, compaction, and tubulogenesis \cite{Dhimolea2010}
to determine the functional consequences of the hybrid phenotype (Fig. \ref{fg:F6}).
MCF10A and DLD1 cells were aggregated via hanging drop, placed on the surface of a collagen gel, and cultured for 72 hrs under various biochemical treatments.
TGF$-\beta$1/2 stimulation significantly enhanced cell matrix invasion and matrix compaction, while in contrast VEGF-A stimulation promoted surface migration but no invasion or compaction (Fig. \ref{fg:F6}B-D).
Interestingly, combined TGF$-\beta$1/2 and VEGF-A stimulation significantly increased cell migration potential above that of VEGF-A alone while maintaining 3D matrix compaction, though with decreased magnitude compared to TGF$-\beta$1/2 alone.
Inhibition of NFATc transcriptional activity by VIVIT decreased migration following treatment with VEGF-A alone (Fig. \ref{fg:F6}B).
Co-treatment of VIVIT significantly decreased migration, while complementarily increasing invasion and compaction, when MCF10A cells were stimulated with both VEGF-A and TGF$-\beta$1/2 (Fig. \ref{fg:F6}B-D).
The responses of DLD1 cells followed a similar trend to MCF10A, although the magnitudes of migration, invasion, and compaction were less.
Cell circularity within 3D gels strongly and negatively correlated with both invasion and compaction regardless of treatment (Fig. \ref{fg:F6}E).
Circularity refers to the morphology of the cells. In general, a quiescent epithelial cells assumes a circular morphology in culture, while an active mesenchymal cell is highly elongated.
The circularity index, a common means of quantifying cell morphology, relates cell area to perimeter.
A perfect circle has a circularity index equal to 1.0, while a straight line has a circularity index equal to 0.0, see Butcher et al. \cite{Butcher:2004aa}.
TGF$-\beta$1/2 treatment alone resulted in irregular and spindle shaped morphology, while VEGF-A exposure promoted round quiescent cells (Fig. \ref{fg:F6}A).
Combined VEGF-A and TGF$-\beta$1/2 promoted morphology between these extremes.
VIVIT mediated NFATc inhibition significantly reduced the circularity index, similar to TGF$-\beta$1/2 treatment (Fig. \ref{fg:F6}F).
VEGF-A treatment also induced the formation of tubular structures (acini), but the number of tubular branches relative to total acini was significantly increased upon combined TGF$-\beta$1/2 and VEGF-A.
No tubular structures were identified within the DLD1 constructs during the 7 day tubulogenesis endpoints, supporting that MCF10A and DLD1 cells have some cell-type specific EMT sensitivity despite their underlying competency for acquiring a heterogeneous phenotype. This suggests that initial EMT sensitivity of a cell influences downstream functional response from TGF-b and VEGFA stimulation.
Together, these results establish that VEGF-A and TGF$-\beta$1/2 ligand concentrations potentiate between acini and ductal branch formation in 3D culture, and are dependent upon NFATc activity.

\clearpage

\section*{Discussion}
In this study, we developed a family of mechanistic models describing the induction of EMT by TGF-$\beta$ isoforms in the presence and absence of VEGF-A.
The signaling architecture encoded in the model, which contained 995 molecular species interconnected by 1700 interactions,
described the expression of 80 genes in response to growth factor stimulation.
This simulation incorporates an unprecedented level of detail compared to previous models, but as a consequence created a large number of unknown model parameters.
Because these parameters could not be estimated uniquely apriori, we estimated an ensemble of likely parameters using the POETs multiobjective optimization framework.
The model population was trained and cross-validated to prescribe biological significance using 41 data sets generated in DLD1 colon carcinoma, MDCKII, and A375 melanoma cell lines \cite{Medici:2008fk}.
POETs generated $>$ 15,000 probable parameter sets using this data, from which we selected N $\simeq$ 1100 for subsequent analysis.
Analysis of this population predicted possible phenotypic modes (and their associated signaling) that cells could exhibit when stimulated with TGF-$\beta$ and/or VEGF-A.
The most novel hypothesis generated from the analysis was that cells could operate in a hybrid state defined by both epithelial and mesenchymal traits when stimulated simultaneously with TGF-$\beta$ and VEGF-A.
We tested this hypothesis in MCF10A and DLD1 cells stimulated with combinations of TGF-$\beta$ and VEGF-A.
As expected, in the presence of TGF-$\beta$ or VEGF-A alone, MCF10A and DLD1 cells were either mesenchymal or epithelial, respectively.
However, with both TGF-$\beta$ and VEGF-A, MCF10A and DLD1 cells exhibited a hybrid phenotype, having both epithelial and mesenchymal characteristics.
Furthermore, we found that functional traits such as tubulogenesis and ductal branching were different for cells in this hybrid phenotype.
Together, this study established a predictive model of EMT induction, determined that deterministic model ensembles could predict population heterogeneity, and proved the existence of a unique hybrid phenotype resulting from the simultaneous integration of extracellular growth factor signals.

Cells routinely process a multitude of signals simultaneously, especially when coordinating developmental or pathological programs.
For example, oncogenic cells integrate both mechanical and chemical cues in their local microenvironment during tumorigenesis, including cytokines VEGF and TGF$-\beta$ \cite{Hong:2013mi}.
VEGF-A mediates pathological angiogenic remodeling of tumors \cite{Nagy:2007lh}, while TGF$-\beta$ can elicit both protective and oncogenic responses \cite{Ferrara:2002kx,Willis:2007uq}.
While much research has tested signaling pathways individually, far less is understood about combinatorial stimulation, such as with both VEGF-A and TGF$-\beta$.
Recent \emph{in vitro} and \emph{in vivo} evidence has suggested that epithelial cells can exhibit heterogeneous phenotypes in addition to classically defined epithelial or mesenchymal states \cite{Polyak2009,Strauss:2011fk}.
For example, expression profiling in human epithelial cancer cell lines demonstrated a spectrum of phenotypes, including some that expressed both E-cadherin and Vimentin simultaneously \cite{Neve:2006uq,Welch-Reardon:2014aa}.
Zajchowski \emph{et al.}, speculated that these expression profiles were somehow important for maintaining epithelial properties, while simultaneously allowing other functional behavior such as proliferation and migration \cite{Zajchowski:2001kx}.
Whether and how heterogeneous phenotypes arise and participate in cancer progression, as well as their response to pharmacological inhibition are fundamental questions
that should receive increased attention.
In this study, we determined that a hybrid phenotype could be obtained through combined treatment with VEGF-A and TGF$-\beta$, both common factors localized in the tumor microenvironment.
Furthermore, our systematic simulation-experimentation strategy identified that the transcriptional activity of Sp1 and NFATc were the critical factors controlling this phenotypic heterogeneity.
Several studies have highlighted the importance of NFATc as a key transcription factor involved in cell growth, survival, invasion, angiogenesis and cancer \cite{Mancini2009}.
For example, proliferation and anchorage-independent growth of pancreatic tumor cells is dependent on calcineurin and NFATc1 activity, consistent with the high levels of nuclear NFATc1 found in pancreatic tumors \cite{Singh:2010jx}.
Likewise, our results found that VEGF-A was a potent inducer of NFATc1 expression, which may be required for epithelial cell migration and tubulogenesis.
Although specific NFATc isoforms were not distinguished in the model, our simulations suggested that NFATc transcriptional activity was capable of maintaining epithelial traits, even during TGF$-\beta$ induced EMT.
Experimentally, we found that E-cadherin expression was dependent upon NFATc dephosphorylation in response to simultaneous VEGF-A and TGF$-\beta$1/2 treatment.
Thus, these results support the hypothesis that NFATc activity plays a critical role in maintaining cell-cell contacts, even during partial EMT.

Epithelial cells reproduce tissue-like organization when grown in a three-dimensional extracellular matrix (ECM) environment, and therefore are an attractive model to study morphogenic mechanisms.
It is well established that MCF10A cells form structures that closely resemble acini (multi-lobed cluster of cells) in three-dimensional \emph{in vitro} cultures \cite{Debnath2003}.
It has been postulated that a cellular response reminiscent of partial EMT underlies this process, stimulating further branching and formation of acini \cite{Pearson:2007qq}.
Normally well controlled process such as tubulogenesis can be co-opted by cancer cells to break away from a primary lesion and invade through the surrounding stroma \cite{OBrien2004}.
However, by retaining a transient hybrid EMT-like state, clusters of these tube-forming tumor cells can reform at a high rate after invasion, possibly explaining why invasive human carcinomas frequently appear to be cellular collections with varying degrees of gland-like differentiation \cite{Debnath:2005qc}.
In this study, we showed that our predicted hybrid phenotype generated by simultaneous treatment of epithelial cells with VEGF-A and TGF$-\beta$ possessed altered migration and invasion, which enhanced tubular branching.
A salient feature of this behavior, however, was the retention of cell-cell contacts that allowed cells to migrate without completely dissociating from their neighbors.
Thus, our results support a mechanism in which hybrid cells can maintain some functional characteristics of epithelial cells such as cell-cell adhesion, which are normally lost in a fully differentiated mesenchymal state. The tumor microenvironment contains many soluble signals simultaneously, including VEGF and TGF$-\beta$.
Thus, it is likely that some cancerous epithelial cells could exhibit hybrid EMT phenotypic states.
This may explain why fibroblastoid morphology, a classical feature of EMT, is not commonly observed in human carcinomas \cite{Debnath:2005qc}.
This study focused on the combinatorial effects of two very different ligand families present together in the tumor environment.
Additional modeling studies are required to unravel the global response of epithelial cells to the full spectrum of chemical, substrate, and mechanical cues.
The simulation strategy presented here is readily adaptable to larger species sets, with the major advantage that experimentally testable hypotheses can be generated regarding how signals get integrated to produce global cellular response. Furthermore, by simulating multiple ensembles of parameter sets, subpopulations across a constellation of phenotypes can be created and mined for common and/or divergent signaling characteristics.
This is a significant advantage over forced convergence to a single unique solution and thereby generating a potentially non-physiological homogeneous population.

The deterministic population of EMT models predicted heterogeneous behavior that was qualitatively consistent with experimental studies.
There is a diversity of algorithmic approaches to estimate model parameters \cite{Moles:2003aa}, as well as many strategies to integrate model identification with experimental design \cite{Rodriguez-Fernandez:2013aa,Villaverde:2014aa}.
However, despite these advances, the identification of models describing intracellular network behavior remains challenging.
There are different schools of thought to deal with this challenge.
One school has focused on model reduction.
Data-driven approaches \cite{Cirit:2012kx}, boolean \cite{Choi:2012bs} or other logical model formulations \cite{Morris:2011ys,Terfve:2012kq}
are emerging paradigms that constrain model complexity by the availability of the training and validation data.
Other techniques such as constraints based modeling, which is commonly used to model metabolic networks, have also been applied to model transcriptional networks, although primarily in lower eukaryotes and prokaryotes \cite{Hyduke:2010rt}. These techniques (and many others, see review \cite{Wayman2013}) are certainly exciting, with many interesting properties.
However, we used the traditional approach of mass action kinetics within an ordinary differential equation framework.
The identification problem for the EMT model was massively underdetermined.
This is not uncommon for differential equation models, especially those that are highly mechanistic.
Of course, we could have discarded mechanism or reduced the model scope to decrease the complexity of the identification problem.
However, a central criticism leveled by biologists is that model simplification is often done at the cost of biological reality, or done for reasons of computational expediency \cite{Sainani2012}.
To avoid this criticism, we systematically identified an ensemble of likely models each consistent with the training data, instead of a single but uncertain best fit model.
Previously, we (and others) have suggested that deterministic ensembles could model heterogeneous populations in situations where stochastic computation was not feasible \cite{Lequieu:2011fj}.
Population heterogeneity using deterministic model families has previously been explored for bacterial growth in batch cultures \cite{Lee:2009fk}.
In that case, distributions were generated because the model parameters varied over the ensemble, i.e., extrinsic noise led to population heterogeneity.
In this study, parameters controlling physical interactions such as disassociation rates or the rate of assembly or degradation of macromolecular machinery such as ribosomes were widely distributed over the ensemble. Population heterogeneity can also arise from intrinsic thermal fluctuations, which are not captured by a deterministic population of models \cite{Swain:2002ve}.
Thus, deterministic ensembles, provide a coarse-grained or extrinsic-only ability to simulate population diversity. Despite this limitation, our prediction of phenotypic heterogeneity
(and the underlying signaling events responsible for the heterogeneity) was consistent with experimental observations. This suggested that deterministic ensembles could simulate
disease or developmental processes in which heterogeneity plays an important role, without having to resort to stochastic simulation.

A common criticism of ODE modeling has been the poorly characterized effect of structural and parametric uncertainty.
In this study, parametric uncertainty was addressed by developing an ensemble of probable models instead of a single best-fit but uncertain model using multiobjective optimization.
While computationally complex, multiobjective optimization is an important tool to address qualitative conflicts in training data that arise from experimental error or cell line artifacts \cite{Handl:2007fk}. On the other hand, structural uncertainty is defined as uncertainty in the biological connectivity.
The EMT model connectivity was assembled from an extensive literature review. However, several potentially important signaling mechanisms were not included.
First, we identified a potential gap in biological knowledge surrounding the regulation of LEF-1 expression, that was filled by the addition of the hypothetical YREG1 transcriptional repressor.
The LEF-1 transcription factor is expressed in tissues that undergo EMT during embryogenesis \cite{Vega:2004qf,Nawshad:2003ff},
and has been suggested to promote an invasive phenotype in cancer cells \cite{Cano:2000kh,Kim:2002lh}.
Low levels of YREG1 were important for stabilizing the interaction between LEF-1 and $\beta$-catenin, while elevated levels inhibited EMT by downregulating LEF-1 transcriptional activity.
Recent evidence has established a complex role of Amino terminal Enhancer of Split (AES) and Groucho/TLE on suppressing LEF-1 activity.
AES opposes LEF-1 transcriptional activation while Groucho/TLE binds with LEF-1 for a histone deacetylase repression.
In addition, $\beta$-catenin directly displaces Groucho/TLE repressors from TCF/LEF-1 in Wnt-mediated transcription activation \cite{Arce:2009pr, Grumolato2013}.
Our model agrees with this newly discovered feedback system, as YREG1 regulates LEF-1 activity leading to EMT stabilization.
Secondly, we should revisit the role of GSK-3$\beta$. GSK-3$\beta$ is an important regulator which controls the abundance of both Snail and $\beta$-catenin through the ubiquitin-proteasome pathway \cite{Zhou:2004kx,Larue:2005vn}. Specifically to our model, expression of Snail increases through 72 hrs. In contrast, experimental data has shown that activity of Snail peaks at 24 hrs which may be controlled by the GSK-3$\beta$ complex \cite{Medici:2006qa}.  Recent evidence has also suggested an essential role of NF-$\kappa$B in epithelial transformation.  NF-$\kappa$B may influence Snail expression through the AKT pathway and directly stabilize Snail activity \cite{Wu:2009cr}.  This is particularly important for integrating inflammation pathways, such as interleukin-6 (IL-6) and tumor necrosis factor-$\alpha$ (TNF-$\alpha$), which have been linked to EMT in pathological conditions \cite{Sullivan:2009bh}.
Other pathways such as Notch have also been shown to act synergistically with TGF-$\beta$ to express Slug in the developing embryo \cite{Niessen:2008nx}.
Lastly, while we have modeled classical protein signaling, we have not considered the role of regulatory RNAs on EMT.
There is growing evidence that microRNAs (miRNAs) play a strong role in EMT, where several miRNAs, for example miR-21 and miR-31 are strongly associated with TGF-$\beta$ exposure \cite{Bullock:2012aa}.
Addressing missing structural components like these, could generate more insight into TGF-$\beta$ signaling and its role in phenotypic transformation.

\clearpage

\section*{Materials and Methods}
The simulation code and parameter ensemble used in this study can be downloaded from GitHub (https://github.com/jeffreyvarner/TGFb-VEGFA-Model\_v1.git).

\subsection*{Signaling network connectivity}
The EMT model described the gene expression program resulting from TGF$-\beta$ and VEGF-A signaling in a prototypical epithelial cell.
The TGF$-\beta$-EMT network contained 995 nodes (proteins, mRNA or genes) interconnected by 1700 interactions.
The network connectivity was curated from more than 40 primary literature sources in combination with on-line databases \cite{Linding:2007fk,Jensen:2009fk}.
The model interactome was not specific to a single epithelial cell line.
Rather, we assembled canonical pathways involved in TGF$-\beta$ and VEGF-A signaling, defaulting to human connectivity when possible.
Using a canonical architecture allowed us to explore general features of TGF$-\beta$ induced EMT without cell line specific artifacts.
On the other hand, because of the canonical architecture, we evaluated the simulation conclusions in several cell lines to test the generality of our conclusions.

Our signaling network reconstruction was based on Medici et al. who identified the pathways though which MDCKII, DLD1 colon carcinoma,
and A375 melanoma cells transition towards a mesenchymal phenotype \cite{Medici:2008fk}.
Sequential activation of MAPK and Smad pathways were initiated upon addition of TGF$\beta$1/2.  Briefly, TGF$\beta$2 signals through the RAS-RAF-MEK-ERK pathway to up-regulate Snail and Slug expression \cite{Medici:2006qa}.  Snail, a known repressor of junctional proteins, inhibits the expression of E-cadherin \cite{Cano:2000kh}. This initial repression of E-cadherin leads to a release of $\beta$-catenin from the cell membrane. Cystolic $\beta$-catenin can then translocate to the nucleus and form transcriptional complexes with TCF-4 to drive TGF$\beta$3 expression \cite{Medici:2008fk}. TGF$\beta$3 signals to the cells interior by binding to type II receptors, which form heterodimers with type I receptors (ALK5) \cite{Derynck:2003fc}. This activates the receptors serine/threonine kinase activity to phosphorylate and activate the receptor Smads 2/3 \cite{Massague:2005qc}.
Phosphorylated Smads 2/3 (pSmad2/3) form heterodimers with partner Smad4 and translocate to the nucleus. pSmads complexes up-regulate other transcription factors, such as LEF-1. The pSmad2/4/LEF-1 has been shown to directly repress the E-cadherin gene \cite{Nawshad:2007kl}. LEF-1 also binds with $\beta$-catenin to upregulate mesenchymal proteins such as fibronectin \cite{Medici2011}.
The EMT gene expression program was initiated by the binding of TGF$-\beta$ isoforms to TGF$-\beta$ surface receptors.
Binding of extracellular TGF$-\beta$1/2 with TGF$-\beta$ surface receptors I/II (TGF$-\beta$R-I/II) initiates the assembly of adapter complexes which starts the
downstream signaling program. In the model, TGF$-\beta$1/2 binds TGF$-\beta$R-I/II followed by the recruitment of activin receptor-like kinase 1 (ALK1) and
TGF$-\beta$ surface receptor III (TGF$-\beta$R-III) to form the activated receptor complex \cite{Derynck:2003fc}.
Alternatively, we also included activin receptor-like kinase 5 (ALK5) recruitment in combination with Endoglin and TGF$-\beta$R-III
as a second (redundant) activated receptor complex \cite{Gatza:2010if}.
Complex assembly activates the serine/threonine kinase activity on the receptor, leading to the recruitment and phosphorylation of Smad partners \cite{Massague:2005qc}.
Phosphorylated Smads2/3 (pSmad2/3) form heterodimers with partner Smad4 and then translocate to the nucleus where they act as both transcriptional activators and repressors.
Nuclear pSmad2/3-Smad4 form transcriptional complexes with several genes in the model including lymphoid enhancer-binding factor 1 (\emph{LEF-1}), Nuclear factor of activated T-cells, cytoplasmic 1 (\emph{NFACT1}), and Specificity Protein 1 (\emph{SP1}).  On the other hand, nuclear pSmad2/3-Smad4 represses (in combination with the LEF-1 protein) the expression of E-cadherin (\emph{Cdh1}) \cite{Nawshad:2007kl} and Cadherin 5, type 2 (VE-Cadherin encoded by \emph{Cdh5}).
Repression of E-cadherin expression is the central event in the transition from an epithelial to a mesenchymal phenotype \cite{Cano:2000kh}.
However, this transition is not solely driven by transcriptional events.
At the protein level, the repression of E-cadherin leads to a release of $\beta$-catenin from cell membrane.
Cystolic $\beta$-catenin then translocates to the nucleus and forms transcriptionally-active complexes with immunoglobulin transcription factor 2 (TCF-4)
to drive TGF$-\beta$3 expression \cite{Medici:2008fk}.
Simultaneously, ERK1/2-mediated phosphorylation of the AP1 and Sp1 transcription factors can also regulate transcriptional complexes involving NFAT, Slug, and Smads.
Lastly, canonical pathways for processing extracellular VEGF-A, BMP and Wnt signals, in addition to the PI3K pathway
were also included in the model. Additional information about the interactions included in the model,
along with the Systems Biology Markup Language (SBML) file encoding these interactions are included in the supplemental materials.

\subsection*{Formulation, solution and analysis of the EMT model equations}
EMT was modeled using mass-action kinetics within an ordinary differential equation (ODE) framework:
\begin{equation}\label{asseassertain_massbalance}
\frac{d\mathbf{x}}{dt}=\mathbf{S}\cdot\mathbf{r}\left(\mathbf{x},\mathbf{k}\right)\qquad\mathbf{x}\left(t_{o}\right)=\mathbf{x}_{o}
\end{equation}
The quantity $\mathbf{x}$ denotes the vector describing the abundance of protein, mRNA, and other species in the model ($995\times 1$).
The stoichiometric matrix $\mathbf{S}$ encodes the signaling architecture considered in the model ($995\times 1700$).
Each row of $\mathbf{S}$ describes a signaling component while each column describes a particular interaction.
The $(i,j)$ element of $\mathbf{S}$, denoted by $\sigma_{ij}$, describes how species $i$ is involved with interaction $j$.
If $\sigma_{ij}>0$, species $i$ is produced by interaction $j$. Conversely, If $\sigma_{ij}<0$, then species $i$ is consumed in interaction $j$.
Lastly, if $\sigma_{ij}=0$, then species $i$ is not involved in interaction $j$.
The term  $\mathbf{r}\left(\mathbf{x},\mathbf{k}\right)$ denotes the vector of interactions rates ($1700\times{1}$).
We modeled each network interaction (gene expression, translation and biochemical transformations)
using elementary rate laws where all reversible interactions were split into two irreversible steps (supplemental materials).
Thus, the rate expression for interaction $q$ was given by:
\begin{equation}\label{eq-mass-action}
r_{q}\left(\mathbf{x},k_{q}\right)=k_{q}\prod_{j\in\left\{\mathbf{R}_{q}\right\}}x_{j}^{-\sigma_{jq}}
\end{equation}
The set $\left\{\mathbf{R}_{q}\right\}$ denotes reactants for reaction $q$, while $\sigma_{jq}$ denotes the stoichiometric coefficient (element of the matrix $\mathbf{S}$) governing species $j$ in reaction $q$. The quantity $k_{q}$ denotes the rate constant (unknown) governing reaction $q$.
Model equations were generated in the C-programming language using the UNIVERSAL code generator, starting from an text-based input file (supplemental materials).
UNIVERSAL, an open source Objective-C/Java code generator, is available as a Google Code project (http://code.google.com/p/universal-code-generator/).
Model equations were solved using the CVODE solver in the SUNDIALS library \cite{Hindmarsh2005} on an Apple workstation (Apple, Cupertino, CA)
as previously described \cite{Tasseff:2011ve}.

\subsubsection*{Estimation of model parameters using multiobjective optimization.}
The EMT model had 1756 unknown parameters (1700 kinetic constants and 56 non-zero initial conditions) which were not uniquely identifiable given the training data.
Instead, we estimated a population of likely models (each consistent with the training data)
using 41 data sets generated in DLD1 colon carcinoma, MDCKII, and A375 melanoma cells taken from Medici \emph{et al.} \cite{Medici:2008fk}.
We used the Pareto Optimal Ensemble Technique (POETs) multiobjective optimization framework in combination
with leave-one-out cross-validation to estimate an ensemble of model parameters \cite{Song:2010fk}.
Cross-validation was used to calculate both training and prediction error during the parameter estimation procedure \cite{kohavi1995study}.
The 41 intracellular protein and mRNA data-sets used for identification were organized into 11 objective functions.
These 11 objective functions were then partitioned, where each partition contained ten training objectives and one validation objective.
The training and validation data were Western blots. Thus, all model simulations were in arbitrary units.
However, POETs does allow a soft constraint on the order of magnitude of the model concentration scale.
In this study, we assumed the natural model concentration scale was pmol/L. We did not place a lower bound on model states.
However, based on the pmol/L natural scale, we treated all values less than 10$^{-3}$ as zero (or no expression).

\subsubsection*{Robustness coefficients.} Robustness coefficients were calculated as shown previously \cite{Tasseff:2011ve,Lequieu:2011fj}.
Robustness coefficients denoted by $\alpha\left(i,j,t_{o},t_{f}\right)$:
\begin{equation}
\alpha\left(i,j,t_{o},t_{f}\right)=\left(\displaystyle\int_{t_{o}}^{t_{f}}x_{i}\left(t\right)dt\right)^{-1}\left(\displaystyle\int_{t_{o}}^{t_{f}}x^{(j)}_{i}\left(t\right)dt\right)
\end{equation}
quantify the response of a marker to a structural or operational perturbation to the network architecture.
Here $t_o$ and $t_f$ denote the initial and final simulation time respectively, while
$i$ and $j$ denote the indices for the marker and the perturbation respectively. A value of $\alpha\left(i,j,t_{o},t_{f}\right)>1$, indicates increased marker abundance, while $\alpha\left(i,j,t_{o},t_{f}\right)<1$ indicates decreased marker abundance following perturbation $j$. If
$\alpha\left(i,j,t_{o},t_{f}\right)\sim{1}$ the $j$th perturbation does not influence the abundance of marker $i$.
Robustness coefficients were calculated for each member of the ensemble (N $\simeq$ 1100).

\subsection*{Cell culture and experimental interrogation}
DLD1 colon carcinoma, MCF10A, and HUVEC were acquired from the American Tissue Culture Collection (Manassas, VA). Cells were grown in culture with RPMI 1640 medium with 10\% fetal bovine serum and 1\% penicillin/streptomycin for DLD1, EBM-2 supplemented with EGM-2, 5\% fetal bovine serum, and 1\% penicillin/streptomycin for HUVEC, or MGEM 2 supplemented with insulin, bovine pituitary extract, cholera toxin, hEGF, hydrocortisone, 5\% horse serum, and 1\% penicillin/streptomycin for MCF10A. Cells were serum starved for 24 hours and removed from all experimental conditions.  Recombinant VEGFA165 was also removed from culture medium prior to experimentation.  Recombinant human TGF$-\beta$2 (R$\And$D Systems, Minneapolis, MN) was added to the culture medium at a concentration of 10 ng/ml and recombinant VEGFA165 at a concentration of (5ng/ml, 50ng/ml) for all relative experiments.  NFAT inhibitor (VIVIT peptide) (EMDBiosciences, Darmstadt, Germany), was added to the culture medium at a concentration of 10$\mu$M for all relative experiments.   Cells were passaged 1:3 or 1:4 every 3-6 d and used between passages 4 and 8.

\subsubsection*{RT-PCR}
RNA extractions were performed using a Qiagen total RNA purification kit (Qiagen, Valencia, CA) and RNA was reverse transcribed to cDNA using the SuperScript III RT-PCR kit with oligo(dT) primer (Invitrogen).  Sufficient quality RNA was determined by an absorbance ratio A260/A280 of 1.8-2.1, while the quantity of RNA was determined by measuring the absorbance at 260nm (A260).  Real-time PCR experiments were conducted using the SYBR Green PCR system (Biorad, Hercules, CA) on a Biorad CFX96 cycler, with 40 cycles per sample. Cycling temperatures were as follows: denaturing, 95C; annealing, 60C; and extension, 70C.  Primers were designed to detect GAPDH, E-cadherin, vimentin, Slug, Sp1, and NFATc1 in cDNA clones: Sp1 (F-TTG AAA AAG GAG TTG GTG GC, R-TGC TGG TTC TGT AAG TTG GG, Accession NG030361.1), NFATc1 (F-GCA TCA CAG GGA AGA CCG TGT C, R-GAA GTT CAA TGT CGG AGT TTC TGA G, Accession  NG029226.1).  GAPDH, E-cadherin, vimentin, and Slug primers were taken from previously published literature \cite{Medici:2008fk}.

\subsubsection*{Antibody Staining}
Samples were fixed in 4\% PFA overnight at 4C.   Samples were then washed for 15 minutes on a rocker 3 times with PBS, permeabilized with 0.2\% Triton-X 100 (VWR International, Radnor, PA) for 10 minutes, and washed another 3 times with PBS.  Samples were incubated overnight at 4C in a 1\% BSA (Rockland Immunochemicals, Inc., Gilbertsville, PA) blocking solution followed by another 4C overnight incubation with either rabbit anti-human E-cadherin 1:100 (Abcam, ab53033), mouse anti-human phospho-Sp1 1:100 (Abcam, ab37707), mouse anti-human vimentin 1:100  (Invitrogen, V9), and rabbit anti-human NFATc1 (Santa Cruz, sc-7294) 1:100.   After 3 washes for 15 minutes with PBS, samples were exposed to Alexa Fluor 488 or 568 conjugated (Invitrogen), species specific secondary antibodies at 1:100 in 1\% BSA for 2 hours at room temperature. Three more washes with PBS for 15 minutes were followed by incubation with either DRAQ5 far red nuclear stain (Enzo Life Sciences, Plymouth Meeting, PA) at 1:1000.

\subsubsection*{FACS}
Flow cytometry for E-cadherin 1:100 (Abcam) and vimentin 1:100 expressing cells was performed.  Briefly, cells were trypsinized, fixed with 4\% PFA for 10 min and then preserved in 50\% methanol/PBS.  Cells were kept in the -20C until antibody staining was preformed.   Samples were divided into multiple aliquots in order to stain the proteins separately and compensate for secondary antibody non-specific binding. Cells were incubated for 24 hrs at 4 C in primary antibody diluted in either PBS (extracellular) or 0.2\% saponin-PBS (intracellular). Cells were then washed 3 times with PBS and incubated with appropriate secondary antibodies and imaged using a Coulter Epics XL-MCL Flow Cytometer (Coulter).  All samples were compensated using appropriate background subtraction and all samples were normalized using 7500 cells per flow condition.

\subsubsection*{Three-Dimensional Culture and Tubulogenesis Assays}
For invasion/migration assays, cells were resuspended in culture media, and allowed to aggregate overnight in hanging drop culture (20$\mu$L; 20,000 cells). The spherical aggregates were placed on the surface of neutralized type I collagen hydrogels (1.5mg/mL) and allowed to adhere for 2~hrs before adding treatments. Cultures were maintained for 72~hrs, after which they were fixed in 4\% PFA and slowly rehydrated using PBS. For compaction assays, cells were pelleted via centrifugation and resuspended within a neutralized collagen hydrogel (1.5mg/mL) solution at a density of 400,000 cells/mL.  250$\mu$L of gel was inoculated into culture wells, which solidified after 60min. Treatments were then added within 800$\mu$L of the culture medium without serum. Gels were liberated from the surfaces of the culture wells the next day and cultured free floating for an additional 3-7 days, exchanging serum free media with appropriate factors every 48~hrs.

Tubulogenesis was defined as a typical nonmalignant acini structure.
This includes a polarized epithelial cell, hollow lumen, and the basal sides of the cell are surrounded by ECM proteins (Fig. \ref{fg:F6}A, Controls or VEGF treated). Previous work has shown that change in the morphological
characteristics of nontumorigenic MCF10A epithelial acini occur over time and exploiting them to growth in 3D culture can be quantified.
For example, using image segmentation, Chang et al. \cite{ChangPP07} examined the elongation of the MCF10A acini at 6, 12, and 96 hours after a particular treatment.
Polizzotti et al. \cite{Polizzotti} also suggested a computational method to quantify acini structure based on morphological characteristics in nonmalignant, noninvasive, and invasive conditions.
Adapted from these approaches, we first fluorescently labeled our cultures and captured the acini structures by 3D confocal microscopy.
Next individual acini structures in the images were segmented by imageJ and labeled.
We then extracted the number of ductal branches. Ductal branching was defined as any elongated cell cluster extending away from the total acini structure, which was manually segmented and counted using ImageJ.
A total of 5 images for each condition were used, and approximately 12 acini were analyzed in each image.
Total branching was normalized to the amount of acini present, and provides an overall general assessment to the extent of acini remodeling.

\subsubsection*{Statistics}
Results are expressed as mean $\pm$ standard error, n$\geq$6. Data was analyzed with the GraphPad Prism version 4.00 for Windows (GraphPad Software, San Diego, CA) and SAS (Statistical Analysis Software, Cary, NC). A one-way ANOVA with Tukey's post hoc was used to compare differences between means and data was transformed when necessary to obtain equal sample variances. Differences between means were considered significant at p$\prec$0.05.

%\section*{Acknowledgments}
%The project described was supported by Award Number \#U54CA143876 from the National Cancer Institute to JDV, National Science Foundation CBET-0955172 to JTB, and National Institute of Health HL110328 to JTB. The content is solely the responsibility of the authors and does not necessarily represent the official views of the National Cancer Institute or the National Institutes of Health. Lastly, we thank the anonymous reviewers for their helpful suggestions.

\clearpage
% \bibliographystyle{plain}
%\bibliographystyle{IEEEbib}

\bibliography{References_v1}

\clearpage

\begin{figure}
%\includegraphics [width=1.0\linewidth] {./figs/Fig-1-Network.pdf}
\caption{Model connectivity recreates the core architecture during EMT. The EMT network contains 995 nodes (proteins, mRNA, and genes) interconnected by 1700 interactions.
Central to EMT induction, activation of the MAPK cascade occurs through TGF-$\beta$1/2 binding which activates the AP-1/Sp1 transcriptional axis. AP-1/Sp1 drives an autocrine response of TGF-$\beta$3, which activates the Smad cascade, leading to phenotypic change. Conversely, VEGF-A binding promotes an epithelial phenotype through NFAT activation.
Other important signaling pathways not shown but also incorporated include the BMP, Wnt, and PI3K pathways.
The complete list of molecular interactions that comprise the model is given in the supplement.}\label{fg:F1}
\end{figure}

%\clearpage

\begin{figure}
%\includegraphics [width=1.0\linewidth] {./figs/Fig-2-Training.pdf}
\caption{Training and validation simulations. The population of EMT models qualitatively captured TGF$\beta$-induced EMT signaling.
(A-I) The population was generated using POETs and trained using 11 different objective functions (41 data sets) taken from Medici \emph{et al.} \cite{Medici:2008fk}.
The model captured the simulated experiments for 78\% of the cases.
(J-L)  The model populations were also compared against untrained temporal data to measure the effectiveness as a pure prediction.
The high predictability can be contributed to the leave-one-out cross validation scheme, objective functions with overlapping data, and multi-objective optimization algorithm.}\label{fg:F2}
\end{figure}

%\clearpage

\begin{figure}
%\includegraphics [width=1.0\linewidth] {./figs/Fig-3-HypotheticalPhenotype.pdf}
\caption{Simulated TGF$\beta$1/2 exposure promoted phenotype heterogeneity.
Robustness coefficients were used to quantify the effect of perturbations after steady state across the population of data sets.
Coefficients with values $>$ 1 ($<$ 1) indicated a marker increased (decreased) compared to a base state, while a value of 1 indicated approximately no change following a perturbation.
E-cadherin and Vimentin robustness coefficients were used as phenotypic markers.
(A) TGF$\beta$1/2 perturbation, we isolated 4 distinct parameter sets known to have phenotypically different behaviors.
(B) Each region has been represented by a small signaling network.
From our analysis, we determined that the differences were a function of downstream transcription factors (phosphorylated-Sp1, and NFAT) within the TGF$\beta$2 and VEGFA pathway, respectively.
(C) We hypothesized that elevated phosphorylated-Sp1 and NFAT levels could drive phenotype heterogeneity through simultaneous TGF$\beta$2 and VEGFA treatment, similar to region three (R3).}\label{fg:F3}
\end{figure}

%\clearpage

\begin{figure}
%\includegraphics [width=1.0\linewidth] {./figs/Fig-4-Robustness.pdf}
\caption{Simulated response to TGF$\beta$1/2 and VEGF-A exposure with and without axis specific inhibitors.
Robustness coefficients were used to quantify the shift in population at 48 hrs.
(A-C) VEGF-A (50 a.u.) treatment resulted in a population with enhanced epithelial (Q4) properties.
This was contrary to the addition of TGF$\beta$2 (10 a.u.), which shifted the population towards a mesenchymal phenotype (Q1).
Interestingly, the combined effects of TGF$\beta$2 and VEGFA was found to increase both ecadherin and vimentin levels, creating a heterogeneous population (Q2).
(D-F) To isolate the effect of NFAT, we inhibited NFAT de-phosphorylation in combination with VEGFA.  This negated the increase in ecadherin expression and shifted the population towards a mesenchymal phenotype (Q1,Q3).  Likewise, combining NFAT inhibition with TGF$\beta$ mitigated all ecadherin expression (Q2) confirming their importance for population heterogeneity.  Lastly, combination of TGF$\beta$2, VEGFA, and NFAT inhibition nearly mitigated all effects of VEGFA, shifting the heterogeneous population (Q2) towards a mesenchymal phenotype (Q1).
In whole, high levels of phosphorylated-Sp1 correlated with vimentin expression, while NFAT was responsible for maintaining E-cadherin expression, although neither were mutually exclusive.}\label{fg:F4}
\end{figure}

%\clearpage

\begin{figure}
%\includegraphics [width=1.0\linewidth] {./figs/Fig-5-MCF10A-Validation.pdf}
\caption{Simultaneous TGF-$\beta$1/2 and VEGF-A treatment induced phenotype heterogeneity and is dependent upon NFAT activity \emph{in-vitro}.  (A) In MCF10A, treatment with (10ng/ml) TGF$\beta$2 increased Slug and vimentin, while ecadherin expression was inhibited at both the gene and protein level at 48 hrs.  Conversely, VEGFA alone increased both NFATc1 and ecadherin gene expression.  Simultaneous TGF$\beta$2 (10ng/ml) and VEGFA (50ng/ml) treatment increased Slug, NFATc1, and vimentin expression, while also increasing ecadherin levels via qPCR.  (B-C)  Immunofluorescence confirmed these results and nuclear co-localization of both phospho-Sp1 and NFAT were found dependent upon TGF$\beta$2 and VEGFA, respectively. (D) To isolate the effect of NFAT, treatment of VEGFA (50ng/ml) and VIVIT (10$\mu$M) reduced ecadherin expression at 48hrs (control-dashed line).  Similarly, combined TGF$\beta$2, VEGFA and VIVIT treatment increased Slug and vimentin expression, while inhibiting ecadherin levels via qPCR.  (E) These findings were confirmed via immunofluorescence as the VIVIT peptide inhibited ecadherin and nuclear localization of NFATc1 in all three cases.  (F)  Quantitative flow cytometry also confirmed this trend.  Similar experiments in DLD1 followed a similar trend (supplement).  Magnification, 40x. Scale bars: 50$\mu$m.  C=Control, T=TGF$\beta$2 , V=VEGFA, VI= NFAT inhibitor (VIVIT).  Asterisks signify statistical differences from each other according to a one-way ANOVA with Tukey's post hoc (p$\prec$0.05).}\label{fg:F5}
\end{figure}

%\clearpage

\begin{figure}
%\includegraphics[width=0.5\linewidth] {./figs/Fig-6-Phenotyping.pdf}
\caption{Ductal branching is dependent upon phenotype heterogeneity within MCF10A in 3-D culture.  MCF10A and DLD1 were formed into spheroids overnight and explanted to a collagen gel for 72 hrs.  For compaction and tubular assays, cells were embedded into collagen gels for 72 hrs, and the extent of tubulogenesis was measured at 7 days.  (A-D) Within MCF10A, TGF$\beta$2 (10ng/ml) enhanced invasion and contractile properties while, VEGFA (50ng/ml) promoted increased migration.  TGF$\beta$2 with VEGFA significantly increased migration, while limiting with compaction.  VIVIT (10$\mu$M) in combination with VEGFA and TGF$\beta$2 decreased migration and compaction, while increasing invasion.  (D) Likewise, cell morphology (circularity index) correlated with both invasion and compaction in MCF10A.  (E-F) The size of tubular structures (acini) also increased significantly upon addition of VEGFA, while the number of ductal branches was most significant upon simultaneous TGF$\beta$2 and VEGFA treatment (Red-Ecadherin, Green-Factin, Blue-Nuclear).   DLD1 cells followed a similar trend, although the degree of migration, invasion, and compaction was less significant.  In addition, no tubular structures were identified during the 7 day tubulogenesis endpoints.  Scale bars: 500$\mu$m, 1000$\mu$m, 250$\mu$m, and 80$\mu$m, respectively.  C=Control, T=TGF$\beta$2 , V=VEGFA, VI= NFAT inhibitor (VIVIT).  Asterisks signify statistical differences from each other according to a one-way ANOVA with Tukey's post hoc (p$\prec$0.05).
Boxes in the left-most panel identify regions identified by arrows that were then imaged in greater zoom in the panel immediately below.
The box diagram was not repeated for arrows in the other panels for clarity, but the same method was applied.}\label{fg:F6}
\end{figure}

%\clearpage

% Supplemental figures -
% Set the S-
\renewcommand\thefigure{S\arabic{figure}}
\renewcommand\thetable{T\arabic{table}}
\renewcommand\thepage{S-\arabic{page}}
\renewcommand\theequation{S\arabic{equation}}

% Reset the counters -
\setcounter{equation}{0}
\setcounter{table}{0}
\setcounter{figure}{0}
\setcounter{page}{1}

\clearpage
\input{Supplemental-JB}

\end{document}

### 2016-09-28 Paper_v7.tex
First commit

--- 2016-09-28 Paper_v5.tex
@@
-\documentclass[11pt,letterpaper]{article}
-\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
+\documentclass[12pt]{article}
+% Load packages
+\usepackage{url}  % Formatting web addresses
+\usepackage{ifthen}  % Conditional
+\usepackage{multicol}   %Columns
+\usepackage[utf8]{inputenc} %unicode support
+\usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{epsfig}
+\usepackage{epstopdf}
+\usepackage{graphicx}
+\usepackage[margin=0.1pt,font=footnotesize,labelfont=bf]{caption}
+\usepackage{setspace}
+%\usepackage{longtable}
+\usepackage{colortbl}
+%\usepackage{palatino,lettrine}
+%\usepackage{times}
+%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
+%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
+\usepackage[wide]{sidecap}
+\usepackage[authoryear,round,comma,sort&compress]{natbib}
+\usepackage[authoryear,round]{natbib}
+\usepackage{supertabular}
+\usepackage{simplemargins}
+\usepackage{lineno}
 
-% Use adjustwidth environment to exceed column width (see example table in text)
-\usepackage{changepage}
+\urlstyle{rm}
 
-% Use Unicode characters when possible
-\usepackage[utf8]{inputenc}
+%\textwidth = 6.50 in
+%\textheight = 9.5 in
+%\oddsidemargin =  0.0 in
+%\evensidemargin = 0.0 in
+%\topmargin = -0.50 in
+%\headheight = 0.0 in
+%\headsep = 0.25 in
+%\parskip = 0.15in
+%\linespread{1.75}
+\doublespace
 
-% textcomp package and marvosym package for additional characters
-\usepackage{textcomp,marvosym}
+\setallmargins{1in}
 
-% fixltx2e package for \textsubscript
-\usepackage{fixltx2e}
+%\bibliographystyle{plain}
+\bibliographystyle{msb}
 
-% amsmath and amssymb packages, useful for mathematical formulas and symbols
-\usepackage{amsmath,amssymb}
-
-% cite package, to clean up citations in the main text. Do not remove.
-\usepackage{cite}
-
-% Use nameref to cite supporting information files (see Supporting Information section for more info)
-\usepackage{nameref,hyperref}
-
-% line numbers
-\usepackage[right]{lineno}
-
-% ligatures disabled
-\usepackage{microtype}
-\DisableLigatures[f]{encoding = *, family = * }
-
-% rotating package for sideways tables
-\usepackage{rotating}
-
-% Remove comment for double spacing
-%\usepackage{setspace}
-%\doublespacing
-
-% Text layout
-\raggedright
-\setlength{\parindent}{0.5cm}
-\textwidth 5.25in
-\textheight 8.75in
-
-% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
-% Captions will be left justified
-\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
-
-% Use the PLoS provided BiBTeX style
-\bibliographystyle{plos2009}
-
-% Remove brackets from numbering in List of References
 \makeatletter
-\renewcommand{\@biblabel}[1]{\quad#1.}
+\renewcommand\subsection{\@startsection
+	{subsection}{2}{0mm}
+	{-0.05in}
+	{-0.5\baselineskip}
+	{\normalfont\normalsize\bfseries}}
+\renewcommand\subsubsection{\@startsection
+	{subsubsection}{2}{0mm}
+	{-0.05in}
+	{-0.5\baselineskip}
+	{\normalfont\normalsize\itshape}}
+\renewcommand\section{\@startsection
+	{subsection}{2}{0mm}
+	{-0.2in}
+	{0.05\baselineskip}
+	{\normalfont\large\bfseries}}
+\renewcommand\paragraph{\@startsection
+	{paragraph}{2}{0mm}
+	{-0.05in}
+	{-0.5\baselineskip}
+	{\normalfont\normalsize\itshape}}
 \makeatother
 
-% Leave date blank
-\date{}
+%Review style settings
+%\newenvironment{bmcformat}{\begin{raggedright}\baselineskip20pt\sloppy\setboolean{publ}{false}}{\end{raggedright}\baselineskip20pt\sloppy}
 
-% Header and Footer with logo
-\usepackage{lastpage,fancyhdr,graphicx}
-\pagestyle{myheadings}
-\pagestyle{fancy}
-\fancyhf{}
-\lhead{\includegraphics[natwidth=1.3in,natheight=0.4in]{PLOSlogo.png}}
-\rfoot{\thepage/\pageref{LastPage}}
-\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
-\fancyheadoffset[L]{2.25in}
-\fancyfootoffset[L]{2.25in}
-\lfoot{\sf PLOS}
+%Publication style settings
 
-%% Include all macros below
+% Single space'd bib -
+\setlength\bibsep{0pt}
 
-\newcommand{\lorem}{{\bf LOREM}}
-\newcommand{\ipsum}{{\bf IPSUM}}
+\renewcommand{\rmdefault}{phv}\renewcommand{\sfdefault}{phv}
 
-%% END MACROS SECTION
+% Change the number format in the ref list -
+\renewcommand{\bibnumfmt}[1]{#1.}
+
+% Change Figure to Fig.
+\renewcommand{\figurename}{Fig.}
 
 \begin{document}
-\vspace*{0.35in}
-\begin{flushleft}
-{\Large
-\textbf\newline{Population heterogeneity in the epithelial to mesenchymal transition is controlled by NFAT and phosphorylated Sp1}}
-\newline
-% Insert Author names, affiliations and corresponding author email.
-\\
-Russell Gould\textsuperscript{1},
-Anirikh Chakrabarti\textsuperscript{2},
-Jeffrey D. Varner\textsuperscript{2},
-Jonathan Butcher\textsuperscript{1,*},
-\\
-\bf{1} Department of Biomedical Engineering, Cornell University, Ithaca, NY, USA
-\\
-\bf{2} School of Chemical and Biomolecular Engineering, Cornell University, Ithaca, NY, USA
-
-% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
-%
-% Remove or comment out the author notes below if they aren't used.
-%
-% Primary Equal Contribution Note
-%\Yinyang These authors contributed equally to this work.
-
-% Additional Equal Contribution Note
-%\ddag These authors also contributed equally to this work.
-
-% Current address notes
-%\textcurrency a Insert current address of first author with an address update
-% \textcurrency b Insert current address of second author with an address update
-% \textcurrency c Insert current address of third author with an address update
-
-% Deceased author note
-%\dag Deceased
-
-% Group/Consortium Author Note
-%\textpilcrow Insert Collaborative Author line here
-
-* E-mail: jtb47@cornell.edu
-\end{flushleft}
-
-%\linenumbers
-
-%{\noindent Keywords:~ | Morphogenesis | Plasticity | Cancer | Systems Biology| Tubulogenesis}
-
+\begin{titlepage}
+{\par\centering\textbf{\Large Population heterogeneity in the epithelial to mesenchymal transition is controlled by NFAT and phosphorylated Sp1}}
+\vspace{0.05in}
+{\par \centering Russell Gould$^{\ddag}$, Anirikh Chakrabarti, Jeffrey D. Varner and Jonathan Butcher$^{\ddag *}$}
+\vspace{0.05in}
+{\par \centering School of Chemical and Biomolecular Engineering and $^{\ddag}$Department of Biomedical Engineering, Cornell University, Ithaca NY 14853}
+\vspace{0.1in}
+{\par \centering \textbf{Running Title:}~ Modeling of TGF-$\beta$ induced EMT}
+\vspace{0.1in}
+% {\par \centering \textbf{To be submitted:}~\emph{Molecular Systems Biology (60,000 Character Limit)}}
+%\vspace{0.1in}
+{\par \centering \textbf{Character count:}~\emph{56,659}}
+\vspace{0.1in}
+{\par \centering $^{*}$Corresponding author:}
+\vspace{0.1in}
+{\par \centering Jonathan T. Butcher,}
+{\par \centering Associate Professor, Department of Biomedical Engineering,}
+{\par \centering 304 Weill Hall, Cornell University, Ithaca NY, 14853}
+{\par \centering Email: jtb47@cornell.edu}
+{\par \centering Phone: (607) 255 - 3575}
+{\par \centering Fax: (607) 255 - 7330}
+\end{titlepage}
+\date{}
+\thispagestyle{empty}
+\pagebreak
 \section*{Abstract}
@@
 
-%as seen by the initiating the formation of endocardial cushions in the atrioventricular valves and mediating palate fusion of medial epithelial cells \cite{Zavadil:2005fu}.
-In this context, EMT terminates and has been suggested as a highly controlled process \cite{Lee:2006tw}.
-EMT is also important under pathological conditions, prominently in fibrosis and in invasion and metastasis of carcinomas \cite{Thiery:2003il}.
-This culminates decades of work within the EMT field appears shared in developmental and pathological conditions \cite{Xu:2009oq}.
+%as seen by the initiating the formation of endocardial cushions in the atrioventricular valves and mediating palate fusion of medial epithelial cells \citep{Zavadil:2005fu}.
+In this context, EMT terminates and has been suggested as a highly controlled process \citep{Lee:2006tw}.
+EMT is also important under pathological conditions, prominently in fibrosis and in invasion and metastasis of carcinomas \citep{Thiery:2003il}.
+This culminates decades of work within the EMT field appears shared in developmental and pathological conditions \citep{Xu:2009oq}.
 %In this study, we determined the extent of EMT in normal epithelial cells exposed to pathological conditions by (1) implementing and subsequently interrogating a population based model containing canonical EMT regulatory networks, and (2) verify our mechanistic targets using established cell lines with differing induction capacities for EMT.
 Toward this objective, we modeled the core molecular interactions of EMT using mass-action kinetics within an ordinary differential equation (ODE) framework.
-A population approach was used to estimate a family or ensemble of EMT models that was consistent with the previous literature \cite{Medici:2006qa,Medici:2008fk}.
+A population approach was used to estimate a family or ensemble of EMT models that was consistent with the previous literature \citep{Medici:2006qa,Medici:2008fk}.
 Using the newly predicted operational paradigms, we sought to capture the role of phospho-Sp1 and NFAT on phenotype heterogeneity during TGF$\beta$ induced EMT \emph{in vitro} via key marker expression profiles and fate processes.
@@
 \section*{Introduction}
-The epithelial to mesenchymal transition (EMT) is a broadly participating, evolutionarily conserved differentiation program essential for tissue morphogenesis, remodeling and pathological processes such as cancer \cite{Thiery:2003il}.
-During EMT polarized, tightly adhered epithelial cell monolayers are transformed into non-interacting motile mesenchymal cells that simultaneously degrade and synthesize extracellular matrix (ECM) components and invade into the underlying tissue space \cite{Stahl:2001pd}.
-EMT is the fundamental initiator of developmental processes such as embryonic gastrulation and valvulogenesis \cite{Eisenberg:1995ai} (also Kalluri J Clin Invest 2009, Thiery Cell 2009).
-Transforming growth factor $\beta$ (TGF-$\beta$) family members are important inducers of both developmental and pathological EMT \cite{Zavadil:2005fu,Xu:2009oq}.
+The epithelial to mesenchymal transition (EMT) is a broadly participating, evolutionarily conserved differentiation program essential for tissue morphogenesis, remodeling and pathological processes such as cancer \citep{Thiery:2003il}.
+During EMT polarized, tightly adhered epithelial cell monolayers are transformed into non-interacting motile mesenchymal cells that simultaneously degrade and synthesize extracellular matrix (ECM) components and invade into the underlying tissue space \citep{Stahl:2001pd}.
+EMT is the fundamental initiator of developmental processes such as embryonic gastrulation and valvulogenesis \citep{Eisenberg:1995ai} (also Kalluri J Clin Invest 2009, Thiery Cell 2009).
+Transforming growth factor $\beta$ (TGF-$\beta$) family members are important inducers of both developmental and pathological EMT \citep{Zavadil:2005fu,Xu:2009oq}.
 Decades of research has focused on identifying molecular regulators of EMT, but almost all on a single gene and in a nearly binary yes/no level of qualitative understanding.
-Medici and coworkers recently identified a core signaling program by which TGF-$\beta$ isoforms induce EMT across a variety of cell lines \cite{Medici:2006qa,Medici:2008fk}.
+Medici and coworkers recently identified a core signaling program by which TGF-$\beta$ isoforms induce EMT across a variety of cell lines \citep{Medici:2006qa,Medici:2008fk}.
 This program involves carefully orchestrated rounds of gene expression driven by the Smad and Snail families of transcription factors as well as
 other key factors such as lymphoid enhancer-binding factor 1 (LEF-1), nuclear factor of activated T-cells, cytoplasmic 1 (NFATc1), and specificity protein 1 (Sp1).
-Coregulators such as $\beta$-catenin, NF-$\kappa$B, and the ErbB family of receptor tyrosine kinases however also participate in EMT regulation, but the degree of each's influence is difficult to ascertain in isolation \cite{Kim:2002lh,Jiang:2007aa,Huber:2004aa,Hardy:2010aa}.
+Coregulators such as $\beta$-catenin, NF-$\kappa$B, and the ErbB family of receptor tyrosine kinases however also participate in EMT regulation, but the degree of each's influence is difficult to ascertain in isolation \citep{Kim:2002lh,Jiang:2007aa,Huber:2004aa,Hardy:2010aa}.
 EMT also exhibits complex temporal dynamics that are often intractable in gain/loss of function studies.
@@
 
-Systems biology and mathematical modeling are essential tools for understanding complex developmental programs like EMT \cite{Ahmed:2007mi}.
+Systems biology and mathematical modeling are essential tools for understanding complex developmental programs like EMT \citep{Ahmed:2007mi}.
 Previous computational models of TGF-$\beta$ induced differentiation focused on single biological factors or EMT in single cells.
 For example, Chung \emph{et al.}, constructed a model of TGF-$\beta$ receptor activation and Smad signaling using ordinary differential equations and mass-action kinetics.
-Their model suggested that a reduction of functional TGF-$\beta$ receptors in cancer cells may lead to an attenuated Smad2 signal \cite{Chung:2009jl}.
-Similarly, Vilar \emph{et al.} suggested that specific changes in receptor trafficking patterns could lead to phenotypes that favor tumorigenesis \cite{Vilar:2006gb}.
+Their model suggested that a reduction of functional TGF-$\beta$ receptors in cancer cells may lead to an attenuated Smad2 signal \citep{Chung:2009jl}.
+Similarly, Vilar \emph{et al.} suggested that specific changes in receptor trafficking patterns could lead to phenotypes that favor tumorigenesis \citep{Vilar:2006gb}.
 Although these models provided insight into the role of receptor dynamics, EMT induction involves many other components, including competing second messengers and interconnected transcriptional regulatory loops.
 Integrating these additional scales of molecular signaling while maintaining the capacity for robust prediction requires a new and expanded computational and experimental strategy.
-Data-driven systems approaches \cite{Cirit:2012kx} or logical model formulations \cite{Morris:2011ys} are emerging paradigms that constrain model complexity through the incorporation of training and validation data.
+Data-driven systems approaches \citep{Cirit:2012kx} or logical model formulations \citep{Morris:2011ys} are emerging paradigms that constrain model complexity through the incorporation of training and validation data.
 These are interesting techniques because the data informs model structure (which can be expanded as more data becomes available).
-Alternatively, Bailey proposed more than a decade ago that a qualitative understanding of a complex biological system should not require complete definition of its structural and parametric content \cite{2001_bailey_NatBiotech}.
-Shortly thereafter, Sethna and coworkers showed that complex model behavior is often controlled by only a few parameter combinations, a characteristic seemingly universal to multi-parameter models referred to as ``sloppiness'' \cite{Machta:2013by}.
+Alternatively, Bailey proposed more than a decade ago that a qualitative understanding of a complex biological system should not require complete definition of its structural and parametric content \citep{2001_bailey_NatBiotech}.
+Shortly thereafter, Sethna and coworkers showed that complex model behavior is often controlled by only a few parameter combinations, a characteristic seemingly universal to multi-parameter models referred to as ``sloppiness'' \citep{Machta:2013by}.
 Thus, reasonable model predictions are often possible with only limited parameter information.
-Taking advantage of this property, we developed sloppy techniques for parameter identification using ensembles of deterministic models \cite{Song:2010fk}.
-Furthermore, we proposed that the sloppy behavior of biological networks may also be seen as a source of cell-to-cell \cite{Lequieu:2011fj} or even patient-to-patient heterogeneity \cite{2010_luan_varner_MolBioSys}.
-Recently, Bayesian parameter identification techniques have also been used to explore cell-to-cell heterogeneity  \cite{2011_kalita_brasier_JBC,2011_hasenauer_allgower_BMCBioinfo},
-where a population of cells could be viewed as a dynamic ensemble of context-specific biochemical networks \cite{2012_creixell_linding_NatBiotech}.
+Taking advantage of this property, we developed sloppy techniques for parameter identification using ensembles of deterministic models \citep{Song:2010fk}.
+Furthermore, we proposed that the sloppy behavior of biological networks may also be seen as a source of cell-to-cell \citep{Lequieu:2011fj} or even patient-to-patient heterogeneity \citep{2010_luan_varner_MolBioSys}.
+Recently, Bayesian parameter identification techniques have also been used to explore cell-to-cell heterogeneity  \citep{2011_kalita_brasier_JBC,2011_hasenauer_allgower_BMCBioinfo},
+where a population of cells could be viewed as a dynamic ensemble of context-specific biochemical networks \citep{2012_creixell_linding_NatBiotech}.
 
@@
 Model equations were formulated using mass-action kinetics within an ordinary differential equation (ODE) framework.
-ODEs and mass action kinetics are common tools to model biochemical pathways \cite{Schoeberl:2002rv,Chen:2009uq,Tasseff:2011ve}.
+ODEs and mass action kinetics are common tools to model biochemical pathways \citep{Schoeberl:2002rv,Chen:2009uq,Tasseff:2011ve}.
 However, while ODE models can simulate complex intracellular behavior,
@@
 The EMT model had 1756 unknown model parameters, 1700 kinetic constants and 56 non-zero initial conditions.
-As expected, these parameters were not uniquely identifiable given the training data \cite{Gadkar:2005vn}.
+As expected, these parameters were not uniquely identifiable given the training data \citep{Gadkar:2005vn}.
 Thus, instead of identifying a single best fit (but uncertain) model,
@@
 The training data were generated in DLD1 colon carcinoma, MDCKII, and A375 melanoma cells following
-exposure to TGF-$\beta$ isoforms \cite{Medici:2008fk}. We organized these data sets into 11 objective functions which were simultaneously minimized by POETs.
-Additionally, we used 12 molecular data sets generated in HK-2 cells following VEGF-A exposure to train VEGF-A responsive model processes \cite{Lian:2011bs}.
+exposure to TGF-$\beta$ isoforms \citep{Medici:2008fk}. We organized these data sets into 11 objective functions which were simultaneously minimized by POETs.
+Additionally, we used 12 molecular data sets generated in HK-2 cells following VEGF-A exposure to train VEGF-A responsive model processes \citep{Lian:2011bs}.
 To guard against overfitting, we augmented the multiobjective optimization with leave-one-out cross validation to independently estimate both the training and
@@
 The high predictability can be attributed to the combination of the leave-one-out cross validation scheme, diverse objective functions, and robustness of the POETs algorithm.
-Taken together, the model captured the key signaling events revealed by Medici \emph{et al.} \cite{Medici:2008fk} that drive the phenotypic conversion.
+Taken together, the model captured the key signaling events revealed by Medici \emph{et al.} \citep{Medici:2008fk} that drive the phenotypic conversion.
 A listing of data used for training is included in the supplement (Fig. \ref{fig:S5} and Fig. \ref{fig:S6}).
@@
 and the upregulation of LEF-1 expression following TGF-$\beta$1/2 exposure.
-SNAIL/SLUG are well known transcriptional repressors \cite{Hemavathy:2000aa,Hemavathy:2000ab,Dhasarathy:2011aa}, although there are a few studies
-which suggest that at least SNAIL can also act as a transcriptional activator \cite{Guaita:2002aa}.
-In the model, we assumed the expression of SNAIL/SLUG was likely regulated by AP1/SP1 \cite{Jackstadt:2013aa}.
+SNAIL/SLUG are well known transcriptional repressors \citep{Hemavathy:2000aa,Hemavathy:2000ab,Dhasarathy:2011aa}, although there are a few studies
+which suggest that at least SNAIL can also act as a transcriptional activator \citep{Guaita:2002aa}.
+In the model, we assumed the expression of SNAIL/SLUG was likely regulated by AP1/SP1 \citep{Jackstadt:2013aa}.
 Thus, upon receiving a TGF-$\beta$1/2 signal, the model predicted enhanced SNAIL/SLUG expression, consistent with experimental observations.
-TGF-$\beta$1/2 stimulation also induces LEF-1 expression. However, literature evidence suggested that LEF-1 expression was not strongly dependent upon AP1/SP1 activity \cite{Eastman:1999aa}.
+TGF-$\beta$1/2 stimulation also induces LEF-1 expression. However, literature evidence suggested that LEF-1 expression was not strongly dependent upon AP1/SP1 activity \citep{Eastman:1999aa}.
 Thus, either SNAIL/SLUG are acting as inducers (contrary to substantial biochemical evidence) or, they are repressing the expression of an intermediate repressor.
@@
 By adding this regulator, our simulations became consistent with training and literature data.
-Medici et al. suggested a similar idea where feedback between $\beta$-catenin and LEF-1 was likely, although this feedback had yet to be identified \cite{Medici:2008fk}.
+Medici et al. suggested a similar idea where feedback between $\beta$-catenin and LEF-1 was likely, although this feedback had yet to be identified \citep{Medici:2008fk}.
 Low levels of YREG1 expression were used in all simulations to regulate the formation of the $\beta$-catenin-LEF-1 complex.
@@
 While we captured the central tendency of many of the molecular features of EMT induction following TGF$-\beta$1/2 exposure, an often neglected but important emergent
-feature of developmental and pathological programs is population heterogeneity \cite{Park:2010fu}.
-We (and others) previously hypothesized that deterministic model ensembles can interrogate population behavior, at least at a course grained level \cite{Lequieu:2011fj}.
+feature of developmental and pathological programs is population heterogeneity \citep{Park:2010fu}.
+We (and others) previously hypothesized that deterministic model ensembles can interrogate population behavior, at least at a course grained level \citep{Lequieu:2011fj}.
 We tested this hypothesis by analyzing the response of the population of EMT models to extracellular cues and then comparing this response to flow cytometry studies.
@@
 We have previously used robustness coefficients to systematically quantify response of a system to structural or operational perturbations,
-for example gene deletions or the addition of a growth factor or hormone  \cite{Song:2010fk,Tasseff:2010pi,Lequieu:2011fj,Tasseff:2011ve}.
+for example gene deletions or the addition of a growth factor or hormone  \citep{Song:2010fk,Tasseff:2010pi,Lequieu:2011fj,Tasseff:2011ve}.
 Robustness coefficients quantify shifts in molecular marker abundance resulting from molecular of environmental perturbations relative to an unperturbed control state.
@@
 \ref{fg:S4}) to combinations of VEGF-A and TGF$-\beta$1/2 in the presence or absence of VIVIT,
-a soluble peptide inhibitor of NFATc transcriptional activity \cite{Aramburu:1999if}.
+a soluble peptide inhibitor of NFATc transcriptional activity \citep{Aramburu:1999if}.
 Treatment with VEGF-A (50ng/ml) and VIVIT (10$\mu$M) in MCF10A cells significantly reduced E-cadherin expression compared to VEGF-A alone (Fig 5D,E).
@@
 \subsection*{Ductal branching during acini formation is dependent upon phenotype heterogeneity in MCF10A and DLD1 cells}
-We finally employed established three-dimensional (3D) \emph{in vitro} models of invasion, migration, compaction, and tubulogenesis \cite{Dhimolea2010}
+We finally employed established three-dimensional (3D) \emph{in vitro} models of invasion, migration, compaction, and tubulogenesis \citep{Dhimolea2010}
 to determine the functional consequences of the hybrid phenotype (Fig. \ref{fg:F6}).
@@
 The circularity index, a common means of quantifying cell morphology, relates cell area to perimeter.
-A perfect circle has a circularity index equal to 1.0, while a straight line has a circularity index equal to 0.0, see Butcher et al. \cite{Butcher:2004aa}.
+A perfect circle has a circularity index equal to 1.0, while a straight line has a circularity index equal to 0.0, see Butcher et al. \citep{Butcher:2004aa}.
 TGF$-\beta$1/2 treatment alone resulted in irregular and spindle shaped morphology, while VEGF-A exposure promoted round quiescent cells (Fig. \ref{fg:F6}A).
@@
 Because these parameters could not be estimated uniquely apriori, we estimated an ensemble of likely parameters using the POETs multiobjective optimization framework.
-The model population was trained and cross-validated to prescribe biological significance using 41 data sets generated in DLD1 colon carcinoma, MDCKII, and A375 melanoma cell lines \cite{Medici:2008fk}.
+The model population was trained and cross-validated to prescribe biological significance using 41 data sets generated in DLD1 colon carcinoma, MDCKII, and A375 melanoma cell lines \citep{Medici:2008fk}.
 POETs generated $>$ 15,000 probable parameter sets using this data, from which we selected N $\simeq$ 1100 for subsequent analysis.
@@
 Cells routinely process a multitude of signals simultaneously, especially when coordinating developmental or pathological programs.
-For example, oncogenic cells integrate both mechanical and chemical cues in their local microenvironment during tumorigenesis, including cytokines VEGF and TGF$-\beta$ \cite{Hong:2013mi}.
-VEGF-A mediates pathological angiogenic remodeling of tumors \cite{Nagy:2007lh}, while TGF$-\beta$ can elicit both protective and oncogenic responses \cite{Ferrara:2002kx,Willis:2007uq}.
+For example, oncogenic cells integrate both mechanical and chemical cues in their local microenvironment during tumorigenesis, including cytokines VEGF and TGF$-\beta$ \citep{Hong:2013mi}.
+VEGF-A mediates pathological angiogenic remodeling of tumors \citep{Nagy:2007lh}, while TGF$-\beta$ can elicit both protective and oncogenic responses \citep{Ferrara:2002kx,Willis:2007uq}.
 While much research has tested signaling pathways individually, far less is understood about combinatorial stimulation, such as with both VEGF-A and TGF$-\beta$.
-Recent \emph{in vitro} and \emph{in vivo} evidence has suggested that epithelial cells can exhibit heterogeneous phenotypes in addition to classically defined epithelial or mesenchymal states \cite{Polyak2009,Strauss:2011fk}.
-For example, expression profiling in human epithelial cancer cell lines demonstrated a spectrum of phenotypes, including some that expressed both E-cadherin and Vimentin simultaneously \cite{Neve:2006uq,Welch-Reardon:2014aa}.
-Zajchowski \emph{et al.}, speculated that these expression profiles were somehow important for maintaining epithelial properties, while simultaneously allowing other functional behavior such as proliferation and migration \cite{Zajchowski:2001kx}.
+Recent \emph{in vitro} and \emph{in vivo} evidence has suggested that epithelial cells can exhibit heterogeneous phenotypes in addition to classically defined epithelial or mesenchymal states \citep{Polyak2009,Strauss:2011fk}.
+For example, expression profiling in human epithelial cancer cell lines demonstrated a spectrum of phenotypes, including some that expressed both E-cadherin and Vimentin simultaneously \citep{Neve:2006uq,Welch-Reardon:2014aa}.
+Zajchowski \emph{et al.}, speculated that these expression profiles were somehow important for maintaining epithelial properties, while simultaneously allowing other functional behavior such as proliferation and migration \citep{Zajchowski:2001kx}.
 Whether and how heterogeneous phenotypes arise and participate in cancer progression, as well as their response to pharmacological inhibition are fundamental questions
@@
 Furthermore, our systematic simulation-experimentation strategy identified that the transcriptional activity of Sp1 and NFATc were the critical factors controlling this phenotypic heterogeneity.
-Several studies have highlighted the importance of NFATc as a key transcription factor involved in cell growth, survival, invasion, angiogenesis and cancer \cite{Mancini2009}.
-For example, proliferation and anchorage-independent growth of pancreatic tumor cells is dependent on calcineurin and NFATc1 activity, consistent with the high levels of nuclear NFATc1 found in pancreatic tumors \cite{Singh:2010jx}.
+Several studies have highlighted the importance of NFATc as a key transcription factor involved in cell growth, survival, invasion, angiogenesis and cancer \citep{Mancini2009}.
+For example, proliferation and anchorage-independent growth of pancreatic tumor cells is dependent on calcineurin and NFATc1 activity, consistent with the high levels of nuclear NFATc1 found in pancreatic tumors \citep{Singh:2010jx}.
 Likewise, our results found that VEGF-A was a potent inducer of NFATc1 expression, which may be required for epithelial cell migration and tubulogenesis.
@@
 Epithelial cells reproduce tissue-like organization when grown in a three-dimensional extracellular matrix (ECM) environment, and therefore are an attractive model to study morphogenic mechanisms.
-It is well established that MCF10A cells form structures that closely resemble acini (multi-lobed cluster of cells) in three-dimensional \emph{in vitro} cultures \cite{Debnath2003}.
-It has been postulated that a cellular response reminiscent of partial EMT underlies this process, stimulating further branching and formation of acini \cite{Pearson:2007qq}.
-Normally well controlled process such as tubulogenesis can be co-opted by cancer cells to break away from a primary lesion and invade through the surrounding stroma \cite{OBrien2004}.
-However, by retaining a transient hybrid EMT-like state, clusters of these tube-forming tumor cells can reform at a high rate after invasion, possibly explaining why invasive human carcinomas frequently appear to be cellular collections with varying degrees of gland-like differentiation \cite{Debnath:2005qc}.
+It is well established that MCF10A cells form structures that closely resemble acini (multi-lobed cluster of cells) in three-dimensional \emph{in vitro} cultures \citep{Debnath2003}.
+It has been postulated that a cellular response reminiscent of partial EMT underlies this process, stimulating further branching and formation of acini \citep{Pearson:2007qq}.
+Normally well controlled process such as tubulogenesis can be co-opted by cancer cells to break away from a primary lesion and invade through the surrounding stroma \citep{OBrien2004}.
+However, by retaining a transient hybrid EMT-like state, clusters of these tube-forming tumor cells can reform at a high rate after invasion, possibly explaining why invasive human carcinomas frequently appear to be cellular collections with varying degrees of gland-like differentiation \citep{Debnath:2005qc}.
 In this study, we showed that our predicted hybrid phenotype generated by simultaneous treatment of epithelial cells with VEGF-A and TGF$-\beta$ possessed altered migration and invasion, which enhanced tubular branching.
@@
 Thus, it is likely that some cancerous epithelial cells could exhibit hybrid EMT phenotypic states.
-This may explain why fibroblastoid morphology, a classical feature of EMT, is not commonly observed in human carcinomas \cite{Debnath:2005qc}.
+This may explain why fibroblastoid morphology, a classical feature of EMT, is not commonly observed in human carcinomas \citep{Debnath:2005qc}.
 This study focused on the combinatorial effects of two very different ligand families present together in the tumor environment.
@@
 The deterministic population of EMT models predicted heterogeneous behavior that was qualitatively consistent with experimental studies.
-There is a diversity of algorithmic approaches to estimate model parameters \cite{Moles:2003aa}, as well as many strategies to integrate model identification with experimental design \cite{Rodriguez-Fernandez:2013aa,Villaverde:2014aa}.
+There is a diversity of algorithmic approaches to estimate model parameters \citep{Moles:2003aa}, as well as many strategies to integrate model identification with experimental design \citep{Rodriguez-Fernandez:2013aa,Villaverde:2014aa}.
 However, despite these advances, the identification of models describing intracellular network behavior remains challenging.
@@
 One school has focused on model reduction.
-Data-driven approaches \cite{Cirit:2012kx}, boolean \cite{Choi:2012bs} or other logical model formulations \cite{Morris:2011ys,Terfve:2012kq}
+Data-driven approaches \citep{Cirit:2012kx}, boolean \citep{Choi:2012bs} or other logical model formulations \citep{Morris:2011ys,Terfve:2012kq}
 are emerging paradigms that constrain model complexity by the availability of the training and validation data.
-Other techniques such as constraints based modeling, which is commonly used to model metabolic networks, have also been applied to model transcriptional networks, although primarily in lower eukaryotes and prokaryotes \cite{Hyduke:2010rt}.
-These techniques (and many others, see review \cite{Wayman2013}) are certainly exciting, with many interesting properties.
+Other techniques such as constraints based modeling, which is commonly used to model metabolic networks, have also been applied to model transcriptional networks, although primarily in lower eukaryotes and prokaryotes \citep{Hyduke:2010rt}.
+These techniques (and many others, see review \citep{Wayman2013}) are certainly exciting, with many interesting properties.
 However, we used the traditional approach of mass action kinetics within an ordinary differential equation framework.
@@
 Of course, we could have discarded mechanism or reduced the model scope to decrease the complexity of the identification problem.
-However, a central criticism leveled by biologists is that model simplification is often done at the cost of biological reality, or done for reasons of computational expediency \cite{Sainani2012}.
+However, a central criticism leveled by biologists is that model simplification is often done at the cost of biological reality, or done for reasons of computational expediency \citep{Sainani2012}.
 To avoid this criticism, we systematically identified an ensemble of likely models each consistent with the training data, instead of a single but uncertain best fit model.
-Previously, we (and others) have suggested that deterministic ensembles could model heterogeneous populations in situations where stochastic computation was not feasible \cite{Lequieu:2011fj}.
-Population heterogeneity using deterministic model families has previously been explored for bacterial growth in batch cultures \cite{Lee:2009fk}.
+Previously, we (and others) have suggested that deterministic ensembles could model heterogeneous populations in situations where stochastic computation was not feasible \citep{Lequieu:2011fj}.
+Population heterogeneity using deterministic model families has previously been explored for bacterial growth in batch cultures \citep{Lee:2009fk}.
 In that case, distributions were generated because the model parameters varied over the ensemble, i.e., extrinsic noise led to population heterogeneity.
 In this study, parameters controlling physical interactions such as disassociation rates or the rate of assembly or degradation of macromolecular machinery such as ribosomes were widely distributed over the ensemble.
-Population heterogeneity can also arise from intrinsic thermal fluctuations, which are not captured by a deterministic population of models \cite{Swain:2002ve}.
+Population heterogeneity can also arise from intrinsic thermal fluctuations, which are not captured by a deterministic population of models \citep{Swain:2002ve}.
 Thus, deterministic ensembles, provide a coarse-grained or extrinsic-only ability to simulate population diversity. Despite this limitation, our prediction of phenotypic heterogeneity
@@
 In this study, parametric uncertainty was addressed by developing an ensemble of probable models instead of a single best-fit but uncertain model using multiobjective optimization.
-While computationally complex, multiobjective optimization is an important tool to address qualitative conflicts in training data that arise from experimental error or cell line artifacts \cite{Handl:2007fk}.
+While computationally complex, multiobjective optimization is an important tool to address qualitative conflicts in training data that arise from experimental error or cell line artifacts \citep{Handl:2007fk}.
 On the other hand, structural uncertainty is defined as uncertainty in the biological connectivity.
@@
 First, we identified a potential gap in biological knowledge surrounding the regulation of LEF-1 expression, that was filled by the addition of the hypothetical YREG1 transcriptional repressor.
-The LEF-1 transcription factor is expressed in tissues that undergo EMT during embryogenesis \cite{Vega:2004qf,Nawshad:2003ff},
-and has been suggested to promote an invasive phenotype in cancer cells \cite{Cano:2000kh,Kim:2002lh}.
+The LEF-1 transcription factor is expressed in tissues that undergo EMT during embryogenesis \citep{Vega:2004qf,Nawshad:2003ff},
+and has been suggested to promote an invasive phenotype in cancer cells \citep{Cano:2000kh,Kim:2002lh}.
 Low levels of YREG1 were important for stabilizing the interaction between LEF-1 and $\beta$-catenin, while elevated levels inhibited EMT by downregulating LEF-1 transcriptional activity.
@@
 AES opposes LEF-1 transcriptional activation while Groucho/TLE binds with LEF-1 for a histone deacetylase repression.
-In addition, $\beta$-catenin directly displaces Groucho/TLE repressors from TCF/LEF-1 in Wnt-mediated transcription activation \cite{Arce:2009pr, Grumolato2013}.
+In addition, $\beta$-catenin directly displaces Groucho/TLE repressors from TCF/LEF-1 in Wnt-mediated transcription activation \citep{Arce:2009pr, Grumolato2013}.
 Our model agrees with this newly discovered feedback system, as YREG1 regulates LEF-1 activity leading to EMT stabilization.
 Secondly, we should revisit the role of GSK-3$\beta$.
-GSK-3$\beta$ is an important regulator which controls the abundance of both Snail and $\beta$-catenin through the ubiquitin-proteasome pathway \cite{Zhou:2004kx,Larue:2005vn}.
+GSK-3$\beta$ is an important regulator which controls the abundance of both Snail and $\beta$-catenin through the ubiquitin-proteasome pathway \citep{Zhou:2004kx,Larue:2005vn}.
 Specifically to our model, expression of Snail increases through 72 hrs.
-In contrast, experimental data has shown that activity of Snail peaks at 24 hrs which may be controlled by the GSK-3$\beta$ complex \cite{Medici:2006qa}.
+In contrast, experimental data has shown that activity of Snail peaks at 24 hrs which may be controlled by the GSK-3$\beta$ complex \citep{Medici:2006qa}.
 Recent evidence has also suggested an essential role of NF-$\kappa$B in epithelial transformation.
-NF-$\kappa$B may influence Snail expression through the AKT pathway and directly stabilize Snail activity \cite{Wu:2009cr}.
-This is particularly important for integrating inflammation pathways, such as interleukin-6 (IL-6) and tumor necrosis factor-$\alpha$ (TNF-$\alpha$), which have been linked to EMT in pathological conditions \cite{Sullivan:2009bh}.
-Other pathways such as Notch have also been shown to act synergistically with TGF-$\beta$ to express Slug in the developing embryo \cite{Niessen:2008nx}.
+NF-$\kappa$B may influence Snail expression through the AKT pathway and directly stabilize Snail activity \citep{Wu:2009cr}.
+This is particularly important for integrating inflammation pathways, such as interleukin-6 (IL-6) and tumor necrosis factor-$\alpha$ (TNF-$\alpha$), which have been linked to EMT in pathological conditions \citep{Sullivan:2009bh}.
+Other pathways such as Notch have also been shown to act synergistically with TGF-$\beta$ to express Slug in the developing embryo \citep{Niessen:2008nx}.
 Lastly, while we have modeled classical protein signaling, we have not considered the role of regulatory RNAs on EMT.
-There is growing evidence that microRNAs (miRNAs) play a strong role in EMT, where several miRNAs, for example miR-21 and miR-31 are strongly associated with TGF-$\beta$ exposure \cite{Bullock:2012aa}.
+There is growing evidence that microRNAs (miRNAs) play a strong role in EMT, where several miRNAs, for example miR-21 and miR-31 are strongly associated with TGF-$\beta$ exposure \citep{Bullock:2012aa}.
 Addressing missing structural components like these, could generate more insight into TGF-$\beta$ signaling and its role in phenotypic transformation.
@@
 The TGF$-\beta$-EMT network contained 995 nodes (proteins, mRNA or genes) interconnected by 1700 interactions.
-The network connectivity was curated from more than 40 primary literature sources in combination with on-line databases \cite{Linding:2007fk,Jensen:2009fk}.
+The network connectivity was curated from more than 40 primary literature sources in combination with on-line databases \citep{Linding:2007fk,Jensen:2009fk}.
 The model interactome was not specific to a single epithelial cell line.
@@
 Our signaling network reconstruction was based on Medici et al. who identified the pathways though which MDCKII, DLD1 colon carcinoma,
-and A375 melanoma cells transition towards a mesenchymal phenotype \cite{Medici:2008fk}.
+and A375 melanoma cells transition towards a mesenchymal phenotype \citep{Medici:2008fk}.
 Sequential activation of MAPK and Smad pathways were initiated upon addition of TGF$\beta$1/2.
-Briefly, TGF$\beta$2 signals through the RAS-RAF-MEK-ERK pathway to up-regulate Snail and Slug expression \cite{Medici:2006qa}.
-Snail, a known repressor of junctional proteins, inhibits the expression of E-cadherin \cite{Cano:2000kh}.
+Briefly, TGF$\beta$2 signals through the RAS-RAF-MEK-ERK pathway to up-regulate Snail and Slug expression \citep{Medici:2006qa}.
+Snail, a known repressor of junctional proteins, inhibits the expression of E-cadherin \citep{Cano:2000kh}.
 This initial repression of E-cadherin leads to a release of $\beta$-catenin from the cell membrane.
-Cystolic $\beta$-catenin can then translocate to the nucleus and form transcriptional complexes with TCF-4 to drive TGF$\beta$3 expression \cite{Medici:2008fk}.
-TGF$\beta$3 signals to the cells interior by binding to type II receptors, which form heterodimers with type I receptors (ALK5) \cite{Derynck:2003fc}.
-This activates the receptors serine/threonine kinase activity to phosphorylate and activate the receptor Smads 2/3 \cite{Massague:2005qc}.
+Cystolic $\beta$-catenin can then translocate to the nucleus and form transcriptional complexes with TCF-4 to drive TGF$\beta$3 expression \citep{Medici:2008fk}.
+TGF$\beta$3 signals to the cells interior by binding to type II receptors, which form heterodimers with type I receptors (ALK5) \citep{Derynck:2003fc}.
+This activates the receptors serine/threonine kinase activity to phosphorylate and activate the receptor Smads 2/3 \citep{Massague:2005qc}.
 Phosphorylated Smads 2/3 (pSmad2/3) form heterodimers with partner Smad4 and translocate to the nucleus.
 pSmads complexes up-regulate other transcription factors, such as LEF-1.
-The pSmad2/4/LEF-1 has been shown to directly repress the E-cadherin gene \cite{Nawshad:2007kl}.
-LEF-1 also binds with $\beta$-catenin to upregulate mesenchymal proteins such as fibronectin \cite{Medici2011}.
+The pSmad2/4/LEF-1 has been shown to directly repress the E-cadherin gene \citep{Nawshad:2007kl}.
+LEF-1 also binds with $\beta$-catenin to upregulate mesenchymal proteins such as fibronectin \citep{Medici2011}.
 The EMT gene expression program was initiated by the binding of TGF$-\beta$ isoforms to TGF$-\beta$ surface receptors.
@@
 downstream signaling program. In the model, TGF$-\beta$1/2 binds TGF$-\beta$R-I/II followed by the recruitment of activin receptor-like kinase 1 (ALK1) and
-TGF$-\beta$ surface receptor III (TGF$-\beta$R-III) to form the activated receptor complex \cite{Derynck:2003fc}.
+TGF$-\beta$ surface receptor III (TGF$-\beta$R-III) to form the activated receptor complex \citep{Derynck:2003fc}.
 Alternatively, we also included activin receptor-like kinase 5 (ALK5) recruitment in combination with Endoglin and TGF$-\beta$R-III
-as a second (redundant) activated receptor complex \cite{Gatza:2010if}.
-Complex assembly activates the serine/threonine kinase activity on the receptor, leading to the recruitment and phosphorylation of Smad partners \cite{Massague:2005qc}.
+as a second (redundant) activated receptor complex \citep{Gatza:2010if}.
+Complex assembly activates the serine/threonine kinase activity on the receptor, leading to the recruitment and phosphorylation of Smad partners \citep{Massague:2005qc}.
 Phosphorylated Smads2/3 (pSmad2/3) form heterodimers with partner Smad4 and then translocate to the nucleus where they act as both transcriptional activators and repressors.
 Nuclear pSmad2/3-Smad4 form transcriptional complexes with several genes in the model including lymphoid enhancer-binding factor 1 (\emph{LEF-1}), Nuclear factor of activated T-cells, cytoplasmic 1 (\emph{NFACT1}), and Specificity Protein 1 (\emph{SP1}).
-On the other hand, nuclear pSmad2/3-Smad4 represses (in combination with the LEF-1 protein) the expression of E-cadherin (\emph{Cdh1}) \cite{Nawshad:2007kl} and Cadherin 5, type 2 (VE-Cadherin encoded by \emph{Cdh5}).
-Repression of E-cadherin expression is the central event in the transition from an epithelial to a mesenchymal phenotype \cite{Cano:2000kh}.
+On the other hand, nuclear pSmad2/3-Smad4 represses (in combination with the LEF-1 protein) the expression of E-cadherin (\emph{Cdh1}) \citep{Nawshad:2007kl} and Cadherin 5, type 2 (VE-Cadherin encoded by \emph{Cdh5}).
+Repression of E-cadherin expression is the central event in the transition from an epithelial to a mesenchymal phenotype \citep{Cano:2000kh}.
 However, this transition is not solely driven by transcriptional events.
@@
 Cystolic $\beta$-catenin then translocates to the nucleus and forms transcriptionally-active complexes with immunoglobulin transcription factor 2 (TCF-4)
-to drive TGF$-\beta$3 expression \cite{Medici:2008fk}.
+to drive TGF$-\beta$3 expression \citep{Medici:2008fk}.
 Simultaneously, ERK1/2-mediated phosphorylation of the AP1 and Sp1 transcription factors can also regulate transcriptional complexes involving NFAT, Slug, and Smads.
@@
 UNIVERSAL, an open source Objective-C/Java code generator, is available as a Google Code project (http://code.google.com/p/universal-code-generator/).
-Model equations were solved using the CVODE solver in the SUNDIALS library \cite{Hindmarsh2005} on an Apple workstation (Apple, Cupertino, CA)
-as previously described \cite{Tasseff:2011ve}.
+Model equations were solved using the CVODE solver in the SUNDIALS library \citep{Hindmarsh2005} on an Apple workstation (Apple, Cupertino, CA)
+as previously described \citep{Tasseff:2011ve}.
 
@@
 Instead, we estimated a population of likely models (each consistent with the training data)
-using 41 data sets generated in DLD1 colon carcinoma, MDCKII, and A375 melanoma cells taken from Medici \emph{et al.} \cite{Medici:2008fk}.
+using 41 data sets generated in DLD1 colon carcinoma, MDCKII, and A375 melanoma cells taken from Medici \emph{et al.} \citep{Medici:2008fk}.
 We used the Pareto Optimal Ensemble Technique (POETs) multiobjective optimization framework in combination
-with leave-one-out cross-validation to estimate an ensemble of model parameters \cite{Song:2010fk}.
-Cross-validation was used to calculate both training and prediction error during the parameter estimation procedure \cite{kohavi1995study}.
+with leave-one-out cross-validation to estimate an ensemble of model parameters \citep{Song:2010fk}.
+Cross-validation was used to calculate both training and prediction error during the parameter estimation procedure \citep{kohavi1995study}.
 The 41 intracellular protein and mRNA data-sets used for identification were organized into 11 objective functions.
@@
 
-\subsubsection*{Robustness coefficients.} Robustness coefficients were calculated as shown previously \cite{Tasseff:2011ve,Lequieu:2011fj}.
+\subsubsection*{Robustness coefficients.} Robustness coefficients were calculated as shown previously \citep{Tasseff:2011ve,Lequieu:2011fj}.
 Robustness coefficients denoted by $\alpha\left(i,j,t_{o},t_{f}\right)$:
@@
 Primers were designed to detect GAPDH, E-cadherin, vimentin, Slug, Sp1, and NFATc1 in cDNA clones: Sp1 (F-TTG AAA AAG GAG TTG GTG GC, R-TGC TGG TTC TGT AAG TTG GG, Accession NG030361.1), NFATc1 (F-GCA TCA CAG GGA AGA CCG TGT C, R-GAA GTT CAA TGT CGG AGT TTC TGA G, Accession  NG029226.1).
-GAPDH, E-cadherin, vimentin, and Slug primers were taken from previously published literature \cite{Medici:2008fk}.
+GAPDH, E-cadherin, vimentin, and Slug primers were taken from previously published literature \citep{Medici:2008fk}.
 
@@
 characteristics of nontumorigenic MCF10A epithelial acini occur over time and exploiting them to growth in 3D culture can be quantified.
-For example, using image segmentation, Chang et al. \cite{ChangPP07} examined the elongation of the MCF10A acini at 6, 12, and 96 hours after a particular treatment.
-Polizzotti et al. \cite{Polizzotti} also suggested a computational method to quantify acini structure based on morphological characteristics in nonmalignant, noninvasive, and invasive conditions.
+For example, using image segmentation, Chang et al. \citep{ChangPP07} examined the elongation of the MCF10A acini at 6, 12, and 96 hours after a particular treatment.
+Polizzotti et al. \citep{Polizzotti} also suggested a computational method to quantify acini structure based on morphological characteristics in nonmalignant, noninvasive, and invasive conditions.
 Adapted from these approaches, we first fluorescently labeled our cultures and captured the acini structures by 3D confocal microscopy.
@@
 \begin{figure}
-%\includegraphics [width=1.0\linewidth] {./figs/Fig-1-Network.pdf}
+\includegraphics [width=1.0\linewidth] {./figs/Fig-1-Network.pdf}
 \caption{Model connectivity recreates the core architecture during EMT. The EMT network contains 995 nodes (proteins, mRNA, and genes) interconnected by 1700 interactions.
@@
 \begin{figure}
-%\includegraphics [width=1.0\linewidth] {./figs/Fig-2-Training.pdf}
+\includegraphics [width=1.0\linewidth] {./figs/Fig-2-Training.pdf}
 \caption{Training and validation simulations. The population of EMT models qualitatively captured TGF$\beta$-induced EMT signaling.
-(A-I) The population was generated using POETs and trained using 11 different objective functions (41 data sets) taken from Medici \emph{et al.} \cite{Medici:2008fk}.
+(A-I) The population was generated using POETs and trained using 11 different objective functions (41 data sets) taken from Medici \emph{et al.} \citep{Medici:2008fk}.
 The model captured the simulated experiments for 78\% of the cases.
@@
 \begin{figure}
-%\includegraphics [width=1.0\linewidth] {./figs/Fig-3-HypotheticalPhenotype.pdf}
+\includegraphics [width=1.0\linewidth] {./figs/Fig-3-HypotheticalPhenotype.pdf}
 \caption{Simulated TGF$\beta$1/2 exposure promoted phenotype heterogeneity.
@@
 \begin{figure}
-%\includegraphics [width=1.0\linewidth] {./figs/Fig-4-Robustness.pdf}
+\includegraphics [width=1.0\linewidth] {./figs/Fig-4-Robustness.pdf}
 \caption{Simulated response to TGF$\beta$1/2 and VEGF-A exposure with and without axis specific inhibitors.
@@
 \begin{figure}
-%\includegraphics [width=1.0\linewidth] {./figs/Fig-5-MCF10A-Validation.pdf}
+\includegraphics [width=1.0\linewidth] {./figs/Fig-5-MCF10A-Validation.pdf}
 \caption{Simultaneous TGF-$\beta$1/2 and VEGF-A treatment induced phenotype heterogeneity and is dependent upon NFAT activity \emph{in-vitro}.
@@
 \begin{figure}
-%\includegraphics[width=0.5\linewidth] {./figs/Fig-6-Phenotyping.pdf}
+\includegraphics[width=0.5\linewidth] {./figs/Fig-6-Phenotyping.pdf}
 \caption{Ductal branching is dependent upon phenotype heterogeneity within MCF10A in 3-D culture.

### 2016-09-28 Supplemental-JB.tex
First commit

\section*{Supplemental Materials and Methods}
\subsection*{EMT model network architecture.} TGF$\beta$ is a major inducer of EMT in development, fibrosis, and carcinogenesis with different isoforms mediating various effects depending on specific cellular context \cite{Nawshad:2005pi}.
TGF$\beta$1 was first described as an inducer of EMT in normal mammary epithelial cells \cite{Miettinen1994} and has since been shown to mediate EMT in vitro in a number of different epithelial cells, including renal proximal tubular, DLD1 colon carcinoma, and most recently alveolar epithelial cells \cite{Fan1999, Hales1994,Kasai2005,Willis2005}.
TGF$\beta$ signaling occurs through the Smad pathway in which signals are transduced by transmembrane serine/threonine kinase type I (ALK5) and type II (TGF$\beta$RII) receptors.
To increase ligand affinity, betaglycan (TGF$\beta$RIII) can also interact with TGF$\beta$RI,II \cite{Gatza:2010if}.
Upon TGF$\beta$ stimulation, the receptors are internalized into early endosomes where Smad anchor for receptor activation (SARA) modulates formation of complexes with (R-Smad) Smad2 or Smad3.
Smad2 and Smad3 are then phosphorylated at serine residues by the type I receptor \cite{Massague:2005qc}.
Phosphorylation induces their association with (Co-Smad) Smad4 and translocation to the nucleus where they interact with other transcription factors to regulate the transcription of TGF$\beta$ responsive genes, including alpha-smooth muscle actin, collagen1A2, vimentin, fibronectin, and plasminogen activator inhibitor-1 (PAI-1) by interacting with Smad-binding elements \cite{Dennler:2002hl,Derynck:2003fc}.
To regulate TGF$\beta$ signaling, Smurf2 (a ubiquitin E3 ligase) can become activated to mediate proteasome dependent degradation of Smad2 or bind with Smad 7 to target TGF$\beta$ receptor for degradation \cite{Kavsak2000,Bonni2001}.

\subsubsection*{Cell Type Dependency} Interestingly, differential roles for Smad2 and Smad3 in TGF$\beta$ induced EMT have been demonstrated.
For example, using primary cells from mice with hepatocyte-specific double knockout of Smad2 and Smad3, it was demonstrated that Smad3 but not Smad2 was required for a key morphological changes and induction of EMT \cite{Ju:2006so}.
In contrast, using human proximal tubular epithelial cells, increased colony growth factor and decreased E-cadherin were Smad3 dependent, increased MMP-2 was Smad2 dependent, while alpha-smooth muscle actin was dependent on both \cite{Phanish2006}.
Furthermore, Smad2 signaling has been demonstrated in cancerous lung epithelial A549 cells \cite{Kasai2005}.
Inhibition of Smad3 signaling decreases the metastatic potential of xenografted breast cancer cell lines \cite{Tian2003,Tian2004}.
Together, these results suggest that the precise Smad pathway activated may depend on the particular cellular context.
Regardless, a recent transcriptomic analysis of TGF$\beta$ induced EMT in normal mouse and human epithelial cells using a dominant negative approach demonstrated that Smad signaling was critical for regulation of all tested target genes \cite{Valcourt2005}.

\subsubsection*{Smad-Independent EMT Induction Pathways. }Although less well established than the Smad-dependent pathways in the induction of EMT, there is substantial evidence for TGF$\beta$ activation of Smad-independent signaling in some aspects of this process.
These include RhoA, MAPK, PI3 kinase, and Notch signaling pathway, which have mostly been studied in-vitro.
For example, the small GTPase RhoA is involved in TGF$\beta$ induced EMT in a number of cell types including NMuMG mammary epithelial cells and mink lung epithelial (Mv1Lu) cells particularly in the regulation of cytoskeletal and adherens junction rearrangement.
In addition to its role in cytoskeletal remodeling, Rho has been shown to activate the alpha-smooth muscle promoter during EMT in kidney proximal tubular (LLCPK1) cells \cite{Masszi:2003hb}.
TGF$\beta$ has been shown to elevate ERK activity in culture models of EMT (human keratinocytes, NMuMG mammary epithelial cells, and mouse cortical tubule epithelial cells) \cite{Zavadil2001,Xie2004,Davies2005}.
This ERK activity was required for disassembly of adherens junctions and induction of cell motility \cite{Xie2004}.
Inhibition of ERK blocked these key morphological changes of EMT in mammary gland epithelial cells \cite{Xie2003}.
Recent studies in MDCKII and DLD1 colon carcinoma cells revealed that TGF$\beta$3 can signal through PI3 kinase.
Upon activation, PI3 kinase activates molecules such as ILK and AKT, which can phosphorylate and inactivate GSK3$\beta$, a protein that targets both Snail and $\beta$-catenin for degradation through the ubiquitin proteosome pathway \cite{Medici:2008fk}.
Lastly, there is increasing evidence for a role of Notch pathways in regulating EMT, particularly during development \cite{Nawshad:2005pi}.
In cardiac cushion EMT, it was found that that Slug is directly up-regulated by Notch in endothelial cells repression of the vascular endothelial cadherin promoter and for promoting migration of transformed endothelial cells.
In whole, the distinction between Smad-dependent and Smad-independent mechanisms remains difficult to interpret due to the significant cross talk between these pathways.
In most cases, stimulation of these cooperative pathways provides the context for induction and specification of EMT within a particular tissue/cell type, with Smads representing the dominant pathway, which in some instances may be necessary but not sufficient for induction of full EMT \cite{Zavadil:2005fu}.

\subsubsection*{Master Transcription Regulators}
Master transcription factors of EMT are thought to be regulated by Snail1 (SNAI1) and Snail2 (SNAI2) (previously known as Snail and Slug, respectively).
These transcription factors are zinc finger proteins that function as repressors of E-cadherin transcription in cultured epithelial cells \cite{Cano:2000kh}, repression of E-cadherin leads to dissolution of adherens junctions.
Snail and Slug can be activated by TGF$\beta$ via both Smad-dependent and -independent pathways in a cell type dependent fashion in cultured cells \cite{Peinado2003}.
Differential expression of Snail and Slug is observed in TGF$\beta$ induced EMT in keratinocytes, renal proximal tubular, and mammary epithelial cells, suggesting that they are regulated in a cell specific.
Interestingly, key regulatory units of Snail and Slug appear to be regulated by upstream transcription factors AP1 and SP1, respectively.
Peinado et al. clearly established a necessary role of Snail in mediating TGF$\beta$1 induced EMT in MDCK cells.
They demonstrated that the Ras-Raf-MEK-ERK-AP1 signaling pathway could up-regulate synthesis of the E-cadherin repressor molecule Snail (directly binds to Snail promoter) \cite{Peinado:2004pt}.
Likewise, TGF$\beta$1 induced the expression of Slug in both lens and other epithelial cells in vitro.
The Sp1 binding site in the Slug promoter is largely responsible for TGF$\beta$1 induced Slug expression and upstream of MAPK signaling.
In addition, the TGF$\beta$1 mediated repression of E-cadherin was significantly inhibited by Slug siRNA \cite{Choi2007}.

\subsubsection*{ERK/MAPK Crosstalk}
The ERK-MAPK pathway has been shown to modify TGF$\beta$ signaling at multiple levels.
One way is that the activated Ras pathway inhibits the TGF$\beta$ induced nuclear accumulation of Smad2/3, as seen in epithelial cells \cite{Zhang2009a}.
ERK kinases have been shown to phosphorylate Smad2 and Smad3 at specific sites in the region linking the MH1 and MH2 domains.
These sites are separate from the TGF$\beta$ receptor phosphorylation sites which activate Smad nuclear translocation \cite{Massague2003}.
The effect of interaction between ERKs and Smads is the subject of some controversy, with data suggesting that such an interaction either enhances or inhibits downstream events.
A hyperactive Ras pathway has been shown to effectively counteract the antiproliferative activity of TGF$\beta$ through attenuation of Smad accumulation in the nucleus \cite{Kretzschmar:1999wb}.
In contrast, Ras signals strongly cooperated with Smads for invasion of human carcinoma cells \cite{Oft1996}.
A second way is through transcriptional regulation.
Recently, it was found that Sp1, via transcriptional induction of Vimentin, cooperates with activated Smad complexes in mesenchymal transition and migration of pancreatic cancer cells upon TGF$\beta$ stimulation \cite{Jungert:2007ri}.
Likewise, binding sites at the SP1/Smad3 complex was found to regulate the betaglycan receptors \cite{Lopez-Casillas2003}.
Taken together, multiple levels of cross-talk (both positive and negative) exist within MAPK and TGF$\beta$, and may also be cell type dependent.

\subsection*{Analysis of the signal flow through the EMT architecture.}
To investigate temporal shifts for key species dominating the EMT response, we calculated the scaled flux through the signaling architecture (Fig. \ref{fg:S1}).
Three modes of operation were simulated to identify distinct behavioral differences: (a) no TGF-$\beta$2 (-TGF-$\beta$1/2), (b) TGF-$\beta$1/2 stimulation (+TGF-$\beta$2),
and (c) TGF-$\beta$1/2 stimulation while blocking the autocrine response of TGF-$\beta$3 (+TGF-$\beta$1/2 + TGF-$\beta$3Ab).
Following TGF-$\beta$1/2 stimulation, AP1/SP1 rapidly stimulates Snail/Slug expression (Fig. \ref{fg:S1}A) within 1 hr.
During this phase, Snail/Slug repress E-cadherin expression, which in turn reduces the interaction between E-cadherin and $\beta$-catenin, and the accumulation of free cytosolic $\beta$-catenin ensues.
Conversely, in the presences of a TGF-$\beta$3 neutralizing antibody, membrane bound $\beta$-catenin increased after 10 hrs,
while low levels of free cytosolic $\beta$-catenin accumulated between 1-5 hrs (Fig. \ref{fg:S2}A,B).  Accumulation of free $\beta$-catenin is critical for complexing with TCF4 to rapidly produce the autocrine response of TGF-$\beta$3 within 1-10 hrs.  Following TGF-$\beta$3 expression, formation and spatial relocation of the Smad complexes occurs in a time dependent manner.  Smad2 is phosphorylated within 1 hr and nuclear localization of the pSmad2/4 complex dramatically increases after 10 hours (Fig. \ref{fg:S1}C).  This timely formation is critical for completion of the EMT process.  At steady state, the E-cadherin complex was maintained at a basal epithelial level with no expression of Vimentin (Fig. \ref{fg:S1}D). During TGF-$\beta$1/2 stimulation, both MAPK and Smad act synergistically to repress the E-cadherin complex within 1 hr.
This is followed by increase of Vimentin at $\sim$ 1hr, while exponentially increasing around 10 hrs.
Both species elevate within 1 hr, however the complex does not significantly form until $\sim$ 10 hrs.
When blocking TGF-$\beta$3, Snail/Slug downregulates the E-cadherin complex between 5-10 hrs,
but cannot complete the transformation due to the loss of Smad2 phosphorylation (Fig. \ref{fg:S1}D).

\begin{figure}
\includegraphics [width=1.0\linewidth] {./figs/Fig-1-Supplemental-FluxAnalysis.pdf}
\caption{Signal flow analysis of key species at steady state, TGF-$\beta$1/2 stimulation, and blocking the TGF-$\beta$3 autocrine response.
(A)  The MAPK cascade is directly responsible for rapid expression of Snail and downstream TGF-$\beta$3 formation (1~hrs).
(B) TGF-$\beta$1/2 reduces  $\beta$-catenin, allowing rapid free-cytosolic $\beta$-catenin to accumulate (1hr).
Blocking TGF$\beta$3 increases membrane bound $\beta$-catenin (10hr). (C)  TGF-$\beta$3 activates the Smad cascade.
Nuclear localization of the pSmad2/4 complex (10~hrs) is dependent upon both the phosphorylation of Smad2 (1~hrs) and complexing with Smad4 (5~hrs).
(D)  TGF-$\beta$1/2 rapidly reduces the E-cadherin complex, while upregulating Vimentin (5-10 hrs).
Blocking TGF$\beta$3 increases E-cadherin (10~hrs) and Vimentin is significantly reduced.}\label{fg:S1}
\end{figure}

\begin{figure}
\includegraphics [width=0.5\linewidth] {./figs/Fig-2-Supplemental-Robustness.pdf}
%\includegraphics [width=1.0\linewidth] {./ManuFigures/Figs/F2.pdf}
%\epsfig{file=./figures/Figures_EMT_Manuscript/Final/ManuFigures/F2.eps,,width=1.0\textwidth}
\caption{Robustness analysis for key molecular species at t = 48 hrs for combinations of TGF-$\beta$1/2, VEGF-A and NFATc1 inhibitors. Robustness coefficients for the indicated species were calculated
for N $\sim{1100}$ ensemble members for 48 hrs following the addition of TGF-$\beta$1/2 (T), TGF-$\beta$1/2 + NFATc1 inhibitor (T + I), VEGF-A (V), VEGF-A + NFATc1 inhibitor (V + I) and TGF-$\beta$1/2 + VEGF-A (T+V) + NFATc1 inhibitor (T + V + I).
(A) Robustness coefficients for E-cadherin and Vimentin as a function of condition.
(B) Robustness coefficients for nuclear localized phosphorylated Sp1 and NFATc1 as a function of condition.
(C) Robustness coefficients for nuclear localized Slug and $\alpha$-smooth muscle actin ($\alpha$-SMA) as a function of condition.
In each case the error bars denote one-standard deviation of robustness coefficient calculated over the model ensemble.
C=Control, T=TGF$\beta$2 , V=VEGFA, VI= NFAT inhibitor (VIVIT).}\label{fg:S2}
\end{figure}

\begin{figure}
\includegraphics [width=1.0\linewidth] {./figs/Fig-3-Supplemental-DLD1.pdf}
\caption{VEGF-A attenuates TGF-$\beta$1/2 to induce phenotype heterogeneity in DLD1.
(A) In DLD1, we found that 5ng/ml of VEGFA increased NFATc1 and E-cadherin gene expression via qPCR and 50ng/ml potentiated this effect at 48 hrs.
(B - C) These findings were confirmed at the protein level via immunofluorescence, as ecadherin levels and nuclear localization of NFATc1 increased.
(D) Treatment with (10ng/ml) TGF$\beta$2 resulted in mesenchymal transformation as measured via qPCR against target genes Slug, ecadherin, vimentin, Sp1, and NFATc1.
(E - F) Immunofluorescence and nuclear localization revealed a strong presence of phospho-Sp1.
(G) Combination of VEGFA (50ng/ml) and TGF$\beta$2 (10ng/ml) treatment resulted in increased Slug, NFATc1, and vimentin expression, while also increasing ecadherin levels compared to control.
(H) Immunofluorescence confirmed these results, as both ecadherin and vimentin levels were elevated.
(I) A significant increase in nuclear localization of both NFATc1 and phospho-Sp1 were also found.
Magnification, 40x. Scale bars: 50$\mu$m.  C=Control, T=TGF$\beta$2 , V=VEGFA, VI=NFAT inhibitor (VIVIT).
Asterisks signify statistical differences from each other according to a one-way ANOVA with Tukey's post hoc (p$\prec$0.05).}\label{fg:S3}
\end{figure}

\begin{figure}
\includegraphics [width=1.0\linewidth] {./figs/Fig-4-Supplemental-DLD1-FlowCyto.pdf}
\caption{E-cadherin expression is dependent upon NFAT activity in DLD1.
(A) Treatment with VEGFA (50ng/ml) and NFAT inhibitory peptide VIVIT (10$\mu$M) resulted in significantly reduced ecadherin expression (qRT-PCR at 48hrs).
Addition of TGF$\beta$2 (10ng/ml) and VIVIT resulted in increased Slug and vimentin expression, while inhibiting ecadherin levels.
Combined TGF$\beta$2, VEGFA, and VIVIT treatment resulted in target genes Slug and vimentin expression increased, while inhibiting ecadherin levels.
No change in Sp1 or NFATc1 expression was found.
(B) These findings were confirmed via immunofluorescence as the VIVIT inhibitors was shown to inhibit ecadherin levels in all three cases.
We also found no change in gene or nuclear localization of NFATc1 in all three cases, while phospho-Sp1 was found to increase in both TGF$\beta$ conditions.
(C) Quantitative flow cytometry also confirmed this trend.
(D,E)  TGF$\beta$2, VEGFA and VIVIT treatment in DLD1 and MCF10A resulted in no change of Sp1 expression or NFATc1 expression.
(F)  Likewise, no change in nuclear localization of NFAT in all three cases, however phospho-Sp1 was found to increase in both TGF$\beta$ conditions.
Magnification, 40x. Scale bars: 50$\mu$m.  C=Control, T=TGF$\beta$2 , V=VEGFA, VI= NFAT inhibitor (VIVIT).
Asterisks signify statistical differences from each other according to a one-way ANOVA with Tukey's post hoc (p$\prec$0.05).}\label{fg:S4}
\end{figure}

\newpage

\subsection*{Estimation and cross-validation of EMT model parameters.}
We used the Pareto Optimal Ensemble Technique (POETs) multiobjective optimization framework in combination with leave-one-out cross-validation to estimate an ensemble of TGF$\beta$/EMT models.
Cross-validation was used to calculate both training and prediction error during the parameter estimation procedure \cite{kohavi1995study}.
The 41 intracellular protein and mRNA data-sets used for identification were organized into 11 objective functions.
These 11 objective functions were then partitioned, where each partition contained ten training objectives and one validation objective.
POETs integrates standard search strategies e.g., Simulated Annealing (SA) or Pattern Search (PS)
with a Pareto-rank fitness assignment \cite{Song:2010fk}.
Denote a candidate parameter set at iteration $i+1$ as $\mathbf{k}_{i+1}$.
The squared error for $\mathbf{k}_{i+1}$ for training set $j$ was defined as:
\begin{equation}\label{eqn_cost2}
	E_{j}(\mathbf{k}) = \sum_{i=1}^{\mathcal{T}_{j}}\biggl(\hat{\mathcal{M}}_{ij}-\hat{y}_{ij}(\mathbf{k})\biggr)^2
\end{equation}
The symbol $\hat{\mathcal{M}}_{ij}$ denotes scaled experimental observations (from training set $j$) while $\hat{y}_{ij}$ denotes the scaled simulation output (from training set $j$).
The quantity $i$ denotes the sampled time-index and $\mathcal{T}_{j}$ denotes the number of time points for experiment $j$.
In this study, the experimental data used for model training was typically the band intensity from Western or Northern blots.
Band intensity was estimated using the ImageJ software package \cite{IMAGEJ}.
The scaled measurement for species $x$ at time $i=\{t_{1},t_{2},..,t_{n}\}$ in condition $j$ is given by:
\begin{equation}\label{norm_exp_data}
\hat{\mathcal{M}}_{ij} = \frac{\mathcal{M}_{ij} - \min_{i}\mathcal{M}_{ij}}{\max_{i}{\mathcal{M}_{ij}}-\min_{i}{\mathcal{M}_{ij}}}
\end{equation}
Under this scaling, the lowest intensity band equaled zero while the highest intensity band equaled one.
A similar scaling was defined for the simulation output. By doing this scaling, we trained the model on the relative change in blot intensity, over conditions or time (depending upon the experiment).
Thus, when using multiple data sets (possibly from different sources) that were qualitatively similar but quantitatively different e.g., slightly different blot intensities over time or condition, we captured the underlying trends in the scaled data. Additionally, we applied a universal convention of identifying the no expression case as protein value below $<$10$^{-3}$. This is similar to previously published models from our lab. In our studies, Figure 2 identifies experimental data extracted from published Western blots and our simulation results. It is clear from these that 1) the training data included a variety of Western blot data treatments that were effectively zero, and 2) our simulations matched the training data virtually perfectly over time and across multiple biological species. These results validate the power of our simulation scheme, which necessarily includes numerical interpretations of zero. In our Supplemental Figure S9, we present the raw concentrations of our simulations (previously validated from the training data) over hundreds of parameter ensembles, with no “zero” threshold applied.  While these values do not have a threshold, we only interpret values below 10$^{-3}$ equivalent to zero. Likewise, our robustness coefficients (Figure 3) (which represent the ratio of integrated areas of the treatment effect over the baseline effect) identify no difference in model output for vimentin (or e-cadherin for that matter) less than 10$^{-3}$, confirming this interpretation.

We computed the Pareto rank of $\mathbf{k}_{i+1}$ by comparing the simulation error at iteration $i+1$ against an archive of accepted parameter sets $\mathbf{K}_{i}$.
We used the Fonseca and Fleming ranking scheme \cite{fonseca1993genetic} to estimate the number of parameter sets in the archive that dominate
$\mathbf{k}_{i+1}$. Parameter sets with increasing rank were progressively further away from the optimal trade-off surface.
The parameter set $\mathbf{k}_{i+1}$ was accepted or rejected by POETs with probability $\mathcal{P}\left(\mathbf{k}_{i+1}\right)$:
\begin{equation}\label{eqn_costMOSA}
\mathcal{P}(\mathbf{k}_{i+1}) \equiv \exp{\{-rank\left(\mathbf{k}_{i+1} \mid \mathbf{K}_{i} \right)/T\}}
\end{equation}
where $T$ is the annealing temperature and $rank\left(\mathbf{k}_{i+1} \mid \mathbf{K}_{i} \right)$ denotes the Pareto rank for $\mathbf{k}_{i+1}$.
The annealing temperature was discretized into 10 quanta between $T_{o}$ and $T_{f}$ and adjusted according to the schedule
$T_k = \beta^k T_0$ where $\beta$ was defined as $\left({T_{f}}/{T_{o}}\right)^{1/10}$.
The initial temperature was $T_{o} = \mathit{n}/log(2)$, where $n = 4$ in this study and the final temperature was $T_{f} = 0.1$.
The epoch-counter $k$ was incremented after the addition of 100 members to the ensemble.
Thus, as the ensemble grew, the likelihood of accepting parameter sets with a large Pareto rank decreased.
To generate parameter diversity, we randomly perturbed each parameter by $\leq\pm25\%$ at iteration of the search.
In addition, we performed a local pattern search every $q$-iterations to minimize the residual for a single random or the worst performing objective function.
The local pattern-search algorithm has been described previously \cite{gadkar2003cybernetic}.
From the 15,000 probable EMT models, we selected N = 1093 models with Pareto rank $\leq{1}$ for subsequent analysis.
A quick estimate of the set to set correlation showed that we could expect on order 25\% correlation between parameter sets in the ensemble.

\begin{figure}
\includegraphics [width=1.0\linewidth] {./figs/Fig-5-Supplemental-ErrorTable.pdf}
\caption{Training and prediction values for the 11 TGF-$\beta$ objective functions versus a random parameter control.}\label{fig:S5}
\end{figure}

\begin{figure}
\includegraphics [width=1.0\linewidth] {./figs/Fig-6-Supplemental-CompeteErrorTable.pdf}
\caption{Training and prediction values as a function of condition for the 11 TGF-$\beta$ objective functions versus a random parameter control.}\label{fig:S6}
\end{figure}

\begin{figure}
\includegraphics [width=1.0\linewidth] {./figs/Fig-7-Supplemental-Correlation.pdf}
\caption{Parameter set to set correlation for 55 random parameter sets selected from the ensemble.
Of the 55 sets selected, the average correlation between sets was less than 25\% for greater than 80\% of the parameter sets.}\label{fig:S7}
\end{figure}

\begin{figure}
\includegraphics [width=1.0\linewidth] {./figs/Fig-8-Supplemental-YREG1.pdf}
\caption{Robustness of E-cadherin and Vimentin expression to a knockout (A) and overexpression (B) of the hypothetical regulator 1 (YREG1) protein.
Robustness coefficients were calculated for each member of the ensemble. Each point represents the response of a single model in the ensemble to either a knockout or
overexpression of YREG1.}\label{fig:S8}
\end{figure}

\begin{figure}
\includegraphics [width=1.0\linewidth] {./figs/Fig-9-Supplemental-RawExpression_v2.pdf}
\caption{Steady state protein abundance for E-cadherin and Vimentin (AU) as a function of TGF-$\beta$1/2 exposure.
Top: Overlay of the model population for Vimentin (AU) and E-cadherin (AU) expression in the presence (blue) and absence (gray) of TGF-$\beta$1/2.
Midddle: Vimentin (AU) and E-cadherin (AU) expression in the absence of TGF-$\beta$1/2 showed exhinibted population heterogenity.
Bottom: Vimentin (AU) and E-cadherin (AU) expression in the presence of TGF-$\beta$1/2 moved the centriod of the population toward Vimentin (AU) and away from E-cadherin (AU) expression.}\label{fig:S9}
\end{figure}

%\bibliography{References_v1}

### 2016-10-03 Paper_v7.tex
update edits DB1

@@
 \vspace{0.05in}
-{\par \centering Russell Gould$^{\ddag}$, Anirikh Chakrabarti, Jeffrey D. Varner and Jonathan Butcher$^{\ddag *}$}
+{\par \centering Russell Gould$^{\ddag}$ and David M. Bassen, Anirikh Chakrabarti, Jeffrey D. Varner and Jonathan Butcher$^{\ddag *}$}
 \vspace{0.05in}
@@
 In this study, we developed a family of mechanistic models describing the induction of EMT by TGF-$\beta$ isoforms in the presence and absence of vascular endothelial growth factor A (VEGF-A).
-We incorporated mass action kinetics within an ordinary differential equation (ODE) framework to describe the EMT interaction network containing 995 gene, protein or mRNA components interconnected through 1700 interactions.
+We incorporated mass action kinetics within an ordinary differential equation (ODE) framework to describe the EMT interaction network containing 97 gene, protein or mRNA components interconnected through 169 interactions.
 A family of model parameters was estimated using 41 molecular data sets generated in DLD1 colon carcinoma, MDCKII and A375 melanoma cells using the Pareto optimal ensemble technique (POETs) multiobjective optimization algorithm.
-POETs identified more than 15,000 likely TGF-$\beta$ induced EMT models, from which we selected approximately 1100 models for further analysis.
+POETs generated an ensemble of approximately 1400 models for analysis.
 Analysis of the model population suggested that both MCF10A and DLD1 cells could exhibit phenotypic heterogeneity if treated simultaneously with TGF-$\beta$1/2 and VEGF-A.
@@
 The EMT model architecture, based upon curated molecular connectivity,
-described the expression of 80 genes following exposure to TGF-$\beta$ isoforms and VEGF-A (Fig. \ref{fg:F1}).
-The EMT model contained 995 molecular species interconnected by 1700 interactions.
+described the expression of 23 genes following exposure to TGF-$\beta$ isoforms and VEGF-A (Fig. \ref{fg:F1}).
+The EMT model contained 74 molecular species interconnected by 169 interactions.
 Model equations were formulated using mass-action kinetics within an ordinary differential equation (ODE) framework.
@@
 they require estimates for model parameters which are often difficult to obtain.
-The EMT model had 1756 unknown model parameters, 1700 kinetic constants and 56 non-zero initial conditions.
+The EMT model had 251 unknown model parameters, 169 kinetic constants and 45 non-zero initial conditions.
 As expected, these parameters were not uniquely identifiable given the training data \citep{Gadkar:2005vn}.
@@
 POETs generated a population of probable signaling models which captured the multiple phases of EMT induction (Fig. \ref{fg:F2}).
-POETs sampled well over 10$^6$ probable models during each stage of the cross-validation, using a combination of both local and global random sampling.
-This sampling generated approximately 15,000 highly probable models from which we selected N $\simeq$ 1100 models for further analysis.
-The selected models all had the same possible molecular connectivity, but different values for model parameters and extrinsic factors such as RNA polymerase or ribosome abundance.
+POETs sampled well over 10$^4$ probable models during each stage of the cross-validation #####, using a combination of both local and global random sampling.
+From this analysis, N $\simeq$ 1400 models were selected for further analysis.
+The selected models all had the same possible molecular connectivity, but different values for model parameters.
+Trascription and translation rates, as well as mRNA and protein degradation terms, were set using physical values from the literature, and allowed to vary by a scaling factor, see methods ####citebionumbers####.
 Model selection was based upon Pareto rank, the prediction and training error across all objectives and model to model correlation (supplemental materials).
@@
 Conversely, over-expression of E-cadherin inhibited the TGF$-\beta$3 autocrine production by sequestering cytosolic $\beta$-catenin, thereby blocking EMT (Fig. \ref{fg:F2}C, Lane 4 or 5).
-TGF$-\beta$3 signaled through through the Smad pathway to regulate LEF-1 expression and downstream target EMT genes (Fig. \ref{fg:F2}G).
+TGF$-\beta$3 signaled through the Smad pathway to regulate LEF-1 expression and downstream target EMT genes (Fig. \ref{fg:F2}G).
 TGF$-\beta$3 (10 a.u.) in combination with downstream inhibitors (DN-Smad4 and DN-LEF-1) completely inhibited Vimentin expression, while elevating E-cadherin expression (Fig. \ref{fg:F2}H,I).
@@
 The predictive power of the ensemble was tested using both cross validation and by comparing simulations with data sets not used for model training.
-In whole, 78\% of our training objectives were statistically significant (at a 95\% confidence interval) compared to a randomized parameter family (N = 100)
-generated from the best-fit nominal set (starting point for the optimization).
-Conversely, we \emph{predicted} approximately 60\% of the training objectives, at a 95\% confidence interval compared to randomized parameters.
+In whole, 72\% of our training objectives were statistically significant (at a 95\% confidence interval) compared to a randomized parameter family (N = 100)
+generated from a random starting point.
+Conversely, we \emph{predicted} approximately 72\% of the training objectives, at a 95\% confidence interval compared to randomized parameters.
 The model also captured the temporal gene expression responses of E-cadherin, pSmad2, and LEF-1 to within one-standard deviation (up to the 48 hr time-point) (Fig. \ref{fg:F2}J-L).
@@
 Medici et al. suggested a similar idea where feedback between $\beta$-catenin and LEF-1 was likely, although this feedback had yet to be identified \citep{Medici:2008fk}.
-Low levels of YREG1 expression were used in all simulations to regulate the formation of the $\beta$-catenin-LEF-1 complex.
+Low levels of YREG1 expression were present in all simulations to regulate the formation of the $\beta$-catenin-LEF-1 complex.
 To test the potency of YREG1, we conducted knockdown and over-expression simulations following the addition of TGF-$\beta$1/2 (Fig. \ref{fig:S8}).
@@
 thereby not allowing free $\beta$-catenin to form the $\beta$-catenin-LEF-1 complex which promotes mesenchymal gene expression, or SMAD activity following from autocrine TGF-$\beta$3 signaling.
-Likewise, the abundance of the pSmad2/4-LEF-1 complex was also reduced in cells overexpression YREG1, which blocked the repression of E-cadherin.
 Taken together, we found that low YREG1 expression was necessary for stabilizing EMT,
@@
 We tested this hypothesis by analyzing the response of the population of EMT models to extracellular cues and then comparing this response to flow cytometry studies.
-We used robustness coefficients to quantify the response of the individual members of the ensemble to TGF$-\beta$1/2 stimulation.
-We have previously used robustness coefficients to systematically quantify response of a system to structural or operational perturbations,
-for example gene deletions or the addition of a growth factor or hormone  \citep{Song:2010fk,Tasseff:2010pi,Lequieu:2011fj,Tasseff:2011ve}.
-Robustness coefficients quantify shifts in molecular marker abundance resulting from molecular of environmental perturbations relative to an unperturbed control state.
-Robustness coefficients $\gg$ 1 indicate that marker abundance increased, while robustness coefficients $\ll$ 1 indicates marker abundance decreased relative to an unperturbed control.
-A value of $\sim$ 1 indicates approximately no change in marker abundance following the perturbation.
-We calculated robustness coefficients for each member of the ensemble (N $\simeq$ 1100) for two downstream phenotypic markers, Vimentin (mesenchymal) and E-cadherin (epithelial)
+We quantiied the phenotypic response of the individual members of the ensemble to TGF$-\beta$1/2 stimulation as a function of physical measurements of abundance of markers in system.
+We evlautated these quantities for each member of the ensemble (N $\simeq$ 1400) for two downstream phenotypic markers, Vimentin (mesenchymal) and E-cadherin (epithelial)
 following the addition of TGF$-\beta$1/2 alone (Fig. \ref{fg:F3}), and VEGF-A in combination with NFATc inhibitors (Fig. \ref{fg:F4}).
-The absence of TGF$-\beta$1/2 or VEGF-A stimulation was used as the baseline for the robustness calculations.
+We were further able to use this analysis to examine the activity and effects of underlying signaling molecules.
 
@@
 
-Our robustness analysis predicted that NFATc transcriptional activity was critical to maintaining E-cadherin expression in the presence of both VEGF-A and TGF$-\beta$1/2.
+Our phenotypic analysis predicted that NFATc transcriptional activity was critical to maintaining E-cadherin expression in the presence of both VEGF-A and TGF$-\beta$1/2.
 We experimentally tested this hypothesis by exposing both MCF10A (Fig.
@@
 In this study, we developed a family of mechanistic models describing the induction of EMT by TGF-$\beta$ isoforms in the presence and absence of VEGF-A.
-The signaling architecture encoded in the model, which contained 995 molecular species interconnected by 1700 interactions,
-described the expression of 80 genes in response to growth factor stimulation.
+The signaling architecture encoded in the model, which contained 74 molecular species interconnected by 169 interactions,
+described the expression of 23 genes in response to growth factor stimulation.
 This simulation incorporates an unprecedented level of detail compared to previous models, but as a consequence created a large number of unknown model parameters.
@@
 The model population was trained and cross-validated to prescribe biological significance using 41 data sets generated in DLD1 colon carcinoma, MDCKII, and A375 melanoma cell lines \citep{Medici:2008fk}.
-POETs generated $>$ 15,000 probable parameter sets using this data, from which we selected N $\simeq$ 1100 for subsequent analysis.
+The absence of TGF$-\beta$1/2 or VEGF-A stimulation was used as the baseline for the robustness calculations.
 Analysis of this population predicted possible phenotypic modes (and their associated signaling) that cells could exhibit when stimulated with TGF-$\beta$ and/or VEGF-A.
@@
 Our model agrees with this newly discovered feedback system, as YREG1 regulates LEF-1 activity leading to EMT stabilization.
-Secondly, we should revisit the role of GSK-3$\beta$.
+####Secondly, we should revisit the role of GSK-3$\beta$.
 GSK-3$\beta$ is an important regulator which controls the abundance of both Snail and $\beta$-catenin through the ubiquitin-proteasome pathway \citep{Zhou:2004kx,Larue:2005vn}.
@@
 The EMT model described the gene expression program resulting from TGF$-\beta$ and VEGF-A signaling in a prototypical epithelial cell.
-The TGF$-\beta$-EMT network contained 995 nodes (proteins, mRNA or genes) interconnected by 1700 interactions.
+The TGF$-\beta$-EMT network contained 97 nodes (proteins, mRNA or genes) interconnected by 251 interactions.
 The network connectivity was curated from more than 40 primary literature sources in combination with on-line databases \citep{Linding:2007fk,Jensen:2009fk}.

### 2016-10-03 Paper_v7.tex
update edits DB2

@@
-\documentclass[12pt]{article}
+y\documentclass[12pt]{article}
 % Load packages
@@
 mesenchymal markers such as Vimentin and key upstream transcription factors such as LEF-1 in a SMAD dependent manner.
+TGF-$\beta$3 expression was also able to sustain $/beta$-catenin release by inhibiting its sequestration by the APC complex through PI3K mediated GSK3, which was captured by the model (Fig.
+\ref{fg:F4B})
 Each phase involved the hierarchal expression and/or post-translational modification of several key transcription factors.
@@
 Taken together, the model captured the key signaling events revealed by Medici \emph{et al.} \citep{Medici:2008fk} that drive the phenotypic conversion.
-A listing of data used for training is included in the supplement (Fig. \ref{fig:S5} and Fig. \ref{fig:S6}).
+A listing of objective function data resulting from training and cross validation is included in the supplement (Fig. \ref{fig:SX}).
 
@@
 Low levels of YREG1 expression were present in all simulations to regulate the formation of the $\beta$-catenin-LEF-1 complex.
-To test the potency of YREG1, we conducted knockdown and over-expression simulations following the addition of TGF-$\beta$1/2 (Fig. \ref{fig:S8}).
+To test the potency of YREG1, we conducted over-expression and knockdown simulations following the addition of TGF-$\beta$1/2 #### (Fig. \ref{fig:4}C and \ref{fig:4}D).
 In the absence of YREG1, most of the population failed to consistently respond to TGF-$\beta$1/2 exposure compared to the wild-type (Fig. \ref{fig:S8}A).
-Conversely, YREG1 overexpression revealed an exclusively epithelial phenotype following TGF-$\beta$1/2 stimulation (Fig. \ref{fig:S8}B).
+Conversely, YREG1 overexpression revealed an exclusively epithelial phenotype following TGF-$\beta$1/2 stimulation (Fig. \ref{fig:S8}B).####
 Overexpression of YREG1 repressed LEF-1 and TGF-$\beta$3 expression,
@@
 We evlautated these quantities for each member of the ensemble (N $\simeq$ 1400) for two downstream phenotypic markers, Vimentin (mesenchymal) and E-cadherin (epithelial)
-following the addition of TGF$-\beta$1/2 alone (Fig. \ref{fg:F3}), and VEGF-A in combination with NFATc inhibitors (Fig. \ref{fg:F4}).
+following the addition of TGF$-\beta$1/2 alone (Fig. \ref{fg:F3}), and VEGF-A in combination with NFATc inhibitors (Fig. \ref{fg:F3}).
 We were further able to use this analysis to examine the activity and effects of underlying signaling molecules.
 
-We identified model subpopulations that exhibited different behaviors following exposure to TGF$-\beta$1/2 (Fig. \ref{fg:F3}A, labeled P1-P4).
+We identified model subpopulations that exhibited different behaviors following exposure to TGF$-\beta$1/2 (Fig. \ref{fg:F3}A).
 Analysis of the molecular signatures of these subpopulations suggested the abundance, localization and state of the Sp1, AP-1 and NFATc transcription factors controlled population heterogeneity.
-The behavior of the majority of models ($>$70\%) was similar to subpopulation one (P1) or subpopulation two (P2) in Fig \ref{fg:F3}.
+The behavior of the majority of models ($>$80\%) respond to treatment in Fig \ref{fg:F3}A-C, moving off of the untreated population set (gray).
 These models showed the classically expected behavior, a switch from an epithelial to mesenchymal phenotype following TGF$-\beta$1/2 exposure.
-Models near P1 had evaluated nuclear localized phosphorylated Sp1, relative to non-induced cells (and models near P2).
+Models near resembling untreated cells  had evaluated nuclear localized phosphorylated Sp1, relative to non-induced cells.
 Elevated Sp1 activity decreased E-cadherin expression through Slug-mediated inhibition, which in turn increased Vimentin expression through TGF$-\beta$3 autocrine signaling and the liberation of $\beta$-catenin.
-Near P2, Sp1 transcriptional activity was lower than P1, leading to only modestly increased Vimentin expression and E-cadherin repression following TGF$-\beta$1/2 stimulation.
-Near subpopulation three (P3), reduced levels of nuclear phosphorylated AP-1, Sp1, and NFAT (resulting from the loss of ERK kinase activity)
-were responsible for Vimentin \emph{repression} relative to the control.
-However, the most biologically interesting behavior was exhibited by subpopulation four (P4).
-Models near P4 had elevated Sp1 and NFAT transcriptional activity, which increased \emph{both} Vimentin and E-cadherin expression.
+
+However, the most biologically interesting behavior was exhibited by cells achieving a hybrid phenotype, most notable in a dual treatment condition (black arrow), but also present in a small percentage of untreated cells (Fig.
+\ref{fg:F3}A) .
+Models with this hybrid phenotype had elevated Sp1 and NFAT transcriptional activity, resulting in increased \emph{both} Vimentin and E-cadherin expression, respectively (Fig. \ref{fg:F4}A).
 Analysis of these hypothetical cells suggested they had \emph{abnormal} signaling; deregulated NFAT expression and nuclear localization promoted E-cadherin expression while TGF$-\beta$1/2 induced Sp1 action promoted Vimentin expression.
-Analysis of the connectivity and information flow through the signaling architecture suggested that Sp1 and NFAT action could be manipulated \emph{independently} by simultaneous TGF$-\beta$1/2 and VEGF-A stimulation (Fig.
+####Analysis of the connectivity and information flow through the signaling architecture suggested that Sp1 and NFAT action could be manipulated \emph{independently} by simultaneous TGF$-\beta$1/2 and VEGF-A stimulation (Fig.
 \ref{fg:S1}).
@@
 To test this hypothesis, we simulated the response of the network to TGF$-\beta$1/2 and VEGF-A treatment with and without NFATc inhibitors (Fig. \ref{fg:F4}).
-As expected, stimulation with VEGF-A (50 a.u.) maintained an epithelial population (Q4-43.6\%), while TGF$-\beta$1/2 (10 a.u.) exposure shifted
-the population from an epithelial (Q4-5.5\%) to a mesenchymal (Q1-45.6\%) phenotype (Fig. \ref{fg:F4}A and Fig. \ref{fg:F4}B).
-On the other hand, combined stimulation with TGF$-\beta$1/2 (10 a.u.) and VEGF-A (50 a.u.) increased both E-cadherin and Vimentin expression (Q2-45.3\%), resulting in a hybrid phenotype with both epithelial and mesenchymal characteristics (Fig.
-\ref{fg:F4}C).
-To better understand this hybrid response, we quantified the simulated protein levels for E-cadherin, Vimentin, phosphorylated nuclear Sp1, nuclear NFATc1, $\alpha$-smooth muscle actin ($\alpha$-SMA) and Slug as a function of condition (Fig.
-\ref{fg:S2}A-C).
+As expected, stimulation with VEGF-A (50 a.u.) maintained an epithelial population, while TGF$-\beta$1/2 (10 a.u.) exposure shifted
+the population from an epithelial to a mesenchymal phenotype (Fig. \ref{fg:F3}A and Fig. \ref{fg:F3}B).
+On the other hand, combined stimulation with TGF$-\beta$1/2 (10 a.u.) and VEGF-A (50 a.u.) increased both E-cadherin and Vimentin expression, resulting in a hybrid phenotype with both epithelial and mesenchymal characteristics (Fig.
+\ref{fg:F3}C).
 Vimentin expression was correlated with high levels of nuclear phosphorylated Sp1, following TGF$-\beta$1/2 exposure.
@@
 To further isolate the role of NFAT on this hybrid state, we simulated the inhibition of NFAT transcriptional activity across all conditions (all else being equal).
-NFAT inhibition in combination with VEGF-A treatment blocked all E-cadherin positive sets (Fig. \ref{fg:F4}D).
-Likewise, TGF$-\beta$1/2 treatment in combination with NFATc inhibition also resulted in the loss of E-cadherin expression (Fig. \ref{fg:F4}E).
+NFAT inhibition in combination with VEGF-A and TGF$-\beta$1/2 treatments blocks increased activation of E-cadherin (Fig. \ref{fg:3}D and Fig. \ref{fg:3}E).
 Lastly, NFATc inhibition in combination with simultaneous TGF$-\beta$1/2 and VEGF-A exposure repressed nearly all E-cadherin expression, shifting nearly the entire population towards a mesenchymal phenotype (Fig.
 \ref{fg:F4}F).
-Taken together, high levels of nuclear localized phosphorylated Sp1 correlated with Vimentin expression, while NFATc transcriptional activity was predicted to be critical for maintaining E-cadherin expression.
+Taken together, high levels of nuclear localized phosphorylated Sp1 correlated with Vimentin expression, while NFATc transcriptional activity was predicted to be critical for maintaining E-cadherin expression in the prescence of competing signals.
+In particular, when present, VEGF-A signaling through NFATc was predicted to provide a protective up-regulation of E-cadherin.
 
@@
 NFATc1 expression increased, while Sp1 expression was similar to the TGF$-\beta$1/2 case alone (Fig. \ref{fg:F5}A-B,  Fig S3D,E; Fig S4C)), supporting their independent regulation.
-The expression of Slug, and Vimentin significantly increased, while E-cadherin levels were increased in MCF10A cells (Fig 5A) and maintained at control levels in DLD1 cells (Fig. S3D).
+The expression of Slug, and Vimentin significantly increased, while E-cadherin levels were increased in MCF10A cells (Fig 5A) and maintained at control levels in DLD1 cells (Fig. S3D).y
 As further predicted, nuclear co-localization of both NFATc1 and phosphorylated Sp1 were apparent in MCF10A and DLD1 cells treated with both ligands (Fig. \ref{fg:F5}B,C Fig S3E,F).
@@
 VIVIT mediated NFATc inhibition significantly reduced the circularity index, similar to TGF$-\beta$1/2 treatment (Fig. \ref{fg:F6}F).
-VEGF-A treatment also induced the formation of tubular structures (acini), but the number of tubular branches relative to total acini was significantly increased upon combined TGF$-\beta$1/2 and VEGF-A.
+VEGF-A treatment also induced the formation of tubular structures (acini), but the number of tubular branches relative to total acini was significantly increased upon combined TGF$-\betya$1/2 and VEGF-A.
 No tubular structures were identified within the DLD1 constructs during the 7 day tubulogenesis endpoints, supporting that MCF10A and DLD1 cells have some cell-type specific EMT sensitivity despite their underlying competency for acquiring a heterogeneous phenotype.
@@
 Our model agrees with this newly discovered feedback system, as YREG1 regulates LEF-1 activity leading to EMT stabilization.
-####Secondly, we should revisit the role of GSK-3$\beta$.
-GSK-3$\beta$ is an important regulator which controls the abundance of both Snail and $\beta$-catenin through the ubiquitin-proteasome pathway \citep{Zhou:2004kx,Larue:2005vn}.
-Specifically to our model, expression of Snail increases through 72 hrs.
-In contrast, experimental data has shown that activity of Snail peaks at 24 hrs which may be controlled by the GSK-3$\beta$ complex \citep{Medici:2006qa}.
+
 Recent evidence has also suggested an essential role of NF-$\kappa$B in epithelial transformation.
@@
 This activates the receptors serine/threonine kinase activity to phosphorylate and activate the receptor Smads 2/3 \citep{Massague:2005qc}.
-Phosphorylated Smads 2/3 (pSmad2/3) form heterodimers with partner Smad4 and translocate to the nucleus.
+Phosphorylated Smads 2/3 (pSmad2/3) form heterodimers and translocate to the nucleus.
 pSmads complexes up-regulate other transcription factors, such as LEF-1.
-The pSmad2/4/LEF-1 has been shown to directly repress the E-cadherin gene \citep{Nawshad:2007kl}.
+The pSmad2/4-LEF-1 complex has been shown to directly repress the E-cadherin gene \citep{Nawshad:2007kl}.
 LEF-1 also binds with $\beta$-catenin to upregulate mesenchymal proteins such as fibronectin \citep{Medici2011}.
+In the model, Smad signaling is lumped into a single smad species that can act in a co-dependent fashion with LEF1 to downregulate E-cadherin. 
 The EMT gene expression program was initiated by the binding of TGF$-\beta$ isoforms to TGF$-\beta$ surface receptors.
@@
 Repression of E-cadherin expression is the central event in the transition from an epithelial to a mesenchymal phenotype \citep{Cano:2000kh}.
+####
 However, this transition is not solely driven by transcriptional events.
@@
 to drive TGF$-\beta$3 expression \citep{Medici:2008fk}.
-Simultaneously, ERK1/2-mediated phosphorylation of the AP1 and Sp1 transcription factors can also regulate transcriptional complexes involving NFAT, Slug, and Smads.
-Lastly, canonical pathways for processing extracellular VEGF-A, BMP and Wnt signals, in addition to the PI3K pathway
-were also included in the model. Additional information about the interactions included in the model,
+The PI3K to GSK3 pathway was included and acted as an activating mechanism of $/beta$-catenin signaling through TGF$-\beta$3 signaling \citep{Medici:2008fk}.
+GSK3 is known to act on $\beta$-catenin signaling through the ubiquitin-proteasome pathway \citep{\
+Zhou:2004kx,Larue:2005vn}
+VEGF-A activation of NFATc1 takes place through calcineurin signaling leading to an enhancement of E-cadherin expression, as supported by our VEGF-A experimental data (fig.
+\ref{fgSX2}) \cite{Suehiro2014}.
+Additional information about the interactions included in the model,
 along with the Systems Biology Markup Language (SBML) file encoding these interactions are included in the supplemental materials.
@@
 \subsubsection*{Estimation of model parameters using multiobjective optimization.}
-The EMT model had 1756 unknown parameters (1700 kinetic constants and 56 non-zero initial conditions) which were not uniquely identifiable given the training data.
+The EMT model had 296 unknown parameters (169 kinetic constants, 44 saturation constants, 38 control logic paramters, and 45 non-zero initial conditions) which were not uniquely identifiable given the training data.
 Instead, we estimated a population of likely models (each consistent with the training data)
@@
 These 11 objective functions were then partitioned, where each partition contained ten training objectives and one validation objective.
-The training and validation data were Western blots. Thus, all model simulations were in arbitrary units.
-However, POETs does allow a soft constraint on the order of magnitude of the model concentration scale.
-In this study, we assumed the natural model concentration scale was pmol/L. We did not place a lower bound on model states.
-However, based on the pmol/L natural scale, we treated all values less than 10$^{-3}$ as zero (or no expression).
+The training and validation data were Western blots.
+We achived a physical simulation scale by establishing characeteristic rates of transcription, translation, mRNA and protein degradation, as well as characteristic concentrations of ribosomes and RNAPs ####cite bionumbers####.
+The concentration scale is in nM, with proteins ranging from 10-1000nM and mRNA ranging from 0.01 to 1nM, reflecting the true abundances and ratios between each species.
 
@@
 Cells were passaged 1:3 or 1:4 every 3-6 d and used between passages 4 and 8.
+####
+DLD1 and MCF10A cells were suspended in culture media (with RPMI 1640 medium with 10\% fetal bovine serum and 1\% penicillin/streptomycin for DLD1 or MGEM 2 supplemented with insulin, bovine pituitary extract, cholera toxin, hEGF, hydrocortisone, 5\% horse serum, and 1\% penicillin/streptomycin for MCF10A), and allowed to aggregate overnight in hanging drop culture (20μL; 20,000 cells).
+The spherical aggregates were placed on the surface of neutralized type I collagen hydrogels (1.5mg/mL) and allowed to adhere.
+Cultures were then serum starved (1\% serum) for 24 hours.
+Recombinant VEGFA165 was then added to the media (5ng/ml, 50ng/ml) and mRNA was harvested after 3hr and 24hr timepoint.
+####
 
@@
 \begin{figure}
-\includegraphics [width=1.0\linewidth] {./figs/Fig-1-Network.pdf}
-\caption{Model connectivity recreates the core architecture during EMT. The EMT network contains 995 nodes (proteins, mRNA, and genes) interconnected by 1700 interactions.
+\includegraphics [width=1.0\linewidth] {./figs/Fig1_coarse_network.pdf}
+\caption{Model connectivity recreates the core architecture during EMT. The EMT network contains 97 nodes (proteins, mRNA, and genes) interconnected by 169 interactions.
 Central to EMT induction, activation of the MAPK cascade occurs through TGF-$\beta$1/2 binding which activates the AP-1/Sp1 transcriptional axis.
 AP-1/Sp1 drives an autocrine response of TGF-$\beta$3, which activates the Smad cascade, leading to phenotypic change.
-Conversely, VEGF-A binding promotes an epithelial phenotype through NFAT activation.
-Other important signaling pathways not shown but also incorporated include the BMP, Wnt, and PI3K pathways.
+Conversely, VEGF-A binding can stabalize an epithelial phenotype through NFAT activation.
+Downstream activation of $/beta$-catenin signaling due to E-cadherin loss and GSK3 inactivation of $/beta$-cateinin confinement is critical to the complete activation of the EMT program.
 The complete list of molecular interactions that comprise the model is given in the supplement.}\label{fg:F1}
@@
 \begin{figure}
-\includegraphics [width=1.0\linewidth] {./figs/Fig-2-Training.pdf}
+\includegraphics [width=1.0\linewidth] {./figs/Fig2_training.pdf}
 \caption{Training and validation simulations. The population of EMT models qualitatively captured TGF$\beta$-induced EMT signaling.
 (A-I) The population was generated using POETs and trained using 11 different objective functions (41 data sets) taken from Medici \emph{et al.} \citep{Medici:2008fk}.
-The model captured the simulated experiments for 78\% of the cases.
-(J-L)  The model populations were also compared against untrained temporal data to measure the effectiveness as a pure prediction.
-The high predictability can be contributed to the leave-one-out cross validation scheme, objective functions with overlapping data, and multi-objective optimization algorithm.}\label{fg:F2}
+The model captured the simulated experiments for 72\% of the cases.
+(J-L)  The model populations were also compared against untrained temporal data to measure the effectiveness as a pure prediction.}\label{fg:F2}
 \end{figure}
@@
 \begin{figure}
-\includegraphics [width=1.0\linewidth] {./figs/Fig-3-HypotheticalPhenotype.pdf}
-\caption{Simulated TGF$\beta$1/2 exposure promoted phenotype heterogeneity.
-Robustness coefficients were used to quantify the effect of perturbations after steady state across the population of data sets.
-Coefficients with values $>$ 1 ($<$ 1) indicated a marker increased (decreased) compared to a base state, while a value of 1 indicated approximately no change following a perturbation.
-E-cadherin and Vimentin robustness coefficients were used as phenotypic markers.
-(A) TGF$\beta$1/2 perturbation, we isolated 4 distinct parameter sets known to have phenotypically different behaviors.
-(B) Each region has been represented by a small signaling network.
-From our analysis, we determined that the differences were a function of downstream transcription factors (phosphorylated-Sp1, and NFAT) within the TGF$\beta$2 and VEGFA pathway, respectively.
-(C) We hypothesized that elevated phosphorylated-Sp1 and NFAT levels could drive phenotype heterogeneity through simultaneous TGF$\beta$2 and VEGFA treatment, similar to region three (R3).}\label{fg:F3}
+\includegraphics [width=1.0\linewidth] {./figs/Fig3_phenotype.pdf}
+\caption{Simulated VEGF-A and TGF$\beta$1/2 exposure promoted phenotype heterogeneity.
+Simulated response to TGF$\beta$1/2 and VEGF-A exposure with and without axis specific inhibitors.
+Vimentin and E-cadherin abundances (in nM) were used to quantify the shift in population at 48 hrs.
+(A-C) VEGF-A (50 a.u.) treatment resulted in a population with enhanced epithelial properties.
+This was contrary to the addition of TGF$\beta$2 (10 a.u.), which shifted the population towards a mesenchymal phenotype.
+Interestingly, the combined effects of TGF$\beta$2 and VEGFA was found to increase both ecadherin and vimentin levels, creating a heterogeneous population (black arrow), which can also be seen in a minority of untreated cells (gray arrow).
+(D-F) To isolate the effect of NFAT, we inhibited NFAT de-phosphorylation in combination with VEGFA.
+This negated the increase in ecadherin expression and shifted the population towards a mesenchymal phenotype (Q1,Q3).
+Likewise, combining NFAT inhibition with TGF$\beta$ mitigated all VEGF enhanced ecadherin expression.
+Lastly, combination of TGF$\beta$2, VEGFA, and NFAT inhibition nearly mitigated all effects of VEGFA, shifting the heterogeneous population towards a mesenchymal phenotype.
+In whole, high levels of phosphorylated-Sp1 correlated with vimentin expression, while NFAT was responsible for maintaining E-cadherin expression, although neither were mutually exclusive.}
 \end{figure}
@@
 \begin{figure}
-\includegraphics [width=1.0\linewidth] {./figs/Fig-4-Robustness.pdf}
-\caption{Simulated response to TGF$\beta$1/2 and VEGF-A exposure with and without axis specific inhibitors.
-Robustness coefficients were used to quantify the shift in population at 48 hrs.
-(A-C) VEGF-A (50 a.u.) treatment resulted in a population with enhanced epithelial (Q4) properties.
-This was contrary to the addition of TGF$\beta$2 (10 a.u.), which shifted the population towards a mesenchymal phenotype (Q1).
-Interestingly, the combined effects of TGF$\beta$2 and VEGFA was found to increase both ecadherin and vimentin levels, creating a heterogeneous population (Q2).
-(D-F) To isolate the effect of NFAT, we inhibited NFAT de-phosphorylation in combination with VEGFA.
-This negated the increase in ecadherin expression and shifted the population towards a mesenchymal phenotype (Q1,Q3).
-Likewise, combining NFAT inhibition with TGF$\beta$ mitigated all ecadherin expression (Q2) confirming their importance for population heterogeneity.
-Lastly, combination of TGF$\beta$2, VEGFA, and NFAT inhibition nearly mitigated all effects of VEGFA, shifting the heterogeneous population (Q2) towards a mesenchymal phenotype (Q1).
-In whole, high levels of phosphorylated-Sp1 correlated with vimentin expression, while NFAT was responsible for maintaining E-cadherin expression, although neither were mutually exclusive.}\label{fg:F4}
+\includegraphics [width=1.0\linewidth] {./figs/Fig4_mechanism.pdf}
+\caption{Analysis of underlying signaling responses.
+(A) We examined the distribution of NFATc1 and AP1/SP1 in cells containing the hybrid phenotype (VEGF-A + TGF$\beta$2 case), showing the potential for cells to express both SP1 and NFATc1 in a non exclusive manner.
+(B) We were able to show a fit to an additional objective demonstrating the actiation of GSK3 through PI3K.
+Our model captures this activation through TGF$\beta$3 signaling.
+LY294002 is a PI3K inhibitor.
+(C) We identified a novel regulator of LEF1 called YREG1 that allows Snail/Slug to emulate an inducer by repressing YREG1.
+Over expression of YREG1 increases the degree of E-cadherin expression, stabalizing the epithelial phenotype.
+(D) Kockout of YREG1 leads to a loss of both E-cadherin and vimentin for cells that were inherently activated in the untreated population. ####
+}\label{fg:F4}
 \end{figure}

### 2016-10-03 Supplemental.tex
update edits DB2

--- 2016-09-28 Supplemental-JB.tex
@@
+\documentclass[12pt]{article}
+% Load packages
+\usepackage{url}  % Formatting web addresses
+\usepackage{ifthen}  % Conditional
+\usepackage{multicol}   %Columns
+\usepackage[utf8]{inputenc} %unicode support
+\usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{epsfig}
+\usepackage{epstopdf}
+\usepackage{graphicx}
+\usepackage[margin=0.1pt,font=footnotesize,labelfont=bf]{caption}
+\usepackage{setspace}
+%\usepackage{longtable}
+\usepackage{colortbl}
+%\usepackage{palatino,lettrine}
+%\usepackage{times}
+%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
+%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
+\usepackage[wide]{sidecap}
+\usepackage[authoryear,round,comma,sort&compress]{natbib}
+\usepackage[authoryear,round]{natbib}
+\usepackage{supertabular}
+\usepackage{simplemargins}
+\usepackage{lineno}
+
+\urlstyle{rm}
+
+%\textwidth = 6.50 in
+%\textheight = 9.5 in
+%\oddsidemargin =  0.0 in
+%\evensidemargin = 0.0 in
+%\topmargin = -0.50 in
+%\headheight = 0.0 in
+%\headsep = 0.25 in
+%\parskip = 0.15in
+%\linespread{1.75}
+\doublespace
+
+\setallmargins{1in}
+
+%\bibliographystyle{plain}
+\bibliographystyle{msb}
+
+\makeatletter
+\renewcommand\subsection{\@startsection
+	{subsection}{2}{0mm}
+	{-0.05in}
+	{-0.5\baselineskip}
+	{\normalfont\normalsize\bfseries}}
+\renewcommand\subsubsection{\@startsection
+	{subsubsection}{2}{0mm}
+	{-0.05in}
+	{-0.5\baselineskip}
+	{\normalfont\normalsize\itshape}}
+\renewcommand\section{\@startsection
+	{subsection}{2}{0mm}
+	{-0.2in}
+	{0.05\baselineskip}
+	{\normalfont\large\bfseries}}
+\renewcommand\paragraph{\@startsection
+	{paragraph}{2}{0mm}
+	{-0.05in}
+	{-0.5\baselineskip}
+	{\normalfont\normalsize\itshape}}
+\makeatother
+
+%Review style settings
+%\newenvironment{bmcformat}{\begin{raggedright}\baselineskip20pt\sloppy\setboolean{publ}{false}}{\end{raggedright}\baselineskip20pt\sloppy}
+
+%Publication style settings
+
+% Single space'd bib -
+\setlength\bibsep{0pt}
+
+\renewcommand{\rmdefault}{phv}\renewcommand{\sfdefault}{phv}
+
+% Change the number format in the ref list -
+\renewcommand{\bibnumfmt}[1]{#1.}
+
+% Change Figure to Fig.
+\renewcommand{\figurename}{Fig.}
+
+\begin{document}
 \section*{Supplemental Materials and Methods}
@@
 \begin{figure}
-\includegraphics [width=1.0\linewidth] {./figs/Fig-5-Supplemental-ErrorTable.pdf}
-\caption{Training and prediction values for the 11 TGF-$\beta$ objective functions versus a random parameter control.}\label{fig:S5}
+\includegraphics [width=1.0\linewidth] {./figs/FigSX_error_table.pdf}
+\caption{Training and prediction values for the 11 TGF-$\beta$ objective functions versus a random parameter control.}\label{fig:SX}
+\end{figure}
+
+\begin{figure}
+\includegraphics [width=1.0\linewidth] {./figs/FigSX2_Handfit_VEGFA.pdf}
+\caption{VEGF-A qPCR data used to hand fit VEGF enhancement of E-cadherin expression. mRNA was harvested after 3hr and 24hr timepoint.}\label{fig:SX2}
 \end{figure}
@@
 
-%\bibliography{References_v1}
+\bibliography{References_v1}
+
+\end{document}

### 2016-10-03 Paper_v7.tex
First pass through by JV

@@
-y\documentclass[12pt]{article}
+\documentclass[12pt]{article}
 % Load packages
@@
 \vspace{0.05in}
-{\par \centering School of Chemical and Biomolecular Engineering and $^{\ddag}$Department of Biomedical Engineering, Cornell University, Ithaca NY 14853}
+{\par \centering Robert Frederick Smith School of Chemical and Biomolecular Engineering and $^{\ddag}$Nancy E.
+and Peter C.
+Meinig School of Biomedical Engineering, Cornell University, Ithaca NY 14853}
 \vspace{0.1in}
@@
 {\par \centering Jonathan T. Butcher,}
-{\par \centering Associate Professor, Department of Biomedical Engineering,}
+{\par \centering Associate Professor, Nancy E. and Peter C. Meinig School of Biomedical Engineering,}
 {\par \centering 304 Weill Hall, Cornell University, Ithaca NY, 14853}
@@
 We incorporated mass action kinetics within an ordinary differential equation (ODE) framework to describe the EMT interaction network containing 97 gene, protein or mRNA components interconnected through 169 interactions.
-A family of model parameters was estimated using 41 molecular data sets generated in DLD1 colon carcinoma, MDCKII and A375 melanoma cells using the Pareto optimal ensemble technique (POETs) multiobjective optimization algorithm.
-POETs generated an ensemble of approximately 1400 models for analysis.
+A family of model parameters was estimated using 41 molecular data sets generated in DLD1 colon carcinoma, MDCKII and A375 melanoma cells using the Pareto optimal ensemble technique (JuPOETs) multiobjective optimization algorithm.
+JuPOETs generated an ensemble of approximately 1400 models for analysis.
 Analysis of the model population suggested that both MCF10A and DLD1 cells could exhibit phenotypic heterogeneity if treated simultaneously with TGF-$\beta$1/2 and VEGF-A.
@@
 The EMT model contained 74 molecular species interconnected by 169 interactions.
-Model equations were formulated using mass-action kinetics within an ordinary differential equation (ODE) framework.
+Model equations were formulated using either saturation or mass-action kinetics within an ordinary differential equation (ODE) framework.
 ODEs and mass action kinetics are common tools to model biochemical pathways \citep{Schoeberl:2002rv,Chen:2009uq,Tasseff:2011ve}.
@@
 we estimated a sloppy population of models (each consistent with the training data) by simultaneously minimizing the difference between model simulations and 41 molecular
-data sets using the Pareto Optimal Ensemble Technique (POETs).
+data sets using the Pareto Optimal Ensemble Technique (JuPOETs).
 The training data were generated in DLD1 colon carcinoma, MDCKII, and A375 melanoma cells following
-exposure to TGF-$\beta$ isoforms \citep{Medici:2008fk}. We organized these data sets into 11 objective functions which were simultaneously minimized by POETs.
+exposure to TGF-$\beta$ isoforms \citep{Medici:2008fk}. We organized these data sets into 11 objective functions which were simultaneously minimized by JuPOETs.
 Additionally, we used 12 molecular data sets generated in HK-2 cells following VEGF-A exposure to train VEGF-A responsive model processes \citep{Lian:2011bs}.
@@
 levels) to evaluate the predictive power of the parameter ensemble.
-Additional details of the signaling architecture included in the model are presented in the materials and methods and the supplement.
 
-POETs generated a population of probable signaling models which captured the multiple phases of EMT induction (Fig. \ref{fg:F2}).
-POETs sampled well over 10$^4$ probable models during each stage of the cross-validation #####, using a combination of both local and global random sampling.
+JuPOETs generated a population of probable signaling models which captured the multiple phases of EMT induction (Fig. \ref{fg:F2}).
+JuPOETs sampled well over 10$^4$ probable models during each stage of the cross-validation using global random sampling.
 From this analysis, N $\simeq$ 1400 models were selected for further analysis.
 The selected models all had the same possible molecular connectivity, but different values for model parameters.
-Trascription and translation rates, as well as mRNA and protein degradation terms, were set using physical values from the literature, and allowed to vary by a scaling factor, see methods ####citebionumbers####.
-Model selection was based upon Pareto rank, the prediction and training error across all objectives and model to model correlation (supplemental materials).
+Transcription and translation rates, as well as mRNA and protein degradation terms, were set using physical values from the literature \citep{Milo:2010aa},
+and allowed to vary by a scaling factor, see methods.
+Model selection was based upon Pareto rank, the prediction and training error across all objectives.
 The model population recapitulated key signaling events following TGF-$\beta$ exposure.
@@
 mesenchymal markers such as Vimentin and key upstream transcription factors such as LEF-1 in a SMAD dependent manner.
-TGF-$\beta$3 expression was also able to sustain $/beta$-catenin release by inhibiting its sequestration by the APC complex through PI3K mediated GSK3, which was captured by the model (Fig.
-\ref{fg:F4B})
+TGF-$\beta$3 expression was also able to sustain $\beta$-catenin release by inhibiting its sequestration by the APC complex through PI3K mediated GSK3,
+which was captured by the model (Fig. \ref{fg:F4}B).
 Each phase involved the hierarchal expression and/or post-translational modification of several key transcription factors.
@@
 In whole, 72\% of our training objectives were statistically significant (at a 95\% confidence interval) compared to a randomized parameter family (N = 100)
-generated from a random starting point.
-Conversely, we \emph{predicted} approximately 72\% of the training objectives, at a 95\% confidence interval compared to randomized parameters.
-The model also captured the temporal gene expression responses of E-cadherin, pSmad2, and LEF-1 to within one-standard deviation (up to the 48 hr time-point) (Fig. \ref{fg:F2}J-L).
-This data was not used for model training.
-The high predictability can be attributed to the combination of the leave-one-out cross validation scheme, diverse objective functions, and robustness of the POETs algorithm.
+generated from a random starting point. Conversely, we \emph{predicted} approximately 72\% of the training objectives, at a 95\% confidence interval compared to randomized parameters.
+The model also captured the temporal gene expression responses of E-cadherin, pSmad2, and LEF-1 (not used for model training) to within one-standard deviation (up to the 48 hr time-point) (Fig.
+\ref{fg:F2}J-L).
 Taken together, the model captured the key signaling events revealed by Medici \emph{et al.} \citep{Medici:2008fk} that drive the phenotypic conversion.
-A listing of objective function data resulting from training and cross validation is included in the supplement (Fig. \ref{fig:SX}).
+A listing of objective function values resulting from training, cross validation and the random parameter control is given in the supplement (Fig. \ref{fg:ObjTable}).
 
@@
 By adding this regulator, our simulations became consistent with training and literature data.
-Medici et al. suggested a similar idea where feedback between $\beta$-catenin and LEF-1 was likely, although this feedback had yet to be identified \citep{Medici:2008fk}.
+Medici et al. suggested that feedback between $\beta$-catenin and LEF-1 was likely, although this feedback had yet to be identified \citep{Medici:2008fk}.
 Low levels of YREG1 expression were present in all simulations to regulate the formation of the $\beta$-catenin-LEF-1 complex.
-To test the potency of YREG1, we conducted over-expression and knockdown simulations following the addition of TGF-$\beta$1/2 #### (Fig. \ref{fig:4}C and \ref{fig:4}D).
-In the absence of YREG1, most of the population failed to consistently respond to TGF-$\beta$1/2 exposure compared to the wild-type (Fig. \ref{fig:S8}A).
-Conversely, YREG1 overexpression revealed an exclusively epithelial phenotype following TGF-$\beta$1/2 stimulation (Fig. \ref{fig:S8}B).####
-Overexpression of YREG1 repressed LEF-1 and TGF-$\beta$3 expression,
-thereby not allowing free $\beta$-catenin to form the $\beta$-catenin-LEF-1 complex which promotes mesenchymal gene expression, or SMAD activity following from autocrine TGF-$\beta$3 signaling.
-Taken together, we found that low YREG1 expression was necessary for stabilizing EMT,
-while elevated YREG1 levels limited the extent of EMT induction.
+To test the potency of YREG1, we conducted over-expression and knockdown simulations following the addition of TGF-$\beta$1/2 (Fig. \ref{fg:F4}C and \ref{fg:F4}D).
+
+% TODO: REVISE -----
+% In the absence of YREG1, most of the population failed to consistently respond to TGF-$\beta$1/2 exposure compared to the wild-type (Fig. \ref{fig:S8}A).
+% Conversely, YREG1 overexpression revealed an exclusively epithelial phenotype following TGF-$\beta$1/2 stimulation (Fig. \ref{fig:S8}B).]
+% Overexpression of YREG1 repressed LEF-1 and TGF-$\beta$3 expression,
+% thereby not allowing free $\beta$-catenin to form the $\beta$-catenin-LEF-1 complex which promotes mesenchymal gene expression, or SMAD activity following from autocrine TGF-$\beta$3 signaling.
+% Taken together, we found that low YREG1 expression was necessary for stabilizing EMT,
+% while elevated YREG1 levels limited the extent of EMT induction.
 
@@
 However, the most biologically interesting behavior was exhibited by cells achieving a hybrid phenotype, most notable in a dual treatment condition (black arrow), but also present in a small percentage of untreated cells (Fig.
-\ref{fg:F3}A) .
-Models with this hybrid phenotype had elevated Sp1 and NFAT transcriptional activity, resulting in increased \emph{both} Vimentin and E-cadherin expression, respectively (Fig. \ref{fg:F4}A).
+\ref{fg:F3}A).
+Models with this hybrid phenotype had elevated Sp1 and NFAT transcriptional activity, resulting in the simultaneous increase of \emph{both} Vimentin and E-cadherin expression (Fig. \ref{fg:F4}A).
 Analysis of these hypothetical cells suggested they had \emph{abnormal} signaling; deregulated NFAT expression and nuclear localization promoted E-cadherin expression while TGF$-\beta$1/2 induced Sp1 action promoted Vimentin expression.
-####Analysis of the connectivity and information flow through the signaling architecture suggested that Sp1 and NFAT action could be manipulated \emph{independently} by simultaneous TGF$-\beta$1/2 and VEGF-A stimulation (Fig.
+
+% ####Analysis of the connectivity and information flow through the signaling architecture suggested that Sp1 and NFAT action could be manipulated \emph{independently} by simultaneous TGF$-\beta$1/2 and VEGF-A stimulation (Fig.
 \ref{fg:S1}).
@@
 To further isolate the role of NFAT on this hybrid state, we simulated the inhibition of NFAT transcriptional activity across all conditions (all else being equal).
-NFAT inhibition in combination with VEGF-A and TGF$-\beta$1/2 treatments blocks increased activation of E-cadherin (Fig. \ref{fg:3}D and Fig. \ref{fg:3}E).
+NFAT inhibition in combination with VEGF-A and TGF$-\beta$1/2 treatments blocks increased activation of E-cadherin (Fig. \ref{fg:F3}D and Fig. \ref{fg:F3}E).
 Lastly, NFATc inhibition in combination with simultaneous TGF$-\beta$1/2 and VEGF-A exposure repressed nearly all E-cadherin expression, shifting nearly the entire population towards a mesenchymal phenotype (Fig.
@@
 The EMT model simulations suggested the transcriptional activity of NFATc and Sp1 could be independently tuned to generate a hybrid cell population with both epithelial and mesenchymal characteristics.
-To test this hypothesis, we exposed either quiescent epithelial (MCFA10, (Fig. \ref{fg:F5})) or transformed epithelial cells (DLD1, (Fig. \ref{fg:S3})) to combinations of TGF$-\beta$1/2 and/or VEGF-A.
+To test this hypothesis, we exposed either quiescent epithelial (MCFA10, (Fig.
+\ref{fg:F5})) or transformed epithelial cells (DLD1, (Fig.
+\ref{fg:S3})) to combinations of TGF$-\beta$1/2 and/or VEGF-A.
 As expected, treatment with TGF$-\beta$1/2 (10ng/ml) increased Slug and Vimentin expression, while repressing E-cadherin expression both at the transcript and protein levels in MCF10A (Fig.
@@
 \ref{fg:F5}E,F) and DLD1 cells (Fig.
-\ref{fg:S4}) to combinations of VEGF-A and TGF$-\beta$1/2 in the presence or absence of VIVIT,
-a soluble peptide inhibitor of NFATc transcriptional activity \citep{Aramburu:1999if}.
+\ref{fg:S4}) to combinations of VEGF-A and TGF$-\beta$1/2 in the presence or absence of VIVIT, a soluble peptide inhibitor of NFATc transcriptional activity \citep{Aramburu:1999if}.
 Treatment with VEGF-A (50ng/ml) and VIVIT (10$\mu$M) in MCF10A cells significantly reduced E-cadherin expression compared to VEGF-A alone (Fig 5D,E).
@@
 VIVIT mediated NFATc inhibition significantly reduced the circularity index, similar to TGF$-\beta$1/2 treatment (Fig. \ref{fg:F6}F).
-VEGF-A treatment also induced the formation of tubular structures (acini), but the number of tubular branches relative to total acini was significantly increased upon combined TGF$-\betya$1/2 and VEGF-A.
+VEGF-A treatment also induced the formation of tubular structures (acini), but the number of tubular branches relative to total acini was significantly increased upon combined TGF$-\beta$1/2 and VEGF-A.
 No tubular structures were identified within the DLD1 constructs during the 7 day tubulogenesis endpoints, supporting that MCF10A and DLD1 cells have some cell-type specific EMT sensitivity despite their underlying competency for acquiring a heterogeneous phenotype.
@@
 This simulation incorporates an unprecedented level of detail compared to previous models, but as a consequence created a large number of unknown model parameters.
-Because these parameters could not be estimated uniquely apriori, we estimated an ensemble of likely parameters using the POETs multiobjective optimization framework.
+Because these parameters could not be estimated uniquely apriori, we estimated an ensemble of likely parameters using the JuPOETs multiobjective optimization framework.
 The model population was trained and cross-validated to prescribe biological significance using 41 data sets generated in DLD1 colon carcinoma, MDCKII, and A375 melanoma cell lines \citep{Medici:2008fk}.
@@
 \section*{Materials and Methods}
-The simulation code and parameter ensemble used in this study can be downloaded from GitHub (https://github.com/jeffreyvarner/TGFb-VEGFA-Model\_v1.git).
+All model code and parameter ensemble is freely available under an MIT software license and can be downloaded from http://www.varnerlab.org.
 
@@
 Using a canonical architecture allowed us to explore general features of TGF$-\beta$ induced EMT without cell line specific artifacts.
-On the other hand, because of the canonical architecture, we evaluated the simulation conclusions in several cell lines to test the generality of our conclusions.
+On the other hand, because of the canonical architecture, we can test the model against several cell lines to test the generality of our conclusions.
 
@@
 and A375 melanoma cells transition towards a mesenchymal phenotype \citep{Medici:2008fk}.
-Sequential activation of MAPK and Smad pathways were initiated upon addition of TGF$\beta$1/2.
-Briefly, TGF$\beta$2 signals through the RAS-RAF-MEK-ERK pathway to up-regulate Snail and Slug expression \citep{Medici:2006qa}.
+Sequential activation of MAPK and Smad pathways were initiated upon addition of TGF$-\beta$1/2.
+Briefly, TGF$-\beta$2 signals through the RAS-RAF-MEK-ERK pathway to up-regulate Snail and Slug expression \citep{Medici:2006qa}.
 Snail, a known repressor of junctional proteins, inhibits the expression of E-cadherin \citep{Cano:2000kh}.
 This initial repression of E-cadherin leads to a release of $\beta$-catenin from the cell membrane.
-Cystolic $\beta$-catenin can then translocate to the nucleus and form transcriptional complexes with TCF-4 to drive TGF$\beta$3 expression \citep{Medici:2008fk}.
-TGF$\beta$3 signals to the cells interior by binding to type II receptors, which form heterodimers with type I receptors (ALK5) \citep{Derynck:2003fc}.
+Cystolic $\beta$-catenin can then translocate to the nucleus and form transcriptional complexes with TCF-4 to drive TGF$-\beta$3 expression \citep{Medici:2008fk}.
+TGF$-\beta$3 signals to the cells interior by binding to type II receptors, which form heterodimers with type I receptors (ALK5) \citep{Derynck:2003fc}.
 This activates the receptors serine/threonine kinase activity to phosphorylate and activate the receptor Smads 2/3 \citep{Massague:2005qc}.
@@
 Repression of E-cadherin expression is the central event in the transition from an epithelial to a mesenchymal phenotype \citep{Cano:2000kh}.
-####
 However, this transition is not solely driven by transcriptional events.
@@
 to drive TGF$-\beta$3 expression \citep{Medici:2008fk}.
-The PI3K to GSK3 pathway was included and acted as an activating mechanism of $/beta$-catenin signaling through TGF$-\beta$3 signaling \citep{Medici:2008fk}.
-GSK3 is known to act on $\beta$-catenin signaling through the ubiquitin-proteasome pathway \citep{\
-Zhou:2004kx,Larue:2005vn}
-VEGF-A activation of NFATc1 takes place through calcineurin signaling leading to an enhancement of E-cadherin expression, as supported by our VEGF-A experimental data (fig.
-\ref{fgSX2}) \cite{Suehiro2014}.
-Additional information about the interactions included in the model,
-along with the Systems Biology Markup Language (SBML) file encoding these interactions are included in the supplemental materials.
+The PI3K to GSK3 pathway was included and acted as an activating mechanism of $\beta$-catenin signaling through TGF$-\beta$3 signaling \citep{Medici:2008fk}.
+GSK3 is known to act on $\beta$-catenin signaling through the ubiquitin-proteasome pathway
+\citep{Zhou:2004kx,Larue:2005vn}.
+VEGF-A activation of NFATc1 takes place through calcineurin signaling leading to an enhancement of E-cadherin expression \citep{Suehiro:2014aa},
+as supported by our VEGF-A experimental data (Fig. \ref{fg:VEGFA-Data}).
+
+%Additional information about the interactions included in the model,
+%along with the Systems Biology Markup Language (SBML) file encoding these interactions are included in the supplemental materials.
 
 \subsection*{Formulation, solution and analysis of the EMT model equations}
-EMT was modeled using mass-action kinetics within an ordinary differential equation (ODE) framework:
-\begin{equation}\label{asseassertain_massbalance}
-\frac{d\mathbf{x}}{dt}=\mathbf{S}\cdot\mathbf{r}\left(\mathbf{x},\mathbf{k}\right)\qquad\mathbf{x}\left(t_{o}\right)=\mathbf{x}_{o}
+\subsubsection*{EMT signaling events.}
+EMT signaling events were modeled using either saturation or mass-action kinetics within an ordinary differential equation (ODE) framework:
+\begin{equation}
+	\frac{1}{\tau_{i}}\frac{dx_{i}}{dt}  =  \sum_{j = 1}^{\mathcal{R}}\sigma_{ij}r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)-\mu x_{i}\qquad{i=1,2,\hdots,\mathcal{M}}\\
 \end{equation}
-The quantity $\mathbf{x}$ denotes the vector describing the abundance of protein, mRNA, and other species in the model ($995\times 1$).
-The stoichiometric matrix $\mathbf{S}$ encodes the signaling architecture considered in the model ($995\times 1700$).
-Each row of $\mathbf{S}$ describes a signaling component while each column describes a particular interaction.
-The $(i,j)$ element of $\mathbf{S}$, denoted by $\sigma_{ij}$, describes how species $i$ is involved with interaction $j$.
-If $\sigma_{ij}>0$, species $i$ is produced by interaction $j$. Conversely, If $\sigma_{ij}<0$, then species $i$ is consumed in interaction $j$.
-Lastly, if $\sigma_{ij}=0$, then species $i$ is not involved in interaction $j$.
-The term  $\mathbf{r}\left(\mathbf{x},\mathbf{k}\right)$ denotes the vector of interactions rates ($1700\times{1}$).
-We modeled each network interaction (gene expression, translation and biochemical transformations)
-using elementary rate laws where all reversible interactions were split into two irreversible steps (supplemental materials).
-Thus, the rate expression for interaction $q$ was given by:
-\begin{equation}\label{eq-mass-action}
-r_{q}\left(\mathbf{x},k_{q}\right)=k_{q}\prod_{j\in\left\{\mathbf{R}_{q}\right\}}x_{j}^{-\sigma_{jq}}
+where $\mathcal{R}$ denotes the number of signaling reactions and $\mathcal{M}$ denotes the number of proteins in the model.
+The quantity $\tau_{i}$ denotes a time scale parameter for species $i$ which captures unmodeled effects. In the current study $\tau_{i} = 1$ for all species.
+The quantity $r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)$ denotes the rate of reaction $j$.
+Typically, reaction $j$ is a non-linear function of biochemical and enzyme species abundance, as well as unknown model parameters $\mathbf{k}$ ($\mathcal{K}\times{1}$).
+The quantity $\sigma_{ij}$ denotes the stoichiometric coefficient for species $i$ in reaction $j$.
+If $\sigma_{ij}>0$, species $i$ is produced by reaction $j$.
+Conversely, if $\sigma_{ij}<0$, species $i$ is consumed by reaction $j$, while $\sigma_{ij} = 0$ indicates species $i$ is not connected with reaction $j$.
+Species balances were subject to the initial conditions $\mathbf{x}\left(t_{o}\right) = \mathbf{x}_{o}$.
+
+Rate processes were written as the product of a kinetic term ($\bar{r}_{j}$) and a control term ($v_{j}$) in the EMT model.
+The rate of enzyme catalyzed reactions was modeled using saturation kinetics:
+\begin{equation}\label{eqn:rate-saturation}
+	\bar{r}_{j} = k_{j}^{cat}\epsilon_{i}\left(\frac{x_{s}}{K_{js} + x_{s}}\right)
 \end{equation}
-The set $\left\{\mathbf{R}_{q}\right\}$ denotes reactants for reaction $q$, while $\sigma_{jq}$ denotes the stoichiometric coefficient (element of the matrix $\mathbf{S}$) governing species $j$ in reaction $q$.
+where $k_{j}^{cat}$ denotes the catalytic rate constant for reaction $j$, $\epsilon_{i}$ denotes the abundance of the enzyme catalyzing reaction $j$,
+and $K_{js}$ denotes the saturation constant for species $s$ in reaction $j$.
+On the other hand, mass action kinetics were used to model protein-protein binding interactions within the network:
+\begin{equation}\label{eqn:rate-action}
+	\bar{r}_{j} = k_{j}^{max}\prod_{s\in{m_{j}^{-}}}x_{s}^{-\sigma_{sj}}
+\end{equation}
+where $k_{j}^{max}$ denotes the maximum rate for reaction $j$, $\sigma_{sj}$ denotes the stoichiometric coefficient for species $s$ in reaction $j$,
+and $s\in{m_{j}}$ denotes the set of \textit{reactants} for reaction $j$. We assumed all binding interactions were irreversible.
+
+The control terms $0\leq v_{j}\leq 1$ depended upon the combination of factors which influenced rate process $j$.
+For each rate, we used a rule-based approach to select from competing control factors.
+If rate $j$ was influenced by $1,\dots,m$ factors, we modeled this relationship as
+$v_{j}=\mathcal{I}_{j}\left(f_{1j}\left(\cdot\right),\hdots,f_{mj}\left(\cdot\right)\right)$
+where $0\leq f_{ij}\left(\cdot\right)\leq 1$ denotes a regulatory transfer function quantifying the influence of factor $i$ on rate $j$.
+The function $\mathcal{I}_{j}\left(\cdot\right)$ is an integration rule which maps the output of regulatory transfer functions into a control
+variable. In this study, we used $\mathcal{I}_{j}\in\left\{min,max\right\}$ \citep{pr3010178}. If a process has no modifying factors, $v_{j}=1$.
+
+\subsubsection*{EMT gene expression processes.}
+The EMT model described both signal transduction and gene expression events following the addition of TGF$-\beta$ and VEGF-A.
+For each gene, we modeled both the resulting mRNA ($m_{j}$) and protein ($p_{j}$):
+\begin{eqnarray}
+	\frac{dm_{j}}{dt} &=& r_{T,j} - \left(\mu+\delta_{m,j}\right)m_{j}+\lambda_{j}\\
+	\frac{dp_{j}}{dt} &=& r_{X,j} - \left(\mu+\delta_{p,j}\right)p_{j}
+\end{eqnarray}where $j=1,2,\dots,\mathcal{G}$.
+The terms $r_{T,j}$ and $r_{X,j}$ denote the specific rate of transcription, and translation while
+the terms $\delta_{m,j}$ and $\delta_{p,j}$ denote degradation constants for mRNA and protein, respectively.
+Lastly, $\mu$ denotes the specific growth rate, and $\lambda_{j}$ denotes the constitutive rate of gene expression for gene $j$.
+The specific transcription rate was modeled as the product of a kinetic term $\bar{r}_{T,j}$ and a control term $u_{j}$ which described how the
+abundance of transcription factors, or other regulators influenced the expression of gene $j$. The kinetic rate of transcription was modeled as:
+\begin{equation}
+	\bar{r}_{T,j} = \alpha_{j}\left[V^{max}_{T}\left(\frac{G_{j}}{K_{T}+G_{j}}\right)\right]
+\end{equation}where the maximum gene expression rate was defined as the product of a characteristic transcription rate constant ($k_{T}$)
+and the abundance of RNA polymerase, $V^{max}_{T} = k_{T}\left(RNAP\right)$. The parameter $\alpha_{j}$ was used to adjust the transcription to that of gene $j$
+(estimated in this study), while $k_{T}$, $G_{j}$ and $RNAP$ were estimated from literature \citep{Milo:2010aa}.
+Similar to the signaling processes, the gene expression control term $0\leq u_{j}\leq 1$ depended upon the combination of factors which influenced rate process $j$.
+For each rate, we used a rule-based approach to select from competing control factors.
+If the expression of gene $j$ was influenced by $1,\dots,m$ factors, we modeled this relationship as
+$u_{j}=\mathcal{I}_{j}\left(f_{1j}\left(\cdot\right),\hdots,f_{mj}\left(\cdot\right)\right)$
+where $0\leq f_{ij}\left(\cdot\right)\leq 1$ denotes a regulatory transfer function quantifying the influence of factor $i$ on the expression of gene $j$.
+The function $\mathcal{I}_{j}\left(\cdot\right)$ is an integration rule which maps the output of regulatory transfer functions into a control
+variable. In this study, we used $\mathcal{I}_{j}\in\left\{min,max\right\}$ \citep{pr3010178}.
+If a gene expression process has no modifying factors, $u_{j}=1$.
+Lastly, the specific translation rate was modeled as:
+\begin{equation}
+	r_{X,j} = \beta_{j}\left[V^{max}_{X}\left(\frac{m_{j}}{K_{X}+m_{j}}\right)\right]
+\end{equation}where $V^{max}_{X}$ denotes a characteristic maximum translation rate estimated from literature, $\beta_{j}$ denotes the transcript specific correction the characteristic translation rate,
+and $K_{X}$ denotes a translation saturation constant.
+The characteristic maximum translation rate was defined as the product of a characteristic translation rate constant ($k_{X}$)
+and the abundance of Ribosomes ($RIBO$), $V^{max}_{X} = k_{X}\left(RIBO\right)$, where both $k_{X}$ and $RIBO$ abundance were estimated from literature \citep{Milo:2010aa}.
+
+The signaling and gene expression model equations were implemented in Julia and solved using the \texttt{CVODE} routine of the Sundials package \citep{Julia,Hindmarsh2005}.
+The model code and parameter ensemble is freely available under an MIT software license and can be downloaded from http://www.varnerlab.org.
+
+% \begin{equation}\label{asseassertain_massbalance}
+% \frac{d\mathbf{x}}{dt}=\mathbf{S}\cdot\mathbf{r}\left(\mathbf{x},\mathbf{k}\right)\qquad\mathbf{x}\left(t_{o}\right)=\mathbf{x}_{o}
+% \end{equation}
+% The quantity $\mathbf{x}$ denotes the vector describing the abundance of protein, mRNA, and other species in the model ($995\times 1$).
+% The stoichiometric matrix $\mathbf{S}$ encodes the signaling architecture considered in the model ($995\times 1700$).
+% Each row of $\mathbf{S}$ describes a signaling component while each column describes a particular interaction.
+% The $(i,j)$ element of $\mathbf{S}$, denoted by $\sigma_{ij}$, describes how species $i$ is involved with interaction $j$.
+% If $\sigma_{ij}>0$, species $i$ is produced by interaction $j$. Conversely, If $\sigma_{ij}<0$, then species $i$ is consumed in interaction $j$.
+% Lastly, if $\sigma_{ij}=0$, then species $i$ is not involved in interaction $j$.
+% The term  $\mathbf{r}\left(\mathbf{x},\mathbf{k}\right)$ denotes the vector of interactions rates ($1700\times{1}$).
+% We modeled each network interaction (gene expression, translation and biochemical transformations)
+% using elementary rate laws where all reversible interactions were split into two irreversible steps (supplemental materials).
+% Thus, the rate expression for interaction $q$ was given by:
+% \begin{equation}\label{eq-mass-action}
+% r_{q}\left(\mathbf{x},k_{q}\right)=k_{q}\prod_{j\in\left\{\mathbf{R}_{q}\right\}}x_{j}^{-\sigma_{jq}}
+% \end{equation}
+% The set $\left\{\mathbf{R}_{q}\right\}$ denotes reactants for reaction $q$, while $\sigma_{jq}$ denotes the stoichiometric coefficient (element of the matrix $\mathbf{S}$) governing species $j$ in reaction $q$.
 The quantity $k_{q}$ denotes the rate constant (unknown) governing reaction $q$.
-Model equations were generated in the C-programming language using the UNIVERSAL code generator, starting from an text-based input file (supplemental materials).
-UNIVERSAL, an open source Objective-C/Java code generator, is available as a Google Code project (http://code.google.com/p/universal-code-generator/).
-Model equations were solved using the CVODE solver in the SUNDIALS library \citep{Hindmarsh2005} on an Apple workstation (Apple, Cupertino, CA)
-as previously described \citep{Tasseff:2011ve}.
+% Model equations were generated in the C-programming language using the UNIVERSAL code generator, starting from an text-based input file (supplemental materials).
+% UNIVERSAL, an open source Objective-C/Java code generator, is available as a Google Code project (http://code.google.com/p/universal-code-generator/).
+% Model equations were solved using the CVODE solver in the SUNDIALS library \citep{Hindmarsh2005} on an Apple workstation (Apple, Cupertino, CA)
+% as previously described \citep{Tasseff:2011ve}.
 
@@
 using 41 data sets generated in DLD1 colon carcinoma, MDCKII, and A375 melanoma cells taken from Medici \emph{et al.} \citep{Medici:2008fk}.
-We used the Pareto Optimal Ensemble Technique (POETs) multiobjective optimization framework in combination
+We used the Pareto Optimal Ensemble Technique (JuPOETs) multiobjective optimization framework in combination
 with leave-one-out cross-validation to estimate an ensemble of model parameters \citep{Song:2010fk}.
@@
 The training and validation data were Western blots.
-We achived a physical simulation scale by establishing characeteristic rates of transcription, translation, mRNA and protein degradation, as well as characteristic concentrations of ribosomes and RNAPs ####cite bionumbers####.
+We achived a physical simulation scale by establishing characteristic rates of transcription, translation, mRNA and protein degradation, as well as characteristic concentrations of ribosomes and RNAPs \citep{Milo:2010aa}.
 The concentration scale is in nM, with proteins ranging from 10-1000nM and mRNA ranging from 0.01 to 1nM, reflecting the true abundances and ratios between each species.
 
-\subsubsection*{Robustness coefficients.} Robustness coefficients were calculated as shown previously \citep{Tasseff:2011ve,Lequieu:2011fj}.
-Robustness coefficients denoted by $\alpha\left(i,j,t_{o},t_{f}\right)$:
-\begin{equation}
-\alpha\left(i,j,t_{o},t_{f}\right)=\left(\displaystyle\int_{t_{o}}^{t_{f}}x_{i}\left(t\right)dt\right)^{-1}\left(\displaystyle\int_{t_{o}}^{t_{f}}x^{(j)}_{i}\left(t\right)dt\right)
-\end{equation}
-quantify the response of a marker to a structural or operational perturbation to the network architecture.
-Here $t_o$ and $t_f$ denote the initial and final simulation time respectively, while
-$i$ and $j$ denote the indices for the marker and the perturbation respectively.
+% \subsubsection*{Robustness coefficients.} Robustness coefficients were calculated as shown previously \citep{Tasseff:2011ve,Lequieu:2011fj}.
+% Robustness coefficients denoted by $\alpha\left(i,j,t_{o},t_{f}\right)$:
+% \begin{equation}
+% \alpha\left(i,j,t_{o},t_{f}\right)=\left(\displaystyle\int_{t_{o}}^{t_{f}}x_{i}\left(t\right)dt\right)^{-1}\left(\displaystyle\int_{t_{o}}^{t_{f}}x^{(j)}_{i}\left(t\right)dt\right)
+% \end{equation}
+% quantify the response of a marker to a structural or operational perturbation to the network architecture.
+% Here $t_o$ and $t_f$ denote the initial and final simulation time respectively, while
+% $i$ and $j$ denote the indices for the marker and the perturbation respectively.
 A value of $\alpha\left(i,j,t_{o},t_{f}\right)>1$, indicates increased marker abundance, while $\alpha\left(i,j,t_{o},t_{f}\right)<1$ indicates decreased marker abundance following perturbation $j$.
 If
-$\alpha\left(i,j,t_{o},t_{f}\right)\sim{1}$ the $j$th perturbation does not influence the abundance of marker $i$.
-Robustness coefficients were calculated for each member of the ensemble (N $\simeq$ 1100).
+% $\alpha\left(i,j,t_{o},t_{f}\right)\sim{1}$ the $j$th perturbation does not influence the abundance of marker $i$.
+% Robustness coefficients were calculated for each member of the ensemble (N $\simeq$ 1100).
 
@@
 Cells were passaged 1:3 or 1:4 every 3-6 d and used between passages 4 and 8.
-####
-DLD1 and MCF10A cells were suspended in culture media (with RPMI 1640 medium with 10\% fetal bovine serum and 1\% penicillin/streptomycin for DLD1 or MGEM 2 supplemented with insulin, bovine pituitary extract, cholera toxin, hEGF, hydrocortisone, 5\% horse serum, and 1\% penicillin/streptomycin for MCF10A), and allowed to aggregate overnight in hanging drop culture (20μL; 20,000 cells).
+
+\subsubsection*{VEGF treatment}
+DLD1 and MCF10A cells were suspended in culture media (with RPMI 1640 medium with 10\% fetal bovine serum and 1\% penicillin/streptomycin for DLD1 or MGEM 2 supplemented with insulin, bovine pituitary extract, cholera toxin, hEGF, hydrocortisone, 5\% horse serum, and 1\% penicillin/streptomycin for MCF10A), and allowed to aggregate overnight in hanging drop culture (20$\mu$L; 20,000 cells).
 The spherical aggregates were placed on the surface of neutralized type I collagen hydrogels (1.5mg/mL) and allowed to adhere.
@@
 Recombinant VEGFA165 was then added to the media (5ng/ml, 50ng/ml) and mRNA was harvested after 3hr and 24hr timepoint.
-####
 
@@
 Conversely, VEGF-A binding can stabalize an epithelial phenotype through NFAT activation.
-Downstream activation of $/beta$-catenin signaling due to E-cadherin loss and GSK3 inactivation of $/beta$-cateinin confinement is critical to the complete activation of the EMT program.
+Downstream activation of $\beta$-catenin signaling due to E-cadherin loss and GSK3 inactivation of $\beta$-cateinin confinement is critical to the complete activation of the EMT program.
 The complete list of molecular interactions that comprise the model is given in the supplement.}\label{fg:F1}
@@
 
-%\clearpage
+\clearpage
 
@@
 \includegraphics [width=1.0\linewidth] {./figs/Fig2_training.pdf}
-\caption{Training and validation simulations. The population of EMT models qualitatively captured TGF$\beta$-induced EMT signaling.
-(A-I) The population was generated using POETs and trained using 11 different objective functions (41 data sets) taken from Medici \emph{et al.} \citep{Medici:2008fk}.
+\caption{Training and validation simulations. The population of EMT models qualitatively captured TGF$-\beta$-induced EMT signaling.
+(A-I) The population was generated using JuPOETs and trained using 11 different objective functions (41 data sets) taken from Medici \emph{et al.} \citep{Medici:2008fk}.
 The model captured the simulated experiments for 72\% of the cases.
@@
 \includegraphics [width=1.0\linewidth] {./figs/Fig3_phenotype.pdf}
-\caption{Simulated VEGF-A and TGF$\beta$1/2 exposure promoted phenotype heterogeneity.
-Simulated response to TGF$\beta$1/2 and VEGF-A exposure with and without axis specific inhibitors.
+\caption{Simulated VEGF-A and TGF$-\beta$1/2 exposure promoted phenotype heterogeneity.
+Simulated response to TGF$-\beta$1/2 and VEGF-A exposure with and without axis specific inhibitors.
 Vimentin and E-cadherin abundances (in nM) were used to quantify the shift in population at 48 hrs.
 (A-C) VEGF-A (50 a.u.) treatment resulted in a population with enhanced epithelial properties.
-This was contrary to the addition of TGF$\beta$2 (10 a.u.), which shifted the population towards a mesenchymal phenotype.
-Interestingly, the combined effects of TGF$\beta$2 and VEGFA was found to increase both ecadherin and vimentin levels, creating a heterogeneous population (black arrow), which can also be seen in a minority of untreated cells (gray arrow).
+This was contrary to the addition of TGF$-\beta$2 (10 a.u.), which shifted the population towards a mesenchymal phenotype.
+Interestingly, the combined effects of TGF$-\beta$2 and VEGFA was found to increase both ecadherin and vimentin levels, creating a heterogeneous population (black arrow), which can also be seen in a minority of untreated cells (gray arrow).
 (D-F) To isolate the effect of NFAT, we inhibited NFAT de-phosphorylation in combination with VEGFA.
 This negated the increase in ecadherin expression and shifted the population towards a mesenchymal phenotype (Q1,Q3).
-Likewise, combining NFAT inhibition with TGF$\beta$ mitigated all VEGF enhanced ecadherin expression.
-Lastly, combination of TGF$\beta$2, VEGFA, and NFAT inhibition nearly mitigated all effects of VEGFA, shifting the heterogeneous population towards a mesenchymal phenotype.
+Likewise, combining NFAT inhibition with TGF$-\beta$ mitigated all VEGF enhanced ecadherin expression.
+Lastly, combination of TGF$-\beta$2, VEGFA, and NFAT inhibition nearly mitigated all effects of VEGFA, shifting the heterogeneous population towards a mesenchymal phenotype.
 In whole, high levels of phosphorylated-Sp1 correlated with vimentin expression, while NFAT was responsible for maintaining E-cadherin expression, although neither were mutually exclusive.}
+\label{fg:F3}
 \end{figure}
 
-%\clearpage
+\clearpage
 
@@
 \caption{Analysis of underlying signaling responses.
-(A) We examined the distribution of NFATc1 and AP1/SP1 in cells containing the hybrid phenotype (VEGF-A + TGF$\beta$2 case), showing the potential for cells to express both SP1 and NFATc1 in a non exclusive manner.
+(A) We examined the distribution of NFATc1 and AP1/SP1 in cells containing the hybrid phenotype (VEGF-A + TGF$-\beta$2 case), showing the potential for cells to express both SP1 and NFATc1 in a non exclusive manner.
 (B) We were able to show a fit to an additional objective demonstrating the actiation of GSK3 through PI3K.
-Our model captures this activation through TGF$\beta$3 signaling.
+Our model captures this activation through TGF$-\beta$3 signaling.
 LY294002 is a PI3K inhibitor.
@@
 Over expression of YREG1 increases the degree of E-cadherin expression, stabalizing the epithelial phenotype.
-(D) Kockout of YREG1 leads to a loss of both E-cadherin and vimentin for cells that were inherently activated in the untreated population. ####
-}\label{fg:F4}
+(D) Kockout of YREG1 leads to a loss of both E-cadherin and vimentin for cells that were inherently activated in the untreated population.}\label{fg:F4}
 \end{figure}
 
-%\clearpage
+\clearpage
 
@@
 \caption{Simultaneous TGF-$\beta$1/2 and VEGF-A treatment induced phenotype heterogeneity and is dependent upon NFAT activity \emph{in-vitro}.
-(A) In MCF10A, treatment with (10ng/ml) TGF$\beta$2 increased Slug and vimentin, while ecadherin expression was inhibited at both the gene and protein level at 48 hrs.
+(A) In MCF10A, treatment with (10ng/ml) TGF$-\beta$2 increased Slug and vimentin, while ecadherin expression was inhibited at both the gene and protein level at 48 hrs.
 Conversely, VEGFA alone increased both NFATc1 and ecadherin gene expression.
-Simultaneous TGF$\beta$2 (10ng/ml) and VEGFA (50ng/ml) treatment increased Slug, NFATc1, and vimentin expression, while also increasing ecadherin levels via qPCR.
-(B-C)  Immunofluorescence confirmed these results and nuclear co-localization of both phospho-Sp1 and NFAT were found dependent upon TGF$\beta$2 and VEGFA, respectively.
+Simultaneous TGF$-\beta$2 (10ng/ml) and VEGFA (50ng/ml) treatment increased Slug, NFATc1, and vimentin expression, while also increasing ecadherin levels via qPCR.
+(B-C)  Immunofluorescence confirmed these results and nuclear co-localization of both phospho-Sp1 and NFAT were found dependent upon TGF$-\beta$2 and VEGFA, respectively.
 (D) To isolate the effect of NFAT, treatment of VEGFA (50ng/ml) and VIVIT (10$\mu$M) reduced ecadherin expression at 48hrs (control-dashed line).
-Similarly, combined TGF$\beta$2, VEGFA and VIVIT treatment increased Slug and vimentin expression, while inhibiting ecadherin levels via qPCR.
+Similarly, combined TGF$-\beta$2, VEGFA and VIVIT treatment increased Slug and vimentin expression, while inhibiting ecadherin levels via qPCR.
 (E) These findings were confirmed via immunofluorescence as the VIVIT peptide inhibited ecadherin and nuclear localization of NFATc1 in all three cases.
@@
 Scale bars: 50$\mu$m.
-C=Control, T=TGF$\beta$2 , V=VEGFA, VI= NFAT inhibitor (VIVIT).
+C=Control, T=TGF$-\beta$2 , V=VEGFA, VI= NFAT inhibitor (VIVIT).
 Asterisks signify statistical differences from each other according to a one-way ANOVA with Tukey's post hoc (p$\prec$0.05).}\label{fg:F5}
@@
 For compaction and tubular assays, cells were embedded into collagen gels for 72 hrs, and the extent of tubulogenesis was measured at 7 days.
-(A-D) Within MCF10A, TGF$\beta$2 (10ng/ml) enhanced invasion and contractile properties while, VEGFA (50ng/ml) promoted increased migration.
-TGF$\beta$2 with VEGFA significantly increased migration, while limiting with compaction.
-VIVIT (10$\mu$M) in combination with VEGFA and TGF$\beta$2 decreased migration and compaction, while increasing invasion.
+(A-D) Within MCF10A, TGF$-\beta$2 (10ng/ml) enhanced invasion and contractile properties while, VEGFA (50ng/ml) promoted increased migration.
+TGF$-\beta$2 with VEGFA significantly increased migration, while limiting with compaction.
+VIVIT (10$\mu$M) in combination with VEGFA and TGF$-\beta$2 decreased migration and compaction, while increasing invasion.
 (D) Likewise, cell morphology (circularity index) correlated with both invasion and compaction in MCF10A.
-(E-F) The size of tubular structures (acini) also increased significantly upon addition of VEGFA, while the number of ductal branches was most significant upon simultaneous TGF$\beta$2 and VEGFA treatment (Red-Ecadherin, Green-Factin, Blue-Nuclear).
+(E-F) The size of tubular structures (acini) also increased significantly upon addition of VEGFA, while the number of ductal branches was most significant upon simultaneous TGF$-\beta$2 and VEGFA treatment (Red-Ecadherin, Green-Factin, Blue-Nuclear).
 DLD1 cells followed a similar trend, although the degree of migration, invasion, and compaction was less significant.
@@
 Scale bars: 500$\mu$m, 1000$\mu$m, 250$\mu$m, and 80$\mu$m, respectively.
-C=Control, T=TGF$\beta$2 , V=VEGFA, VI= NFAT inhibitor (VIVIT).
+C=Control, T=TGF$-\beta$2 , V=VEGFA, VI= NFAT inhibitor (VIVIT).
 Asterisks signify statistical differences from each other according to a one-way ANOVA with Tukey's post hoc (p$\prec$0.05).

### 2016-10-03 Paper_v7.tex
DB edits 3

@@
 \subsection*{Identification of a novel LEF-1 regulator}
-During model identification, we found that consistent TGF-$\beta$ induced EMT required an additional regulatory protein.
+During model identification, we found that consistent TGF-$\beta$ induced EMT from a stable epithelial cell population required an additional regulatory protein.
 This protein, which we called hypothetical regulator 1 (YREG1), was required to mediate between SNAIL/SLUG transcriptional activity
@@
 In the model, we assumed the expression of SNAIL/SLUG was likely regulated by AP1/SP1 \citep{Jackstadt:2013aa}.
-Thus, upon receiving a TGF-$\beta$1/2 signal, the model predicted enhanced SNAIL/SLUG expression, consistent with experimental observations.
+Thus, upon receiving direct Snail/Slug and TGF-$\beta$3 signals, the model predicted enhanced SNAIL/SLUG expression, consistent with experimental observations.
 TGF-$\beta$1/2 stimulation also induces LEF-1 expression. However, literature evidence suggested that LEF-1 expression was not strongly dependent upon AP1/SP1 activity \citep{Eastman:1999aa}.
@@
 Low levels of YREG1 expression were present in all simulations to regulate the formation of the $\beta$-catenin-LEF-1 complex.
-To test the potency of YREG1, we conducted over-expression and knockdown simulations following the addition of TGF-$\beta$1/2 (Fig. \ref{fg:F4}C and \ref{fg:F4}D).
-
-% TODO: REVISE -----
-% In the absence of YREG1, most of the population failed to consistently respond to TGF-$\beta$1/2 exposure compared to the wild-type (Fig. \ref{fig:S8}A).
-% Conversely, YREG1 overexpression revealed an exclusively epithelial phenotype following TGF-$\beta$1/2 stimulation (Fig. \ref{fig:S8}B).]
-% Overexpression of YREG1 repressed LEF-1 and TGF-$\beta$3 expression,
-% thereby not allowing free $\beta$-catenin to form the $\beta$-catenin-LEF-1 complex which promotes mesenchymal gene expression, or SMAD activity following from autocrine TGF-$\beta$3 signaling.
-% Taken together, we found that low YREG1 expression was necessary for stabilizing EMT,
-% while elevated YREG1 levels limited the extent of EMT induction.
+To test the effect of YREG1 on the epithelial population, we conducted over-expression and knockdown simulations on untreated cells (Fig. \ref{fg:F4}C and \ref{fg:F4}D).
+In the absence of YREG1, most of the population failed to consistently to retain a stable epithelial state (Fig. \ref{fig:F4}D).
+Conversely, YREG1 overexpression revealed an enhanced epithelial phenotype, while some inherently transformed cells moved towards a hybrid phenotype  (Fig. \ref{fig:F4}C).
+Elevation of YREG1 repressed LEF-1 and TGF-$\beta$3 expression, thereby not allowing free $\beta$-catenin to form the $\beta$-catenin-LEF-1 complex which promotes mesenchymal gene expression, or SMAD activity following from autocrine TGF-$\beta$3 signaling.
+Taken together, we found that low YREG1 expression was necessary for producing a stabile the model with an epithelial phentoype that is inducable across multiple stimuli such as TGF-$\beta$1/2, TGF-$\beta$3 and SNAIl/SLUG transfection, as seen in the training objectives.
 
@@
 Analysis of these hypothetical cells suggested they had \emph{abnormal} signaling; deregulated NFAT expression and nuclear localization promoted E-cadherin expression while TGF$-\beta$1/2 induced Sp1 action promoted Vimentin expression.
-
-% ####Analysis of the connectivity and information flow through the signaling architecture suggested that Sp1 and NFAT action could be manipulated \emph{independently} by simultaneous TGF$-\beta$1/2 and VEGF-A stimulation (Fig.
-\ref{fg:S1}).
 
@@
 This initial repression of E-cadherin leads to a release of $\beta$-catenin from the cell membrane.
-Cystolic $\beta$-catenin can then translocate to the nucleus and form transcriptional complexes with TCF-4 to drive TGF$-\beta$3 expression \citep{Medici:2008fk}.
-TGF$-\beta$3 signals to the cells interior by binding to type II receptors, which form heterodimers with type I receptors (ALK5) \citep{Derynck:2003fc}.
+This release of $\beta$-catenin can then translocate to the nucleus and form transcriptional complexes with TCF-4 to drive TGF$-\beta$3 expression \citep{Medici:2008fk}.
+The PI3K to GSK3 pathway was included and acted as an activating mechanism of $\beta$-catenin signaling through TGF$-\beta$3 signaling \c\
+itep{Medici:2008fk}. GSK3 is known to act on $\beta$-catenin signaling through the ubiquitin-proteasome pathway
+\citep{Zhou:2004kx,Larue:2005vn}.
+Thereby, further $\beta$-catenin release also resulted from by TGF$-\beta$3 signals to the cells interior by binding to type II receptors, which form heterodimers with type I receptors (ALK5) \citep{Derynck:2003fc}.
 This activates the receptors serine/threonine kinase activity to phosphorylate and activate the receptor Smads 2/3 \citep{Massague:2005qc}.
+In the model, receptors are represented as either bound or unbound complexes.
 Phosphorylated Smads 2/3 (pSmad2/3) form heterodimers and translocate to the nucleus.
@@
 GSK3 is known to act on $\beta$-catenin signaling through the ubiquitin-proteasome pathway
-\citep{Zhou:2004kx,Larue:2005vn}.
-VEGF-A activation of NFATc1 takes place through calcineurin signaling leading to an enhancement of E-cadherin expression \citep{Suehiro:2014aa},
+\citep{Zhou:2004kx,Larue:2005vn}. Lastly, VEGF-A activation of NFATc1 takes place through calcineurin signaling leading to an enhancement of E-cadherin expression \citep{Suehiro:2014aa},
 as supported by our VEGF-A experimental data (Fig. \ref{fg:VEGFA-Data}).
@@
 
-\begin{figure}
+\begin{figurey}
 \includegraphics [width=1.0\linewidth] {./figs/Fig4_mechanism.pdf}
@@
 LY294002 is a PI3K inhibitor.
-(C) We identified a novel regulator of LEF1 called YREG1 that allows Snail/Slug to emulate an inducer by repressing YREG1.
-Over expression of YREG1 increases the degree of E-cadherin expression, stabalizing the epithelial phenotype.
-(D) Kockout of YREG1 leads to a loss of both E-cadherin and vimentin for cells that were inherently activated in the untreated population.}\label{fg:F4}
+(C) We identified a novel regulator of LEF1 called YREG1 that allows Snail/Slug to emulate an inducer by repressing YREG1, which was required to stabalize the untreated population. 
+YREG1 overexpression revealed an enhanced epithelial phenotype, while some inherently transformed cells moved towards a hybrid phenotype.
+(D) In the absence of YREG1, most of the population failed to consistently to retain a stable epithelial state.}\label{fg:F4}
 \end{figure}

### 2016-10-04 Paper_v7.tex
Fixing spelling, adding some polish

@@
 In the model, we assumed the expression of SNAIL/SLUG was likely regulated by AP1/SP1 \citep{Jackstadt:2013aa}.
-Thus, upon receiving direct Snail/Slug and TGF-$\beta$3 signals, the model predicted enhanced SNAIL/SLUG expression, consistent with experimental observations.
+Thus, upon receiving direct SNAIL/SLUG and TGF-$\beta$3 signals, the model predicted enhanced SNAIL/SLUG expression, consistent with experimental observations.
 TGF-$\beta$1/2 stimulation also induces LEF-1 expression. However, literature evidence suggested that LEF-1 expression was not strongly dependent upon AP1/SP1 activity \citep{Eastman:1999aa}.
 Thus, either SNAIL/SLUG are acting as inducers (contrary to substantial biochemical evidence) or, they are repressing the expression of an intermediate repressor.
-Given the biochemical evidence supporting SNAIL/SLUG as repressors, we created YREG1 a hypothetical intermediate repressor whose
+Given the biochemical evidence supporting SNAIL/SLUG as repressors, we created the hypothetical YREG1 repressor whose
 expression is downregulated by SNAIL/SLUG. The literature data therefore suggested that YREG1 had two transcriptional targets, LEF-1 and TGF-$\beta$3.
@@
 To test the effect of YREG1 on the epithelial population, we conducted over-expression and knockdown simulations on untreated cells (Fig. \ref{fg:F4}C and \ref{fg:F4}D).
-In the absence of YREG1, most of the population failed to consistently to retain a stable epithelial state (Fig. \ref{fig:F4}D).
-Conversely, YREG1 overexpression revealed an enhanced epithelial phenotype, while some inherently transformed cells moved towards a hybrid phenotype  (Fig. \ref{fig:F4}C).
-Elevation of YREG1 repressed LEF-1 and TGF-$\beta$3 expression, thereby not allowing free $\beta$-catenin to form the $\beta$-catenin-LEF-1 complex which promotes mesenchymal gene expression, or SMAD activity following from autocrine TGF-$\beta$3 signaling.
-Taken together, we found that low YREG1 expression was necessary for producing a stabile the model with an epithelial phentoype that is inducable across multiple stimuli such as TGF-$\beta$1/2, TGF-$\beta$3 and SNAIl/SLUG transfection, as seen in the training objectives.
+In the absence of YREG1, the population of models failed to consistently to retain a stable epithelial state (Fig. \ref{fg:F4}D).
+Conversely, YREG1 amplification revealed an enhanced epithelial phenotype, while some inherently transformed cells moved towards a hybrid phenotype  (Fig. \ref{fg:F4}C).
+Elevated YREG1 repressed LEF-1 and TGF-$\beta$3 expression, thereby not allowing free $\beta$-catenin to form the $\beta$-catenin-LEF-1 complex, or TGF-$\beta$3 induced SMAD activation.
+Taken together, low YREG1 expression was required for the maintenance of a stable epithelial phenotype that was simultaneously inducible across TGF-$\beta$1/2, TGF-$\beta$3 and SNAIl/SLUG transfection, as seen in the training objectives.
 
@@
 feature of developmental and pathological programs is population heterogeneity \citep{Park:2010fu}.
-We (and others) previously hypothesized that deterministic model ensembles can interrogate population behavior, at least at a course grained level \citep{Lequieu:2011fj}.
+We (and others) have previously hypothesized that deterministic model ensembles can simulate population behavior, at least at a course grained level \citep{Lequieu:2011fj}.
 We tested this hypothesis by analyzing the response of the population of EMT models to extracellular cues and then comparing this response to flow cytometry studies.
-We quantiied the phenotypic response of the individual members of the ensemble to TGF$-\beta$1/2 stimulation as a function of physical measurements of abundance of markers in system.
-We evlautated these quantities for each member of the ensemble (N $\simeq$ 1400) for two downstream phenotypic markers, Vimentin (mesenchymal) and E-cadherin (epithelial)
-following the addition of TGF$-\beta$1/2 alone (Fig. \ref{fg:F3}), and VEGF-A in combination with NFATc inhibitors (Fig. \ref{fg:F3}).
-We were further able to use this analysis to examine the activity and effects of underlying signaling molecules.
+We quantified the phenotypic response of the individual members of the ensemble to TGF$-\beta$1/2 stimulation for two downstream phenotypic markers, Vimentin (mesenchymal) and E-cadherin (epithelial) following the addition of TGF$-\beta$1/2 alone (Fig.
+\ref{fg:F3}), and VEGF-A in combination with NFATc inhibitors (Fig.
+\ref{fg:F3}).
+%We were further able to use this analysis to examine the activity and effects of underlying signaling molecules.
 
@@
 Analysis of the molecular signatures of these subpopulations suggested the abundance, localization and state of the Sp1, AP-1 and NFATc transcription factors controlled population heterogeneity.
-The behavior of the majority of models ($>$80\%) respond to treatment in Fig \ref{fg:F3}A-C, moving off of the untreated population set (gray).
+The majority of models ($>$80\%) responded to treatment, moving off of the untreated population (Fig \ref{fg:F3}A-C, gray).
 These models showed the classically expected behavior, a switch from an epithelial to mesenchymal phenotype following TGF$-\beta$1/2 exposure.
@@
 Elevated Sp1 activity decreased E-cadherin expression through Slug-mediated inhibition, which in turn increased Vimentin expression through TGF$-\beta$3 autocrine signaling and the liberation of $\beta$-catenin.
-
 However, the most biologically interesting behavior was exhibited by cells achieving a hybrid phenotype, most notable in a dual treatment condition (black arrow), but also present in a small percentage of untreated cells (Fig.
@@
 Models with this hybrid phenotype had elevated Sp1 and NFAT transcriptional activity, resulting in the simultaneous increase of \emph{both} Vimentin and E-cadherin expression (Fig. \ref{fg:F4}A).
-Analysis of these hypothetical cells suggested they had \emph{abnormal} signaling; deregulated NFAT expression and nuclear localization promoted E-cadherin expression while TGF$-\beta$1/2 induced Sp1 action promoted Vimentin expression.
+Analysis of these hypothetical cells suggested they had deregulated NFAT expression and nuclear localization promoted E-cadherin expression while TGF$-\beta$1/2 induced Sp1 action promoted Vimentin expression.
 
@@
 
-\begin{figurey}
+\begin{figure}
 \includegraphics [width=1.0\linewidth] {./figs/Fig4_mechanism.pdf}

### 2016-10-04 Paper_v7.tex
DB edits 4

@@
 The predictive power of the ensemble was tested using both cross validation and by comparing simulations with data sets not used for model training.
-In whole, 72\% of our training objectives were statistically significant (at a 95\% confidence interval) compared to a randomized parameter family (N = 100)
-generated from a random starting point. Conversely, we \emph{predicted} approximately 72\% of the training objectives, at a 95\% confidence interval compared to randomized parameters.
+In whole, 100\% of our training objectives were statistically significant (at a 95\% confidence interval) compared to a randomized parameter family (N = 100)
+generated from a random starting point. Conversely, we \emph{predicted} 100\% of the training objectives, at a 95\% confidence interval compared to randomized parameters (Wicoxon non-parametric test).
 The model also captured the temporal gene expression responses of E-cadherin, pSmad2, and LEF-1 (not used for model training) to within one-standard deviation (up to the 48 hr time-point) (Fig.
@@
 We quantified the phenotypic response of the individual members of the ensemble to TGF$-\beta$1/2 stimulation for two downstream phenotypic markers, Vimentin (mesenchymal) and E-cadherin (epithelial) following the addition of TGF$-\beta$1/2 alone (Fig.
-\ref{fg:F3}), and VEGF-A in combination with NFATc inhibitors (Fig.
+\ref{fg:F3}), and/or VEGF-A in combination with NFATc inhibitors (Fig.
 \ref{fg:F3}).
@@
 
-We identified model subpopulations that exhibited different behaviors following exposure to TGF$-\beta$1/2 (Fig. \ref{fg:F3}A).
+We identified model subpopulations that exhibited different behaviors following exposure to TGF$-\beta$1/2 (Fig. \ref{fg:F3}B).
 Analysis of the molecular signatures of these subpopulations suggested the abundance, localization and state of the Sp1, AP-1 and NFATc transcription factors controlled population heterogeneity.
-The majority of models ($>$80\%) responded to treatment, moving off of the untreated population (Fig \ref{fg:F3}A-C, gray).
+The majority of models ($>$80\%) responded to treatment, moving away from the untreated population (Fig \ref{fg:F3}A-C, gray).
 These models showed the classically expected behavior, a switch from an epithelial to mesenchymal phenotype following TGF$-\beta$1/2 exposure.
-Models near resembling untreated cells had evaluated nuclear localized phosphorylated Sp1, relative to non-induced cells.
-Elevated Sp1 activity decreased E-cadherin expression through Slug-mediated inhibition, which in turn increased Vimentin expression through TGF$-\beta$3 autocrine signaling and the liberation of $\beta$-catenin.
-However, the most biologically interesting behavior was exhibited by cells achieving a hybrid phenotype, most notable in a dual treatment condition (black arrow), but also present in a small percentage of untreated cells (Fig.
+Some models resembled untreated cells; they had evaluated nuclear localized phosphorylated Sp1, relative to non-induced cells,
+which decreased E-cadherin expression through Slug-mediated inhibition, which in turn increased Vimentin expression through TGF$-\beta$3 autocrine signaling and the liberation of $\beta$-catenin.
+However, the most biologically interesting behavior was exhibited by cells achieving a hybrid phenotype, most notable in a dual treatment condition (\ref{fg:F3}C, black arrow), but also present in a small percentage of untreated cells (Fig.
 \ref{fg:F3}A).
-Models with this hybrid phenotype had elevated Sp1 and NFAT transcriptional activity, resulting in the simultaneous increase of \emph{both} Vimentin and E-cadherin expression (Fig. \ref{fg:F4}A).
-Analysis of these hypothetical cells suggested they had deregulated NFAT expression and nuclear localization promoted E-cadherin expression while TGF$-\beta$1/2 induced Sp1 action promoted Vimentin expression.
+Models with this hybrid phenotype had elevated Sp1 and NFAT transcriptional activity, resulting in simultaneously increased Vimentin and E-cadherin expression (Fig. \ref{fg:F4}A).
 
-To test this hypothesis, we simulated the response of the network to TGF$-\beta$1/2 and VEGF-A treatment with and without NFATc inhibitors (Fig. \ref{fg:F4}).
-As expected, stimulation with VEGF-A (50 a.u.) maintained an epithelial population, while TGF$-\beta$1/2 (10 a.u.) exposure shifted
-the population from an epithelial to a mesenchymal phenotype (Fig. \ref{fg:F3}A and Fig. \ref{fg:F3}B).
+%Analysis of these hypothetical cells suggested they had deregulated NFAT expression and nuclear localization promoted E-cadherin expression while TGF$-\beta$1/2 induced Sp1 action promoted Vimentin expression.
+
+To better understand the hybrid phenotype, we simulated the response of the model population to TGF$-\beta$1/2 and VEGF-A treatment with and without NFATc inhibitors (Fig. \ref{fg:F3}).
+As expected, stimulation with VEGF-A (50 a.u.) maintained an epithelial population (Fig. \ref{fg:F3}A), while TGF$-\beta$1/2 (10 a.u.) exposure shifted
+the population from an epithelial to a mesenchymal phenotype (Fig. \ref{fg:F3}B).
 On the other hand, combined stimulation with TGF$-\beta$1/2 (10 a.u.) and VEGF-A (50 a.u.) increased both E-cadherin and Vimentin expression, resulting in a hybrid phenotype with both epithelial and mesenchymal characteristics (Fig.
@@
 To further isolate the role of NFAT on this hybrid state, we simulated the inhibition of NFAT transcriptional activity across all conditions (all else being equal).
-NFAT inhibition in combination with VEGF-A and TGF$-\beta$1/2 treatments blocks increased activation of E-cadherin (Fig. \ref{fg:F3}D and Fig. \ref{fg:F3}E).
+NFAT inhibition in combination with VEGF-A or TGF$-\beta$1/2 treatments blocked increased E-cadherin expression in the case of VEGF-A (Fig.
+\ref{fg:F3}D), but did not influence TGF$-\beta$1/2 signaling (Fig.
+\ref{fg:F3}E).
 Lastly, NFATc inhibition in combination with simultaneous TGF$-\beta$1/2 and VEGF-A exposure repressed nearly all E-cadherin expression, shifting nearly the entire population towards a mesenchymal phenotype (Fig.
-\ref{fg:F4}F).
-Taken together, high levels of nuclear localized phosphorylated Sp1 correlated with Vimentin expression, while NFATc transcriptional activity was predicted to be critical for maintaining E-cadherin expression in the prescence of competing signals.
-In particular, when present, VEGF-A signaling through NFATc was predicted to provide a protective up-regulation of E-cadherin.
+\ref{fg:F3}F).
+Taken together, high levels of nuclear localized phosphorylated Sp1 correlated with Vimentin expression,
+while NFATc transcriptional activity was critical for maintaining E-cadherin expression in the presence of competing signals.
+
+%In particular, when present, VEGF-A signaling through NFATc was predicted to provide a protective up-regulation of E-cadherin.
 
@@
 \section*{Materials and Methods}
-All model code and parameter ensemble is freely available under an MIT software license and can be downloaded from http://www.varnerlab.org.
+The model code and parameter ensemble is freely available under an MIT software license and can be downloaded from http://www.varnerlab.org.
 
@@
 Binding of extracellular TGF$-\beta$1/2 with TGF$-\beta$ surface receptors I/II (TGF$-\beta$R-I/II) initiates the assembly of adapter complexes which starts the
-downstream signaling program. In the model, TGF$-\beta$1/2 binds TGF$-\beta$R-I/II followed by the recruitment of activin receptor-like kinase 1 (ALK1) and
-TGF$-\beta$ surface receptor III (TGF$-\beta$R-III) to form the activated receptor complex \citep{Derynck:2003fc}.
-Alternatively, we also included activin receptor-like kinase 5 (ALK5) recruitment in combination with Endoglin and TGF$-\beta$R-III
-as a second (redundant) activated receptor complex \citep{Gatza:2010if}.
+downstream signaling program.
 Complex assembly activates the serine/threonine kinase activity on the receptor, leading to the recruitment and phosphorylation of Smad partners \citep{Massague:2005qc}.
 Phosphorylated Smads2/3 (pSmad2/3) form heterodimers with partner Smad4 and then translocate to the nucleus where they act as both transcriptional activators and repressors.
-Nuclear pSmad2/3-Smad4 form transcriptional complexes with several genes in the model including lymphoid enhancer-binding factor 1 (\emph{LEF-1}), Nuclear factor of activated T-cells, cytoplasmic 1 (\emph{NFACT1}), and Specificity Protein 1 (\emph{SP1}).
-On the other hand, nuclear pSmad2/3-Smad4 represses (in combination with the LEF-1 protein) the expression of E-cadherin (\emph{Cdh1}) \citep{Nawshad:2007kl} and Cadherin 5, type 2 (VE-Cadherin encoded by \emph{Cdh5}).
 Repression of E-cadherin expression is the central event in the transition from an epithelial to a mesenchymal phenotype \citep{Cano:2000kh}.
@@
 The PI3K to GSK3 pathway was included and acted as an activating mechanism of $\beta$-catenin signaling through TGF$-\beta$3 signaling \citep{Medici:2008fk}.
-GSK3 is known to act on $\beta$-catenin signaling through the ubiquitin-proteasome pathway
+GSK3 is known to act on $\beta$-catenin signaling through the ubiquitin-proteasome pathway.
 \citep{Zhou:2004kx,Larue:2005vn}. Lastly, VEGF-A activation of NFATc1 takes place through calcineurin signaling leading to an enhancement of E-cadherin expression \citep{Suehiro:2014aa},

### 2016-10-04 Supplemental-JB.tex
DB edits 4

@@
 \section*{Supplemental Materials and Methods}
-\subsection*{EMT model network architecture.} TGF$\beta$ is a major inducer of EMT in development, fibrosis, and carcinogenesis with different isoforms mediating various effects depending on specific cellular context \cite{Nawshad:2005pi}.
-TGF$\beta$1 was first described as an inducer of EMT in normal mammary epithelial cells \cite{Miettinen1994} and has since been shown to mediate EMT in vitro in a number of different epithelial cells, including renal proximal tubular, DLD1 colon carcinoma, and most recently alveolar epithelial cells \cite{Fan1999, Hales1994,Kasai2005,Willis2005}.
-TGF$\beta$ signaling occurs through the Smad pathway in which signals are transduced by transmembrane serine/threonine kinase type I (ALK5) and type II (TGF$\beta$RII) receptors.
-To increase ligand affinity, betaglycan (TGF$\beta$RIII) can also interact with TGF$\beta$RI,II \cite{Gatza:2010if}.
-Upon TGF$\beta$ stimulation, the receptors are internalized into early endosomes where Smad anchor for receptor activation (SARA) modulates formation of complexes with (R-Smad) Smad2 or Smad3.
-Smad2 and Smad3 are then phosphorylated at serine residues by the type I receptor \cite{Massague:2005qc}.
-Phosphorylation induces their association with (Co-Smad) Smad4 and translocation to the nucleus where they interact with other transcription factors to regulate the transcription of TGF$\beta$ responsive genes, including alpha-smooth muscle actin, collagen1A2, vimentin, fibronectin, and plasminogen activator inhibitor-1 (PAI-1) by interacting with Smad-binding elements \cite{Dennler:2002hl,Derynck:2003fc}.
-To regulate TGF$\beta$ signaling, Smurf2 (a ubiquitin E3 ligase) can become activated to mediate proteasome dependent degradation of Smad2 or bind with Smad 7 to target TGF$\beta$ receptor for degradation \cite{Kavsak2000,Bonni2001}.
 
-\subsubsection*{Cell Type Dependency} Interestingly, differential roles for Smad2 and Smad3 in TGF$\beta$ induced EMT have been demonstrated.
-For example, using primary cells from mice with hepatocyte-specific double knockout of Smad2 and Smad3, it was demonstrated that Smad3 but not Smad2 was required for a key morphological changes and induction of EMT \cite{Ju:2006so}.
-In contrast, using human proximal tubular epithelial cells, increased colony growth factor and decreased E-cadherin were Smad3 dependent, increased MMP-2 was Smad2 dependent, while alpha-smooth muscle actin was dependent on both \cite{Phanish2006}.
-Furthermore, Smad2 signaling has been demonstrated in cancerous lung epithelial A549 cells \cite{Kasai2005}.
-Inhibition of Smad3 signaling decreases the metastatic potential of xenografted breast cancer cell lines \cite{Tian2003,Tian2004}.
-Together, these results suggest that the precise Smad pathway activated may depend on the particular cellular context.
-Regardless, a recent transcriptomic analysis of TGF$\beta$ induced EMT in normal mouse and human epithelial cells using a dominant negative approach demonstrated that Smad signaling was critical for regulation of all tested target genes \cite{Valcourt2005}.
+\subsection*{Estimation and cross-validation of EMT model parameters.}
+We used the Pareto Optimal Ensemble Technique (POETs) multiobjective optimization framework in combination with leave-one-out cross-validation to estimate an ensemble of TGF$-\beta$/EMT models.
+Cross-validation was used to calculate both training and prediction error during the parameter estimation procedure \citep{kohavi1995study}.
+The 41 intracellular protein and mRNA data-sets used for identification were organized into 11 objective functions.
+These 11 objective functions were then partitioned, where each partition contained ten training objectives and one validation objective.
+POETs integrates standard search strategies e.g., Simulated Annealing (SA) or Pattern Search (PS)
+with a Pareto-rank fitness assignment \citep{Song:2010fk,JuPOETs-BioArXiv}.
+Denote a candidate parameter set at iteration $i+1$ as $\mathbf{k}_{i+1}$.
+The squared error for $\mathbf{k}_{i+1}$ for training set $j$ was defined as:
+\begin{equation}\label{eqn_cost2}
+	E_{j}(\mathbf{k}) = \sum_{i=1}^{\mathcal{T}_{j}}\biggl(\hat{\mathcal{M}}_{ij}-\hat{y}_{ij}(\mathbf{k})\biggr)^2
+\end{equation}
+The symbol $\hat{\mathcal{M}}_{ij}$ denotes scaled experimental observations (from training set $j$) while $\hat{y}_{ij}$ denotes the scaled simulation output (from training set $j$).
+The quantity $i$ denotes the sampled time-index and $\mathcal{T}_{j}$ denotes the number of time points for experiment $j$.
+In this study, the experimental data used for model training was typically the band intensity from Western or Northern blots.
+Band intensity was estimated using the ImageJ software package \cite{IMAGEJ}.
+The scaled measurement for species $x$ at time $i=\{t_{1},t_{2},..,t_{n}\}$ in condition $j$ is given by:
+\begin{equation}\label{norm_exp_data}
+\hat{\mathcal{M}}_{ij} = \frac{\mathcal{M}_{ij} - \min_{i}\mathcal{M}_{ij}}{\max_{i}{\mathcal{M}_{ij}}-\min_{i}{\mathcal{M}_{ij}}}
+\end{equation}
+Under this scaling, the lowest intensity band equaled zero while the highest intensity band equaled one.
+A similar scaling was defined for the simulation output.
+By doing this scaling, we trained the model on the relative change in blot intensity, over conditions or time (depending upon the experiment).
+Thus, when using multiple data sets (possibly from different sources) that were qualitatively similar but quantitatively different e.g., slightly different blot intensities over time or condition, we captured the underlying trends in the scaled data.
 
-\subsubsection*{Smad-Independent EMT Induction Pathways.
+% Additionally, we applied a universal convention of identifying the no expression case as protein value below $<$10$^{-3}$.
+This is similar to previously published models from our lab.
+In our studies, Figure 2 identifies experimental data extracted from published Western blots and our simulation results.
+It is clear from these that 1) the training data included a variety of Western blot data treatments that were effectively zero, and 2) our simulations matched the training data virtually perfectly over time and across multiple biological species.
+These results validate the power of our simulation scheme, which necessarily includes numerical interpretations of zero.
+In our Supplemental Figure S9, we present the raw concentrations of our simulations (previously validated from the training data) over hundreds of parameter ensembles, with no “zero” threshold applied.
+While these values do not have a threshold, we only interpret values below 10$^{-3}$ equivalent to zero.
+Likewise, our robustness coefficients (Figure 3) (which represent the ratio of integrated areas of the treatment effect over the baseline effect) identify no difference in model output for vimentin (or e-cadherin for that matter) less than 10$^{-3}$, confirming this interpretation.
+
+\begin{figure}
+	\center
+	\includegraphics [width=1.0\linewidth] {./figs/Fig-Supplemental-Error-Table.pdf}
+	\caption{Training and prediction values as a function of condition for the 11 TGF-$\beta$ objective functions versus a random parameter control.}\label{fg:ObjTable}
+\end{figure}
+
+% We computed the Pareto rank of $\mathbf{k}_{i+1}$ by comparing the simulation error at iteration $i+1$ against an archive of accepted parameter sets $\mathbf{K}_{i}$.
+% We used the Fonseca and Fleming ranking scheme \cite{fonseca1993genetic} to estimate the number of parameter sets in the archive that dominate
+% $\mathbf{k}_{i+1}$. Parameter sets with increasing rank were progressively further away from the optimal trade-off surface.
+% The parameter set $\mathbf{k}_{i+1}$ was accepted or rejected by POETs with probability $\mathcal{P}\left(\mathbf{k}_{i+1}\right)$:
+% \begin{equation}\label{eqn_costMOSA}
+% \mathcal{P}(\mathbf{k}_{i+1}) \equiv \exp{\{-rank\left(\mathbf{k}_{i+1} \mid \mathbf{K}_{i} \right)/T\}}
+% \end{equation}
+% where $T$ is the annealing temperature and $rank\left(\mathbf{k}_{i+1} \mid \mathbf{K}_{i} \right)$ denotes the Pareto rank for $\mathbf{k}_{i+1}$.
+% The annealing temperature was discretized into 10 quanta between $T_{o}$ and $T_{f}$ and adjusted according to the schedule
+% $T_k = \beta^k T_0$ where $\beta$ was defined as $\left({T_{f}}/{T_{o}}\right)^{1/10}$.
+% The initial temperature was $T_{o} = \mathit{n}/log(2)$, where $n = 4$ in this study and the final temperature was $T_{f} = 0.1$.
+% The epoch-counter $k$ was incremented after the addition of 100 members to the ensemble.
+% Thus, as the ensemble grew, the likelihood of accepting parameter sets with a large Pareto rank decreased.
+% To generate parameter diversity, we randomly perturbed each parameter by $\leq\pm25\%$ at iteration of the search.
+% In addition, we performed a local pattern search every $q$-iterations to minimize the residual for a single random or the worst performing objective function.
+% The local pattern-search algorithm has been described previously \cite{gadkar2003cybernetic}.
+% From the 15,000 probable EMT models, we selected N = 1093 models with Pareto rank $\leq{1}$ for subsequent analysis.
+% A quick estimate of the set to set correlation showed that we could expect on order 25\% correlation between parameter sets in the ensemble.
+
+% \begin{figure}
+% \includegraphics [width=1.0\linewidth] {./figs/Fig-5-Supplemental-ErrorTable.pdf}
+% \caption{Training and prediction values for the 11 TGF-$\beta$ objective functions versus a random parameter control.}\label{fig:S5}
+% \end{figure}
+
+% \begin{figure}
+% \includegraphics [width=1.0\linewidth] {./figs/Fig-7-Supplemental-Correlation.pdf}
+% \caption{Parameter set to set correlation for 55 random parameter sets selected from the ensemble.
+% Of the 55 sets selected, the average correlation between sets was less than 25\% for greater than 80\% of the parameter sets.}\label{fig:S7}
+% \end{figure}
+
+% \begin{figure}
+% \includegraphics [width=1.0\linewidth] {./figs/Fig-8-Supplemental-YREG1.pdf}
+% \caption{Robustness of E-cadherin and Vimentin expression to a knockout (A) and overexpression (B) of the hypothetical regulator 1 (YREG1) protein.
+% Robustness coefficients were calculated for each member of the ensemble. Each point represents the response of a single model in the ensemble to either a knockout or
+% overexpression of YREG1.}\label{fig:S8}
+% \end{figure}
+
+% \begin{figure}
+% \includegraphics [width=1.0\linewidth] {./figs/Fig-9-Supplemental-RawExpression_v2.pdf}
+% \caption{Steady state protein abundance for E-cadherin and Vimentin (AU) as a function of TGF-$\beta$1/2 exposure.
+% Top: Overlay of the model population for Vimentin (AU) and E-cadherin (AU) expression in the presence (blue) and absence (gray) of TGF-$\beta$1/2.
+% Midddle: Vimentin (AU) and E-cadherin (AU) expression in the absence of TGF-$\beta$1/2 showed exhinibted population heterogenity.
+% Bottom: Vimentin (AU) and E-cadherin (AU) expression in the presence of TGF-$\beta$1/2 moved the centriod of the population toward Vimentin (AU) and away from E-cadherin (AU) expression.}\label{fig:S9}
+% \end{figure}
+
+% \subsection*{EMT model network architecture.} TGF$\beta$ is a major inducer of EMT in development, fibrosis, and carcinogenesis with different isoforms mediating various effects depending on specific cellular context \cite{Nawshad:2005pi}.
+% TGF$\beta$1 was first described as an inducer of EMT in normal mammary epithelial cells \cite{Miettinen1994} and has since been shown to mediate EMT in vitro in a number of different epithelial cells, including renal proximal tubular, DLD1 colon carcinoma, and most recently alveolar epithelial cells \cite{Fan1999, Hales1994,Kasai2005,Willis2005}.
+% TGF$\beta$ signaling occurs through the Smad pathway in which signals are transduced by transmembrane serine/threonine kinase type I (ALK5) and type II (TGF$\beta$RII) receptors.
+% To increase ligand affinity, betaglycan (TGF$\beta$RIII) can also interact with TGF$\beta$RI,II \cite{Gatza:2010if}.
+% Upon TGF$\beta$ stimulation, the receptors are internalized into early endosomes where Smad anchor for receptor activation (SARA) modulates formation of complexes with (R-Smad) Smad2 or Smad3.
+% Smad2 and Smad3 are then phosphorylated at serine residues by the type I receptor \cite{Massague:2005qc}.
+% Phosphorylation induces their association with (Co-Smad) Smad4 and translocation to the nucleus where they interact with other transcription factors to regulate the transcription of TGF$\beta$ responsive genes, including alpha-smooth muscle actin, collagen1A2, vimentin, fibronectin, and plasminogen activator inhibitor-1 (PAI-1) by interacting with Smad-binding elements \cite{Dennler:2002hl,Derynck:2003fc}.
+% To regulate TGF$\beta$ signaling, Smurf2 (a ubiquitin E3 ligase) can become activated to mediate proteasome dependent degradation of Smad2 or bind with Smad 7 to target TGF$\beta$ receptor for degradation \cite{Kavsak2000,Bonni2001}.
+%
+% \subsubsection*{Cell Type Dependency} Interestingly, differential roles for Smad2 and Smad3 in TGF$\beta$ induced EMT have been demonstrated.
+% For example, using primary cells from mice with hepatocyte-specific double knockout of Smad2 and Smad3, it was demonstrated that Smad3 but not Smad2 was required for a key morphological changes and induction of EMT \cite{Ju:2006so}.
+% In contrast, using human proximal tubular epithelial cells, increased colony growth factor and decreased E-cadherin were Smad3 dependent, increased MMP-2 was Smad2 dependent, while alpha-smooth muscle actin was dependent on both \cite{Phanish2006}.
+% Furthermore, Smad2 signaling has been demonstrated in cancerous lung epithelial A549 cells \cite{Kasai2005}.
+% Inhibition of Smad3 signaling decreases the metastatic potential of xenografted breast cancer cell lines \cite{Tian2003,Tian2004}.
+% Together, these results suggest that the precise Smad pathway activated may depend on the particular cellular context.
+% Regardless, a recent transcriptomic analysis of TGF$\beta$ induced EMT in normal mouse and human epithelial cells using a dominant negative approach demonstrated that Smad signaling was critical for regulation of all tested target genes \cite{Valcourt2005}.
+%
+% \subsubsection*{Smad-Independent EMT Induction Pathways.
 }Although less well established than the Smad-dependent pathways in the induction of EMT, there is substantial evidence for TGF$\beta$ activation of Smad-independent signaling in some aspects of this process.
-These include RhoA, MAPK, PI3 kinase, and Notch signaling pathway, which have mostly been studied in-vitro.
-For example, the small GTPase RhoA is involved in TGF$\beta$ induced EMT in a number of cell types including NMuMG mammary epithelial cells and mink lung epithelial (Mv1Lu) cells particularly in the regulation of cytoskeletal and adherens junction rearrangement.
-In addition to its role in cytoskeletal remodeling, Rho has been shown to activate the alpha-smooth muscle promoter during EMT in kidney proximal tubular (LLCPK1) cells \cite{Masszi:2003hb}.
-TGF$\beta$ has been shown to elevate ERK activity in culture models of EMT (human keratinocytes, NMuMG mammary epithelial cells, and mouse cortical tubule epithelial cells) \cite{Zavadil2001,Xie2004,Davies2005}.
-This ERK activity was required for disassembly of adherens junctions and induction of cell motility \cite{Xie2004}.
-Inhibition of ERK blocked these key morphological changes of EMT in mammary gland epithelial cells \cite{Xie2003}.
-Recent studies in MDCKII and DLD1 colon carcinoma cells revealed that TGF$\beta$3 can signal through PI3 kinase.
-Upon activation, PI3 kinase activates molecules such as ILK and AKT, which can phosphorylate and inactivate GSK3$\beta$, a protein that targets both Snail and $\beta$-catenin for degradation through the ubiquitin proteosome pathway \cite{Medici:2008fk}.
-Lastly, there is increasing evidence for a role of Notch pathways in regulating EMT, particularly during development \cite{Nawshad:2005pi}.
-In cardiac cushion EMT, it was found that that Slug is directly up-regulated by Notch in endothelial cells repression of the vascular endothelial cadherin promoter and for promoting migration of transformed endothelial cells.
-In whole, the distinction between Smad-dependent and Smad-independent mechanisms remains difficult to interpret due to the significant cross talk between these pathways.
-In most cases, stimulation of these cooperative pathways provides the context for induction and specification of EMT within a particular tissue/cell type, with Smads representing the dominant pathway, which in some instances may be necessary but not sufficient for induction of full EMT \cite{Zavadil:2005fu}.
-
-\subsubsection*{Master Transcription Regulators}
-Master transcription factors of EMT are thought to be regulated by Snail1 (SNAI1) and Snail2 (SNAI2) (previously known as Snail and Slug, respectively).
-These transcription factors are zinc finger proteins that function as repressors of E-cadherin transcription in cultured epithelial cells \cite{Cano:2000kh}, repression of E-cadherin leads to dissolution of adherens junctions.
-Snail and Slug can be activated by TGF$\beta$ via both Smad-dependent and -independent pathways in a cell type dependent fashion in cultured cells \cite{Peinado2003}.
-Differential expression of Snail and Slug is observed in TGF$\beta$ induced EMT in keratinocytes, renal proximal tubular, and mammary epithelial cells, suggesting that they are regulated in a cell specific.
-Interestingly, key regulatory units of Snail and Slug appear to be regulated by upstream transcription factors AP1 and SP1, respectively.
-Peinado et al. clearly established a necessary role of Snail in mediating TGF$\beta$1 induced EMT in MDCK cells.
-They demonstrated that the Ras-Raf-MEK-ERK-AP1 signaling pathway could up-regulate synthesis of the E-cadherin repressor molecule Snail (directly binds to Snail promoter) \cite{Peinado:2004pt}.
-Likewise, TGF$\beta$1 induced the expression of Slug in both lens and other epithelial cells in vitro.
-The Sp1 binding site in the Slug promoter is largely responsible for TGF$\beta$1 induced Slug expression and upstream of MAPK signaling.
-In addition, the TGF$\beta$1 mediated repression of E-cadherin was significantly inhibited by Slug siRNA \cite{Choi2007}.
-
-\subsubsection*{ERK/MAPK Crosstalk}
-The ERK-MAPK pathway has been shown to modify TGF$\beta$ signaling at multiple levels.
-One way is that the activated Ras pathway inhibits the TGF$\beta$ induced nuclear accumulation of Smad2/3, as seen in epithelial cells \cite{Zhang2009a}.
-ERK kinases have been shown to phosphorylate Smad2 and Smad3 at specific sites in the region linking the MH1 and MH2 domains.
-These sites are separate from the TGF$\beta$ receptor phosphorylation sites which activate Smad nuclear translocation \cite{Massague2003}.
-The effect of interaction between ERKs and Smads is the subject of some controversy, with data suggesting that such an interaction either enhances or inhibits downstream events.
-A hyperactive Ras pathway has been shown to effectively counteract the antiproliferative activity of TGF$\beta$ through attenuation of Smad accumulation in the nucleus \cite{Kretzschmar:1999wb}.
-In contrast, Ras signals strongly cooperated with Smads for invasion of human carcinoma cells \cite{Oft1996}.
-A second way is through transcriptional regulation.
-Recently, it was found that Sp1, via transcriptional induction of Vimentin, cooperates with activated Smad complexes in mesenchymal transition and migration of pancreatic cancer cells upon TGF$\beta$ stimulation \cite{Jungert:2007ri}.
-Likewise, binding sites at the SP1/Smad3 complex was found to regulate the betaglycan receptors \cite{Lopez-Casillas2003}.
-Taken together, multiple levels of cross-talk (both positive and negative) exist within MAPK and TGF$\beta$, and may also be cell type dependent.
-
-\subsection*{Analysis of the signal flow through the EMT architecture.}
-To investigate temporal shifts for key species dominating the EMT response, we calculated the scaled flux through the signaling architecture (Fig. \ref{fg:S1}).
-Three modes of operation were simulated to identify distinct behavioral differences: (a) no TGF-$\beta$2 (-TGF-$\beta$1/2), (b) TGF-$\beta$1/2 stimulation (+TGF-$\beta$2),
-and (c) TGF-$\beta$1/2 stimulation while blocking the autocrine response of TGF-$\beta$3 (+TGF-$\beta$1/2 + TGF-$\beta$3Ab).
-Following TGF-$\beta$1/2 stimulation, AP1/SP1 rapidly stimulates Snail/Slug expression (Fig. \ref{fg:S1}A) within 1 hr.
-During this phase, Snail/Slug repress E-cadherin expression, which in turn reduces the interaction between E-cadherin and $\beta$-catenin, and the accumulation of free cytosolic $\beta$-catenin ensues.
-Conversely, in the presences of a TGF-$\beta$3 neutralizing antibody, membrane bound $\beta$-catenin increased after 10 hrs,
-while low levels of free cytosolic $\beta$-catenin accumulated between 1-5 hrs (Fig.
+% These include RhoA, MAPK, PI3 kinase, and Notch signaling pathway, which have mostly been studied in-vitro.
+% For example, the small GTPase RhoA is involved in TGF$\beta$ induced EMT in a number of cell types including NMuMG mammary epithelial cells and mink lung epithelial (Mv1Lu) cells particularly in the regulation of cytoskeletal and adherens junction rearrangement.
+% In addition to its role in cytoskeletal remodeling, Rho has been shown to activate the alpha-smooth muscle promoter during EMT in kidney proximal tubular (LLCPK1) cells \cite{Masszi:2003hb}.
+% TGF$\beta$ has been shown to elevate ERK activity in culture models of EMT (human keratinocytes, NMuMG mammary epithelial cells, and mouse cortical tubule epithelial cells) \cite{Zavadil2001,Xie2004,Davies2005}.
+% This ERK activity was required for disassembly of adherens junctions and induction of cell motility \cite{Xie2004}.
+% Inhibition of ERK blocked these key morphological changes of EMT in mammary gland epithelial cells \cite{Xie2003}.
+% Recent studies in MDCKII and DLD1 colon carcinoma cells revealed that TGF$\beta$3 can signal through PI3 kinase.
+% Upon activation, PI3 kinase activates molecules such as ILK and AKT, which can phosphorylate and inactivate GSK3$\beta$, a protein that targets both Snail and $\beta$-catenin for degradation through the ubiquitin proteosome pathway \cite{Medici:2008fk}.
+% Lastly, there is increasing evidence for a role of Notch pathways in regulating EMT, particularly during development \cite{Nawshad:2005pi}.
+% In cardiac cushion EMT, it was found that that Slug is directly up-regulated by Notch in endothelial cells repression of the vascular endothelial cadherin promoter and for promoting migration of transformed endothelial cells.
+% In whole, the distinction between Smad-dependent and Smad-independent mechanisms remains difficult to interpret due to the significant cross talk between these pathways.
+% In most cases, stimulation of these cooperative pathways provides the context for induction and specification of EMT within a particular tissue/cell type, with Smads representing the dominant pathway, which in some instances may be necessary but not sufficient for induction of full EMT \cite{Zavadil:2005fu}.
+%
+% \subsubsection*{Master Transcription Regulators}
+% Master transcription factors of EMT are thought to be regulated by Snail1 (SNAI1) and Snail2 (SNAI2) (previously known as Snail and Slug, respectively).
+% These transcription factors are zinc finger proteins that function as repressors of E-cadherin transcription in cultured epithelial cells \cite{Cano:2000kh}, repression of E-cadherin leads to dissolution of adherens junctions.
+% Snail and Slug can be activated by TGF$\beta$ via both Smad-dependent and -independent pathways in a cell type dependent fashion in cultured cells \cite{Peinado2003}.
+% Differential expression of Snail and Slug is observed in TGF$\beta$ induced EMT in keratinocytes, renal proximal tubular, and mammary epithelial cells, suggesting that they are regulated in a cell specific.
+% Interestingly, key regulatory units of Snail and Slug appear to be regulated by upstream transcription factors AP1 and SP1, respectively.
+% Peinado et al. clearly established a necessary role of Snail in mediating TGF$\beta$1 induced EMT in MDCK cells.
+% They demonstrated that the Ras-Raf-MEK-ERK-AP1 signaling pathway could up-regulate synthesis of the E-cadherin repressor molecule Snail (directly binds to Snail promoter) \cite{Peinado:2004pt}.
+% Likewise, TGF$\beta$1 induced the expression of Slug in both lens and other epithelial cells in vitro.
+% The Sp1 binding site in the Slug promoter is largely responsible for TGF$\beta$1 induced Slug expression and upstream of MAPK signaling.
+% In addition, the TGF$\beta$1 mediated repression of E-cadherin was significantly inhibited by Slug siRNA \cite{Choi2007}.
+%
+% \subsubsection*{ERK/MAPK Crosstalk}
+% The ERK-MAPK pathway has been shown to modify TGF$\beta$ signaling at multiple levels.
+% One way is that the activated Ras pathway inhibits the TGF$\beta$ induced nuclear accumulation of Smad2/3, as seen in epithelial cells \cite{Zhang2009a}.
+% ERK kinases have been shown to phosphorylate Smad2 and Smad3 at specific sites in the region linking the MH1 and MH2 domains.
+% These sites are separate from the TGF$\beta$ receptor phosphorylation sites which activate Smad nuclear translocation \cite{Massague2003}.
+% The effect of interaction between ERKs and Smads is the subject of some controversy, with data suggesting that such an interaction either enhances or inhibits downstream events.
+% A hyperactive Ras pathway has been shown to effectively counteract the antiproliferative activity of TGF$\beta$ through attenuation of Smad accumulation in the nucleus \cite{Kretzschmar:1999wb}.
+% In contrast, Ras signals strongly cooperated with Smads for invasion of human carcinoma cells \cite{Oft1996}.
+% A second way is through transcriptional regulation.
+% Recently, it was found that Sp1, via transcriptional induction of Vimentin, cooperates with activated Smad complexes in mesenchymal transition and migration of pancreatic cancer cells upon TGF$\beta$ stimulation \cite{Jungert:2007ri}.
+% Likewise, binding sites at the SP1/Smad3 complex was found to regulate the betaglycan receptors \cite{Lopez-Casillas2003}.
+% Taken together, multiple levels of cross-talk (both positive and negative) exist within MAPK and TGF$\beta$, and may also be cell type dependent.
+%
+% \subsection*{Analysis of the signal flow through the EMT architecture.}
+% To investigate temporal shifts for key species dominating the EMT response, we calculated the scaled flux through the signaling architecture (Fig. \ref{fg:S1}).
+% Three modes of operation were simulated to identify distinct behavioral differences: (a) no TGF-$\beta$2 (-TGF-$\beta$1/2), (b) TGF-$\beta$1/2 stimulation (+TGF-$\beta$2),
+% and (c) TGF-$\beta$1/2 stimulation while blocking the autocrine response of TGF-$\beta$3 (+TGF-$\beta$1/2 + TGF-$\beta$3Ab).
+% Following TGF-$\beta$1/2 stimulation, AP1/SP1 rapidly stimulates Snail/Slug expression (Fig. \ref{fg:S1}A) within 1 hr.
+% During this phase, Snail/Slug repress E-cadherin expression, which in turn reduces the interaction between E-cadherin and $\beta$-catenin, and the accumulation of free cytosolic $\beta$-catenin ensues.
+% Conversely, in the presences of a TGF-$\beta$3 neutralizing antibody, membrane bound $\beta$-catenin increased after 10 hrs,
+% while low levels of free cytosolic $\beta$-catenin accumulated between 1-5 hrs (Fig.
 \ref{fg:S2}A,B).
@@
 During TGF-$\beta$1/2 stimulation, both MAPK and Smad act synergistically to repress the E-cadherin complex within 1 hr.
-This is followed by increase of Vimentin at $\sim$ 1hr, while exponentially increasing around 10 hrs.
-Both species elevate within 1 hr, however the complex does not significantly form until $\sim$ 10 hrs.
-When blocking TGF-$\beta$3, Snail/Slug downregulates the E-cadherin complex between 5-10 hrs,
-but cannot complete the transformation due to the loss of Smad2 phosphorylation (Fig. \ref{fg:S1}D).
-
-\begin{figure}
-\includegraphics [width=1.0\linewidth] {./figs/Fig-1-Supplemental-FluxAnalysis.pdf}
-\caption{Signal flow analysis of key species at steady state, TGF-$\beta$1/2 stimulation, and blocking the TGF-$\beta$3 autocrine response.
-(A)  The MAPK cascade is directly responsible for rapid expression of Snail and downstream TGF-$\beta$3 formation (1~hrs).
-(B) TGF-$\beta$1/2 reduces  $\beta$-catenin, allowing rapid free-cytosolic $\beta$-catenin to accumulate (1hr).
-Blocking TGF$\beta$3 increases membrane bound $\beta$-catenin (10hr). (C)  TGF-$\beta$3 activates the Smad cascade.
-Nuclear localization of the pSmad2/4 complex (10~hrs) is dependent upon both the phosphorylation of Smad2 (1~hrs) and complexing with Smad4 (5~hrs).
-(D)  TGF-$\beta$1/2 rapidly reduces the E-cadherin complex, while upregulating Vimentin (5-10 hrs).
-Blocking TGF$\beta$3 increases E-cadherin (10~hrs) and Vimentin is significantly reduced.}\label{fg:S1}
-\end{figure}
-
-\begin{figure}
-\includegraphics [width=0.5\linewidth] {./figs/Fig-2-Supplemental-Robustness.pdf}
-%\includegraphics [width=1.0\linewidth] {./ManuFigures/Figs/F2.pdf}
-%\epsfig{file=./figures/Figures_EMT_Manuscript/Final/ManuFigures/F2.eps,,width=1.0\textwidth}
-\caption{Robustness analysis for key molecular species at t = 48 hrs for combinations of TGF-$\beta$1/2, VEGF-A and NFATc1 inhibitors. Robustness coefficients for the indicated species were calculated
-for N $\sim{1100}$ ensemble members for 48 hrs following the addition of TGF-$\beta$1/2 (T), TGF-$\beta$1/2 + NFATc1 inhibitor (T + I), VEGF-A (V), VEGF-A + NFATc1 inhibitor (V + I) and TGF-$\beta$1/2 + VEGF-A (T+V) + NFATc1 inhibitor (T + V + I).
-(A) Robustness coefficients for E-cadherin and Vimentin as a function of condition.
-(B) Robustness coefficients for nuclear localized phosphorylated Sp1 and NFATc1 as a function of condition.
-(C) Robustness coefficients for nuclear localized Slug and $\alpha$-smooth muscle actin ($\alpha$-SMA) as a function of condition.
-In each case the error bars denote one-standard deviation of robustness coefficient calculated over the model ensemble.
-C=Control, T=TGF$\beta$2 , V=VEGFA, VI= NFAT inhibitor (VIVIT).}\label{fg:S2}
-\end{figure}
+% This is followed by increase of Vimentin at $\sim$ 1hr, while exponentially increasing around 10 hrs.
+% Both species elevate within 1 hr, however the complex does not significantly form until $\sim$ 10 hrs.
+% When blocking TGF-$\beta$3, Snail/Slug downregulates the E-cadherin complex between 5-10 hrs,
+% but cannot complete the transformation due to the loss of Smad2 phosphorylation (Fig. \ref{fg:S1}D).
+%
+% \begin{figure}
+% \includegraphics [width=1.0\linewidth] {./figs/Fig-1-Supplemental-FluxAnalysis.pdf}
+% \caption{Signal flow analysis of key species at steady state, TGF-$\beta$1/2 stimulation, and blocking the TGF-$\beta$3 autocrine response.
+% (A)  The MAPK cascade is directly responsible for rapid expression of Snail and downstream TGF-$\beta$3 formation (1~hrs).
+% (B) TGF-$\beta$1/2 reduces  $\beta$-catenin, allowing rapid free-cytosolic $\beta$-catenin to accumulate (1hr).
+% Blocking TGF$\beta$3 increases membrane bound $\beta$-catenin (10hr). (C)  TGF-$\beta$3 activates the Smad cascade.
+% Nuclear localization of the pSmad2/4 complex (10~hrs) is dependent upon both the phosphorylation of Smad2 (1~hrs) and complexing with Smad4 (5~hrs).
+% (D)  TGF-$\beta$1/2 rapidly reduces the E-cadherin complex, while upregulating Vimentin (5-10 hrs).
+% Blocking TGF$\beta$3 increases E-cadherin (10~hrs) and Vimentin is significantly reduced.}\label{fg:S1}
+% \end{figure}
+%
+% \begin{figure}
+% \includegraphics [width=0.5\linewidth] {./figs/Fig-2-Supplemental-Robustness.pdf}
+% %\includegraphics [width=1.0\linewidth] {./ManuFigures/Figs/F2.pdf}
+% %\epsfig{file=./figures/Figures_EMT_Manuscript/Final/ManuFigures/F2.eps,,width=1.0\textwidth}
+% \caption{Robustness analysis for key molecular species at t = 48 hrs for combinations of TGF-$\beta$1/2, VEGF-A and NFATc1 inhibitors.
+Robustness coefficients for the indicated species were calculated
+% for N $\sim{1100}$ ensemble members for 48 hrs following the addition of TGF-$\beta$1/2 (T), TGF-$\beta$1/2 + NFATc1 inhibitor (T + I), VEGF-A (V), VEGF-A + NFATc1 inhibitor (V + I) and TGF-$\beta$1/2 + VEGF-A (T+V) + NFATc1 inhibitor (T + V + I).
+% (A) Robustness coefficients for E-cadherin and Vimentin as a function of condition.
+% (B) Robustness coefficients for nuclear localized phosphorylated Sp1 and NFATc1 as a function of condition.
+% (C) Robustness coefficients for nuclear localized Slug and $\alpha$-smooth muscle actin ($\alpha$-SMA) as a function of condition.
+% In each case the error bars denote one-standard deviation of robustness coefficient calculated over the model ensemble.
+% C=Control, T=TGF$\beta$2 , V=VEGFA, VI= NFAT inhibitor (VIVIT).}\label{fg:S2}
+% \end{figure}
 
@@
 
-\subsection*{Estimation and cross-validation of EMT model parameters.}
-We used the Pareto Optimal Ensemble Technique (POETs) multiobjective optimization framework in combination with leave-one-out cross-validation to estimate an ensemble of TGF$\beta$/EMT models.
-Cross-validation was used to calculate both training and prediction error during the parameter estimation procedure \cite{kohavi1995study}.
-The 41 intracellular protein and mRNA data-sets used for identification were organized into 11 objective functions.
-These 11 objective functions were then partitioned, where each partition contained ten training objectives and one validation objective.
-POETs integrates standard search strategies e.g., Simulated Annealing (SA) or Pattern Search (PS)
-with a Pareto-rank fitness assignment \cite{Song:2010fk}.
-Denote a candidate parameter set at iteration $i+1$ as $\mathbf{k}_{i+1}$.
-The squared error for $\mathbf{k}_{i+1}$ for training set $j$ was defined as:
-\begin{equation}\label{eqn_cost2}
-	E_{j}(\mathbf{k}) = \sum_{i=1}^{\mathcal{T}_{j}}\biggl(\hat{\mathcal{M}}_{ij}-\hat{y}_{ij}(\mathbf{k})\biggr)^2
-\end{equation}
-The symbol $\hat{\mathcal{M}}_{ij}$ denotes scaled experimental observations (from training set $j$) while $\hat{y}_{ij}$ denotes the scaled simulation output (from training set $j$).
-The quantity $i$ denotes the sampled time-index and $\mathcal{T}_{j}$ denotes the number of time points for experiment $j$.
-In this study, the experimental data used for model training was typically the band intensity from Western or Northern blots.
-Band intensity was estimated using the ImageJ software package \cite{IMAGEJ}.
-The scaled measurement for species $x$ at time $i=\{t_{1},t_{2},..,t_{n}\}$ in condition $j$ is given by:
-\begin{equation}\label{norm_exp_data}
-\hat{\mathcal{M}}_{ij} = \frac{\mathcal{M}_{ij} - \min_{i}\mathcal{M}_{ij}}{\max_{i}{\mathcal{M}_{ij}}-\min_{i}{\mathcal{M}_{ij}}}
-\end{equation}
-Under this scaling, the lowest intensity band equaled zero while the highest intensity band equaled one.
-A similar scaling was defined for the simulation output. By doing this scaling, we trained the model on the relative change in blot intensity, over conditions or time (depending upon the experiment).
-Thus, when using multiple data sets (possibly from different sources) that were qualitatively similar but quantitatively different e.g., slightly different blot intensities over time or condition, we captured the underlying trends in the scaled data.
-Additionally, we applied a universal convention of identifying the no expression case as protein value below $<$10$^{-3}$.
-This is similar to previously published models from our lab.
-In our studies, Figure 2 identifies experimental data extracted from published Western blots and our simulation results.
-It is clear from these that 1) the training data included a variety of Western blot data treatments that were effectively zero, and 2) our simulations matched the training data virtually perfectly over time and across multiple biological species.
-These results validate the power of our simulation scheme, which necessarily includes numerical interpretations of zero.
-In our Supplemental Figure S9, we present the raw concentrations of our simulations (previously validated from the training data) over hundreds of parameter ensembles, with no “zero” threshold applied.
-While these values do not have a threshold, we only interpret values below 10$^{-3}$ equivalent to zero.
-Likewise, our robustness coefficients (Figure 3) (which represent the ratio of integrated areas of the treatment effect over the baseline effect) identify no difference in model output for vimentin (or e-cadherin for that matter) less than 10$^{-3}$, confirming this interpretation.
-
-We computed the Pareto rank of $\mathbf{k}_{i+1}$ by comparing the simulation error at iteration $i+1$ against an archive of accepted parameter sets $\mathbf{K}_{i}$.
-We used the Fonseca and Fleming ranking scheme \cite{fonseca1993genetic} to estimate the number of parameter sets in the archive that dominate
-$\mathbf{k}_{i+1}$. Parameter sets with increasing rank were progressively further away from the optimal trade-off surface.
-The parameter set $\mathbf{k}_{i+1}$ was accepted or rejected by POETs with probability $\mathcal{P}\left(\mathbf{k}_{i+1}\right)$:
-\begin{equation}\label{eqn_costMOSA}
-\mathcal{P}(\mathbf{k}_{i+1}) \equiv \exp{\{-rank\left(\mathbf{k}_{i+1} \mid \mathbf{K}_{i} \right)/T\}}
-\end{equation}
-where $T$ is the annealing temperature and $rank\left(\mathbf{k}_{i+1} \mid \mathbf{K}_{i} \right)$ denotes the Pareto rank for $\mathbf{k}_{i+1}$.
-The annealing temperature was discretized into 10 quanta between $T_{o}$ and $T_{f}$ and adjusted according to the schedule
-$T_k = \beta^k T_0$ where $\beta$ was defined as $\left({T_{f}}/{T_{o}}\right)^{1/10}$.
-The initial temperature was $T_{o} = \mathit{n}/log(2)$, where $n = 4$ in this study and the final temperature was $T_{f} = 0.1$.
-The epoch-counter $k$ was incremented after the addition of 100 members to the ensemble.
-Thus, as the ensemble grew, the likelihood of accepting parameter sets with a large Pareto rank decreased.
-To generate parameter diversity, we randomly perturbed each parameter by $\leq\pm25\%$ at iteration of the search.
-In addition, we performed a local pattern search every $q$-iterations to minimize the residual for a single random or the worst performing objective function.
-The local pattern-search algorithm has been described previously \cite{gadkar2003cybernetic}.
-From the 15,000 probable EMT models, we selected N = 1093 models with Pareto rank $\leq{1}$ for subsequent analysis.
-A quick estimate of the set to set correlation showed that we could expect on order 25\% correlation between parameter sets in the ensemble.
-
 \begin{figure}
-\includegraphics [width=1.0\linewidth] {./figs/Fig-5-Supplemental-ErrorTable.pdf}
-\caption{Training and prediction values for the 11 TGF-$\beta$ objective functions versus a random parameter control.}\label{fig:S5}
-\end{figure}
-
-\begin{figure}
-\includegraphics [width=1.0\linewidth] {./figs/Fig-6-Supplemental-CompeteErrorTable.pdf}
-\caption{Training and prediction values as a function of condition for the 11 TGF-$\beta$ objective functions versus a random parameter control.}\label{fig:S6}
-\end{figure}
-
-\begin{figure}
-\includegraphics [width=1.0\linewidth] {./figs/Fig-7-Supplemental-Correlation.pdf}
-\caption{Parameter set to set correlation for 55 random parameter sets selected from the ensemble.
-Of the 55 sets selected, the average correlation between sets was less than 25\% for greater than 80\% of the parameter sets.}\label{fig:S7}
-\end{figure}
-
-\begin{figure}
-\includegraphics [width=1.0\linewidth] {./figs/Fig-8-Supplemental-YREG1.pdf}
-\caption{Robustness of E-cadherin and Vimentin expression to a knockout (A) and overexpression (B) of the hypothetical regulator 1 (YREG1) protein.
-Robustness coefficients were calculated for each member of the ensemble. Each point represents the response of a single model in the ensemble to either a knockout or
-overexpression of YREG1.}\label{fig:S8}
-\end{figure}
-
-\begin{figure}
-\includegraphics [width=1.0\linewidth] {./figs/Fig-9-Supplemental-RawExpression_v2.pdf}
-\caption{Steady state protein abundance for E-cadherin and Vimentin (AU) as a function of TGF-$\beta$1/2 exposure.
-Top: Overlay of the model population for Vimentin (AU) and E-cadherin (AU) expression in the presence (blue) and absence (gray) of TGF-$\beta$1/2.
-Midddle: Vimentin (AU) and E-cadherin (AU) expression in the absence of TGF-$\beta$1/2 showed exhinibted population heterogenity.
-Bottom: Vimentin (AU) and E-cadherin (AU) expression in the presence of TGF-$\beta$1/2 moved the centriod of the population toward Vimentin (AU) and away from E-cadherin (AU) expression.}\label{fig:S9}
+	\center
+	\includegraphics [width=0.75\linewidth] {./figs/Fig-Supplemental-Data-VEGFA.pdf}
+	\caption{VEGF-A qPCR data used to hand fit VEGF enhancement of E-cadherin expression. mRNA was harvested after 3hr and 24hr timepoint.}\label{fg:VEGFA-Data}
 \end{figure}

### 2016-10-04 Paper_v7.tex
DB semi final

@@
 Model parameters were estimated from multiple data sets using multiobjective optimization, in combination with cross-validation.
-TGF-$\beta$ exposure drove the model population toward a mesenchymal phenotype, while an epithelial phenotype was maintained following vascular endothelial growth factor A (VEGF-A) exposure.
+TGF-$\beta$ exposure drove the model population toward a mesenchymal phenotype, while an epithelial phenotype was enhanced following vascular endothelial growth factor A (VEGF-A) exposure.
 Simulations predicted that the transcription factors phosphorylated SP1 and NFAT were master regulators promoting or inhibiting EMT, respectively.
@@
 Analysis of the molecular signatures of these subpopulations suggested the abundance, localization and state of the Sp1, AP-1 and NFATc transcription factors controlled population heterogeneity.
-The majority of models ($>$80\%) responded to treatment, moving away from the untreated population (Fig \ref{fg:F3}A-C, gray).
+The majority of models ($>$80\%) responded to treatment, moving away from the untreated population (Fig \ref{fg:F3}A-F, gray).
 These models showed the classically expected behavior, a switch from an epithelial to mesenchymal phenotype following TGF$-\beta$1/2 exposure.
-Some models resembled untreated cells; they had evaluated nuclear localized phosphorylated Sp1, relative to non-induced cells,
+Some models resembled untreated cells; they had elevated phosphorylated Sp1, relative to non-induced cells,
 which decreased E-cadherin expression through Slug-mediated inhibition, which in turn increased Vimentin expression through TGF$-\beta$3 autocrine signaling and the liberation of $\beta$-catenin.
 However, the most biologically interesting behavior was exhibited by cells achieving a hybrid phenotype, most notable in a dual treatment condition (\ref{fg:F3}C, black arrow), but also present in a small percentage of untreated cells (Fig.
-\ref{fg:F3}A).
+\ref{fg:F3}B, gray arrow).
 Models with this hybrid phenotype had elevated Sp1 and NFAT transcriptional activity, resulting in simultaneously increased Vimentin and E-cadherin expression (Fig. \ref{fg:F4}A).
@@
 The model population was trained and cross-validated to prescribe biological significance using 41 data sets generated in DLD1 colon carcinoma, MDCKII, and A375 melanoma cell lines \citep{Medici:2008fk}.
-The absence of TGF$-\beta$1/2 or VEGF-A stimulation was used as the baseline for the robustness calculations.
 Analysis of this population predicted possible phenotypic modes (and their associated signaling) that cells could exhibit when stimulated with TGF-$\beta$ and/or VEGF-A.
@@
 These techniques (and many others, see review \citep{Wayman2013}) are certainly exciting, with many interesting properties.
-However, we used the traditional approach of mass action kinetics within an ordinary differential equation framework.
+Here, we used a traditional approach of mass action kinetics within an ordinary differential equation framework that also included transfer functions to simplify scenarios where reactions involving one species are controlled by several others (e.g.
+E-cadherin transcription).
 The identification problem for the EMT model was massively underdetermined.
@@
 This release of $\beta$-catenin can then translocate to the nucleus and form transcriptional complexes with TCF-4 to drive TGF$-\beta$3 expression \citep{Medici:2008fk}.
-The PI3K to GSK3 pathway was included and acted as an activating mechanism of $\beta$-catenin signaling through TGF$-\beta$3 signaling \c\
-itep{Medici:2008fk}. GSK3 is known to act on $\beta$-catenin signaling through the ubiquitin-proteasome pathway
-\citep{Zhou:2004kx,Larue:2005vn}.
+The PI3K to GSK3 pathway was included and acted as an activating mechanism of $\beta$-catenin signaling through TGF$-\beta$3 signaling \citep{Medici:2008fk}.
+GSK3 is known to act on $\beta$-catenin signaling through the ubiquitin-proteasome pathway \citep{Zhou:2004kx,Larue:2005vn}.
 Thereby, further $\beta$-catenin release also resulted from by TGF$-\beta$3 signals to the cells interior by binding to type II receptors, which form heterodimers with type I receptors (ALK5) \citep{Derynck:2003fc}.
 This activates the receptors serine/threonine kinase activity to phosphorylate and activate the receptor Smads 2/3 \citep{Massague:2005qc}.
-In the model, receptors are represented as either bound or unbound complexes.
+In the model, receptors are simplified and represented as either bound or unbound complexes with their ligands.
 Phosphorylated Smads 2/3 (pSmad2/3) form heterodimers and translocate to the nucleus.
@@
 LEF-1 also binds with $\beta$-catenin to upregulate mesenchymal proteins such as fibronectin \citep{Medici2011}.
-In the model, Smad signaling is lumped into a single smad species that can act in a co-dependent fashion with LEF1 to downregulate E-cadherin.
-The EMT gene expression program was initiated by the binding of TGF$-\beta$ isoforms to TGF$-\beta$ surface receptors.
-Binding of extracellular TGF$-\beta$1/2 with TGF$-\beta$ surface receptors I/II (TGF$-\beta$R-I/II) initiates the assembly of adapter complexes which starts the
-downstream signaling program.
-Complex assembly activates the serine/threonine kinase activity on the receptor, leading to the recruitment and phosphorylation of Smad partners \citep{Massague:2005qc}.
-Phosphorylated Smads2/3 (pSmad2/3) form heterodimers with partner Smad4 and then translocate to the nucleus where they act as both transcriptional activators and repressors.
-Repression of E-cadherin expression is the central event in the transition from an epithelial to a mesenchymal phenotype \citep{Cano:2000kh}.
-However, this transition is not solely driven by transcriptional events.
+In the model, Smad signaling is consolidated into a single Smad species that can act in a co-dependent fashion with LEF1 to downregulate E-cadherin via a transfer function, eliminating the need for an explicity LEF-1, pSmad complex.
+The EMT gene expression program was initiated by the binding of TGF$-\beta$ isoforms to TGF$-\beta$ surface receptors, starting the downstream signaling program.
+Repression of E-cadherin expression is the central event in the transition from an epithelial to a mesenchymal phenotype \citep{Cano:2000kh}.However, this transition is not solely driven by transcriptional events.
 At the protein level, the repression of E-cadherin leads to a release of $\beta$-catenin from cell membrane.
@@
 The PI3K to GSK3 pathway was included and acted as an activating mechanism of $\beta$-catenin signaling through TGF$-\beta$3 signaling \citep{Medici:2008fk}.
-GSK3 is known to act on $\beta$-catenin signaling through the ubiquitin-proteasome pathway.
-\citep{Zhou:2004kx,Larue:2005vn}. Lastly, VEGF-A activation of NFATc1 takes place through calcineurin signaling leading to an enhancement of E-cadherin expression \citep{Suehiro:2014aa},
+GSK3 is known to act on $\beta$-catenin signaling through APC complex associated ubiquitin-proteasome pathway.
+The APC complex is represented in our model and serves as a second reservoir  of $\beta$-cateinin in untransformed cells whose sequestration is regulated by GSK3 \citep{Zhou:2004kx,Larue:2005vn,Medici:2008fk}.
+Lastly, VEGF-A activation of NFATc1 takes place through calcineurin signaling leading to an enhancement of E-cadherin expression \citep{Suehiro:2014aa},
 as supported by our VEGF-A experimental data (Fig. \ref{fg:VEGFA-Data}).
@@
 (A-I) The population was generated using JuPOETs and trained using 11 different objective functions (41 data sets) taken from Medici \emph{et al.} \citep{Medici:2008fk}.
-The model captured the simulated experiments for 72\% of the cases.
+The model captured the simulated experiments for all cases when compared to randomized controls.
 (J-L)  The model populations were also compared against untrained temporal data to measure the effectiveness as a pure prediction.}\label{fg:F2}
@@
 (D-F) To isolate the effect of NFAT, we inhibited NFAT de-phosphorylation in combination with VEGFA.
-This negated the increase in ecadherin expression and shifted the population towards a mesenchymal phenotype (Q1,Q3).
+This negated the increase in ecadherin expression and shifted the population towards a mesenchymal phenotype.
 Likewise, combining NFAT inhibition with TGF$-\beta$ mitigated all VEGF enhanced ecadherin expression.
 Lastly, combination of TGF$-\beta$2, VEGFA, and NFAT inhibition nearly mitigated all effects of VEGFA, shifting the heterogeneous population towards a mesenchymal phenotype.
-In whole, high levels of phosphorylated-Sp1 correlated with vimentin expression, while NFAT was responsible for maintaining E-cadherin expression, although neither were mutually exclusive.}
+In whole, high levels of phosphorylated-Sp1 correlated with vimentin expression, while NFAT was responsible for maintaining E-cadherin expression in the presence of other factors, although neither were mutually exclusive.}
 \label{fg:F3}

### 2016-10-04 Paper_v7.tex
Small changes

@@
 The predictive power of the ensemble was tested using both cross validation and by comparing simulations with data sets not used for model training.
-In whole, 100\% of our training objectives were statistically significant (at a 95\% confidence interval) compared to a randomized parameter family (N = 100)
-generated from a random starting point. Conversely, we \emph{predicted} 100\% of the training objectives, at a 95\% confidence interval compared to randomized parameters (Wicoxon non-parametric test).
+In whole, all of our training objectives were statistically significant (at a 95\% confidence interval) compared to a randomized parameter family (N = 100)
+generated from a random starting point.
+Conversely, we \emph{predicted} all of the training objectives, at a 95\% confidence interval compared to randomized parameters (Wicoxon non-parametric test).
 The model also captured the temporal gene expression responses of E-cadherin, pSmad2, and LEF-1 (not used for model training) to within one-standard deviation (up to the 48 hr time-point) (Fig.

### 2016-10-04 Paper_v7.tex
Fixed figure numbering ... I think

@@
 The EMT model simulations suggested the transcriptional activity of NFATc and Sp1 could be independently tuned to generate a hybrid cell population with both epithelial and mesenchymal characteristics.
-To test this hypothesis, we exposed either quiescent epithelial (MCFA10, (Fig.
-\ref{fg:F5})) or transformed epithelial cells (DLD1, (Fig.
-\ref{fg:S3})) to combinations of TGF$-\beta$1/2 and/or VEGF-A.
-As expected, treatment with TGF$-\beta$1/2 (10ng/ml) increased Slug and Vimentin expression, while repressing E-cadherin expression both at the transcript and protein levels in MCF10A (Fig.
+To test this hypothesis, we exposed either quiescent epithelial (MCFA10, Fig.
+\ref{fg:F5}) or transformed epithelial cells (DLD1, Fig.
+\ref{fg:S3}) to combinations of TGF$-\beta$1/2 and/or VEGF-A.
+As expected, TGF$-\beta$1/2 treatment (10ng/ml) increased Slug and Vimentin expression, while repressing E-cadherin expression both at the transcript and protein levels in MCF10A (Fig.
 \ref{fg:F5}A-B) and DLD1 cells (Fig.
-\ref{fg:S4}C, Fig S3 D,E).
+\ref{fg:S4}C).
 Both MCF10A (Fig. \ref{fg:F5}C) and DLD1 cells (Fig. \ref{fg:S3}E,G) transitioned from quiescent cobblestone morphology to spread spindle shapes, consistent with EMT.
@@
 Consistent with model predictions, VEGF-A (50ng/ml) treatment increased the abundance of NFATc1 and E-cadherin at both the transcript and protein level in both MCF10A (Fig.
-\ref{fg:F5}A) and DLD1 (Fig.
-\ref{fg:S3}A) cells.
-We also found that NFATc1 nuclear localization significantly increased in both MCF10 and DLD1 treated with VEGF-A independently of the abundance of nuclear localized phosphorylated Sp1 levels (Fig.
-\ref{fg:F5}B,C Fig.S3C,E ).
-Interestingly, combining VEGF-A (50ng/ml) with TGF$-\beta$1/2 (10ng/ml) resulted in significantly elevated expression of both E-cadherin and Vimentin at the transcript and protein levels in both MCF10A and DLD1 cells (Fig 5A,B; Fig S3D,E; Fig S4C).
-NFATc1 expression increased, while Sp1 expression was similar to the TGF$-\beta$1/2 case alone (Fig. \ref{fg:F5}A-B,  Fig S3D,E; Fig S4C)), supporting their independent regulation.
-The expression of Slug, and Vimentin significantly increased, while E-cadherin levels were increased in MCF10A cells (Fig 5A) and maintained at control levels in DLD1 cells (Fig. S3D).y
-As further predicted, nuclear co-localization of both NFATc1 and phosphorylated Sp1 were apparent in MCF10A and DLD1 cells treated with both ligands (Fig. \ref{fg:F5}B,C Fig S3E,F).
+\ref{fg:F5}A) and DLD1 cells (Fig.
+\ref{fg:S3}).
+We also found that NFATc1 nuclear localization significantly increased in both MCF10 (Fig. \ref{fg:F5}B,C) and DLD1 (Fig. \ref{fg:S3}C,E) cells treated with VEGF-A.
+Interestingly, combining VEGF-A (50ng/ml) with TGF$-\beta$1/2 (10ng/ml) resulted in significantly elevated expression of both E-cadherin and Vimentin at the transcript and protein levels in both MCF10A (Fig.
+\ref{fg:F5}A,B) and DLD1 cells (Fig.
+\ref{fg:S3}-\ref{fg:S4}).
+NFATc1 expression increased, while Sp1 expression was similar to the TGF$-\beta$1/2 case alone (Fig. \ref{fg:F5}A-B and Fig. \ref{fg:S3}D,E),
+supporting their independent regulation.
+The expression of Slug, and Vimentin significantly increased, while E-cadherin levels were increased in MCF10A cells (Fig. \ref{fg:F5}A) and maintained at control levels in DLD1 cells
+(Fig. \ref{fg:S3}D).
+As predicted, nuclear co-localization of both NFATc1 and phosphorylated Sp1 were apparent in MCF10A (Fig. \ref{fg:F5}B,C) and DLD1 (Fig. \ref{fg:S3}E,F) cells treated with both ligands.
 Taken together, combined VEGF-A and TGF$-\beta$1/2 treatment elicited a hybrid phenotype expressing both mesenchymal and epithelial characteristics in both MCF10A and DLD1 cells.
@@
 \ref{fg:S4}) to combinations of VEGF-A and TGF$-\beta$1/2 in the presence or absence of VIVIT, a soluble peptide inhibitor of NFATc transcriptional activity \citep{Aramburu:1999if}.
-Treatment with VEGF-A (50ng/ml) and VIVIT (10$\mu$M) in MCF10A cells significantly reduced E-cadherin expression compared to VEGF-A alone (Fig 5D,E).
-Co-treatment with VIVIT and TGF$-\beta$1/2 did not enhance EMT capacity of MCF10A cells above that of TGF$-\beta$1/2 alone (Fig 5A,B,E).
+Treatment with VEGF-A (50ng/ml) and VIVIT (10$\mu$M) in MCF10A cells significantly reduced E-cadherin expression compared to VEGF-A alone (Fig. \ref{fg:F5}D,E).
+Co-treatment with VIVIT and TGF$-\beta$1/2 did not enhance EMT capacity of MCF10A cells above that of TGF$-\beta$1/2 alone (Fig. \ref{fg:F5}A,B,E).
 Likewise, VIVIT in combination with both TGF$-\beta$1/2 and VEGF-A resulted in a loss of E-cadherin gene and protein expression, while Slug and Vimentin levels remained increased (Fig.
-\ref{fg:F5}D,E ).
-Quantitative flow cytometry confirmed these results in both MCF10A (Fig. \ref{fg:F5}F) and DLD1 cells (Fig. S4C).
-Both epithelial cell lines initially had high levels of E-cadherin expression, and low vimentin abundance (Q1-99.5\%), but both MCF10A and DLD1 cells shifted from an epithelial to mesenchymal phenotype (Q1-33.4\%, Q4-42.8\%) following TGF$-\beta$1/2 exposure.
+\ref{fg:F5}D,E).
+Quantitative flow cytometry confirmed these results in both MCF10A (Fig. \ref{fg:F5}F) and DLD1 cells (Fig. \ref{fg:S4}C).
+Both epithelial cell lines initially had high levels of E-cadherin expression, and low Vimentin abundance (Q1-99.5\%), but both MCF10A and DLD1 cells shifted from an epithelial to mesenchymal phenotype (Q1-33.4\%, Q4-42.8\%) following TGF$-\beta$1/2 exposure.
 As expected, NFATc nuclear localization was repressed with VIVIT treatment regardless of ligand stimulation, while the abundance of nuclear phosphorylated Sp1 increased for both TGF$-\beta$1/2 and TGF$-\beta$1/2 + VIVIT conditions (Fig.
-5D,E).
+\ref{fg:F5}C,E).
 Combined TGF$-\beta$1/2 and VEGF-A increased both Vimentin and E-cadherin expression (Q1-42.1\%, Q2-52.3\%) compared to TGF$-\beta$1/2 alone.

### 2016-10-04 Paper_v7.tex
Added table to supplemental

@@
 
+\usepackage{listings}
+
 \urlstyle{rm}
@@
 \bibliographystyle{msb}
+
+\definecolor{mygreen}{rgb}{0,0.6,0}
+\definecolor{mygray}{rgb}{0.5,0.5,0.5}
+\definecolor{mymauve}{rgb}{0.58,0,0.82}
+
+\lstdefinestyle{myCustomMatlabStyle}{
+%belowcaptionskip=\baselineskip,
+breaklines=true,
+%frame=L,
+xleftmargin=\parindent,
+language=Octave,
+showstringspaces=false,
+basicstyle=\tiny\ttfamily,
+keywordstyle=\bfseries\color{black},
+commentstyle=\itshape\color{black},
+identifierstyle=\color{blue},
+stringstyle=\color{orange},
+}
 
@@
 \begin{document}
+
 \begin{titlepage}
@@
 and the abundance of Ribosomes ($RIBO$), $V^{max}_{X} = k_{X}\left(RIBO\right)$, where both $k_{X}$ and $RIBO$ abundance were estimated from literature \citep{Milo:2010aa}.
+The values used for the characteristic transcription and translation parameters are given in the supplemental materials.
 
@@
 \subsubsection*{Estimation of model parameters using multiobjective optimization.}
-The EMT model had 296 unknown parameters (169 kinetic constants, 44 saturation constants, 38 control logic paramters, and 45 non-zero initial conditions) which were not uniquely identifiable given the training data.
+The EMT model had 296 unknown parameters (169 kinetic constants, 44 saturation constants, 38 control logic parameters, and 45 non-zero initial conditions) which were not uniquely identifiable given the training data.
 Instead, we estimated a population of likely models (each consistent with the training data)
@@
 We used the Pareto Optimal Ensemble Technique (JuPOETs) multiobjective optimization framework in combination
-with leave-one-out cross-validation to estimate an ensemble of model parameters \citep{Song:2010fk}.
+with leave-one-out cross-validation to estimate an ensemble of model parameters \citep{Song:2010fk,JuPOETs-BioArXiv}.
 Cross-validation was used to calculate both training and prediction error during the parameter estimation procedure \citep{kohavi1995study}.

### 2016-10-04 Supplemental-JB.tex
Added table to supplemental

@@
+\nolinenumbers
 \section*{Supplemental Materials and Methods}
+
+\subsection*{Characteristic transcription and translation parameters.}
+We used literature based transcription and translation parameters to establish the characteristic synthesis and degradation rates for both mRNA and protein.
+We estimated values for the rate parameters from the Bionumbers database \citep{Milo:2010aa}. These parameters were then used for all gene
+expression calculations:
+
+\lstset{basicstyle=\tiny,style=myCustomMatlabStyle}
+\begin{lstlisting}
+-------------------------------------------------------------------
+# Description
+-------------------------------------------------------------------
+cell_diameter = 12                	# \mum
+number_of_rnapII = 75000          	# copies/cells
+number_of_ribosome = 1e6          	# copies/cells
+mRNA_half_life_TF = 2             	# hrs
+protein_half_life = 10            	# hrs
+doubling_time 	= 19.5         		# hrs
+max_translation_rate = 5          	# aa/sec
+max_transcription_rate = 6.0       	# nt/sec
+average_transcript_length = 15000 	# nt
+average_protein_length = 5000     	# aa
+fraction_nucleus = 0.49           	# dimensionless
+av_number = 6.02e23               	# number/mol
+avg_gene_number = 2               	# number of copies of a gene
+-------------------------------------------------------------------
+
+---------------------------------------------------------------------------------------
+# Description
+---------------------------------------------------------------------------------------
+# Calculate the volume (units: L)
+V = ((1-fraction_nucleus)*(1/6)*(3.14159)*(hl60_diameter)^3)*(1e-15)
+
+# Calculate the rnapII_concentration and ribosome_concentration (units: nM)
+rnapII_concentration = number_of_rnapII*(1/av_number)*(1/V)*1e9
+ribosome_concentration = number_of_ribosome*(1/av_number)*(1/V)*1e9
+
+# degradation rate constants (units: hr^-1)
+degradation_constant_mRNA = -(1/mRNA_half_life_TF)*log(0.5)
+degradation_constant_protein = -(1/protein_half_life)*log(0.5)
+
+# kcats for transcription and translation (units: hr^-1)
+kcat_transcription = max_transcription_rate*(3600/average_transcript_length)
+kcat_translation = max_translation_rate*(3600/average_protein_length)
+
+# Maximum specific growth rate (units: hr^-1)
+maximum_specific_growth_rate = (1/doubling_time)*log(2)
+
+# What is the average gene concentration (units: nM)
+avg_gene_concentration = avg_gene_number*(1/av_number)*(1/V)*1e9
+
+# Cell death constant (units: hr^-1)
+death_rate_constant = 0.2*maximum_specific_growth_rate
+
+# Saturation constants for translation and transcription (units: nM)
+saturation_transcription = 4600*(1/av_number)*(1/V)*1e9
+saturation_translation = 100000*(1/av_number)*(1/V)*1e9
+---------------------------------------------------------------------------------------
+
+\end{lstlisting}
 
@@
 In this study, the experimental data used for model training was typically the band intensity from Western or Northern blots.
-Band intensity was estimated using the ImageJ software package \cite{IMAGEJ}.
+Band intensity was estimated using the ImageJ software package \citep{IMAGEJ}.
 The scaled measurement for species $x$ at time $i=\{t_{1},t_{2},..,t_{n}\}$ in condition $j$ is given by:
@@
 Thus, when using multiple data sets (possibly from different sources) that were qualitatively similar but quantitatively different e.g., slightly different blot intensities over time or condition, we captured the underlying trends in the scaled data.
+JuPOETs is free or charge, open source and available for download under an MIT software license from http://www.varnerlab.org.
+Details of the JuPOETs implementation, including example codes are presented in Bassen et al., \citep{JuPOETs-BioArXiv}.
 

### 2016-10-04 Paper_v7.tex
Ok

@@
 In this study, we created an ensemble of dynamic mathematical models describing TGF-$\beta$ induced EMT to better understand the operational hierarchy of this complex molecular program.
-These models incorporate mass action kinetics within an ordinary differential equation (ODE) framework to describe the transcriptional and post-translational regulatory events driving EMT.
+We used ordinary differential equations (ODEs) to describe the transcriptional and post-translational regulatory events driving EMT.
 Model parameters were estimated from multiple data sets using multiobjective optimization, in combination with cross-validation.
@@
 Much research over the past 20 years has identified many different molecular species that are relevant, but these have mainly been studied one at a time.
-In this study, we developed and implemented a novel computational strategy to interrogate all of the known players in this transformation process to identify which are the major bottlenecks.
+In this study, we developed and implemented a novel computational strategy to interrogate the key players in this transformation process to identify which are the major bottlenecks.
 We determined that NFATc1 and pSP1 are essential for promoting epithelial or mesenchymal differentiation, respectively.
@@
 Decades of research has focused on identifying molecular regulators of EMT, but almost all on a single gene and in a nearly binary yes/no level of qualitative understanding.
-Medici and coworkers recently identified a core signaling program by which TGF-$\beta$ isoforms induce EMT across a variety of cell lines \citep{Medici:2006qa,Medici:2008fk}.
+Medici and coworkers identified a core signaling program by which TGF-$\beta$ isoforms induce EMT across a variety of cell lines \citep{Medici:2006qa,Medici:2008fk}.
 This program involves carefully orchestrated rounds of gene expression driven by the Smad and Snail families of transcription factors as well as
@@
 Furthermore, we proposed that the sloppy behavior of biological networks may also be seen as a source of cell-to-cell \citep{Lequieu:2011fj} or even patient-to-patient heterogeneity \citep{2010_luan_varner_MolBioSys}.
-Recently, Bayesian parameter identification techniques have also been used to explore cell-to-cell heterogeneity  \citep{2011_kalita_brasier_JBC,2011_hasenauer_allgower_BMCBioinfo},
+Bayesian parameter identification techniques have also been used to explore cell-to-cell heterogeneity  \citep{2011_kalita_brasier_JBC,2011_hasenauer_allgower_BMCBioinfo},
 where a population of cells could be viewed as a dynamic ensemble of context-specific biochemical networks \citep{2012_creixell_linding_NatBiotech}.
 
-In this study, we developed a family of mechanistic models describing the induction of EMT by TGF-$\beta$ isoforms in the presence and absence of vascular endothelial growth factor A (VEGF-A).
-We incorporated mass action kinetics within an ordinary differential equation (ODE) framework to describe the EMT interaction network containing 97 gene, protein or mRNA components interconnected through 169 interactions.
+In this study, we developed a family of mathematical models describing the induction of EMT by TGF-$\beta$ isoforms in the presence and absence of vascular endothelial growth factor A (VEGF-A).
+We integrated a simple rule-based description of activity and gene expression regulation with traditional ordinary differential equation (ODE) modeling to describe an EMT interaction network containing 97 gene, protein or mRNA components interconnected through 169 interactions.
+This integration allows the description of complex regulatory interactions in the absence of specific mechanistic information, it also allowed to build a predictive yet compact model.
 A family of model parameters was estimated using 41 molecular data sets generated in DLD1 colon carcinoma, MDCKII and A375 melanoma cells using the Pareto optimal ensemble technique (JuPOETs) multiobjective optimization algorithm.
@@
 The EMT model contained 74 molecular species interconnected by 169 interactions.
-Model equations were formulated using either saturation or mass-action kinetics within an ordinary differential equation (ODE) framework.
-ODEs and mass action kinetics are common tools to model biochemical pathways \citep{Schoeberl:2002rv,Chen:2009uq,Tasseff:2011ve}.
+Model equations were formulated as ordinary differential equations (ODEs) augmented with rule-based descriptions of activity and gene expression regulation.
+ODEs are common tools to model biochemical pathways \citep{Schoeberl:2002rv,Chen:2009uq,Tasseff:2011ve}.
 However, while ODE models can simulate complex intracellular behavior,
 they require estimates for model parameters which are often difficult to obtain.
-The EMT model had 251 unknown model parameters, 169 kinetic constants and 45 non-zero initial conditions.
+The EMT model had 251 unknown model parameters, 169 kinetic constants 38 control constants and 44 saturation constants.
 As expected, these parameters were not uniquely identifiable given the training data \citep{Gadkar:2005vn}.
@@
 exposure to TGF-$\beta$ isoforms \citep{Medici:2008fk}. We organized these data sets into 11 objective functions which were simultaneously minimized by JuPOETs.
-Additionally, we used 12 molecular data sets generated in HK-2 cells following VEGF-A exposure to train VEGF-A responsive model processes \citep{Lian:2011bs}.
+Additionally, we used data generated in this study (Fig.
+\ref{fg:VEGFA-Data}), and 12 molecular data sets generated in HK-2 cells following VEGF-A exposure to train VEGF-A responsive model processes \citep{Lian:2011bs}.
 To guard against overfitting, we augmented the multiobjective optimization with leave-one-out cross validation to independently estimate both the training and
@@
 
-The predictive power of the ensemble was tested using both cross validation and by comparing simulations with data sets not used for model training.
+The predictive power of the ensemble was tested using cross validation and by comparing simulations with data not used for model training.
 In whole, all of our training objectives were statistically significant (at a 95\% confidence interval) compared to a randomized parameter family (N = 100)
@@
 Elevated YREG1 repressed LEF-1 and TGF-$\beta$3 expression, thereby not allowing free $\beta$-catenin to form the $\beta$-catenin-LEF-1 complex, or TGF-$\beta$3 induced SMAD activation.
-Taken together, low YREG1 expression was required for the maintenance of a stable epithelial phenotype that was simultaneously inducible across TGF-$\beta$1/2, TGF-$\beta$3 and SNAIl/SLUG transfection, as seen in the training objectives.
+Taken together, low YREG1 expression was required for the maintenance of a stable epithelial phenotype that was simultaneously inducible across TGF-$\beta$1/2, TGF-$\beta$3 and SNAIl/SLUG transfection.
+
+%as seen in the training objectives.
 
@@
 \section*{Discussion}
-In this study, we developed a family of mechanistic models describing the induction of EMT by TGF-$\beta$ isoforms in the presence and absence of VEGF-A.
-The signaling architecture encoded in the model, which contained 74 molecular species interconnected by 169 interactions,
+In this study, we developed a family of mathematical models describing the induction of EMT by TGF-$\beta$ isoforms in the presence and absence of VEGF-A.
+The model, which contained 74 molecular species interconnected by 169 interactions,
 described the expression of 23 genes in response to growth factor stimulation.
-This simulation incorporates an unprecedented level of detail compared to previous models, but as a consequence created a large number of unknown model parameters.
-Because these parameters could not be estimated uniquely apriori, we estimated an ensemble of likely parameters using the JuPOETs multiobjective optimization framework.
+We estimated an ensemble of likely model parameters using the JuPOETs multiobjective optimization framework.
 The model population was trained and cross-validated to prescribe biological significance using 41 data sets generated in DLD1 colon carcinoma, MDCKII, and A375 melanoma cell lines \citep{Medici:2008fk}.
@@
 While much research has tested signaling pathways individually, far less is understood about combinatorial stimulation, such as with both VEGF-A and TGF$-\beta$.
-Recent \emph{in vitro} and \emph{in vivo} evidence has suggested that epithelial cells can exhibit heterogeneous phenotypes in addition to classically defined epithelial or mesenchymal states \citep{Polyak2009,Strauss:2011fk}.
+Both \emph{in vitro} and \emph{in vivo} studies have suggested that epithelial cells can exhibit heterogeneous phenotypes in addition to classically defined epithelial or mesenchymal states \citep{Polyak2009,Strauss:2011fk}.
 For example, expression profiling in human epithelial cancer cell lines demonstrated a spectrum of phenotypes, including some that expressed both E-cadherin and Vimentin simultaneously \citep{Neve:2006uq,Welch-Reardon:2014aa}.
@@
 E-cadherin transcription).
-The identification problem for the EMT model was massively underdetermined.
-This is not uncommon for differential equation models, especially those that are highly mechanistic.
-Of course, we could have discarded mechanism or reduced the model scope to decrease the complexity of the identification problem.
+The identification problem for the EMT model was underdetermined (not uncommon for differential equation based models).
 However, a central criticism leveled by biologists is that model simplification is often done at the cost of biological reality, or done for reasons of computational expediency \citep{Sainani2012}.
@@
 In that case, distributions were generated because the model parameters varied over the ensemble, i.e., extrinsic noise led to population heterogeneity.
-In this study, parameters controlling physical interactions such as disassociation rates or the rate of assembly or degradation of macromolecular machinery such as ribosomes were widely distributed over the ensemble.
+In this study, parameters controlling physical interactions such as disassociation rates, or processes such as gene expression were distributed over the ensemble.
 Population heterogeneity can also arise from intrinsic thermal fluctuations, which are not captured by a deterministic population of models \citep{Swain:2002ve}.
@@
 
-Recent evidence has also suggested an essential role of NF-$\kappa$B in epithelial transformation.
+NF-$\kappa$B may also play an essential role of in the epithelial transformation.
 NF-$\kappa$B may influence Snail expression through the AKT pathway and directly stabilize Snail activity \citep{Wu:2009cr}.
@@
 Using a canonical architecture allowed us to explore general features of TGF$-\beta$ induced EMT without cell line specific artifacts.
-On the other hand, because of the canonical architecture, we can test the model against several cell lines to test the generality of our conclusions.
+%On the other hand, because of the canonical architecture, we can test the model against several cell lines to test the generality of our conclusions.
 
@@
 \end{equation}
-where $\mathcal{R}$ denotes the number of signaling reactions and $\mathcal{M}$ denotes the number of proteins in the model.
-The quantity $\tau_{i}$ denotes a time scale parameter for species $i$ which captures unmodeled effects. In the current study $\tau_{i} = 1$ for all species.
+where $\mathcal{R}$ denotes the number of signaling reactions and $\mathcal{M}$ denotes the number of signaling proteins in the model.
+The quantity $\tau_{i}$ denotes a time scale parameter for species $i$ which captures un-modeled effects; in the current study $\tau_{i} = 1$ for all species.
 The quantity $r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)$ denotes the rate of reaction $j$.
@@
 where $k_{j}^{max}$ denotes the maximum rate for reaction $j$, $\sigma_{sj}$ denotes the stoichiometric coefficient for species $s$ in reaction $j$,
-and $s\in{m_{j}}$ denotes the set of \textit{reactants} for reaction $j$. We assumed all binding interactions were irreversible.
+and $s\in{m_{j}}$ denotes the set of \textit{reactants} for reaction $j$. Reversible binding was decomposed into two irreversible steps.
+
+%We assumed all binding interactions were irreversible.
 
@@
 The function $\mathcal{I}_{j}\left(\cdot\right)$ is an integration rule which maps the output of regulatory transfer functions into a control
-variable. In this study, we used $\mathcal{I}_{j}\in\left\{min,max\right\}$ \citep{pr3010178}. If a process has no modifying factors, $v_{j}=1$.
+variable. In this study, we used $\mathcal{I}_{j}\in\left\{min,max\right\}$ and hill transfer functions \citep{pr3010178,pr3010138}.
+If a process has no modifying factors, $v_{j}=1$.
 
@@
 The EMT model described both signal transduction and gene expression events following the addition of TGF$-\beta$ and VEGF-A.
-For each gene, we modeled both the resulting mRNA ($m_{j}$) and protein ($p_{j}$):
+For each gene of the $\mathcal{G}=23$ we considered, we modeled both the resulting mRNA ($m_{j}$) and protein ($p_{j}$):
 \begin{eqnarray}
@@
 \end{eqnarray}where $j=1,2,\dots,\mathcal{G}$.
-The terms $r_{T,j}$ and $r_{X,j}$ denote the specific rate of transcription, and translation while
+The terms $r_{T,j}$ and $r_{X,j}$ denote the specific rates of transcription, and translation while
 the terms $\delta_{m,j}$ and $\delta_{p,j}$ denote degradation constants for mRNA and protein, respectively.
@@
 The specific transcription rate was modeled as the product of a kinetic term $\bar{r}_{T,j}$ and a control term $u_{j}$ which described how the
-abundance of transcription factors, or other regulators influenced the expression of gene $j$. The kinetic rate of transcription was modeled as:
+abundance of transcription factors, or other regulators influenced the expression of gene $j$.
+The kinetic rate of transcription was modeled as:
 \begin{equation}
@@
 \end{equation}where the maximum gene expression rate was defined as the product of a characteristic transcription rate constant ($k_{T}$)
-and the abundance of RNA polymerase, $V^{max}_{T} = k_{T}\left(RNAP\right)$. The parameter $\alpha_{j}$ was used to adjust the transcription to that of gene $j$
+and the abundance of RNA polymerase, $V^{max}_{T} = k_{T}\left(RNAP\right)$. The parameter $\alpha_{j}$ denotes the gene specific correction to the characteristic transcription rate.
 (estimated in this study), while $k_{T}$, $G_{j}$ and $RNAP$ were estimated from literature \citep{Milo:2010aa}.
@@
 The function $\mathcal{I}_{j}\left(\cdot\right)$ is an integration rule which maps the output of regulatory transfer functions into a control
-variable. In this study, we used $\mathcal{I}_{j}\in\left\{min,max\right\}$ \citep{pr3010178}.
+variable. In this study, we used $\mathcal{I}_{j}\in\left\{min,max\right\}$ and hill transfer functions \citep{pr3010178,pr3010138}.
 If a gene expression process has no modifying factors, $u_{j}=1$.
@@
 and the abundance of Ribosomes ($RIBO$), $V^{max}_{X} = k_{X}\left(RIBO\right)$, where both $k_{X}$ and $RIBO$ abundance were estimated from literature \citep{Milo:2010aa}.
-The values used for the characteristic transcription and translation parameters are given in the supplemental materials.
+
+In this study, we estimated $k_{T}, k_{X}, RNAP$ and $RIBO$ from literature. We calculated the concentration for gene $j$ by assuming, on average,
+the cell had two copies of each gene at any given time. We also estimated characteristic values for $\delta_{m,j}$ and $\delta_{p,j}$,
+and corrected these values with mRNA/protein specific correction factors. Thus, the bulk of our gene expression parameters were based directly upon literature values.
+The values used for the characteristic transcription/translation parameters are given in the supplemental materials.
 
@@
 with leave-one-out cross-validation to estimate an ensemble of model parameters \citep{Song:2010fk,JuPOETs-BioArXiv}.
+Model parameter values were adjusted to minimize the residual between simulations and experimental measurements for the 11 objective functions (supplemental materials).
 Cross-validation was used to calculate both training and prediction error during the parameter estimation procedure \citep{kohavi1995study}.
@@
 The training and validation data were Western blots.
-We achived a physical simulation scale by establishing characteristic rates of transcription, translation, mRNA and protein degradation, as well as characteristic concentrations of ribosomes and RNAPs \citep{Milo:2010aa}.
-The concentration scale is in nM, with proteins ranging from 10-1000nM and mRNA ranging from 0.01 to 1nM, reflecting the true abundances and ratios between each species.
+We achieved a biologically realistic concentration scale by establishing characteristic rates of transcription, translation, mRNA and protein degradation, as well as characteristic concentrations of ribosomes and RNAPs using the Bionumbers database \citep{Milo:2010aa}, (supplemental materials).
+The overall concentration scale was nM, with proteins ranging from 10-1000nM and mRNA ranging from 0.01 to 1nM, reflecting the true abundances and ratios between each species.
+An initial nominal parameter set was established by inspection. JuPOETs was then allowed to search in a neighborhood of $\pm$30\% of this nominal set.
+The parameter ensemble estimated by JuPOETs is available with the model source code.
+JuPOETs is open source and freely available for download under an MIT software license from http://www.varnerlab.org.
 

### 2016-10-04 Paper_v8.tex
PLoS formatting

--- 2016-10-04 Paper_v7.tex
@@
-\documentclass[12pt]{article}
-% Load packages
-\usepackage{url}  % Formatting web addresses
-\usepackage{ifthen}  % Conditional
-\usepackage{multicol}   %Columns
-\usepackage[utf8]{inputenc} %unicode support
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{epsfig}
-\usepackage{epstopdf}
-\usepackage{graphicx}
-\usepackage[margin=0.1pt,font=footnotesize,labelfont=bf]{caption}
-\usepackage{setspace}
-%\usepackage{longtable}
-\usepackage{colortbl}
-%\usepackage{palatino,lettrine}
-%\usepackage{times}
-%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
-%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
-\usepackage[wide]{sidecap}
-\usepackage[authoryear,round,comma,sort&compress]{natbib}
-\usepackage[authoryear,round]{natbib}
-\usepackage{supertabular}
-\usepackage{simplemargins}
-\usepackage{lineno}
+\documentclass[11pt,letterpaper]{article}
+\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
+
+% Use adjustwidth environment to exceed column width (see example table in text)
+\usepackage{changepage}
+
+% Use Unicode characters when possible
+\usepackage[utf8]{inputenc}
+
+% textcomp package and marvosym package for additional characters
+\usepackage{textcomp,marvosym}
+
+% fixltx2e package for \textsubscript
+\usepackage{fixltx2e}
+
+% amsmath and amssymb packages, useful for mathematical formulas and symbols
+\usepackage{amsmath,amssymb}
+
+% cite package, to clean up citations in the main text. Do not remove.
+\usepackage{cite}
+
+% Use nameref to cite supporting information files (see Supporting Information section for more info)
+\usepackage{nameref,hyperref}
+
+% line numbers
+\usepackage[right]{lineno}
+
+% ligatures disabled
+\usepackage{microtype}
+\DisableLigatures[f]{encoding = *, family = * }
+
+% rotating package for sideways tables
+\usepackage{rotating}
+
+% Remove comment for double spacing
+%\usepackage{setspace}
+%\doublespacing
+
+% Text layout
+\raggedright
+\setlength{\parindent}{0.5cm}
+\textwidth 5.25in
+\textheight 8.75in
+
+% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
+% Captions will be left justified
+\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
+
+% Use the PLoS provided BiBTeX style
+\bibliographystyle{plos2009}
+
+% Remove brackets from numbering in List of References
+\makeatletter
+\renewcommand{\@biblabel}[1]{\quad#1.}
+\makeatother
+
+% Leave date blank
+\date{}
+
+% Header and Footer with logo
+\usepackage{lastpage,fancyhdr,graphicx}
+\pagestyle{myheadings}
+\pagestyle{fancy}
+\fancyhf{}
+\lhead{\includegraphics[natwidth=1.3in,natheight=0.4in]{PLOSlogo.png}}
+\rfoot{\thepage/\pageref{LastPage}}
+\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
+\fancyheadoffset[L]{2.25in}
+\fancyfootoffset[L]{2.25in}
+\lfoot{\sf PLOS}
+
+%% Include all macros below
+
+\newcommand{\lorem}{{\bf LOREM}}
+\newcommand{\ipsum}{{\bf IPSUM}}
 
 \usepackage{listings}
-
-\urlstyle{rm}
-
-%\textwidth = 6.50 in
-%\textheight = 9.5 in
-%\oddsidemargin =  0.0 in
-%\evensidemargin = 0.0 in
-%\topmargin = -0.50 in
-%\headheight = 0.0 in
-%\headsep = 0.25 in
-%\parskip = 0.15in
-%\linespread{1.75}
-\doublespace
-
-\setallmargins{1in}
-
-%\bibliographystyle{plain}
-\bibliographystyle{msb}
-
-\definecolor{mygreen}{rgb}{0,0.6,0}
-\definecolor{mygray}{rgb}{0.5,0.5,0.5}
-\definecolor{mymauve}{rgb}{0.58,0,0.82}
-
+\usepackage{colortbl}
 \lstdefinestyle{myCustomMatlabStyle}{
@@
 basicstyle=\tiny\ttfamily,
-keywordstyle=\bfseries\color{black},
-commentstyle=\itshape\color{black},
+keywordstyle=\bfseries,
+commentstyle=\itshape,
 identifierstyle=\color{blue},
@@
 
-\makeatletter
-\renewcommand\subsection{\@startsection
-	{subsection}{2}{0mm}
-	{-0.05in}
-	{-0.5\baselineskip}
-	{\normalfont\normalsize\bfseries}}
-\renewcommand\subsubsection{\@startsection
-	{subsubsection}{2}{0mm}
-	{-0.05in}
-	{-0.5\baselineskip}
-	{\normalfont\normalsize\itshape}}
-\renewcommand\section{\@startsection
-	{subsection}{2}{0mm}
-	{-0.2in}
-	{0.05\baselineskip}
-	{\normalfont\large\bfseries}}
-\renewcommand\paragraph{\@startsection
-	{paragraph}{2}{0mm}
-	{-0.05in}
-	{-0.5\baselineskip}
-	{\normalfont\normalsize\itshape}}
-\makeatother
-
-%Review style settings
-%\newenvironment{bmcformat}{\begin{raggedright}\baselineskip20pt\sloppy\setboolean{publ}{false}}{\end{raggedright}\baselineskip20pt\sloppy}
-
-%Publication style settings
-
-% Single space'd bib -
-\setlength\bibsep{0pt}
-
-\renewcommand{\rmdefault}{phv}\renewcommand{\sfdefault}{phv}
-
-% Change the number format in the ref list -
-\renewcommand{\bibnumfmt}[1]{#1.}
-
-% Change Figure to Fig.
-\renewcommand{\figurename}{Fig.}
+%% END MACROS SECTION
 
 \begin{document}
+\vspace*{0.35in}
+\begin{flushleft}
+{\Large
+\textbf\newline{Population heterogeneity in the epithelial to mesenchymal transition is controlled by NFAT and phosphorylated Sp1}}
+\newline
+% Insert Author names, affiliations and corresponding author email.
+\\
+Russell Gould\textsuperscript{1},
+David Bassen\textsuperscript{1},
+Anirikh Chakrabarti\textsuperscript{2},
+Jeffrey D. Varner\textsuperscript{2},
+Jonathan Butcher\textsuperscript{1,*},
+\\
+\bf{1} Nancy E. and Peter C. Meinig School of Biomedical Engineering, Cornell University, Ithaca NY
+\\
+\bf{2} Robert Frederick Smith School of Chemical and Biomolecular Engineering, Cornell University, Ithaca, NY, USA
 
-\begin{titlepage}
-{\par\centering\textbf{\Large Population heterogeneity in the epithelial to mesenchymal transition is controlled by NFAT and phosphorylated Sp1}}
-\vspace{0.05in}
-{\par \centering Russell Gould$^{\ddag}$ and David M. Bassen, Anirikh Chakrabarti, Jeffrey D. Varner and Jonathan Butcher$^{\ddag *}$}
-\vspace{0.05in}
-{\par \centering Robert Frederick Smith School of Chemical and Biomolecular Engineering and $^{\ddag}$Nancy E.
-and Peter C.
-Meinig School of Biomedical Engineering, Cornell University, Ithaca NY 14853}
-\vspace{0.1in}
-{\par \centering \textbf{Running Title:}~ Modeling of TGF-$\beta$ induced EMT}
-\vspace{0.1in}
-% {\par \centering \textbf{To be submitted:}~\emph{Molecular Systems Biology (60,000 Character Limit)}}
-%\vspace{0.1in}
-{\par \centering \textbf{Character count:}~\emph{56,659}}
-\vspace{0.1in}
-{\par \centering $^{*}$Corresponding author:}
-\vspace{0.1in}
-{\par \centering Jonathan T. Butcher,}
-{\par \centering Associate Professor, Nancy E. and Peter C. Meinig School of Biomedical Engineering,}
-{\par \centering 304 Weill Hall, Cornell University, Ithaca NY, 14853}
-{\par \centering Email: jtb47@cornell.edu}
-{\par \centering Phone: (607) 255 - 3575}
-{\par \centering Fax: (607) 255 - 7330}
-\end{titlepage}
-\date{}
-\thispagestyle{empty}
-\pagebreak
+% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
+%
+% Remove or comment out the author notes below if they aren't used.
+%
+% Primary Equal Contribution Note
+%\Yinyang These authors contributed equally to this work.
+
+% Additional Equal Contribution Note
+%\ddag These authors also contributed equally to this work.
+
+% Current address notes
+%\textcurrency a Insert current address of first author with an address update
+% \textcurrency b Insert current address of second author with an address update
+% \textcurrency c Insert current address of third author with an address update
+
+% Deceased author note
+%\dag Deceased
+
+% Group/Consortium Author Note
+%\textpilcrow Insert Collaborative Author line here
+
+* E-mail: jtb47@cornell.edu
+\end{flushleft}
+
 \section*{Abstract}
@@
 
-%as seen by the initiating the formation of endocardial cushions in the atrioventricular valves and mediating palate fusion of medial epithelial cells \citep{Zavadil:2005fu}.
-In this context, EMT terminates and has been suggested as a highly controlled process \citep{Lee:2006tw}.
-EMT is also important under pathological conditions, prominently in fibrosis and in invasion and metastasis of carcinomas \citep{Thiery:2003il}.
-This culminates decades of work within the EMT field appears shared in developmental and pathological conditions \citep{Xu:2009oq}.
-%In this study, we determined the extent of EMT in normal epithelial cells exposed to pathological conditions by (1) implementing and subsequently interrogating a population based model containing canonical EMT regulatory networks, and (2) verify our mechanistic targets using established cell lines with differing induction capacities for EMT.
-Toward this objective, we modeled the core molecular interactions of EMT using mass-action kinetics within an ordinary differential equation (ODE) framework.
-A population approach was used to estimate a family or ensemble of EMT models that was consistent with the previous literature \citep{Medici:2006qa,Medici:2008fk}.
-Using the newly predicted operational paradigms, we sought to capture the role of phospho-Sp1 and NFAT on phenotype heterogeneity during TGF$\beta$ induced EMT \emph{in vitro} via key marker expression profiles and fate processes.
-%Infrastructure components include TGF-$\beta$2 ligand binding, TGF-$\beta$RI,II,III transmembrane growth factor receptor phosphorylation, activation of the Ras/MAPK cascade, cytosolic release of $\beta$-catenin from E-cadherin, autocrine response of the TGF-$\beta$3 ligand, phosphorylation of the Smad proteins, and cross-talk between the MAPK/Smad signaling pathways.
-%Furthermore, because these reactions occur with varying initial conditions and unknown parameter values, significant  differences may exist in the overall response of the model.
-%To unravel the operation of this complex architecture, we have used mathematical modeling.
-%In this program, extracellular TGF-$\beta$1/2 induces a series of  that ultimately repress epithelial gene expression
-%while simultaneously activating mesenchymal gene expression.
-%However, this program is not a simple switch. It involves
-%Thus, the molecular architecture responsible for EMT induction is highly complex, involving multiple inputs and overlapping transcriptional networks.
-%However, including these additional mechanisms increases model complexity.
-%Analysis of the model predicted that the NFATc and phosphorylated Sp1 transcription factors were likely the key elements driving this heterogeneity.
-
 \linenumbers
@@
 \section*{Introduction}
-The epithelial to mesenchymal transition (EMT) is a broadly participating, evolutionarily conserved differentiation program essential for tissue morphogenesis, remodeling and pathological processes such as cancer \citep{Thiery:2003il}.
-During EMT polarized, tightly adhered epithelial cell monolayers are transformed into non-interacting motile mesenchymal cells that simultaneously degrade and synthesize extracellular matrix (ECM) components and invade into the underlying tissue space \citep{Stahl:2001pd}.
-EMT is the fundamental initiator of developmental processes such as embryonic gastrulation and valvulogenesis \citep{Eisenberg:1995ai} (also Kalluri J Clin Invest 2009, Thiery Cell 2009).
-Transforming growth factor $\beta$ (TGF-$\beta$) family members are important inducers of both developmental and pathological EMT \citep{Zavadil:2005fu,Xu:2009oq}.
+The epithelial to mesenchymal transition (EMT) is a broadly participating, evolutionarily conserved differentiation program essential for tissue morphogenesis, remodeling and pathological processes such as cancer \cite{Thiery:2003il}.
+During EMT polarized, tightly adhered epithelial cell monolayers are transformed into non-interacting motile mesenchymal cells that simultaneously degrade and synthesize extracellular matrix (ECM) components and invade into the underlying tissue space \cite{Stahl:2001pd}.
+EMT is the fundamental initiator of developmental processes such as embryonic gastrulation and valvulogenesis \cite{Eisenberg:1995ai} (also Kalluri J Clin Invest 2009, Thiery Cell 2009).
+Transforming growth factor $\beta$ (TGF-$\beta$) family members are important inducers of both developmental and pathological EMT \cite{Zavadil:2005fu,Xu:2009oq}.
 Decades of research has focused on identifying molecular regulators of EMT, but almost all on a single gene and in a nearly binary yes/no level of qualitative understanding.
-Medici and coworkers identified a core signaling program by which TGF-$\beta$ isoforms induce EMT across a variety of cell lines \citep{Medici:2006qa,Medici:2008fk}.
+Medici and coworkers identified a core signaling program by which TGF-$\beta$ isoforms induce EMT across a variety of cell lines \cite{Medici:2006qa,Medici:2008fk}.
 This program involves carefully orchestrated rounds of gene expression driven by the Smad and Snail families of transcription factors as well as
 other key factors such as lymphoid enhancer-binding factor 1 (LEF-1), nuclear factor of activated T-cells, cytoplasmic 1 (NFATc1), and specificity protein 1 (Sp1).
-Coregulators such as $\beta$-catenin, NF-$\kappa$B, and the ErbB family of receptor tyrosine kinases however also participate in EMT regulation, but the degree of each's influence is difficult to ascertain in isolation \citep{Kim:2002lh,Jiang:2007aa,Huber:2004aa,Hardy:2010aa}.
+Coregulators such as $\beta$-catenin, NF-$\kappa$B, and the ErbB family of receptor tyrosine kinases however also participate in EMT regulation, but the degree of each's influence is difficult to ascertain in isolation \cite{Kim:2002lh,Jiang:2007aa,Huber:2004aa,Hardy:2010aa}.
 EMT also exhibits complex temporal dynamics that are often intractable in gain/loss of function studies.
@@
 
-Systems biology and mathematical modeling are essential tools for understanding complex developmental programs like EMT \citep{Ahmed:2007mi}.
+Systems biology and mathematical modeling are essential tools for understanding complex developmental programs like EMT \cite{Ahmed:2007mi}.
 Previous computational models of TGF-$\beta$ induced differentiation focused on single biological factors or EMT in single cells.
 For example, Chung \emph{et al.}, constructed a model of TGF-$\beta$ receptor activation and Smad signaling using ordinary differential equations and mass-action kinetics.
-Their model suggested that a reduction of functional TGF-$\beta$ receptors in cancer cells may lead to an attenuated Smad2 signal \citep{Chung:2009jl}.
-Similarly, Vilar \emph{et al.} suggested that specific changes in receptor trafficking patterns could lead to phenotypes that favor tumorigenesis \citep{Vilar:2006gb}.
+Their model suggested that a reduction of functional TGF-$\beta$ receptors in cancer cells may lead to an attenuated Smad2 signal \cite{Chung:2009jl}.
+Similarly, Vilar \emph{et al.} suggested that specific changes in receptor trafficking patterns could lead to phenotypes that favor tumorigenesis \cite{Vilar:2006gb}.
 Although these models provided insight into the role of receptor dynamics, EMT induction involves many other components, including competing second messengers and interconnected transcriptional regulatory loops.
 Integrating these additional scales of molecular signaling while maintaining the capacity for robust prediction requires a new and expanded computational and experimental strategy.
-Data-driven systems approaches \citep{Cirit:2012kx} or logical model formulations \citep{Morris:2011ys} are emerging paradigms that constrain model complexity through the incorporation of training and validation data.
+Data-driven systems approaches \cite{Cirit:2012kx} or logical model formulations \cite{Morris:2011ys} are emerging paradigms that constrain model complexity through the incorporation of training and validation data.
 These are interesting techniques because the data informs model structure (which can be expanded as more data becomes available).
-Alternatively, Bailey proposed more than a decade ago that a qualitative understanding of a complex biological system should not require complete definition of its structural and parametric content \citep{2001_bailey_NatBiotech}.
-Shortly thereafter, Sethna and coworkers showed that complex model behavior is often controlled by only a few parameter combinations, a characteristic seemingly universal to multi-parameter models referred to as ``sloppiness'' \citep{Machta:2013by}.
+Alternatively, Bailey proposed more than a decade ago that a qualitative understanding of a complex biological system should not require complete definition of its structural and parametric content \cite{2001_bailey_NatBiotech}.
+Shortly thereafter, Sethna and coworkers showed that complex model behavior is often controlled by only a few parameter combinations, a characteristic seemingly universal to multi-parameter models referred to as ``sloppiness'' \cite{Machta:2013by}.
 Thus, reasonable model predictions are often possible with only limited parameter information.
-Taking advantage of this property, we developed sloppy techniques for parameter identification using ensembles of deterministic models \citep{Song:2010fk}.
-Furthermore, we proposed that the sloppy behavior of biological networks may also be seen as a source of cell-to-cell \citep{Lequieu:2011fj} or even patient-to-patient heterogeneity \citep{2010_luan_varner_MolBioSys}.
-Bayesian parameter identification techniques have also been used to explore cell-to-cell heterogeneity  \citep{2011_kalita_brasier_JBC,2011_hasenauer_allgower_BMCBioinfo},
-where a population of cells could be viewed as a dynamic ensemble of context-specific biochemical networks \citep{2012_creixell_linding_NatBiotech}.
+Taking advantage of this property, we developed sloppy techniques for parameter identification using ensembles of deterministic models \cite{Song:2010fk}.
+Furthermore, we proposed that the sloppy behavior of biological networks may also be seen as a source of cell-to-cell \cite{Lequieu:2011fj} or even patient-to-patient heterogeneity \cite{2010_luan_varner_MolBioSys}.
+Bayesian parameter identification techniques have also been used to explore cell-to-cell heterogeneity  \cite{2011_kalita_brasier_JBC,2011_hasenauer_allgower_BMCBioinfo},
+where a population of cells could be viewed as a dynamic ensemble of context-specific biochemical networks \cite{2012_creixell_linding_NatBiotech}.
 
@@
 Model equations were formulated as ordinary differential equations (ODEs) augmented with rule-based descriptions of activity and gene expression regulation.
-ODEs are common tools to model biochemical pathways \citep{Schoeberl:2002rv,Chen:2009uq,Tasseff:2011ve}.
+ODEs are common tools to model biochemical pathways \cite{Schoeberl:2002rv,Chen:2009uq,Tasseff:2011ve}.
 However, while ODE models can simulate complex intracellular behavior,
@@
 The EMT model had 251 unknown model parameters, 169 kinetic constants 38 control constants and 44 saturation constants.
-As expected, these parameters were not uniquely identifiable given the training data \citep{Gadkar:2005vn}.
+As expected, these parameters were not uniquely identifiable given the training data \cite{Gadkar:2005vn}.
 Thus, instead of identifying a single best fit (but uncertain) model,
@@
 The training data were generated in DLD1 colon carcinoma, MDCKII, and A375 melanoma cells following
-exposure to TGF-$\beta$ isoforms \citep{Medici:2008fk}. We organized these data sets into 11 objective functions which were simultaneously minimized by JuPOETs.
+exposure to TGF-$\beta$ isoforms \cite{Medici:2008fk}. We organized these data sets into 11 objective functions which were simultaneously minimized by JuPOETs.
 Additionally, we used data generated in this study (Fig.
-\ref{fg:VEGFA-Data}), and 12 molecular data sets generated in HK-2 cells following VEGF-A exposure to train VEGF-A responsive model processes \citep{Lian:2011bs}.
+\ref{fg:VEGFA-Data}), and 12 molecular data sets generated in HK-2 cells following VEGF-A exposure to train VEGF-A responsive model processes \cite{Lian:2011bs}.
 To guard against overfitting, we augmented the multiobjective optimization with leave-one-out cross validation to independently estimate both the training and
@@
 The selected models all had the same possible molecular connectivity, but different values for model parameters.
-Transcription and translation rates, as well as mRNA and protein degradation terms, were set using physical values from the literature \citep{Milo:2010aa},
+Transcription and translation rates, as well as mRNA and protein degradation terms, were set using physical values from the literature \cite{Milo:2010aa},
 and allowed to vary by a scaling factor, see methods.
@@
 \ref{fg:F2}J-L).
-Taken together, the model captured the key signaling events revealed by Medici \emph{et al.} \citep{Medici:2008fk} that drive the phenotypic conversion.
+Taken together, the model captured the key signaling events revealed by Medici \emph{et al.} \cite{Medici:2008fk} that drive the phenotypic conversion.
 A listing of objective function values resulting from training, cross validation and the random parameter control is given in the supplement (Fig. \ref{fg:ObjTable}).
@@
 and the upregulation of LEF-1 expression following TGF-$\beta$1/2 exposure.
-SNAIL/SLUG are well known transcriptional repressors \citep{Hemavathy:2000aa,Hemavathy:2000ab,Dhasarathy:2011aa}, although there are a few studies
-which suggest that at least SNAIL can also act as a transcriptional activator \citep{Guaita:2002aa}.
-In the model, we assumed the expression of SNAIL/SLUG was likely regulated by AP1/SP1 \citep{Jackstadt:2013aa}.
+SNAIL/SLUG are well known transcriptional repressors \cite{Hemavathy:2000aa,Hemavathy:2000ab,Dhasarathy:2011aa}, although there are a few studies
+which suggest that at least SNAIL can also act as a transcriptional activator \cite{Guaita:2002aa}.
+In the model, we assumed the expression of SNAIL/SLUG was likely regulated by AP1/SP1 \cite{Jackstadt:2013aa}.
 Thus, upon receiving direct SNAIL/SLUG and TGF-$\beta$3 signals, the model predicted enhanced SNAIL/SLUG expression, consistent with experimental observations.
-TGF-$\beta$1/2 stimulation also induces LEF-1 expression. However, literature evidence suggested that LEF-1 expression was not strongly dependent upon AP1/SP1 activity \citep{Eastman:1999aa}.
+TGF-$\beta$1/2 stimulation also induces LEF-1 expression. However, literature evidence suggested that LEF-1 expression was not strongly dependent upon AP1/SP1 activity \cite{Eastman:1999aa}.
 Thus, either SNAIL/SLUG are acting as inducers (contrary to substantial biochemical evidence) or, they are repressing the expression of an intermediate repressor.
@@
 By adding this regulator, our simulations became consistent with training and literature data.
-Medici et al. suggested that feedback between $\beta$-catenin and LEF-1 was likely, although this feedback had yet to be identified \citep{Medici:2008fk}.
+Medici et al. suggested that feedback between $\beta$-catenin and LEF-1 was likely, although this feedback had yet to be identified \cite{Medici:2008fk}.
 Low levels of YREG1 expression were present in all simulations to regulate the formation of the $\beta$-catenin-LEF-1 complex.
@@
 While we captured the central tendency of many of the molecular features of EMT induction following TGF$-\beta$1/2 exposure, an often neglected but important emergent
-feature of developmental and pathological programs is population heterogeneity \citep{Park:2010fu}.
-We (and others) have previously hypothesized that deterministic model ensembles can simulate population behavior, at least at a course grained level \citep{Lequieu:2011fj}.
+feature of developmental and pathological programs is population heterogeneity \cite{Park:2010fu}.
+We (and others) have previously hypothesized that deterministic model ensembles can simulate population behavior, at least at a course grained level \cite{Lequieu:2011fj}.
 We tested this hypothesis by analyzing the response of the population of EMT models to extracellular cues and then comparing this response to flow cytometry studies.
@@
 \ref{fg:F5}E,F) and DLD1 cells (Fig.
-\ref{fg:S4}) to combinations of VEGF-A and TGF$-\beta$1/2 in the presence or absence of VIVIT, a soluble peptide inhibitor of NFATc transcriptional activity \citep{Aramburu:1999if}.
+\ref{fg:S4}) to combinations of VEGF-A and TGF$-\beta$1/2 in the presence or absence of VIVIT, a soluble peptide inhibitor of NFATc transcriptional activity \cite{Aramburu:1999if}.
 Treatment with VEGF-A (50ng/ml) and VIVIT (10$\mu$M) in MCF10A cells significantly reduced E-cadherin expression compared to VEGF-A alone (Fig. \ref{fg:F5}D,E).
@@
 \subsection*{Ductal branching during acini formation is dependent upon phenotype heterogeneity in MCF10A and DLD1 cells}
-We finally employed established three-dimensional (3D) \emph{in vitro} models of invasion, migration, compaction, and tubulogenesis \citep{Dhimolea2010}
+We finally employed established three-dimensional (3D) \emph{in vitro} models of invasion, migration, compaction, and tubulogenesis \cite{Dhimolea2010}
 to determine the functional consequences of the hybrid phenotype (Fig. \ref{fg:F6}).
@@
 The circularity index, a common means of quantifying cell morphology, relates cell area to perimeter.
-A perfect circle has a circularity index equal to 1.0, while a straight line has a circularity index equal to 0.0, see Butcher et al. \citep{Butcher:2004aa}.
+A perfect circle has a circularity index equal to 1.0, while a straight line has a circularity index equal to 0.0, see Butcher et al. \cite{Butcher:2004aa}.
 TGF$-\beta$1/2 treatment alone resulted in irregular and spindle shaped morphology, while VEGF-A exposure promoted round quiescent cells (Fig. \ref{fg:F6}A).
@@
 We estimated an ensemble of likely model parameters using the JuPOETs multiobjective optimization framework.
-The model population was trained and cross-validated to prescribe biological significance using 41 data sets generated in DLD1 colon carcinoma, MDCKII, and A375 melanoma cell lines \citep{Medici:2008fk}.
+The model population was trained and cross-validated to prescribe biological significance using 41 data sets generated in DLD1 colon carcinoma, MDCKII, and A375 melanoma cell lines \cite{Medici:2008fk}.
 Analysis of this population predicted possible phenotypic modes (and their associated signaling) that cells could exhibit when stimulated with TGF-$\beta$ and/or VEGF-A.
@@
 Cells routinely process a multitude of signals simultaneously, especially when coordinating developmental or pathological programs.
-For example, oncogenic cells integrate both mechanical and chemical cues in their local microenvironment during tumorigenesis, including cytokines VEGF and TGF$-\beta$ \citep{Hong:2013mi}.
-VEGF-A mediates pathological angiogenic remodeling of tumors \citep{Nagy:2007lh}, while TGF$-\beta$ can elicit both protective and oncogenic responses \citep{Ferrara:2002kx,Willis:2007uq}.
+For example, oncogenic cells integrate both mechanical and chemical cues in their local microenvironment during tumorigenesis, including cytokines VEGF and TGF$-\beta$ \cite{Hong:2013mi}.
+VEGF-A mediates pathological angiogenic remodeling of tumors \cite{Nagy:2007lh}, while TGF$-\beta$ can elicit both protective and oncogenic responses \cite{Ferrara:2002kx,Willis:2007uq}.
 While much research has tested signaling pathways individually, far less is understood about combinatorial stimulation, such as with both VEGF-A and TGF$-\beta$.
-Both \emph{in vitro} and \emph{in vivo} studies have suggested that epithelial cells can exhibit heterogeneous phenotypes in addition to classically defined epithelial or mesenchymal states \citep{Polyak2009,Strauss:2011fk}.
-For example, expression profiling in human epithelial cancer cell lines demonstrated a spectrum of phenotypes, including some that expressed both E-cadherin and Vimentin simultaneously \citep{Neve:2006uq,Welch-Reardon:2014aa}.
-Zajchowski \emph{et al.}, speculated that these expression profiles were somehow important for maintaining epithelial properties, while simultaneously allowing other functional behavior such as proliferation and migration \citep{Zajchowski:2001kx}.
+Both \emph{in vitro} and \emph{in vivo} studies have suggested that epithelial cells can exhibit heterogeneous phenotypes in addition to classically defined epithelial or mesenchymal states \cite{Polyak2009,Strauss:2011fk}.
+For example, expression profiling in human epithelial cancer cell lines demonstrated a spectrum of phenotypes, including some that expressed both E-cadherin and Vimentin simultaneously \cite{Neve:2006uq,Welch-Reardon:2014aa}.
+Zajchowski \emph{et al.}, speculated that these expression profiles were somehow important for maintaining epithelial properties, while simultaneously allowing other functional behavior such as proliferation and migration \cite{Zajchowski:2001kx}.
 Whether and how heterogeneous phenotypes arise and participate in cancer progression, as well as their response to pharmacological inhibition are fundamental questions
@@
 Furthermore, our systematic simulation-experimentation strategy identified that the transcriptional activity of Sp1 and NFATc were the critical factors controlling this phenotypic heterogeneity.
-Several studies have highlighted the importance of NFATc as a key transcription factor involved in cell growth, survival, invasion, angiogenesis and cancer \citep{Mancini2009}.
-For example, proliferation and anchorage-independent growth of pancreatic tumor cells is dependent on calcineurin and NFATc1 activity, consistent with the high levels of nuclear NFATc1 found in pancreatic tumors \citep{Singh:2010jx}.
+Several studies have highlighted the importance of NFATc as a key transcription factor involved in cell growth, survival, invasion, angiogenesis and cancer \cite{Mancini2009}.
+For example, proliferation and anchorage-independent growth of pancreatic tumor cells is dependent on calcineurin and NFATc1 activity, consistent with the high levels of nuclear NFATc1 found in pancreatic tumors \cite{Singh:2010jx}.
 Likewise, our results found that VEGF-A was a potent inducer of NFATc1 expression, which may be required for epithelial cell migration and tubulogenesis.
@@
 Epithelial cells reproduce tissue-like organization when grown in a three-dimensional extracellular matrix (ECM) environment, and therefore are an attractive model to study morphogenic mechanisms.
-It is well established that MCF10A cells form structures that closely resemble acini (multi-lobed cluster of cells) in three-dimensional \emph{in vitro} cultures \citep{Debnath2003}.
-It has been postulated that a cellular response reminiscent of partial EMT underlies this process, stimulating further branching and formation of acini \citep{Pearson:2007qq}.
-Normally well controlled process such as tubulogenesis can be co-opted by cancer cells to break away from a primary lesion and invade through the surrounding stroma \citep{OBrien2004}.
-However, by retaining a transient hybrid EMT-like state, clusters of these tube-forming tumor cells can reform at a high rate after invasion, possibly explaining why invasive human carcinomas frequently appear to be cellular collections with varying degrees of gland-like differentiation \citep{Debnath:2005qc}.
+It is well established that MCF10A cells form structures that closely resemble acini (multi-lobed cluster of cells) in three-dimensional \emph{in vitro} cultures \cite{Debnath2003}.
+It has been postulated that a cellular response reminiscent of partial EMT underlies this process, stimulating further branching and formation of acini \cite{Pearson:2007qq}.
+Normally well controlled process such as tubulogenesis can be co-opted by cancer cells to break away from a primary lesion and invade through the surrounding stroma \cite{OBrien2004}.
+However, by retaining a transient hybrid EMT-like state, clusters of these tube-forming tumor cells can reform at a high rate after invasion, possibly explaining why invasive human carcinomas frequently appear to be cellular collections with varying degrees of gland-like differentiation \cite{Debnath:2005qc}.
 In this study, we showed that our predicted hybrid phenotype generated by simultaneous treatment of epithelial cells with VEGF-A and TGF$-\beta$ possessed altered migration and invasion, which enhanced tubular branching.
@@
 Thus, it is likely that some cancerous epithelial cells could exhibit hybrid EMT phenotypic states.
-This may explain why fibroblastoid morphology, a classical feature of EMT, is not commonly observed in human carcinomas \citep{Debnath:2005qc}.
+This may explain why fibroblastoid morphology, a classical feature of EMT, is not commonly observed in human carcinomas \cite{Debnath:2005qc}.
 This study focused on the combinatorial effects of two very different ligand families present together in the tumor environment.
@@
 The deterministic population of EMT models predicted heterogeneous behavior that was qualitatively consistent with experimental studies.
-There is a diversity of algorithmic approaches to estimate model parameters \citep{Moles:2003aa}, as well as many strategies to integrate model identification with experimental design \citep{Rodriguez-Fernandez:2013aa,Villaverde:2014aa}.
+There is a diversity of algorithmic approaches to estimate model parameters \cite{Moles:2003aa}, as well as many strategies to integrate model identification with experimental design \cite{Rodriguez-Fernandez:2013aa,Villaverde:2014aa}.
 However, despite these advances, the identification of models describing intracellular network behavior remains challenging.
@@
 One school has focused on model reduction.
-Data-driven approaches \citep{Cirit:2012kx}, boolean \citep{Choi:2012bs} or other logical model formulations \citep{Morris:2011ys,Terfve:2012kq}
+Data-driven approaches \cite{Cirit:2012kx}, boolean \cite{Choi:2012bs} or other logical model formulations \cite{Morris:2011ys,Terfve:2012kq}
 are emerging paradigms that constrain model complexity by the availability of the training and validation data.
-Other techniques such as constraints based modeling, which is commonly used to model metabolic networks, have also been applied to model transcriptional networks, although primarily in lower eukaryotes and prokaryotes \citep{Hyduke:2010rt}.
-These techniques (and many others, see review \citep{Wayman2013}) are certainly exciting, with many interesting properties.
+Other techniques such as constraints based modeling, which is commonly used to model metabolic networks, have also been applied to model transcriptional networks, although primarily in lower eukaryotes and prokaryotes \cite{Hyduke:2010rt}.
+These techniques (and many others, see review \cite{Wayman2013}) are certainly exciting, with many interesting properties.
 Here, we used a traditional approach of mass action kinetics within an ordinary differential equation framework that also included transfer functions to simplify scenarios where reactions involving one species are controlled by several others (e.g.
@@
 The identification problem for the EMT model was underdetermined (not uncommon for differential equation based models).
-However, a central criticism leveled by biologists is that model simplification is often done at the cost of biological reality, or done for reasons of computational expediency \citep{Sainani2012}.
+However, a central criticism leveled by biologists is that model simplification is often done at the cost of biological reality, or done for reasons of computational expediency \cite{Sainani2012}.
 To avoid this criticism, we systematically identified an ensemble of likely models each consistent with the training data, instead of a single but uncertain best fit model.
-Previously, we (and others) have suggested that deterministic ensembles could model heterogeneous populations in situations where stochastic computation was not feasible \citep{Lequieu:2011fj}.
-Population heterogeneity using deterministic model families has previously been explored for bacterial growth in batch cultures \citep{Lee:2009fk}.
+Previously, we (and others) have suggested that deterministic ensembles could model heterogeneous populations in situations where stochastic computation was not feasible \cite{Lequieu:2011fj}.
+Population heterogeneity using deterministic model families has previously been explored for bacterial growth in batch cultures \cite{Lee:2009fk}.
 In that case, distributions were generated because the model parameters varied over the ensemble, i.e., extrinsic noise led to population heterogeneity.
 In this study, parameters controlling physical interactions such as disassociation rates, or processes such as gene expression were distributed over the ensemble.
-Population heterogeneity can also arise from intrinsic thermal fluctuations, which are not captured by a deterministic population of models \citep{Swain:2002ve}.
+Population heterogeneity can also arise from intrinsic thermal fluctuations, which are not captured by a deterministic population of models \cite{Swain:2002ve}.
 Thus, deterministic ensembles, provide a coarse-grained or extrinsic-only ability to simulate population diversity. Despite this limitation, our prediction of phenotypic heterogeneity
@@
 In this study, parametric uncertainty was addressed by developing an ensemble of probable models instead of a single best-fit but uncertain model using multiobjective optimization.
-While computationally complex, multiobjective optimization is an important tool to address qualitative conflicts in training data that arise from experimental error or cell line artifacts \citep{Handl:2007fk}.
+While computationally complex, multiobjective optimization is an important tool to address qualitative conflicts in training data that arise from experimental error or cell line artifacts \cite{Handl:2007fk}.
 On the other hand, structural uncertainty is defined as uncertainty in the biological connectivity.
@@
 First, we identified a potential gap in biological knowledge surrounding the regulation of LEF-1 expression, that was filled by the addition of the hypothetical YREG1 transcriptional repressor.
-The LEF-1 transcription factor is expressed in tissues that undergo EMT during embryogenesis \citep{Vega:2004qf,Nawshad:2003ff},
-and has been suggested to promote an invasive phenotype in cancer cells \citep{Cano:2000kh,Kim:2002lh}.
+The LEF-1 transcription factor is expressed in tissues that undergo EMT during embryogenesis \cite{Vega:2004qf,Nawshad:2003ff},
+and has been suggested to promote an invasive phenotype in cancer cells \cite{Cano:2000kh,Kim:2002lh}.
 Low levels of YREG1 were important for stabilizing the interaction between LEF-1 and $\beta$-catenin, while elevated levels inhibited EMT by downregulating LEF-1 transcriptional activity.
@@
 AES opposes LEF-1 transcriptional activation while Groucho/TLE binds with LEF-1 for a histone deacetylase repression.
-In addition, $\beta$-catenin directly displaces Groucho/TLE repressors from TCF/LEF-1 in Wnt-mediated transcription activation \citep{Arce:2009pr, Grumolato2013}.
+In addition, $\beta$-catenin directly displaces Groucho/TLE repressors from TCF/LEF-1 in Wnt-mediated transcription activation \cite{Arce:2009pr, Grumolato2013}.
 Our model agrees with this newly discovered feedback system, as YREG1 regulates LEF-1 activity leading to EMT stabilization.
@@
 NF-$\kappa$B may also play an essential role of in the epithelial transformation.
-NF-$\kappa$B may influence Snail expression through the AKT pathway and directly stabilize Snail activity \citep{Wu:2009cr}.
-This is particularly important for integrating inflammation pathways, such as interleukin-6 (IL-6) and tumor necrosis factor-$\alpha$ (TNF-$\alpha$), which have been linked to EMT in pathological conditions \citep{Sullivan:2009bh}.
-Other pathways such as Notch have also been shown to act synergistically with TGF-$\beta$ to express Slug in the developing embryo \citep{Niessen:2008nx}.
+NF-$\kappa$B may influence Snail expression through the AKT pathway and directly stabilize Snail activity \cite{Wu:2009cr}.
+This is particularly important for integrating inflammation pathways, such as interleukin-6 (IL-6) and tumor necrosis factor-$\alpha$ (TNF-$\alpha$), which have been linked to EMT in pathological conditions \cite{Sullivan:2009bh}.
+Other pathways such as Notch have also been shown to act synergistically with TGF-$\beta$ to express Slug in the developing embryo \cite{Niessen:2008nx}.
 Lastly, while we have modeled classical protein signaling, we have not considered the role of regulatory RNAs on EMT.
-There is growing evidence that microRNAs (miRNAs) play a strong role in EMT, where several miRNAs, for example miR-21 and miR-31 are strongly associated with TGF-$\beta$ exposure \citep{Bullock:2012aa}.
+There is growing evidence that microRNAs (miRNAs) play a strong role in EMT, where several miRNAs, for example miR-21 and miR-31 are strongly associated with TGF-$\beta$ exposure \cite{Bullock:2012aa}.
 Addressing missing structural components like these, could generate more insight into TGF-$\beta$ signaling and its role in phenotypic transformation.
@@
 The TGF$-\beta$-EMT network contained 97 nodes (proteins, mRNA or genes) interconnected by 251 interactions.
-The network connectivity was curated from more than 40 primary literature sources in combination with on-line databases \citep{Linding:2007fk,Jensen:2009fk}.
+The network connectivity was curated from more than 40 primary literature sources in combination with on-line databases \cite{Linding:2007fk,Jensen:2009fk}.
 The model interactome was not specific to a single epithelial cell line.
@@
 Our signaling network reconstruction was based on Medici et al. who identified the pathways though which MDCKII, DLD1 colon carcinoma,
-and A375 melanoma cells transition towards a mesenchymal phenotype \citep{Medici:2008fk}.
+and A375 melanoma cells transition towards a mesenchymal phenotype \cite{Medici:2008fk}.
 Sequential activation of MAPK and Smad pathways were initiated upon addition of TGF$-\beta$1/2.
-Briefly, TGF$-\beta$2 signals through the RAS-RAF-MEK-ERK pathway to up-regulate Snail and Slug expression \citep{Medici:2006qa}.
-Snail, a known repressor of junctional proteins, inhibits the expression of E-cadherin \citep{Cano:2000kh}.
+Briefly, TGF$-\beta$2 signals through the RAS-RAF-MEK-ERK pathway to up-regulate Snail and Slug expression \cite{Medici:2006qa}.
+Snail, a known repressor of junctional proteins, inhibits the expression of E-cadherin \cite{Cano:2000kh}.
 This initial repression of E-cadherin leads to a release of $\beta$-catenin from the cell membrane.
-This release of $\beta$-catenin can then translocate to the nucleus and form transcriptional complexes with TCF-4 to drive TGF$-\beta$3 expression \citep{Medici:2008fk}.
-The PI3K to GSK3 pathway was included and acted as an activating mechanism of $\beta$-catenin signaling through TGF$-\beta$3 signaling \citep{Medici:2008fk}.
-GSK3 is known to act on $\beta$-catenin signaling through the ubiquitin-proteasome pathway \citep{Zhou:2004kx,Larue:2005vn}.
-Thereby, further $\beta$-catenin release also resulted from by TGF$-\beta$3 signals to the cells interior by binding to type II receptors, which form heterodimers with type I receptors (ALK5) \citep{Derynck:2003fc}.
-This activates the receptors serine/threonine kinase activity to phosphorylate and activate the receptor Smads 2/3 \citep{Massague:2005qc}.
+This release of $\beta$-catenin can then translocate to the nucleus and form transcriptional complexes with TCF-4 to drive TGF$-\beta$3 expression \cite{Medici:2008fk}.
+The PI3K to GSK3 pathway was included and acted as an activating mechanism of $\beta$-catenin signaling through TGF$-\beta$3 signaling \cite{Medici:2008fk}.
+GSK3 is known to act on $\beta$-catenin signaling through the ubiquitin-proteasome pathway \cite{Zhou:2004kx,Larue:2005vn}.
+Thereby, further $\beta$-catenin release also resulted from by TGF$-\beta$3 signals to the cells interior by binding to type II receptors, which form heterodimers with type I receptors (ALK5) \cite{Derynck:2003fc}.
+This activates the receptors serine/threonine kinase activity to phosphorylate and activate the receptor Smads 2/3 \cite{Massague:2005qc}.
 In the model, receptors are simplified and represented as either bound or unbound complexes with their ligands.
@@
 pSmads complexes up-regulate other transcription factors, such as LEF-1.
-The pSmad2/4-LEF-1 complex has been shown to directly repress the E-cadherin gene \citep{Nawshad:2007kl}.
-LEF-1 also binds with $\beta$-catenin to upregulate mesenchymal proteins such as fibronectin \citep{Medici2011}.
+The pSmad2/4-LEF-1 complex has been shown to directly repress the E-cadherin gene \cite{Nawshad:2007kl}.
+LEF-1 also binds with $\beta$-catenin to upregulate mesenchymal proteins such as fibronectin \cite{Medici2011}.
 In the model, Smad signaling is consolidated into a single Smad species that can act in a co-dependent fashion with LEF1 to downregulate E-cadherin via a transfer function, eliminating the need for an explicity LEF-1, pSmad complex.
 The EMT gene expression program was initiated by the binding of TGF$-\beta$ isoforms to TGF$-\beta$ surface receptors, starting the downstream signaling program.
-Repression of E-cadherin expression is the central event in the transition from an epithelial to a mesenchymal phenotype \citep{Cano:2000kh}.However, this transition is not solely driven by transcriptional events.
+Repression of E-cadherin expression is the central event in the transition from an epithelial to a mesenchymal phenotype \cite{Cano:2000kh}.However, this transition is not solely driven by transcriptional events.
 At the protein level, the repression of E-cadherin leads to a release of $\beta$-catenin from cell membrane.
 Cystolic $\beta$-catenin then translocates to the nucleus and forms transcriptionally-active complexes with immunoglobulin transcription factor 2 (TCF-4)
-to drive TGF$-\beta$3 expression \citep{Medici:2008fk}.
-The PI3K to GSK3 pathway was included and acted as an activating mechanism of $\beta$-catenin signaling through TGF$-\beta$3 signaling \citep{Medici:2008fk}.
+to drive TGF$-\beta$3 expression \cite{Medici:2008fk}.
+The PI3K to GSK3 pathway was included and acted as an activating mechanism of $\beta$-catenin signaling through TGF$-\beta$3 signaling \cite{Medici:2008fk}.
 GSK3 is known to act on $\beta$-catenin signaling through APC complex associated ubiquitin-proteasome pathway.
-The APC complex is represented in our model and serves as a second reservoir  of $\beta$-cateinin in untransformed cells whose sequestration is regulated by GSK3 \citep{Zhou:2004kx,Larue:2005vn,Medici:2008fk}.
-Lastly, VEGF-A activation of NFATc1 takes place through calcineurin signaling leading to an enhancement of E-cadherin expression \citep{Suehiro:2014aa},
+The APC complex is represented in our model and serves as a second reservoir  of $\beta$-cateinin in untransformed cells whose sequestration is regulated by GSK3 \cite{Zhou:2004kx,Larue:2005vn,Medici:2008fk}.
+Lastly, VEGF-A activation of NFATc1 takes place through calcineurin signaling leading to an enhancement of E-cadherin expression \cite{Suehiro:2014aa},
 as supported by our VEGF-A experimental data (Fig. \ref{fg:VEGFA-Data}).
@@
 \begin{equation}
-	\frac{1}{\tau_{i}}\frac{dx_{i}}{dt}  =  \sum_{j = 1}^{\mathcal{R}}\sigma_{ij}r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)-\mu x_{i}\qquad{i=1,2,\hdots,\mathcal{M}}\\
+	\frac{1}{\tau_{i}}\frac{dx_{i}}{dt}  =  \sum_{j = 1}^{\mathcal{R}}\sigma_{ij}r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)-\mu x_{i}\qquad{i=1,2,\hdots,\mathcal{M}}
 \end{equation}
@@
 The function $\mathcal{I}_{j}\left(\cdot\right)$ is an integration rule which maps the output of regulatory transfer functions into a control
-variable. In this study, we used $\mathcal{I}_{j}\in\left\{min,max\right\}$ and hill transfer functions \citep{pr3010178,pr3010138}.
+variable. In this study, we used $\mathcal{I}_{j}\in\left\{min,max\right\}$ and hill transfer functions \cite{pr3010178,pr3010138}.
 If a process has no modifying factors, $v_{j}=1$.
@@
 and the abundance of RNA polymerase, $V^{max}_{T} = k_{T}\left(RNAP\right)$. The parameter $\alpha_{j}$ denotes the gene specific correction to the characteristic transcription rate.
-(estimated in this study), while $k_{T}$, $G_{j}$ and $RNAP$ were estimated from literature \citep{Milo:2010aa}.
+(estimated in this study), while $k_{T}$, $G_{j}$ and $RNAP$ were estimated from literature \cite{Milo:2010aa}.
 Similar to the signaling processes, the gene expression control term $0\leq u_{j}\leq 1$ depended upon the combination of factors which influenced rate process $j$.
@@
 The function $\mathcal{I}_{j}\left(\cdot\right)$ is an integration rule which maps the output of regulatory transfer functions into a control
-variable. In this study, we used $\mathcal{I}_{j}\in\left\{min,max\right\}$ and hill transfer functions \citep{pr3010178,pr3010138}.
+variable. In this study, we used $\mathcal{I}_{j}\in\left\{min,max\right\}$ and hill transfer functions \cite{pr3010178,pr3010138}.
 If a gene expression process has no modifying factors, $u_{j}=1$.
+
 Lastly, the specific translation rate was modeled as:
@@
 The characteristic maximum translation rate was defined as the product of a characteristic translation rate constant ($k_{X}$)
-and the abundance of Ribosomes ($RIBO$), $V^{max}_{X} = k_{X}\left(RIBO\right)$, where both $k_{X}$ and $RIBO$ abundance were estimated from literature \citep{Milo:2010aa}.
+and the abundance of Ribosomes ($RIBO$), $V^{max}_{X} = k_{X}\left(RIBO\right)$, where both $k_{X}$ and $RIBO$ abundance were estimated from literature \cite{Milo:2010aa}.
 
@@
 
-The signaling and gene expression model equations were implemented in Julia and solved using the \texttt{CVODE} routine of the Sundials package \citep{Julia,Hindmarsh2005}.
+The signaling and gene expression model equations were implemented in Julia and solved using the \texttt{CVODE} routine of the Sundials package \cite{Julia,Hindmarsh2005}.
 The model code and parameter ensemble is freely available under an MIT software license and can be downloaded from http://www.varnerlab.org.
@@
 % UNIVERSAL, an open source Objective-C/Java code generator, is available as a Google Code project (http://code.google.com/p/universal-code-generator/).
-% Model equations were solved using the CVODE solver in the SUNDIALS library \citep{Hindmarsh2005} on an Apple workstation (Apple, Cupertino, CA)
-% as previously described \citep{Tasseff:2011ve}.
+% Model equations were solved using the CVODE solver in the SUNDIALS library \cite{Hindmarsh2005} on an Apple workstation (Apple, Cupertino, CA)
+% as previously described \cite{Tasseff:2011ve}.
 
 \subsubsection*{Estimation of model parameters using multiobjective optimization.}
-The EMT model had 296 unknown parameters (169 kinetic constants, 44 saturation constants, 38 control logic parameters, and 45 non-zero initial conditions) which were not uniquely identifiable given the training data.
-Instead, we estimated a population of likely models (each consistent with the training data)
-using 41 data sets generated in DLD1 colon carcinoma, MDCKII, and A375 melanoma cells taken from Medici \emph{et al.} \citep{Medici:2008fk}.
+We estimated a population of likely EMT model parameter sets (each consistent with the training data)
+using 41 data sets generated in DLD1 colon carcinoma, MDCKII, and A375 melanoma cells taken from Medici \emph{et al.} \cite{Medici:2008fk}.
 We used the Pareto Optimal Ensemble Technique (JuPOETs) multiobjective optimization framework in combination
-with leave-one-out cross-validation to estimate an ensemble of model parameters \citep{Song:2010fk,JuPOETs-BioArXiv}.
+with leave-one-out cross-validation to estimate an ensemble of model parameters \cite{Song:2010fk,JuPOETs-BioArXiv}.
 Model parameter values were adjusted to minimize the residual between simulations and experimental measurements for the 11 objective functions (supplemental materials).
-Cross-validation was used to calculate both training and prediction error during the parameter estimation procedure \citep{kohavi1995study}.
+Cross-validation was used to calculate both training and prediction error during the parameter estimation procedure \cite{kohavi1995study}.
 The 41 intracellular protein and mRNA data-sets used for identification were organized into 11 objective functions.
@@
 The training and validation data were Western blots.
-We achieved a biologically realistic concentration scale by establishing characteristic rates of transcription, translation, mRNA and protein degradation, as well as characteristic concentrations of ribosomes and RNAPs using the Bionumbers database \citep{Milo:2010aa}, (supplemental materials).
+We achieved a biologically realistic concentration scale by establishing characteristic rates of transcription, translation, mRNA and protein degradation, as well as characteristic concentrations of ribosomes and RNAPs using the Bionumbers database \cite{Milo:2010aa}, (supplemental materials).
 The overall concentration scale was nM, with proteins ranging from 10-1000nM and mRNA ranging from 0.01 to 1nM, reflecting the true abundances and ratios between each species.
@@
 
-% \subsubsection*{Robustness coefficients.} Robustness coefficients were calculated as shown previously \citep{Tasseff:2011ve,Lequieu:2011fj}.
+% \subsubsection*{Robustness coefficients.} Robustness coefficients were calculated as shown previously \cite{Tasseff:2011ve,Lequieu:2011fj}.
 % Robustness coefficients denoted by $\alpha\left(i,j,t_{o},t_{f}\right)$:
@@
 Primers were designed to detect GAPDH, E-cadherin, vimentin, Slug, Sp1, and NFATc1 in cDNA clones: Sp1 (F-TTG AAA AAG GAG TTG GTG GC, R-TGC TGG TTC TGT AAG TTG GG, Accession NG030361.1), NFATc1 (F-GCA TCA CAG GGA AGA CCG TGT C, R-GAA GTT CAA TGT CGG AGT TTC TGA G, Accession  NG029226.1).
-GAPDH, E-cadherin, vimentin, and Slug primers were taken from previously published literature \citep{Medici:2008fk}.
+GAPDH, E-cadherin, vimentin, and Slug primers were taken from previously published literature \cite{Medici:2008fk}.
 
@@
 characteristics of nontumorigenic MCF10A epithelial acini occur over time and exploiting them to growth in 3D culture can be quantified.
-For example, using image segmentation, Chang et al. \citep{ChangPP07} examined the elongation of the MCF10A acini at 6, 12, and 96 hours after a particular treatment.
-Polizzotti et al. \citep{Polizzotti} also suggested a computational method to quantify acini structure based on morphological characteristics in nonmalignant, noninvasive, and invasive conditions.
+For example, using image segmentation, Chang et al. \cite{ChangPP07} examined the elongation of the MCF10A acini at 6, 12, and 96 hours after a particular treatment.
+Polizzotti et al. \cite{Polizzotti} also suggested a computational method to quantify acini structure based on morphological characteristics in nonmalignant, noninvasive, and invasive conditions.
 Adapted from these approaches, we first fluorescently labeled our cultures and captured the acini structures by 3D confocal microscopy.
@@
 \caption{Training and validation simulations. The population of EMT models qualitatively captured TGF$-\beta$-induced EMT signaling.
-(A-I) The population was generated using JuPOETs and trained using 11 different objective functions (41 data sets) taken from Medici \emph{et al.} \citep{Medici:2008fk}.
+(A-I) The population was generated using JuPOETs and trained using 11 different objective functions (41 data sets) taken from Medici \emph{et al.} \cite{Medici:2008fk}.
 The model captured the simulated experiments for all cases when compared to randomized controls.

### 2016-10-04 Supplemental-JB.tex
PLoS formatting

@@
 We used literature based transcription and translation parameters to establish the characteristic synthesis and degradation rates for both mRNA and protein.
-We estimated values for the rate parameters from the Bionumbers database \citep{Milo:2010aa}. These parameters were then used for all gene
+We estimated values for the rate parameters from the Bionumbers database \cite{Milo:2010aa}. These parameters were then used for all gene
 expression calculations:
@@
 -------------------------------------------------------------------
-cell_diameter = 12                	# \mum
+cell_diameter = 12                	# mu m
 number_of_rnapII = 75000          	# copies/cells
@@
 We used the Pareto Optimal Ensemble Technique (POETs) multiobjective optimization framework in combination with leave-one-out cross-validation to estimate an ensemble of TGF$-\beta$/EMT models.
-Cross-validation was used to calculate both training and prediction error during the parameter estimation procedure \citep{kohavi1995study}.
+Cross-validation was used to calculate both training and prediction error during the parameter estimation procedure \cite{kohavi1995study}.
 The 41 intracellular protein and mRNA data-sets used for identification were organized into 11 objective functions.
@@
 POETs integrates standard search strategies e.g., Simulated Annealing (SA) or Pattern Search (PS)
-with a Pareto-rank fitness assignment \citep{Song:2010fk,JuPOETs-BioArXiv}.
+with a Pareto-rank fitness assignment \cite{Song:2010fk,JuPOETs-BioArXiv}.
 Denote a candidate parameter set at iteration $i+1$ as $\mathbf{k}_{i+1}$.
@@
 In this study, the experimental data used for model training was typically the band intensity from Western or Northern blots.
-Band intensity was estimated using the ImageJ software package \citep{IMAGEJ}.
+Band intensity was estimated using the ImageJ software package \cite{IMAGEJ}.
 The scaled measurement for species $x$ at time $i=\{t_{1},t_{2},..,t_{n}\}$ in condition $j$ is given by:
@@
 JuPOETs is free or charge, open source and available for download under an MIT software license from http://www.varnerlab.org.
-Details of the JuPOETs implementation, including example codes are presented in Bassen et al., \citep{JuPOETs-BioArXiv}.
+Details of the JuPOETs implementation, including example codes are presented in Bassen et al., \cite{JuPOETs-BioArXiv}.
 

### 2016-10-04 Paper_v7.tex

@@
 %\vspace{0.1in}
-{\par \centering \textbf{Character count:}~\emph{56,659}}
-\vspace{0.1in}
+% {\par \centering \textbf{Character count:}~\emph{56,659}}
+% \vspace{0.1in}
 {\par \centering $^{*}$Corresponding author:}
@@
 If a gene expression process has no modifying factors, $u_{j}=1$.
+
 Lastly, the specific translation rate was modeled as:
@@
 \subsubsection*{Estimation of model parameters using multiobjective optimization.}
-The EMT model had 296 unknown parameters (169 kinetic constants, 44 saturation constants, 38 control logic parameters, and 45 non-zero initial conditions) which were not uniquely identifiable given the training data.
-Instead, we estimated a population of likely models (each consistent with the training data)
+We estimated a population of likely EMT model parameter sets (each consistent with the training data)
 using 41 data sets generated in DLD1 colon carcinoma, MDCKII, and A375 melanoma cells taken from Medici \emph{et al.} \citep{Medici:2008fk}.
@@
 \clearpage
-\input{Supplemental-JB}
+\input{Supplemental-JB-v7}
 

### 2016-10-04 Paper_v8.tex
author ref update
minor fix

@@
 \\
-Russell Gould\textsuperscript{1},
-David Bassen\textsuperscript{1},
+Russell Gould\textsuperscript{1,\Yinyang},
+David M. Bassen\textsuperscript{1,\Yinyang},
 Anirikh Chakrabarti\textsuperscript{2},
@@
 % Primary Equal Contribution Note
-%\Yinyang These authors contributed equally to this work.
+\Yinyang These authors contributed equally to this work.
 
@@
 The EMT model described the gene expression program resulting from TGF$-\beta$ and VEGF-A signaling in a prototypical epithelial cell.
-The TGF$-\beta$-EMT network contained 97 nodes (proteins, mRNA or genes) interconnected by 251 interactions.
+The TGF$-\beta$-EMT network contained 97 nodes (proteins, mRNA or genes) and 169 interactions.
 The network connectivity was curated from more than 40 primary literature sources in combination with on-line databases \cite{Linding:2007fk,Jensen:2009fk}.

### 2016-10-04 Paper_v8.tex
RG JB patches

@@
 If a process has no modifying factors, $v_{j}=1$.
+Factors can be individual metabolite or protein levels or some function, e.g., the product of metabolite levels.
+In the current study, we let factors correspond to the abundance of individual proteins, inhibitors or transcription factors, as well as products of transcription factors where it is necessasry to represent a co-dependent regulatory activity.
 
@@
 \subsection*{Cell culture and experimental interrogation}
-DLD1 colon carcinoma, MCF10A, and HUVEC were acquired from the American Tissue Culture Collection (Manassas, VA).
-Cells were grown in culture with RPMI 1640 medium with 10\% fetal bovine serum and 1\% penicillin/streptomycin for DLD1, EBM-2 supplemented with EGM-2, 5\% fetal bovine serum, and 1\% penicillin/streptomycin for HUVEC, or MGEM 2 supplemented with insulin, bovine pituitary extract, cholera toxin, hEGF, hydrocortisone, 5\% horse serum, and 1\% penicillin/streptomycin for MCF10A.
-Cells were serum starved for 24 hours and removed from all experimental conditions.
-Recombinant VEGFA165 was also removed from culture medium prior to experimentation.
+DLD1 colon carcinoma and MCF10A cells were acquired from the American Tissue Culture Collection (Manassas, VA).
+Cells were grown in culture with RPMI-1640 medium with 10\% fetal bovine serum and 1\% penicillin/streptomycin or MGEM-2 supplemented with insulin, bovine pituitary extract, cholera toxin, hEGF, hydrocortisone, 5\% horse serum, and 1\% penicillin/streptomycin, respectively.
+Cells were passaged 1:3 or 1:4 every 3-6 d and used between passages 4 and 8.
+During treatment regime, DLD1 and MCF10A cells aggregated overnight in hanging drop culture (20$\mu$L; 20,000 cells).
+The spherical aggregates were placed on the surface of neutralized type I collagen hydrogels (1.5mg/mL) and allowed to adhere.
+Cultures were then serum starved (1\% serum) for 24 hours.
 Recombinant human TGF$-\beta$2 (R$\And$D Systems, Minneapolis, MN) was added to the culture medium at a concentration of 10 ng/ml and recombinant VEGFA165 at a concentration of (5ng/ml, 50ng/ml) for all relative experiments.
 NFAT inhibitor (VIVIT peptide) (EMDBiosciences, Darmstadt, Germany), was added to the culture medium at a concentration of 10$\mu$M for all relative experiments.
-Cells were passaged 1:3 or 1:4 every 3-6 d and used between passages 4 and 8.
-
-\subsubsection*{VEGF treatment}
-DLD1 and MCF10A cells were suspended in culture media (with RPMI 1640 medium with 10\% fetal bovine serum and 1\% penicillin/streptomycin for DLD1 or MGEM 2 supplemented with insulin, bovine pituitary extract, cholera toxin, hEGF, hydrocortisone, 5\% horse serum, and 1\% penicillin/streptomycin for MCF10A), and allowed to aggregate overnight in hanging drop culture (20$\mu$L; 20,000 cells).
-The spherical aggregates were placed on the surface of neutralized type I collagen hydrogels (1.5mg/mL) and allowed to adhere.
-Cultures were then serum starved (1\% serum) for 24 hours.
-Recombinant VEGFA165 was then added to the media (5ng/ml, 50ng/ml) and mRNA was harvested after 3hr and 24hr timepoint.
+mRNA was then harvested at both the 3hr and 48hr time-point.
 

### 2016-10-05 Paper_v8.tex
Updated materials and methods

@@
 The EMT model had 251 unknown model parameters, 169 kinetic constants 38 control constants and 44 saturation constants.
-As expected, these parameters were not uniquely identifiable given the training data \cite{Gadkar:2005vn}.
+In addition to constants appearing in the signaling kinetics, characteristic gene expression and translation parameters were estimated from literature,
+and then corrected for each protein and transcript (four correction factors per gene product, see materials and methods).
+As expected, the unknown parameters were not uniquely identifiable given the training data \cite{Gadkar:2005vn}.
 Thus, instead of identifying a single best fit (but uncertain) model,
@@
 In the model, receptors are simplified and represented as either bound or unbound complexes with their ligands.
-Phosphorylated Smads 2/3 (pSmad2/3) form heterodimers and translocate to the nucleus.
-pSmads complexes up-regulate other transcription factors, such as LEF-1.
+Phosphorylated Smads 2/3 (pSmad2/3) form heterodimers and translocate to the nucleus. pSmads complexes up-regulate other transcription factors, such as LEF-1.
 The pSmad2/4-LEF-1 complex has been shown to directly repress the E-cadherin gene \cite{Nawshad:2007kl}.
@@
 The EMT gene expression program was initiated by the binding of TGF$-\beta$ isoforms to TGF$-\beta$ surface receptors, starting the downstream signaling program.
-Repression of E-cadherin expression is the central event in the transition from an epithelial to a mesenchymal phenotype \cite{Cano:2000kh}.However, this transition is not solely driven by transcriptional events.
+Repression of E-cadherin expression is the central event in the transition from an epithelial to a mesenchymal phenotype \cite{Cano:2000kh}.
+However, this transition is not solely driven by transcriptional events.
 At the protein level, the repression of E-cadherin leads to a release of $\beta$-catenin from cell membrane.
@@
 \subsection*{Formulation, solution and analysis of the EMT model equations}
+The EMT model integrated kinetic ODE modeling with a rule-based description of signaling activity and gene expression regulation.
+Regulation of enzyme activity or gene expression processes was modeled by combining individual regulatory contributions, quantified by transfer-functions,
+into a control coefficients using an integration rule. These control coefficients then modified the kinetic rates that appeared in the model.
+This strategy is similar in spirit to the Constrained Fuzzy Logic (cFL) approach of Lauffenburger and coworkers which has been used to effectively model signal transduction pathways important in human health \cite{Morris:2011ys}.
+In our formulation, Hill-like transfer functions $0 < f\left(\mathbf{x}\right) < 1$ were used to calculate the influence of factor abundance upon enzyme activity or a gene expression process.
+In this context, factors can be individual metabolite or protein levels or some function of abundance, e.g., the product of metabolite or proteins levels.
+However, more generally, factors can also correspond to non-modeled influences, categorial variables or other abstract quantities.
+In the current study, we let factors correspond to the abundance of individual proteins, inhibitors or transcription factors, as well as products of transcription factors where it is necessary to represent a co-dependent regulatory activity.
+When a protein or gene was potentially sensitive to more than one regulatory input, logical integration rules were used to select
+which regulatory transfer function influenced enzyme activity at any given time.
+Thus, our EMT network model encoded complex signaling and regulatory features with a relatively small number of equations.
+This model formulation has been used previously to construct reduced order kinetic models of synthetic RNA circuits \cite{Hu:2015aa},
+cell free metabolic models \cite{pr3010138} and reduced-order models of blood coagulation \cite{pr3010178}.
+In this study, we extended this approach to describe gene expression processes in addition to signal transduction events.
+
 \subsubsection*{EMT signaling events.}
@@
 If a process has no modifying factors, $v_{j}=1$.
-Factors can be individual metabolite or protein levels or some function, e.g., the product of metabolite levels.
-In the current study, we let factors correspond to the abundance of individual proteins, inhibitors or transcription factors, as well as products of transcription factors where it is necessasry to represent a co-dependent regulatory activity.
+
+%Factors can be individual metabolite or protein levels or some function, e.g., the product of metabolite levels.
 
@@
 \begin{eqnarray}
-	\frac{dm_{j}}{dt} &=& r_{T,j} - \left(\mu+\delta_{m,j}\right)m_{j}+\lambda_{j}\\
-	\frac{dp_{j}}{dt} &=& r_{X,j} - \left(\mu+\delta_{p,j}\right)p_{j}
+	\frac{dm_{j}}{dt} &=& r_{T,j} - \left(\mu+\theta_{m,j}\right)m_{j}+\lambda_{j}\\
+	\frac{dp_{j}}{dt} &=& r_{X,j} - \left(\mu+\theta_{p,j}\right)p_{j}
 \end{eqnarray}where $j=1,2,\dots,\mathcal{G}$.
 The terms $r_{T,j}$ and $r_{X,j}$ denote the specific rates of transcription, and translation while
-the terms $\delta_{m,j}$ and $\delta_{p,j}$ denote degradation constants for mRNA and protein, respectively.
+the terms $\theta_{m,j}$ and $\theta_{p,j}$ denote first-order degradation constants for mRNA and protein, respectively.
 Lastly, $\mu$ denotes the specific growth rate, and $\lambda_{j}$ denotes the constitutive rate of gene expression for gene $j$.
@@
 \end{equation}where the maximum gene expression rate was defined as the product of a characteristic transcription rate constant ($k_{T}$)
-and the abundance of RNA polymerase, $V^{max}_{T} = k_{T}\left(RNAP\right)$. The parameter $\alpha_{j}$ denotes the gene specific correction to the characteristic transcription rate.
-(estimated in this study), while $k_{T}$, $G_{j}$ and $RNAP$ were estimated from literature \cite{Milo:2010aa}.
+and the abundance of RNA polymerase, $V^{max}_{T} = k_{T}\left(RNAP\right)$.
+The parameter $\alpha_{j}$ denotes the gene specific correction to the characteristic transcription rate (estimated in this study).
 Similar to the signaling processes, the gene expression control term $0\leq u_{j}\leq 1$ depended upon the combination of factors which influenced rate process $j$.
@@
 If a gene expression process has no modifying factors, $u_{j}=1$.
+The degradation rate constants were defined as $\theta_{m,j} = \delta_{j}k_{d,m}$ and $\theta_{p,j} = \gamma_{j}k_{d,p}$
+where $k_{d,m}$ and $k_{d,p}$ denote characteristic degradation constants for mRNA and protein, respectively and $\delta_{j}$ and $\gamma_{j}$ denote transcript and protein specific
+correction factors to these constants.
 
@@
 The characteristic maximum translation rate was defined as the product of a characteristic translation rate constant ($k_{X}$)
-and the abundance of Ribosomes ($RIBO$), $V^{max}_{X} = k_{X}\left(RIBO\right)$, where both $k_{X}$ and $RIBO$ abundance were estimated from literature \cite{Milo:2010aa}.
+and the abundance of Ribosomes ($RIBO$), $V^{max}_{X} = k_{X}\left(RIBO\right)$.
 
-In this study, we estimated $k_{T}, k_{X}, RNAP$ and $RIBO$ from literature. We calculated the concentration for gene $j$ by assuming, on average,
-the cell had two copies of each gene at any given time. We also estimated characteristic values for $\delta_{m,j}$ and $\delta_{p,j}$,
-and corrected these values with mRNA/protein specific correction factors. Thus, the bulk of our gene expression parameters were based directly upon literature values.
-The values used for the characteristic transcription/translation parameters are given in the supplemental materials.
+In this study, we estimated $k_{T}, k_{X}, k_{d,m}, k_{d,p}, RNAP$ and $RIBO$ directly from literature using estimates of transcription and translation rates,
+the half-life of an typical mRNA and protein and a typical value for the copies per cell of RNA polymerase and ribosomes.
+The saturation constants $K_{X}$ and $K_{T}$ were adjusted so that gene expression and translation resulted in gene products on a biologically realistic concentration scale.
+Next, we calculated the concentration for gene $j$ by assuming, on average,
+that a cell had two copies of each gene at any given time.
+Thus, the bulk of our gene expression parameters were based directly upon literature values, and were not adjusted during model identification.
+The values used for the characteristic transcription/translation parameters, degradation constants and macromolecular copy number are given in the supplemental materials along with
+the specific formulas required to calculate all derived constants.
 
@@
 using 41 data sets generated in DLD1 colon carcinoma, MDCKII, and A375 melanoma cells taken from Medici \emph{et al.} \cite{Medici:2008fk}.
+In addition to rate and saturation constants appearing in the signal transduction equations, we estimate the transcript/protein specific
+correction factors $\alpha_{j},\beta_{j},\delta_{j}$ and $\delta_{j}$ and control coefficients from these data.
 We used the Pareto Optimal Ensemble Technique (JuPOETs) multiobjective optimization framework in combination
@@
 An initial nominal parameter set was established by inspection. JuPOETs was then allowed to search in a neighborhood of $\pm$30\% of this nominal set.
+The correction factors estimated by JuPOETs were in the range of 0.1 - 10, with the mean value of 1.68; thus, the characteristic rate parameters estimated from literature were consistent with measurements.
 The parameter ensemble estimated by JuPOETs is available with the model source code.

### 2016-10-05 Paper_v8.tex
Added some references

@@
 Similarly, Vilar \emph{et al.} suggested that specific changes in receptor trafficking patterns could lead to phenotypes that favor tumorigenesis \cite{Vilar:2006gb}.
+Lu \emph{et al.} modeled the role that micro-RNA can play in EMT, suggesting cell-fate determination is regulated by the miR-34/SNAIL and miR-200/ZEB mutual-inhibition feedback circuit \cite{Lu:2013aa}.
+Coarse grained modeling approaches have also been applied to EMT; Steinway \emph{et al.} used discrete dynamic modeling to study developmental EMT and known dysregulations in invasive hepatocellular carcinoma \cite{Steinway:2014aa}.
 Although these models provided insight into the role of receptor dynamics, EMT induction involves many other components, including competing second messengers and interconnected transcriptional regulatory loops.
@@
 Lastly, while we have modeled classical protein signaling, we have not considered the role of regulatory RNAs on EMT.
-There is growing evidence that microRNAs (miRNAs) play a strong role in EMT, where several miRNAs, for example miR-21 and miR-31 are strongly associated with TGF-$\beta$ exposure \cite{Bullock:2012aa}.
+There is growing evidence that microRNAs (miRNAs) play a strong role in EMT, where several miRNAs,
+for example miR-21 and miR-31 are strongly associated with TGF-$\beta$ exposure \cite{Bullock:2012aa}, and Lu et al have suggested that
+EMT is driven by a miR-34/SNAIL and miR-200/ZEB mutual-inhibition feedback circuit \cite{Lu:2013aa}.
 Addressing missing structural components like these, could generate more insight into TGF-$\beta$ signaling and its role in phenotypic transformation.

### 2016-10-05 Paper_v8.tex
Added Zeb1 ref

@@
 Other pathways such as Notch have also been shown to act synergistically with TGF-$\beta$ to express Slug in the developing embryo \cite{Niessen:2008nx}.
+In addition, other important regulatory proteins such as ZEB1 should be added to the next generation model.
+ZEB1 can activate the expression of mesenchymal genes such as collagens, smooth muscle actin and myosin, Vimentin while simultaneously contributing
+to the repression of epithelial genes \cite{Liu:2008aa}.
+Thus, ZEB1 can act in parallel to SNAIL and thus be an alternative to the canonical signaling paradigm presented here.
 Lastly, while we have modeled classical protein signaling, we have not considered the role of regulatory RNAs on EMT.
@@
 Thus, the bulk of our gene expression parameters were based directly upon literature values, and were not adjusted during model identification.
-The values used for the characteristic transcription/translation parameters, degradation constants and macromolecular copy number are given in the supplemental materials along with
-the specific formulas required to calculate all derived constants.
+The values used for the characteristic transcription/translation parameters, degradation constants and macromolecular copy number are given in the supplemental materials along with the specific formulas required to calculate all derived constants.
 

### 2016-10-05 Paper_v8.tex
Updates for the rebuttal

@@
 Models with this hybrid phenotype had elevated Sp1 and NFAT transcriptional activity, resulting in simultaneously increased Vimentin and E-cadherin expression (Fig. \ref{fg:F4}A).
+However, these conclusions are likely sensitivity to the components that we have included in the model, and could also involve proteins that we have not considered.
 
@@
 The model captured the simulated experiments for all cases when compared to randomized controls.
-(J-L)  The model populations were also compared against untrained temporal data to measure the effectiveness as a pure prediction.}\label{fg:F2}
+(J-L)  The model populations were also compared against untrained temporal data to measure the effectiveness as a pure prediction.
+The western blot data was reproduced from Medici \emph{et al.} \cite{Medici:2008fk}. The intensity of each band was estimated using the ImageJ program.
+These blot intensities were then used as the training data for the EMT parameter estimation studies. }\label{fg:F2}
 \end{figure}

### 2016-11-01 Paper_v8.tex
added extra ZEB refs

@@
 Similarly, Vilar \emph{et al.} suggested that specific changes in receptor trafficking patterns could lead to phenotypes that favor tumorigenesis \cite{Vilar:2006gb}.
-Lu \emph{et al.} modeled the role that micro-RNA can play in EMT, suggesting cell-fate determination is regulated by the miR-34/SNAIL and miR-200/ZEB mutual-inhibition feedback circuit \cite{Lu:2013aa}.
+Several studies on ZEB signaling have modeled the role that micro-RNA can play in EMT, suggesting cell-fate determination is regulated by the miR-34/SNAIL and/or miR-200/ZEB mutual-inhibition feedback circuit \cite{Lu:2013aa,Zhang2014,Tian2013}.
 Coarse grained modeling approaches have also been applied to EMT; Steinway \emph{et al.} used discrete dynamic modeling to study developmental EMT and known dysregulations in invasive hepatocellular carcinoma \cite{Steinway:2014aa}.
@@
 There is growing evidence that microRNAs (miRNAs) play a strong role in EMT, where several miRNAs,
-for example miR-21 and miR-31 are strongly associated with TGF-$\beta$ exposure \cite{Bullock:2012aa}, and Lu et al have suggested that
-EMT is driven by a miR-34/SNAIL and miR-200/ZEB mutual-inhibition feedback circuit \cite{Lu:2013aa}.
+for example miR-21 and miR-31 are strongly associated with TGF-$\beta$ exposure \cite{Bullock:2012aa}, and have suggested that
+the EMT process is driven by a miR-34/SNAIL and/or miR-200/ZEB feedback circuit \cite{Lu:2013aa,Zhang2014,Tian2013}.
 Addressing missing structural components like these, could generate more insight into TGF-$\beta$ signaling and its role in phenotypic transformation.
@@
 AP-1/Sp1 drives an autocrine response of TGF-$\beta$3, which activates the Smad cascade, leading to phenotypic change.
-Conversely, VEGF-A binding can stabalize an epithelial phenotype through NFAT activation.
-Downstream activation of $\beta$-catenin signaling due to E-cadherin loss and GSK3 inactivation of $\beta$-cateinin confinement is critical to the complete activation of the EMT program.
+Conversely, VEGF-A binding can stabilize an epithelial phenotype through NFAT activation.
+Downstream activation of $\beta$-catenin signaling due to E-cadherin loss and GSK3 inactivation of $\beta$-catenin confinement is critical to the complete activation of the EMT program.
 The complete list of molecular interactions that comprise the model is given in the supplement.}\label{fg:F1}

### 2016-11-04 Paper_v8.tex
Added text to disucssion re Hong study

@@
 Decades of research has focused on identifying molecular regulators of EMT, but almost all on a single gene and in a nearly binary yes/no level of qualitative understanding.
-Medici and coworkers identified a core signaling program by which TGF-$\beta$ isoforms induce EMT across a variety of cell lines \cite{Medici:2006qa,Medici:2008fk}.
+Medici and coworkers identified a core signaling program by which
+TGF-$\beta$ isoforms induce EMT across a variety of cell lines \cite{Medici:2006qa,Medici:2008fk}.
 This program involves carefully orchestrated rounds of gene expression driven by the Smad and Snail families of transcription factors as well as
@@
 Similarly, Vilar \emph{et al.} suggested that specific changes in receptor trafficking patterns could lead to phenotypes that favor tumorigenesis \cite{Vilar:2006gb}.
-Several studies on ZEB signaling have modeled the role that micro-RNA can play in EMT, suggesting cell-fate determination is regulated by the miR-34/SNAIL and/or miR-200/ZEB mutual-inhibition feedback circuit \cite{Lu:2013aa,Zhang2014,Tian2013}.
+Several studies on ZEB signaling have modeled the role that microRNA can play in EMT, suggesting cell-fate determination is regulated by the miR-34/SNAIL and/or miR-200/ZEB mutual-inhibition feedback circuits \cite{Lu:2013aa,Zhang:2014aa,Tian:2013aa}.
 Coarse grained modeling approaches have also been applied to EMT; Steinway \emph{et al.} used discrete dynamic modeling to study developmental EMT and known dysregulations in invasive hepatocellular carcinoma \cite{Steinway:2014aa}.
@@
 Taken together, the model captured the key signaling events revealed by Medici \emph{et al.} \cite{Medici:2008fk} that drive the phenotypic conversion.
-A listing of objective function values resulting from training, cross validation and the random parameter control is given in the supplement (Fig. \ref{fg:ObjTable}).
+A listing of objective function values resulting from training, cross validation and the random parameter control is given in the supplement (Fig.
+\ref{fg:ObjTable}).
 
@@
 Thus, ZEB1 can act in parallel to SNAIL and thus be an alternative to the canonical signaling paradigm presented here.
+Hong et al. developed an ODE model of ZEB1 signaling in MCF10A cells which also demonstrated intermediate EMT states \cite{Hong:2015aa}.
+However, given that we do not have ZEB1 in our canonical architecture, this suggests multiple types of hybrid phenotypes could exist within the EMT program.
 Lastly, while we have modeled classical protein signaling, we have not considered the role of regulatory RNAs on EMT.
@@
 for example miR-21 and miR-31 are strongly associated with TGF-$\beta$ exposure \cite{Bullock:2012aa}, and have suggested that
-the EMT process is driven by a miR-34/SNAIL and/or miR-200/ZEB feedback circuit \cite{Lu:2013aa,Zhang2014,Tian2013}.
-Addressing missing structural components like these, could generate more insight into TGF-$\beta$ signaling and its role in phenotypic transformation.
+the EMT process is driven by a miR-34/SNAIL and/or miR-200/ZEB feedback circuit \cite{Lu:2013aa,Zhang:2014aa,Tian:2013aa}.
+Addressing missing structural components like these (and potentially others), could generate more insight into TGF-$\beta$ signaling and its role in phenotypic transformation.
 
@@
 At the protein level, the repression of E-cadherin leads to a release of $\beta$-catenin from cell membrane.
-Cystolic $\beta$-catenin then translocates to the nucleus and forms transcriptionally-active complexes with immunoglobulin transcription factor 2 (TCF-4)
+Cytosolic $\beta$-catenin then translocates to the nucleus and forms transcriptionally-active complexes with immunoglobulin transcription factor 2 (TCF-4)
 to drive TGF$-\beta$3 expression \cite{Medici:2008fk}.
@@
 GSK3 is known to act on $\beta$-catenin signaling through APC complex associated ubiquitin-proteasome pathway.
-The APC complex is represented in our model and serves as a second reservoir of $\beta$-cateinin in untransformed cells whose sequestration is regulated by GSK3 \cite{Zhou:2004kx,Larue:2005vn,Medici:2008fk}.
+The APC complex is represented in our model and serves as a second reservoir of $\beta$-catenin in untransformed cells whose sequestration is regulated by GSK3 \cite{Zhou:2004kx,Larue:2005vn,Medici:2008fk}.
 Lastly, VEGF-A activation of NFATc1 takes place through calcineurin signaling leading to an enhancement of E-cadherin expression \cite{Suehiro:2014aa},

### 2016-11-04 Paper_v8.tex
compressed figs and manuscript tweaks

@@
 Similarly, Vilar \emph{et al.} suggested that specific changes in receptor trafficking patterns could lead to phenotypes that favor tumorigenesis \cite{Vilar:2006gb}.
-Several studies on ZEB signaling have modeled the role that microRNA can play in EMT, suggesting cell-fate determination is regulated by the miR-34/SNAIL and/or miR-200/ZEB mutual-inhibition feedback circuits \cite{Lu:2013aa,Zhang:2014aa,Tian:2013aa}.
 Coarse grained modeling approaches have also been applied to EMT; Steinway \emph{et al.} used discrete dynamic modeling to study developmental EMT and known dysregulations in invasive hepatocellular carcinoma \cite{Steinway:2014aa}.
@@
 Other pathways such as Notch have also been shown to act synergistically with TGF-$\beta$ to express Slug in the developing embryo \cite{Niessen:2008nx}.
-In addition, other important regulatory proteins such as ZEB1 should be added to the next generation model.
+In addition, other regulatory proteins such as ZEB1 could be added to the next generation model.
 ZEB1 can activate the expression of mesenchymal genes such as collagens, smooth muscle actin and myosin, Vimentin while simultaneously contributing
@@
 Hong et al. developed an ODE model of ZEB1 signaling in MCF10A cells which also demonstrated intermediate EMT states \cite{Hong:2015aa}.
-However, given that we do not have ZEB1 in our canonical architecture, this suggests multiple types of hybrid phenotypes could exist within the EMT program.
+However, given that we do not have ZEB1 in our model architecture, this suggests multiple types of hybrid phenotypes could exist within the EMT program.
 Lastly, while we have modeled classical protein signaling, we have not considered the role of regulatory RNAs on EMT.

### 2016-11-04 Paper_v8.tex
JB tweaks

@@
 to the repression of epithelial genes \cite{Liu:2008aa}.
-Thus, ZEB1 can act in parallel to SNAIL and thus be an alternative to the canonical signaling paradigm presented here.
+Therefore, ZEB1 can act in parallel to SNAIL and is an alternative to the canonical signaling paradigm presented here.
 Hong et al. developed an ODE model of ZEB1 signaling in MCF10A cells which also demonstrated intermediate EMT states \cite{Hong:2015aa}.
-However, given that we do not have ZEB1 in our model architecture, this suggests multiple types of hybrid phenotypes could exist within the EMT program.
+Thus, there may be alternative modes resulting in a hybrid phenotype that should be included in future model architecture.
 Lastly, while we have modeled classical protein signaling, we have not considered the role of regulatory RNAs on EMT.

### 2016-11-22 Paper_v9.tex
start of reformmating revisions

--- 2016-11-04 Paper_v8.tex
@@
 % Use the PLoS provided BiBTeX style
-\bibliographystyle{plos2009}
+\bibliographystyle{plos2015}
 
@@
 
+\subsection*{TGF$-\beta$1/2 and VEGF-A exposure promotes phenotype heterogeneity through NFATc and phosphorylated Sp1}
+While we captured the central tendency of many of the molecular features of EMT induction following TGF$-\beta$1/2 exposure, an often neglected but important emergent
+feature of developmental and pathological programs is population heterogeneity \cite{Park:2010fu}.
+We (and others) have previously hypothesized that deterministic model ensembles can simulate population behavior, at least at a course grained level \cite{Lequieu:2011fj}.
+We tested this hypothesis by analyzing the response of the population of EMT models to extracellular cues and then comparing this response to flow cytometry studies.
+We quantified the phenotypic response of the individual members of the ensemble to TGF$-\beta$1/2 stimulation for two downstream phenotypic markers, Vimentin (mesenchymal) and E-cadherin (epithelial) following the addition of TGF$-\beta$1/2 alone (Fig.
+\ref{fg:F3}), and/or VEGF-A in combination with NFATc inhibitors (Fig.
+\ref{fg:F3}).
+
+We identified model subpopulations that exhibited different behaviors following exposure to TGF$-\beta$1/2 (Fig. \ref{fg:F3}B).
+Analysis of the molecular signatures of these subpopulations suggested the abundance, localization and state of the Sp1, AP-1 and NFATc transcription factors controlled population heterogeneity.
+The majority of models ($>$80\%) responded to treatment, moving away from the untreated population (Fig \ref{fg:F3}A-F, gray).
+These models showed the classically expected behavior, a switch from an epithelial to mesenchymal phenotype following TGF$-\beta$1/2 exposure.
+Some models resembled untreated cells; they had elevated phosphorylated Sp1, relative to non-induced cells,
+which decreased E-cadherin expression through Slug-mediated inhibition, which in turn increased Vimentin expression through TGF$-\beta$3 autocrine signaling and the liberation of $\beta$-catenin.
+However, the most biologically interesting behavior was exhibited by cells achieving a hybrid phenotype, most notable in a dual treatment condition (\ref{fg:F3}C, black arrow), but also present in a small percentage of untreated cells (Fig.
+\ref{fg:F3}B, gray arrow).
+Models with this hybrid phenotype had elevated Sp1 and NFAT transcriptional activity, resulting in simultaneously increased Vimentin and E-cadherin expression (Fig. \ref{fg:F4}A).
+However, these conclusions are likely sensitive to the components that we have included in the model, and could also involve proteins that we have not considered.
+
+To better understand the hybrid phenotype, we simulated the response of the model population to TGF$-\beta$1/2 and VEGF-A treatment with and without NFATc inhibitors (Fig. \ref{fg:F3}).
+As expected, stimulation with VEGF-A (50 a.u.) maintained an epithelial population (Fig. \ref{fg:F3}A), while TGF$-\beta$1/2 (10 a.u.) exposure shifted
+the population from an epithelial to a mesenchymal phenotype (Fig. \ref{fg:F3}B).
+On the other hand, combined stimulation with TGF$-\beta$1/2 (10 a.u.) and VEGF-A (50 a.u.) increased both E-cadherin and Vimentin expression, resulting in a hybrid phenotype with both epithelial and mesenchymal characteristics (Fig.
+\ref{fg:F3}C).
+Vimentin expression was correlated with high levels of nuclear phosphorylated Sp1, following TGF$-\beta$1/2 exposure.
+Conversely, elevated E-cadherin expression depended upon the activity of NFAT transcription factors downstream of VEGF-A stimulation.
+To further isolate the role of NFAT on this hybrid state, we simulated the inhibition of NFAT transcriptional activity across all conditions (all else being equal).
+NFAT inhibition in combination with VEGF-A or TGF$-\beta$1/2 treatments blocked increased E-cadherin expression in the case of VEGF-A (Fig.
+\ref{fg:F3}D), but did not influence TGF$-\beta$1/2 signaling (Fig.
+\ref{fg:F3}E).
+Lastly, NFATc inhibition in combination with simultaneous TGF$-\beta$1/2 and VEGF-A exposure repressed nearly all E-cadherin expression, shifting nearly the entire population towards a mesenchymal phenotype (Fig.
+\ref{fg:F3}F).
+Taken together, high levels of nuclear localized phosphorylated Sp1 correlated with Vimentin expression,
+while NFATc transcriptional activity was critical for maintaining E-cadherin expression in the presence of competing signals.
+
 \subsection*{Identification of a novel LEF-1 regulator}
@@
 
-%as seen in the training objectives.
-
 \clearpage
-
-\subsection*{TGF$-\beta$1/2 and VEGF-A exposure promotes phenotype heterogeneity through NFATc and phosphorylated Sp1}
-While we captured the central tendency of many of the molecular features of EMT induction following TGF$-\beta$1/2 exposure, an often neglected but important emergent
-feature of developmental and pathological programs is population heterogeneity \cite{Park:2010fu}.
-We (and others) have previously hypothesized that deterministic model ensembles can simulate population behavior, at least at a course grained level \cite{Lequieu:2011fj}.
-We tested this hypothesis by analyzing the response of the population of EMT models to extracellular cues and then comparing this response to flow cytometry studies.
-We quantified the phenotypic response of the individual members of the ensemble to TGF$-\beta$1/2 stimulation for two downstream phenotypic markers, Vimentin (mesenchymal) and E-cadherin (epithelial) following the addition of TGF$-\beta$1/2 alone (Fig.
-\ref{fg:F3}), and/or VEGF-A in combination with NFATc inhibitors (Fig.
-\ref{fg:F3}).
-%We were further able to use this analysis to examine the activity and effects of underlying signaling molecules.
-
-We identified model subpopulations that exhibited different behaviors following exposure to TGF$-\beta$1/2 (Fig. \ref{fg:F3}B).
-Analysis of the molecular signatures of these subpopulations suggested the abundance, localization and state of the Sp1, AP-1 and NFATc transcription factors controlled population heterogeneity.
-The majority of models ($>$80\%) responded to treatment, moving away from the untreated population (Fig \ref{fg:F3}A-F, gray).
-These models showed the classically expected behavior, a switch from an epithelial to mesenchymal phenotype following TGF$-\beta$1/2 exposure.
-Some models resembled untreated cells; they had elevated phosphorylated Sp1, relative to non-induced cells,
-which decreased E-cadherin expression through Slug-mediated inhibition, which in turn increased Vimentin expression through TGF$-\beta$3 autocrine signaling and the liberation of $\beta$-catenin.
-However, the most biologically interesting behavior was exhibited by cells achieving a hybrid phenotype, most notable in a dual treatment condition (\ref{fg:F3}C, black arrow), but also present in a small percentage of untreated cells (Fig.
-\ref{fg:F3}B, gray arrow).
-Models with this hybrid phenotype had elevated Sp1 and NFAT transcriptional activity, resulting in simultaneously increased Vimentin and E-cadherin expression (Fig. \ref{fg:F4}A).
-However, these conclusions are likely sensitivity to the components that we have included in the model, and could also involve proteins that we have not considered.
-
-%Analysis of these hypothetical cells suggested they had deregulated NFAT expression and nuclear localization promoted E-cadherin expression while TGF$-\beta$1/2 induced Sp1 action promoted Vimentin expression.
-
-To better understand the hybrid phenotype, we simulated the response of the model population to TGF$-\beta$1/2 and VEGF-A treatment with and without NFATc inhibitors (Fig. \ref{fg:F3}).
-As expected, stimulation with VEGF-A (50 a.u.) maintained an epithelial population (Fig. \ref{fg:F3}A), while TGF$-\beta$1/2 (10 a.u.) exposure shifted
-the population from an epithelial to a mesenchymal phenotype (Fig. \ref{fg:F3}B).
-On the other hand, combined stimulation with TGF$-\beta$1/2 (10 a.u.) and VEGF-A (50 a.u.) increased both E-cadherin and Vimentin expression, resulting in a hybrid phenotype with both epithelial and mesenchymal characteristics (Fig.
-\ref{fg:F3}C).
-Vimentin expression was correlated with high levels of nuclear phosphorylated Sp1, following TGF$-\beta$1/2 exposure.
-Conversely, elevated E-cadherin expression depended upon the activity of NFAT transcription factors downstream of VEGF-A stimulation.
-To further isolate the role of NFAT on this hybrid state, we simulated the inhibition of NFAT transcriptional activity across all conditions (all else being equal).
-NFAT inhibition in combination with VEGF-A or TGF$-\beta$1/2 treatments blocked increased E-cadherin expression in the case of VEGF-A (Fig.
-\ref{fg:F3}D), but did not influence TGF$-\beta$1/2 signaling (Fig.
-\ref{fg:F3}E).
-Lastly, NFATc inhibition in combination with simultaneous TGF$-\beta$1/2 and VEGF-A exposure repressed nearly all E-cadherin expression, shifting nearly the entire population towards a mesenchymal phenotype (Fig.
-\ref{fg:F3}F).
-Taken together, high levels of nuclear localized phosphorylated Sp1 correlated with Vimentin expression,
-while NFATc transcriptional activity was critical for maintaining E-cadherin expression in the presence of competing signals.
-
-%In particular, when present, VEGF-A signaling through NFATc was predicted to provide a protective up-regulation of E-cadherin.
 
@@
 
-\clearpage
+%\clearpage
 
@@
 
-% \begin{equation}\label{asseassertain_massbalance}
-% \frac{d\mathbf{x}}{dt}=\mathbf{S}\cdot\mathbf{r}\left(\mathbf{x},\mathbf{k}\right)\qquad\mathbf{x}\left(t_{o}\right)=\mathbf{x}_{o}
-% \end{equation}
-% The quantity $\mathbf{x}$ denotes the vector describing the abundance of protein, mRNA, and other species in the model ($995\times 1$).
-% The stoichiometric matrix $\mathbf{S}$ encodes the signaling architecture considered in the model ($995\times 1700$).
-% Each row of $\mathbf{S}$ describes a signaling component while each column describes a particular interaction.
-% The $(i,j)$ element of $\mathbf{S}$, denoted by $\sigma_{ij}$, describes how species $i$ is involved with interaction $j$.
-% If $\sigma_{ij}>0$, species $i$ is produced by interaction $j$. Conversely, If $\sigma_{ij}<0$, then species $i$ is consumed in interaction $j$.
-% Lastly, if $\sigma_{ij}=0$, then species $i$ is not involved in interaction $j$.
-% The term  $\mathbf{r}\left(\mathbf{x},\mathbf{k}\right)$ denotes the vector of interactions rates ($1700\times{1}$).
-% We modeled each network interaction (gene expression, translation and biochemical transformations)
-% using elementary rate laws where all reversible interactions were split into two irreversible steps (supplemental materials).
-% Thus, the rate expression for interaction $q$ was given by:
-% \begin{equation}\label{eq-mass-action}
-% r_{q}\left(\mathbf{x},k_{q}\right)=k_{q}\prod_{j\in\left\{\mathbf{R}_{q}\right\}}x_{j}^{-\sigma_{jq}}
-% \end{equation}
-% The set $\left\{\mathbf{R}_{q}\right\}$ denotes reactants for reaction $q$, while $\sigma_{jq}$ denotes the stoichiometric coefficient (element of the matrix $\mathbf{S}$) governing species $j$ in reaction $q$.
-The quantity $k_{q}$ denotes the rate constant (unknown) governing reaction $q$.
-% Model equations were generated in the C-programming language using the UNIVERSAL code generator, starting from an text-based input file (supplemental materials).
-% UNIVERSAL, an open source Objective-C/Java code generator, is available as a Google Code project (http://code.google.com/p/universal-code-generator/).
-% Model equations were solved using the CVODE solver in the SUNDIALS library \cite{Hindmarsh2005} on an Apple workstation (Apple, Cupertino, CA)
-% as previously described \cite{Tasseff:2011ve}.
-
 \subsubsection*{Estimation of model parameters using multiobjective optimization.}
@@
 JuPOETs is open source and freely available for download under an MIT software license from http://www.varnerlab.org.
-
-% \subsubsection*{Robustness coefficients.} Robustness coefficients were calculated as shown previously \cite{Tasseff:2011ve,Lequieu:2011fj}.
-% Robustness coefficients denoted by $\alpha\left(i,j,t_{o},t_{f}\right)$:
-% \begin{equation}
-% \alpha\left(i,j,t_{o},t_{f}\right)=\left(\displaystyle\int_{t_{o}}^{t_{f}}x_{i}\left(t\right)dt\right)^{-1}\left(\displaystyle\int_{t_{o}}^{t_{f}}x^{(j)}_{i}\left(t\right)dt\right)
-% \end{equation}
-% quantify the response of a marker to a structural or operational perturbation to the network architecture.
-% Here $t_o$ and $t_f$ denote the initial and final simulation time respectively, while
-% $i$ and $j$ denote the indices for the marker and the perturbation respectively.
-A value of $\alpha\left(i,j,t_{o},t_{f}\right)>1$, indicates increased marker abundance, while $\alpha\left(i,j,t_{o},t_{f}\right)<1$ indicates decreased marker abundance following perturbation $j$.
-If
-% $\alpha\left(i,j,t_{o},t_{f}\right)\sim{1}$ the $j$th perturbation does not influence the abundance of marker $i$.
-% Robustness coefficients were calculated for each member of the ensemble (N $\simeq$ 1100).
 
@@
 
-%\section*{Acknowledgments}
-%The project described was supported by Award Number \#U54CA143876 from the National Cancer Institute to JDV, National Science Foundation CBET-0955172 to JTB, and National Institute of Health HL110328 to JTB.
-The content is solely the responsibility of the authors and does not necessarily represent the official views of the National Cancer Institute or the National Institutes of Health.
-Lastly, we thank the anonymous reviewers for their helpful suggestions.
-
 \clearpage
-% \bibliographystyle{plain}
-%\bibliographystyle{IEEEbib}
 

### 2016-11-22 S1_Text.tex
start of reformmating revisions

--- 2016-10-04 Supplemental-JB.tex
@@
 
-% Additionally, we applied a universal convention of identifying the no expression case as protein value below $<$10$^{-3}$.
-This is similar to previously published models from our lab.
-In our studies, Figure 2 identifies experimental data extracted from published Western blots and our simulation results.
-It is clear from these that 1) the training data included a variety of Western blot data treatments that were effectively zero, and 2) our simulations matched the training data virtually perfectly over time and across multiple biological species.
-These results validate the power of our simulation scheme, which necessarily includes numerical interpretations of zero.
-In our Supplemental Figure S9, we present the raw concentrations of our simulations (previously validated from the training data) over hundreds of parameter ensembles, with no “zero” threshold applied.
-While these values do not have a threshold, we only interpret values below 10$^{-3}$ equivalent to zero.
-Likewise, our robustness coefficients (Figure 3) (which represent the ratio of integrated areas of the treatment effect over the baseline effect) identify no difference in model output for vimentin (or e-cadherin for that matter) less than 10$^{-3}$, confirming this interpretation.
-
 \begin{figure}
@@
 \end{figure}
-
-% We computed the Pareto rank of $\mathbf{k}_{i+1}$ by comparing the simulation error at iteration $i+1$ against an archive of accepted parameter sets $\mathbf{K}_{i}$.
-% We used the Fonseca and Fleming ranking scheme \cite{fonseca1993genetic} to estimate the number of parameter sets in the archive that dominate
-% $\mathbf{k}_{i+1}$. Parameter sets with increasing rank were progressively further away from the optimal trade-off surface.
-% The parameter set $\mathbf{k}_{i+1}$ was accepted or rejected by POETs with probability $\mathcal{P}\left(\mathbf{k}_{i+1}\right)$:
-% \begin{equation}\label{eqn_costMOSA}
-% \mathcal{P}(\mathbf{k}_{i+1}) \equiv \exp{\{-rank\left(\mathbf{k}_{i+1} \mid \mathbf{K}_{i} \right)/T\}}
-% \end{equation}
-% where $T$ is the annealing temperature and $rank\left(\mathbf{k}_{i+1} \mid \mathbf{K}_{i} \right)$ denotes the Pareto rank for $\mathbf{k}_{i+1}$.
-% The annealing temperature was discretized into 10 quanta between $T_{o}$ and $T_{f}$ and adjusted according to the schedule
-% $T_k = \beta^k T_0$ where $\beta$ was defined as $\left({T_{f}}/{T_{o}}\right)^{1/10}$.
-% The initial temperature was $T_{o} = \mathit{n}/log(2)$, where $n = 4$ in this study and the final temperature was $T_{f} = 0.1$.
-% The epoch-counter $k$ was incremented after the addition of 100 members to the ensemble.
-% Thus, as the ensemble grew, the likelihood of accepting parameter sets with a large Pareto rank decreased.
-% To generate parameter diversity, we randomly perturbed each parameter by $\leq\pm25\%$ at iteration of the search.
-% In addition, we performed a local pattern search every $q$-iterations to minimize the residual for a single random or the worst performing objective function.
-% The local pattern-search algorithm has been described previously \cite{gadkar2003cybernetic}.
-% From the 15,000 probable EMT models, we selected N = 1093 models with Pareto rank $\leq{1}$ for subsequent analysis.
-% A quick estimate of the set to set correlation showed that we could expect on order 25\% correlation between parameter sets in the ensemble.
-
-% \begin{figure}
-% \includegraphics [width=1.0\linewidth] {./figs/Fig-5-Supplemental-ErrorTable.pdf}
-% \caption{Training and prediction values for the 11 TGF-$\beta$ objective functions versus a random parameter control.}\label{fig:S5}
-% \end{figure}
-
-% \begin{figure}
-% \includegraphics [width=1.0\linewidth] {./figs/Fig-7-Supplemental-Correlation.pdf}
-% \caption{Parameter set to set correlation for 55 random parameter sets selected from the ensemble.
-% Of the 55 sets selected, the average correlation between sets was less than 25\% for greater than 80\% of the parameter sets.}\label{fig:S7}
-% \end{figure}
-
-% \begin{figure}
-% \includegraphics [width=1.0\linewidth] {./figs/Fig-8-Supplemental-YREG1.pdf}
-% \caption{Robustness of E-cadherin and Vimentin expression to a knockout (A) and overexpression (B) of the hypothetical regulator 1 (YREG1) protein.
-% Robustness coefficients were calculated for each member of the ensemble. Each point represents the response of a single model in the ensemble to either a knockout or
-% overexpression of YREG1.}\label{fig:S8}
-% \end{figure}
-
-% \begin{figure}
-% \includegraphics [width=1.0\linewidth] {./figs/Fig-9-Supplemental-RawExpression_v2.pdf}
-% \caption{Steady state protein abundance for E-cadherin and Vimentin (AU) as a function of TGF-$\beta$1/2 exposure.
-% Top: Overlay of the model population for Vimentin (AU) and E-cadherin (AU) expression in the presence (blue) and absence (gray) of TGF-$\beta$1/2.
-% Midddle: Vimentin (AU) and E-cadherin (AU) expression in the absence of TGF-$\beta$1/2 showed exhinibted population heterogenity.
-% Bottom: Vimentin (AU) and E-cadherin (AU) expression in the presence of TGF-$\beta$1/2 moved the centriod of the population toward Vimentin (AU) and away from E-cadherin (AU) expression.}\label{fig:S9}
-% \end{figure}
-
-% \subsection*{EMT model network architecture.} TGF$\beta$ is a major inducer of EMT in development, fibrosis, and carcinogenesis with different isoforms mediating various effects depending on specific cellular context \cite{Nawshad:2005pi}.
-% TGF$\beta$1 was first described as an inducer of EMT in normal mammary epithelial cells \cite{Miettinen1994} and has since been shown to mediate EMT in vitro in a number of different epithelial cells, including renal proximal tubular, DLD1 colon carcinoma, and most recently alveolar epithelial cells \cite{Fan1999, Hales1994,Kasai2005,Willis2005}.
-% TGF$\beta$ signaling occurs through the Smad pathway in which signals are transduced by transmembrane serine/threonine kinase type I (ALK5) and type II (TGF$\beta$RII) receptors.
-% To increase ligand affinity, betaglycan (TGF$\beta$RIII) can also interact with TGF$\beta$RI,II \cite{Gatza:2010if}.
-% Upon TGF$\beta$ stimulation, the receptors are internalized into early endosomes where Smad anchor for receptor activation (SARA) modulates formation of complexes with (R-Smad) Smad2 or Smad3.
-% Smad2 and Smad3 are then phosphorylated at serine residues by the type I receptor \cite{Massague:2005qc}.
-% Phosphorylation induces their association with (Co-Smad) Smad4 and translocation to the nucleus where they interact with other transcription factors to regulate the transcription of TGF$\beta$ responsive genes, including alpha-smooth muscle actin, collagen1A2, vimentin, fibronectin, and plasminogen activator inhibitor-1 (PAI-1) by interacting with Smad-binding elements \cite{Dennler:2002hl,Derynck:2003fc}.
-% To regulate TGF$\beta$ signaling, Smurf2 (a ubiquitin E3 ligase) can become activated to mediate proteasome dependent degradation of Smad2 or bind with Smad 7 to target TGF$\beta$ receptor for degradation \cite{Kavsak2000,Bonni2001}.
-%
-% \subsubsection*{Cell Type Dependency} Interestingly, differential roles for Smad2 and Smad3 in TGF$\beta$ induced EMT have been demonstrated.
-% For example, using primary cells from mice with hepatocyte-specific double knockout of Smad2 and Smad3, it was demonstrated that Smad3 but not Smad2 was required for a key morphological changes and induction of EMT \cite{Ju:2006so}.
-% In contrast, using human proximal tubular epithelial cells, increased colony growth factor and decreased E-cadherin were Smad3 dependent, increased MMP-2 was Smad2 dependent, while alpha-smooth muscle actin was dependent on both \cite{Phanish2006}.
-% Furthermore, Smad2 signaling has been demonstrated in cancerous lung epithelial A549 cells \cite{Kasai2005}.
-% Inhibition of Smad3 signaling decreases the metastatic potential of xenografted breast cancer cell lines \cite{Tian2003,Tian2004}.
-% Together, these results suggest that the precise Smad pathway activated may depend on the particular cellular context.
-% Regardless, a recent transcriptomic analysis of TGF$\beta$ induced EMT in normal mouse and human epithelial cells using a dominant negative approach demonstrated that Smad signaling was critical for regulation of all tested target genes \cite{Valcourt2005}.
-%
-% \subsubsection*{Smad-Independent EMT Induction Pathways.
-}Although less well established than the Smad-dependent pathways in the induction of EMT, there is substantial evidence for TGF$\beta$ activation of Smad-independent signaling in some aspects of this process.
-% These include RhoA, MAPK, PI3 kinase, and Notch signaling pathway, which have mostly been studied in-vitro.
-% For example, the small GTPase RhoA is involved in TGF$\beta$ induced EMT in a number of cell types including NMuMG mammary epithelial cells and mink lung epithelial (Mv1Lu) cells particularly in the regulation of cytoskeletal and adherens junction rearrangement.
-% In addition to its role in cytoskeletal remodeling, Rho has been shown to activate the alpha-smooth muscle promoter during EMT in kidney proximal tubular (LLCPK1) cells \cite{Masszi:2003hb}.
-% TGF$\beta$ has been shown to elevate ERK activity in culture models of EMT (human keratinocytes, NMuMG mammary epithelial cells, and mouse cortical tubule epithelial cells) \cite{Zavadil2001,Xie2004,Davies2005}.
-% This ERK activity was required for disassembly of adherens junctions and induction of cell motility \cite{Xie2004}.
-% Inhibition of ERK blocked these key morphological changes of EMT in mammary gland epithelial cells \cite{Xie2003}.
-% Recent studies in MDCKII and DLD1 colon carcinoma cells revealed that TGF$\beta$3 can signal through PI3 kinase.
-% Upon activation, PI3 kinase activates molecules such as ILK and AKT, which can phosphorylate and inactivate GSK3$\beta$, a protein that targets both Snail and $\beta$-catenin for degradation through the ubiquitin proteosome pathway \cite{Medici:2008fk}.
-% Lastly, there is increasing evidence for a role of Notch pathways in regulating EMT, particularly during development \cite{Nawshad:2005pi}.
-% In cardiac cushion EMT, it was found that that Slug is directly up-regulated by Notch in endothelial cells repression of the vascular endothelial cadherin promoter and for promoting migration of transformed endothelial cells.
-% In whole, the distinction between Smad-dependent and Smad-independent mechanisms remains difficult to interpret due to the significant cross talk between these pathways.
-% In most cases, stimulation of these cooperative pathways provides the context for induction and specification of EMT within a particular tissue/cell type, with Smads representing the dominant pathway, which in some instances may be necessary but not sufficient for induction of full EMT \cite{Zavadil:2005fu}.
-%
-% \subsubsection*{Master Transcription Regulators}
-% Master transcription factors of EMT are thought to be regulated by Snail1 (SNAI1) and Snail2 (SNAI2) (previously known as Snail and Slug, respectively).
-% These transcription factors are zinc finger proteins that function as repressors of E-cadherin transcription in cultured epithelial cells \cite{Cano:2000kh}, repression of E-cadherin leads to dissolution of adherens junctions.
-% Snail and Slug can be activated by TGF$\beta$ via both Smad-dependent and -independent pathways in a cell type dependent fashion in cultured cells \cite{Peinado2003}.
-% Differential expression of Snail and Slug is observed in TGF$\beta$ induced EMT in keratinocytes, renal proximal tubular, and mammary epithelial cells, suggesting that they are regulated in a cell specific.
-% Interestingly, key regulatory units of Snail and Slug appear to be regulated by upstream transcription factors AP1 and SP1, respectively.
-% Peinado et al. clearly established a necessary role of Snail in mediating TGF$\beta$1 induced EMT in MDCK cells.
-% They demonstrated that the Ras-Raf-MEK-ERK-AP1 signaling pathway could up-regulate synthesis of the E-cadherin repressor molecule Snail (directly binds to Snail promoter) \cite{Peinado:2004pt}.
-% Likewise, TGF$\beta$1 induced the expression of Slug in both lens and other epithelial cells in vitro.
-% The Sp1 binding site in the Slug promoter is largely responsible for TGF$\beta$1 induced Slug expression and upstream of MAPK signaling.
-% In addition, the TGF$\beta$1 mediated repression of E-cadherin was significantly inhibited by Slug siRNA \cite{Choi2007}.
-%
-% \subsubsection*{ERK/MAPK Crosstalk}
-% The ERK-MAPK pathway has been shown to modify TGF$\beta$ signaling at multiple levels.
-% One way is that the activated Ras pathway inhibits the TGF$\beta$ induced nuclear accumulation of Smad2/3, as seen in epithelial cells \cite{Zhang2009a}.
-% ERK kinases have been shown to phosphorylate Smad2 and Smad3 at specific sites in the region linking the MH1 and MH2 domains.
-% These sites are separate from the TGF$\beta$ receptor phosphorylation sites which activate Smad nuclear translocation \cite{Massague2003}.
-% The effect of interaction between ERKs and Smads is the subject of some controversy, with data suggesting that such an interaction either enhances or inhibits downstream events.
-% A hyperactive Ras pathway has been shown to effectively counteract the antiproliferative activity of TGF$\beta$ through attenuation of Smad accumulation in the nucleus \cite{Kretzschmar:1999wb}.
-% In contrast, Ras signals strongly cooperated with Smads for invasion of human carcinoma cells \cite{Oft1996}.
-% A second way is through transcriptional regulation.
-% Recently, it was found that Sp1, via transcriptional induction of Vimentin, cooperates with activated Smad complexes in mesenchymal transition and migration of pancreatic cancer cells upon TGF$\beta$ stimulation \cite{Jungert:2007ri}.
-% Likewise, binding sites at the SP1/Smad3 complex was found to regulate the betaglycan receptors \cite{Lopez-Casillas2003}.
-% Taken together, multiple levels of cross-talk (both positive and negative) exist within MAPK and TGF$\beta$, and may also be cell type dependent.
-%
-% \subsection*{Analysis of the signal flow through the EMT architecture.}
-% To investigate temporal shifts for key species dominating the EMT response, we calculated the scaled flux through the signaling architecture (Fig. \ref{fg:S1}).
-% Three modes of operation were simulated to identify distinct behavioral differences: (a) no TGF-$\beta$2 (-TGF-$\beta$1/2), (b) TGF-$\beta$1/2 stimulation (+TGF-$\beta$2),
-% and (c) TGF-$\beta$1/2 stimulation while blocking the autocrine response of TGF-$\beta$3 (+TGF-$\beta$1/2 + TGF-$\beta$3Ab).
-% Following TGF-$\beta$1/2 stimulation, AP1/SP1 rapidly stimulates Snail/Slug expression (Fig. \ref{fg:S1}A) within 1 hr.
-% During this phase, Snail/Slug repress E-cadherin expression, which in turn reduces the interaction between E-cadherin and $\beta$-catenin, and the accumulation of free cytosolic $\beta$-catenin ensues.
-% Conversely, in the presences of a TGF-$\beta$3 neutralizing antibody, membrane bound $\beta$-catenin increased after 10 hrs,
-% while low levels of free cytosolic $\beta$-catenin accumulated between 1-5 hrs (Fig.
-\ref{fg:S2}A,B).
-Accumulation of free $\beta$-catenin is critical for complexing with TCF4 to rapidly produce the autocrine response of TGF-$\beta$3 within 1-10 hrs.
-Following TGF-$\beta$3 expression, formation and spatial relocation of the Smad complexes occurs in a time dependent manner.
-Smad2 is phosphorylated within 1 hr and nuclear localization of the pSmad2/4 complex dramatically increases after 10 hours (Fig.
-\ref{fg:S1}C).
-This timely formation is critical for completion of the EMT process.
-At steady state, the E-cadherin complex was maintained at a basal epithelial level with no expression of Vimentin (Fig.
-\ref{fg:S1}D).
-During TGF-$\beta$1/2 stimulation, both MAPK and Smad act synergistically to repress the E-cadherin complex within 1 hr.
-% This is followed by increase of Vimentin at $\sim$ 1hr, while exponentially increasing around 10 hrs.
-% Both species elevate within 1 hr, however the complex does not significantly form until $\sim$ 10 hrs.
-% When blocking TGF-$\beta$3, Snail/Slug downregulates the E-cadherin complex between 5-10 hrs,
-% but cannot complete the transformation due to the loss of Smad2 phosphorylation (Fig. \ref{fg:S1}D).
-%
-% \begin{figure}
-% \includegraphics [width=1.0\linewidth] {./figs/Fig-1-Supplemental-FluxAnalysis.pdf}
-% \caption{Signal flow analysis of key species at steady state, TGF-$\beta$1/2 stimulation, and blocking the TGF-$\beta$3 autocrine response.
-% (A)  The MAPK cascade is directly responsible for rapid expression of Snail and downstream TGF-$\beta$3 formation (1~hrs).
-% (B) TGF-$\beta$1/2 reduces  $\beta$-catenin, allowing rapid free-cytosolic $\beta$-catenin to accumulate (1hr).
-% Blocking TGF$\beta$3 increases membrane bound $\beta$-catenin (10hr). (C)  TGF-$\beta$3 activates the Smad cascade.
-% Nuclear localization of the pSmad2/4 complex (10~hrs) is dependent upon both the phosphorylation of Smad2 (1~hrs) and complexing with Smad4 (5~hrs).
-% (D)  TGF-$\beta$1/2 rapidly reduces the E-cadherin complex, while upregulating Vimentin (5-10 hrs).
-% Blocking TGF$\beta$3 increases E-cadherin (10~hrs) and Vimentin is significantly reduced.}\label{fg:S1}
-% \end{figure}
-%
-% \begin{figure}
-% \includegraphics [width=0.5\linewidth] {./figs/Fig-2-Supplemental-Robustness.pdf}
-% %\includegraphics [width=1.0\linewidth] {./ManuFigures/Figs/F2.pdf}
-% %\epsfig{file=./figures/Figures_EMT_Manuscript/Final/ManuFigures/F2.eps,,width=1.0\textwidth}
-% \caption{Robustness analysis for key molecular species at t = 48 hrs for combinations of TGF-$\beta$1/2, VEGF-A and NFATc1 inhibitors.
-Robustness coefficients for the indicated species were calculated
-% for N $\sim{1100}$ ensemble members for 48 hrs following the addition of TGF-$\beta$1/2 (T), TGF-$\beta$1/2 + NFATc1 inhibitor (T + I), VEGF-A (V), VEGF-A + NFATc1 inhibitor (V + I) and TGF-$\beta$1/2 + VEGF-A (T+V) + NFATc1 inhibitor (T + V + I).
-% (A) Robustness coefficients for E-cadherin and Vimentin as a function of condition.
-% (B) Robustness coefficients for nuclear localized phosphorylated Sp1 and NFATc1 as a function of condition.
-% (C) Robustness coefficients for nuclear localized Slug and $\alpha$-smooth muscle actin ($\alpha$-SMA) as a function of condition.
-% In each case the error bars denote one-standard deviation of robustness coefficient calculated over the model ensemble.
-% C=Control, T=TGF$\beta$2 , V=VEGFA, VI= NFAT inhibitor (VIVIT).}\label{fg:S2}
-% \end{figure}
 
@@
 	\includegraphics [width=0.75\linewidth] {./figs/Fig-Supplemental-Data-VEGFA.pdf}
-	\caption{VEGF-A qPCR data used to hand fit VEGF enhancement of E-cadherin expression. mRNA was harvested after 3hr and 24hr timepoint.}\label{fg:VEGFA-Data}
+	\caption{VEGF-A qPCR data used to hand fit VEGF enhancement of E-cadherin expression. mRNA was harvested after 3hr and 48hr timepoint.}\label{fg:VEGFA-Data}
 \end{figure}
@@
 %\bibliography{References_v1}
+
+% in place ref %
+
+@article{IMAGEJ,
+	Author = {Abramoff, M.D. and Magelhaes, P.J. and Ram, S.J.},
+	Date-Added = {2011-11-07 13:49:32 -0500},
+	Date-Modified = {2011-11-07 13:49:32 -0500},
+	Journal = {Biophotonics International,},
+	Pages = {36-42},
+	Title = {Image Processing with ImageJ},
+	Volume = {11},
+	Year = {2004}}

### 2016-11-28 Paper_v9.tex
JV changes

@@
 During EMT polarized, tightly adhered epithelial cell monolayers are transformed into non-interacting motile mesenchymal cells that simultaneously degrade and synthesize extracellular matrix (ECM) components and invade into the underlying tissue space \cite{Stahl:2001pd}.
-EMT is the fundamental initiator of developmental processes such as embryonic gastrulation and valvulogenesis \cite{Eisenberg:1995ai} (also Kalluri J Clin Invest 2009, Thiery Cell 2009).
+EMT is the fundamental initiator of developmental processes such as embryonic gastrulation and valvulogenesis \cite{Eisenberg:1995ai}.
 Transforming growth factor $\beta$ (TGF-$\beta$) family members are important inducers of both developmental and pathological EMT \cite{Zavadil:2005fu,Xu:2009oq}.
@@
 Similarly, Vilar \emph{et al.} suggested that specific changes in receptor trafficking patterns could lead to phenotypes that favor tumorigenesis \cite{Vilar:2006gb}.
-Coarse grained modeling approaches have also been applied to EMT; Steinway \emph{et al.} used discrete dynamic modeling to study developmental EMT and known dysregulations in invasive hepatocellular carcinoma \cite{Steinway:2014aa}.
+Coarse grained modeling approaches have also been applied to EMT; Steinway \emph{et al.} used discrete dynamic modeling to study developmental EMT and known dysregulation in invasive hepatocellular carcinoma \cite{Steinway:2014aa}.
 Although these models provided insight into the role of receptor dynamics, EMT induction involves many other components, including competing second messengers and interconnected transcriptional regulatory loops.
@@
 Snail expression was MAPK-dependent; the MEK inhibitor U0126 blocked AP-1 activation and Snail expression following TGF-$\beta$1/2 exposure (Fig. \ref{fg:F2}A, Lane 3).
-Similar results were obtained for Slug expression, confirming initial activation through the MAPK pathway (data not shown).
 Overexpression of either Snail or Slug upregulated TGF-$\beta$3 expression (Fig. \ref{fg:F2}C) while simultaneously downregulating E-cadeherin expression (Fig. \ref{fg:F2}F).
@@
 TGF$-\beta$3 (10 a.u.) in combination with downstream inhibitors (DN-Smad4 and DN-LEF-1) completely inhibited Vimentin expression, while elevating E-cadherin expression (Fig. \ref{fg:F2}H,I).
+
+% JV REMOVED 11/28/16
+%Similar results were obtained for Slug expression, confirming initial activation through the MAPK pathway (data not shown).
 
@@
 This strategy is similar in spirit to the Constrained Fuzzy Logic (cFL) approach of Lauffenburger and coworkers which has been used to effectively model signal transduction pathways important in human health \cite{Morris:2011ys}.
-In our formulation, Hill-like transfer functions $0 < f\left(\mathbf{x}\right) < 1$ were used to calculate the influence of factor abundance upon enzyme activity or a gene expression process.
+In our formulation, Hill-like transfer functions $0 \leq f\left(\mathbf{x}\right) \leq 1$ were used to calculate the influence of factor abundance upon enzyme activity or a gene expression process.
 In this context, factors can be individual metabolite or protein levels or some function of abundance, e.g., the product of metabolite or proteins levels.
@@
 variable. In this study, we used $\mathcal{I}_{j}\in\left\{min,max\right\}$ and hill transfer functions \cite{pr3010178,pr3010138}.
-If a process has no modifying factors, $v_{j}=1$.
+If a process had no modifying factors, $v_{j}=1$.
 
@@
 \end{equation}where the maximum gene expression rate was defined as the product of a characteristic transcription rate constant ($k_{T}$)
-and the abundance of RNA polymerase, $V^{max}_{T} = k_{T}\left(RNAP\right)$.
+and the abundance of RNA polymerase, $V^{max}_{T} = k_{T}\left(\rm RNAP\right)$.
 The parameter $\alpha_{j}$ denotes the gene specific correction to the characteristic transcription rate (estimated in this study).
@@
 The characteristic maximum translation rate was defined as the product of a characteristic translation rate constant ($k_{X}$)
-and the abundance of Ribosomes ($RIBO$), $V^{max}_{X} = k_{X}\left(RIBO\right)$.
-
-In this study, we estimated $k_{T}, k_{X}, k_{d,m}, k_{d,p}, RNAP$ and $RIBO$ directly from literature using estimates of transcription and translation rates,
-the half-life of an typical mRNA and protein and a typical value for the copies per cell of RNA polymerase and ribosomes.
+and the abundance of ribosomes ($\rm RIBO$), $V^{max}_{X} = k_{X}\left(\rm RIBO\right)$.
+
+In this study, we estimated $k_{T}, k_{X}, k_{d,m}, k_{d,p}, \rm RNAP$ and $\rm RIBO$ directly from literature using estimates of transcription and translation rates,
+the half-life of a typical mRNA and protein and a typical value for the copies per cell of RNA polymerase and ribosomes.
 The saturation constants $K_{X}$ and $K_{T}$ were adjusted so that gene expression and translation resulted in gene products on a biologically realistic concentration scale.
@@
 Thus, the bulk of our gene expression parameters were based directly upon literature values, and were not adjusted during model identification.
-The values used for the characteristic transcription/translation parameters, degradation constants and macromolecular copy number are given in the supplemental materials along with the specific formulas required to calculate all derived constants.
+The values used for the characteristic transcription/translation parameters, degradation constants and macromolecular copy number are given in the supplement along with the specific formulas required to calculate all derived constants.
 
@@
 with leave-one-out cross-validation to estimate an ensemble of model parameters \cite{Song:2010fk,JuPOETs-BioArXiv}.
-Model parameter values were adjusted to minimize the residual between simulations and experimental measurements for the 11 objective functions (supplemental materials).
+Model parameter values were adjusted to minimize the residual between simulations and experimental measurements.
 Cross-validation was used to calculate both training and prediction error during the parameter estimation procedure \cite{kohavi1995study}.
@@
 
-\bibliography{References_v1}
+% References directly in *.tex file
+% JV 11/28/16
+%\bibliography{References_v1}
+\begin{thebibliography}{10}
+
+\bibitem{Thiery:2003il}
+Thiery JP.
+\newblock Epithelial-mesenchymal transitions in development and pathologies.
+\newblock Curr Opin Cell Biol. 2003;15(6):740--6.
+
+\bibitem{Stahl:2001pd}
+Stahl PJ, Felsen D.
+\newblock Transforming growth factor-beta, basement membrane, and
+  epithelial-mesenchymal transdifferentiation: implications for fibrosis in
+  kidney disease.
+\newblock Am J Pathol. 2001;159(4):1187--92.
+
+\bibitem{Eisenberg:1995ai}
+Eisenberg LM, Markwald RR.
+\newblock Molecular regulation of atrioventricular valvuloseptal morphogenesis.
+\newblock Circ Res. 1995;77(1):1--6.
+
+\bibitem{Zavadil:2005fu}
+Zavadil J, B{\"o}ttinger EP.
+\newblock TGF-beta and epithelial-to-mesenchymal transitions.
+\newblock Oncogene. 2005;24(37):5764--74.
+\newblock doi:{10.1038/sj.onc.1208927}.
+
+\bibitem{Xu:2009oq}
+Xu J, Lamouille S, Derynck R.
+\newblock TGF-beta-induced epithelial to mesenchymal transition.
+\newblock Cell Res. 2009;19(2):156--72.
+\newblock doi:{10.1038/cr.2009.5}.
+
+\bibitem{Medici:2006qa}
+Medici D, Hay ED, Goodenough DA.
+\newblock Cooperation between snail and LEF-1 transcription factors is
+  essential for TGF-beta1-induced epithelial-mesenchymal transition.
+\newblock Mol Biol Cell. 2006;17(4):1871--9.
+\newblock doi:{10.1091/mbc.E05-08-0767}.
+
+\bibitem{Medici:2008fk}
+Medici D, Hay ED, Olsen BR.
+\newblock Snail and Slug promote epithelial-mesenchymal transition through
+  beta-catenin-T-cell factor-4-dependent expression of transforming growth
+  factor-beta3.
+\newblock Mol Biol Cell. 2008;19(11):4875--87.
+\newblock doi:{10.1091/mbc.E08-05-0506}.
+
+\bibitem{Kim:2002lh}
+Kim K, Lu Z, Hay ED.
+\newblock Direct evidence for a role of beta-catenin/LEF-1 signaling pathway in
+  induction of EMT.
+\newblock Cell Biol Int. 2002;26(5):463--76.
+
+\bibitem{Jiang:2007aa}
+Jiang YG, Luo Y, He Dl, Li X, Zhang Ll, Peng T, et~al.
+\newblock Role of Wnt/beta-catenin signaling pathway in epithelial-mesenchymal
+  transition of human prostate cancer induced by hypoxia-inducible
+  factor-1alpha.
+\newblock Int J Urol. 2007;14(11):1034--9.
+\newblock doi:{10.1111/j.1442-2042.2007.01866.x}.
+
+\bibitem{Huber:2004aa}
+Huber MA, Azoitei N, Baumann B, Gr{\"u}nert S, Sommer A, Pehamberger H, et~al.
+\newblock NF-kappaB is essential for epithelial-mesenchymal transition and
+  metastasis in a model of breast cancer progression.
+\newblock J Clin Invest. 2004;114(4):569--81.
+\newblock doi:{10.1172/JCI21358}.
+
+\bibitem{Hardy:2010aa}
+Hardy KM, Booth BW, Hendrix MJC, Salomon DS, Strizzi L.
+\newblock ErbB/EGF signaling and EMT in mammary development and breast cancer.
+\newblock J Mammary Gland Biol Neoplasia. 2010;15(2):191--9.
+\newblock doi:{10.1007/s10911-010-9172-2}.
+
+\bibitem{Ahmed:2007mi}
+Ahmed S, Nawshad A.
+\newblock Complexity in interpretation of embryonic epithelial-mesenchymal
+  transition in response to transforming growth factor-beta signaling.
+\newblock Cells Tissues Organs. 2007;185(1-3):131--45.
+\newblock doi:{10.1159/000101314}.
+
+\bibitem{Chung:2009jl}
+Chung SW, Miles FL, Sikes RA, Cooper CR, Farach-Carson MC, Ogunnaike BA.
+\newblock Quantitative modeling and analysis of the transforming growth factor
+  beta signaling pathway.
+\newblock Biophys J. 2009;96(5):1733--50.
+\newblock doi:{10.1016/j.bpj.2008.11.050}.
+
+\bibitem{Vilar:2006gb}
+Vilar JMG, Jansen R, Sander C.
+\newblock Signal processing in the TGF-beta superfamily ligand-receptor
+  network.
+\newblock PLoS Comput Biol. 2006;2(1):e3.
+\newblock doi:{10.1371/journal.pcbi.0020003}.
+
+\bibitem{Steinway:2014aa}
+Steinway SN, Za{\~n}udo JGT, Ding W, Rountree CB, Feith DJ, Loughran TP Jr,
+  et~al.
+\newblock Network modeling of TGFB signaling in hepatocellular carcinoma
+  epithelial-to-mesenchymal transition reveals joint sonic hedgehog and Wnt
+  pathway activation.
+\newblock Cancer Res. 2014;74(21):5963--77.
+\newblock doi:{10.1158/0008-5472.CAN-14-0225}.
+
+\bibitem{Cirit:2012kx}
+Cirit M, Haugh JM.
+\newblock Data-driven modelling of receptor tyrosine kinase signalling networks
+  quantifies receptor-specific potencies of PI3K- and Ras-dependent ERK
+  activation.
+\newblock Biochem J. 2012;441(1):77--85.
+\newblock doi:{10.1042/BJ20110833}.
+
+\bibitem{Morris:2011ys}
+Morris MK, Saez-Rodriguez J, Clarke DC, Sorger PK, Lauffenburger DA.
+\newblock Training signaling pathway maps to biochemical data with constrained
+  fuzzy logic: quantitative analysis of liver cell responses to inflammatory
+  stimuli.
+\newblock PLoS Comput Biol. 2011;7(3):e1001099.
+\newblock doi:{10.1371/journal.pcbi.1001099}.
+
+\bibitem{2001_bailey_NatBiotech}
+Bailey JE.
+\newblock Complex biology with no parameters.
+\newblock Nat Biotechnol. 2001;19(6):503--4.
+\newblock doi:{10.1038/89204}.
+
+\bibitem{Machta:2013by}
+Machta BB, Chachra R, Transtrum MK, Sethna JP.
+\newblock Parameter space compression underlies emergent theories and
+  predictive models.
+\newblock Science. 2013;342(6158):604--7.
+\newblock doi:{10.1126/science.1238723}.
+
+\bibitem{Song:2010fk}
+Song SO, Chakrabarti A, Varner JD.
+\newblock Ensembles of signal transduction models using Pareto Optimal Ensemble
+  Techniques (POETs).
+\newblock Biotechnol J. 2010;5(7):768--80.
+\newblock doi:{10.1002/biot.201000059}.
+
+\bibitem{Lequieu:2011fj}
+Lequieu J, Chakrabarti A, Nayak S, Varner JD.
+\newblock Computational modeling and analysis of insulin induced eukaryotic
+  translation initiation.
+\newblock PLoS Comput Biol. 2011;7(11):e1002263.
+\newblock doi:{10.1371/journal.pcbi.1002263}.
+
+\bibitem{2010_luan_varner_MolBioSys}
+Luan D, Szlam F, Tanaka KA, Barie PS, Varner JD.
+\newblock Ensembles of uncertain mathematical models can identify network
+  response to therapeutic interventions.
+\newblock Mol Biosyst. 2010;6(11):2272--86.
+\newblock doi:{10.1039/b920693k}.
+
+\bibitem{2011_kalita_brasier_JBC}
+Kalita MK, Sargsyan K, Tian B, Paulucci-Holthauzen A, Najm HN, Debusschere BJ,
+  et~al.
+\newblock Sources of cell-to-cell variability in canonical nuclear factor-kB
+  (NF-kB) signaling pathway inferred from single cell dynamic images.
+\newblock J Biol Chem. 2011;286(43):37741--57.
+\newblock doi:{10.1074/jbc.M111.280925}.
+
+\bibitem{2011_hasenauer_allgower_BMCBioinfo}
+Hasenauer J, Waldherr S, Doszczak M, Radde N, Scheurich P, Allg{\"o}wer F.
+\newblock Identification of models of heterogeneous cell populations from
+  population snapshot data.
+\newblock BMC Bioinformatics. 2011;12:125.
+\newblock doi:{10.1186/1471-2105-12-125}.
+
+\bibitem{2012_creixell_linding_NatBiotech}
+Creixell P, Schoof EM, Erler JT, Linding R.
+\newblock Navigating cancer network attractors for tumor-specific therapy.
+\newblock Nat Biotechnol. 2012;30(9):842--8.
+\newblock doi:{10.1038/nbt.2345}.
+
+\bibitem{Schoeberl:2002rv}
+Schoeberl B, Eichler-Jonsson C, Gilles ED, Muller G.
+\newblock {C}omputational modeling of the dynamics of the {MAP} kinase cascade
+  activated by surface and internalized {EGF} receptors.
+\newblock Nat Biotechnol. 2002;20(4):370--375.
+\newblock doi:{10.1038/nbt0402-370}.
+
+\bibitem{Chen:2009uq}
+Chen WW, Schoeberl B, Jasper PJ, Niepel M, Nielsen UB, Lauffenburger DA, et~al.
+\newblock {I}nput-output behavior of {E}rb{B} signaling pathways as revealed by
+  a mass action model trained against dynamic data.
+\newblock Mol Syst Biol. 2009;5:239.
+\newblock doi:{10.1038/msb.2008.74}.
+
+\bibitem{Tasseff:2011ve}
+Tasseff R, Nayak S, Song SO, Yen A, Varner JD.
+\newblock {M}odeling and analysis of retinoic acid induced differentiation of
+  uncommitted precursor cells.
+\newblock Integr Biol (Camb). 2011;3(5):578--591.
+\newblock doi:{10.1039/c0ib00141d}.
+
+\bibitem{Gadkar:2005vn}
+Gadkar KG, Varner J, Doyle FJ 3rd.
+\newblock Model identification of signal transduction networks from data using
+  a state regulator problem.
+\newblock Syst Biol (Stevenage). 2005;2(1):17--30.
+
+\bibitem{Lian:2011bs}
+Lian Yg, Zhou Qg, Zhang Yj, Zheng Fl.
+\newblock VEGF ameliorates tubulointerstitial fibrosis in unilateral ureteral
+  obstruction mice via inhibition of epithelial-mesenchymal transition.
+\newblock Acta Pharmacol Sin. 2011;32(12):1513--21.
+\newblock doi:{10.1038/aps.2011.111}.
+
+\bibitem{Milo:2010aa}
+Milo R, Jorgensen P, Moran U, Weber G, Springer M.
+\newblock BioNumbers--the database of key numbers in molecular and cell
+  biology.
+\newblock Nucleic Acids Res. 2010;38(Database issue):D750--3.
+\newblock doi:{10.1093/nar/gkp889}.
+
+\bibitem{Park:2010fu}
+Park SY, Lee HE, Li H, Shipitsin M, Gelman R, Polyak K.
+\newblock Heterogeneity for stem cell-related markers according to tumor
+  subtype and histologic stage in breast cancer.
+\newblock Clin Cancer Res. 2010;16(3):876--87.
+\newblock doi:{10.1158/1078-0432.CCR-09-1532}.
+
+\bibitem{Hemavathy:2000aa}
+Hemavathy K, Ashraf SI, Ip YT.
+\newblock Snail/slug family of repressors: slowly going into the fast lane of
+  development and cancer.
+\newblock Gene. 2000;257(1):1--12.
+
+\bibitem{Hemavathy:2000ab}
+Hemavathy K, Guru SC, Harris J, Chen JD, Ip YT.
+\newblock Human Slug is a repressor that localizes to sites of active
+  transcription.
+\newblock Mol Cell Biol. 2000;20(14):5087--95.
+
+\bibitem{Dhasarathy:2011aa}
+Dhasarathy A, Phadke D, Mav D, Shah RR, Wade PA.
+\newblock The transcription factors Snail and Slug activate the transforming
+  growth factor-beta signaling pathway in breast cancer.
+\newblock PLoS One. 2011;6(10):e26514.
+\newblock doi:{10.1371/journal.pone.0026514}.
+
+\bibitem{Guaita:2002aa}
+Guaita S, Puig I, Franci C, Garrido M, Dominguez D, Batlle E, et~al.
+\newblock Snail induction of epithelial to mesenchymal transition in tumor
+  cells is accompanied by MUC1 repression and ZEB1 expression.
+\newblock J Biol Chem. 2002;277(42):39209--16.
+\newblock doi:{10.1074/jbc.M206400200}.
+
+\bibitem{Jackstadt:2013aa}
+Jackstadt R, R{\"o}h S, Neumann J, Jung P, Hoffmann R, Horst D, et~al.
+\newblock AP4 is a mediator of epithelial-mesenchymal transition and metastasis
+  in colorectal cancer.
+\newblock J Exp Med. 2013;210(7):1331--50.
+\newblock doi:{10.1084/jem.20120812}.
+
+\bibitem{Eastman:1999aa}
+Eastman Q, Grosschedl R.
+\newblock Regulation of LEF-1/TCF transcription factors by Wnt and other
+  signals.
+\newblock Curr Opin Cell Biol. 1999;11(2):233--40.
+
+\bibitem{Aramburu:1999if}
+Aramburu J, Yaffe MB, L{\'o}pez-Rodr{\'\i}guez C, Cantley LC, Hogan PG, Rao A.
+\newblock Affinity-driven peptide selection of an NFAT inhibitor more selective
+  than cyclosporin A.
+\newblock Science. 1999;285(5436):2129--33.
+
+\bibitem{Dhimolea2010}
+Dhimolea E, Maffini MV, Soto AM, Sonnenschein C.
+\newblock The role of collagen reorganization on mammary epithelial
+  morphogenesis in a 3D culture model.
+\newblock Biomaterials. 2010;31(13):3622--3630.
+\newblock doi:{10.1016/j.biomaterials.2010.01.077}.
+
+\bibitem{Butcher:2004aa}
+Butcher JT, Penrod AM, Garc{\'\i}a AJ, Nerem RM.
+\newblock Unique morphology and focal adhesion development of valvular
+  endothelial cells in static and fluid flow environments.
+\newblock Arterioscler Thromb Vasc Biol. 2004;24(8):1429--34.
+\newblock doi:{10.1161/01.ATV.0000130462.50769.5a}.
+
+\bibitem{Hong:2013mi}
+Hong JP, Li XM, Li MX, Zheng FL.
+\newblock VEGF suppresses epithelial-mesenchymal transition by inhibiting the
+  expression of Smad3 and miR-192 a Smad3-dependent microRNA.
+\newblock Int J Mol Med. 2013;31(6):1436--42.
+\newblock doi:{10.3892/ijmm.2013.1337}.
+
+\bibitem{Nagy:2007lh}
+Nagy JA, Dvorak AM, Dvorak HF.
+\newblock VEGF-A and the induction of pathological angiogenesis.
+\newblock Annu Rev Pathol. 2007;2:251--75.
+\newblock doi:{10.1146/annurev.pathol.2.010506.134925}.
+
+\bibitem{Ferrara:2002kx}
+Ferrara N.
+\newblock VEGF and the quest for tumour angiogenesis factors.
+\newblock Nat Rev Cancer. 2002;2(10):795--803.
+\newblock doi:{10.1038/nrc909}.
+
+\bibitem{Willis:2007uq}
+Willis BC, Borok Z.
+\newblock TGF-beta-induced EMT: mechanisms and implications for fibrotic lung
+  disease.
+\newblock Am J Physiol Lung Cell Mol Physiol. 2007;293(3):L525--34.
+\newblock doi:{10.1152/ajplung.00163.2007}.
+
+\bibitem{Polyak2009}
+Polyak K, Weinberg RA.
+\newblock Transitions between epithelial and mesenchymal states: acquisition of
+  malignant and stem cell traits.
+\newblock Nat Rev Cancer. 2009;9(4):265--273.
+\newblock doi:{10.1038/nrc2620}.
+
+\bibitem{Strauss:2011fk}
+Strauss R, Li ZY, Liu Y, Beyer I, Persson J, Sova P, et~al.
+\newblock Analysis of epithelial and mesenchymal markers in ovarian cancer
+  reveals phenotypic heterogeneity and plasticity.
+\newblock PLoS One. 2011;6(1):e16186.
+\newblock doi:{10.1371/journal.pone.0016186}.
+
+\bibitem{Neve:2006uq}
+Neve RM, Chin K, Fridlyand J, Yeh J, Baehner FL, Fevr T, et~al.
+\newblock A collection of breast cancer cell lines for the study of
+  functionally distinct cancer subtypes.
+\newblock Cancer Cell. 2006;10(6):515--27.
+\newblock doi:{10.1016/j.ccr.2006.10.008}.
+
+\bibitem{Welch-Reardon:2014aa}
+Welch-Reardon KM, Wu N, Hughes CCW.
+\newblock A Role for Partial Endothelial-Mesenchymal Transitions in
+  Angiogenesis?
+\newblock Arterioscler Thromb Vasc Biol. 2014;doi:{10.1161/ATVBAHA.114.303220}.
+
+\bibitem{Zajchowski:2001kx}
+Zajchowski DA, Bartholdi MF, Gong Y, Webster L, Liu HL, Munishkin A, et~al.
+\newblock Identification of gene expression profiles that predict the
+  aggressive behavior of breast cancer cells.
+\newblock Cancer Res. 2001;61(13):5168--78.
+
+\bibitem{Mancini2009}
+Mancini M, Toker A.
+\newblock NFAT proteins: emerging roles in cancer progression.
+\newblock Nat Rev Cancer. 2009;9(11):810--820.
+\newblock doi:{10.1038/nrc2735}.
+
+\bibitem{Singh:2010jx}
+Singh G, Singh SK, K{\"o}nig A, Reutlinger K, Nye MD, Adhikary T, et~al.
+\newblock Sequential activation of NFAT and c-Myc transcription factors
+  mediates the TGF-beta switch from a suppressor to a promoter of cancer cell
+  proliferation.
+\newblock J Biol Chem. 2010;285(35):27241--50.
+\newblock doi:{10.1074/jbc.M110.100438}.
+
+\bibitem{Debnath2003}
+Debnath J, Muthuswamy SK, Brugge JS.
+\newblock Morphogenesis and oncogenesis of MCF-10A mammary epithelial acini
+  grown in three-dimensional basement membrane cultures.
+\newblock Methods. 2003;30(3):256--268.
+
+\bibitem{Pearson:2007qq}
+Pearson GW, Hunter T.
+\newblock Real-time imaging reveals that noninvasive mammary epithelial acini
+  can contain motile cells.
+\newblock J Cell Biol. 2007;179(7):1555--67.
+\newblock doi:{10.1083/jcb.200706099}.
+
+\bibitem{OBrien2004}
+O'Brien LE, Tang K, Kats ES, Schutz-Geschwender A, Lipschutz JH, Mostov KE.
+\newblock ERK and MMPs sequentially regulate distinct stages of epithelial
+  tubule development.
+\newblock Dev Cell. 2004;7(1):21--32.
+\newblock doi:{10.1016/j.devcel.2004.06.001}.
+
+\bibitem{Debnath:2005qc}
+Debnath J, Brugge JS.
+\newblock Modelling glandular epithelial cancers in three-dimensional cultures.
+\newblock Nat Rev Cancer. 2005;5(9):675--88.
+\newblock doi:{10.1038/nrc1695}.
+
+\bibitem{Moles:2003aa}
+Moles CG, Mendes P, Banga JR.
+\newblock Parameter estimation in biochemical pathways: a comparison of global
+  optimization methods.
+\newblock Genome Res. 2003;13(11):2467--74.
+\newblock doi:{10.1101/gr.1262503}.
+
+\bibitem{Rodriguez-Fernandez:2013aa}
+Rodriguez-Fernandez M, Rehberg M, Kremling A, Banga JR.
+\newblock Simultaneous model discrimination and parameter estimation in dynamic
+  models of cellular systems.
+\newblock BMC Syst Biol. 2013;7:76.
+\newblock doi:{10.1186/1752-0509-7-76}.
+
+\bibitem{Villaverde:2014aa}
+Villaverde AF, Banga JR.
+\newblock Reverse engineering and identification in systems biology:
+  strategies, perspectives and challenges.
+\newblock J R Soc Interface. 2014;11(91):20130505.
+\newblock doi:{10.1098/rsif.2013.0505}.
+
+\bibitem{Choi:2012bs}
+Choi M, Shi J, Jung SH, Chen X, Cho KH.
+\newblock Attractor landscape analysis reveals feedback loops in the p53
+  network that control the cellular response to DNA damage.
+\newblock Sci Signal. 2012;5(251):ra83.
+\newblock doi:{10.1126/scisignal.2003363}.
+
+\bibitem{Terfve:2012kq}
+Terfve C, Cokelaer T, Henriques D, MacNamara A, Goncalves E, Morris MK, et~al.
+\newblock CellNOptR: a flexible toolkit to train protein signaling networks to
+  data using multiple logic formalisms.
+\newblock BMC Syst Biol. 2012;6:133.
+\newblock doi:{10.1186/1752-0509-6-133}.
+
+\bibitem{Hyduke:2010rt}
+Hyduke DR, Palsson B{\O}.
+\newblock Towards genome-scale signalling network reconstructions.
+\newblock Nat Rev Genet. 2010;11(4):297--307.
+\newblock doi:{10.1038/nrg2750}.
+
+\bibitem{Wayman2013}
+Wayman J, Varner J.
+\newblock Biological systems modeling of metabolic and signaling networks.
+\newblock Curr Opin Chem Eng. 2013;2:365 -- 372.
+
+\bibitem{Sainani2012}
+Sainani KL.
+\newblock Meet the Skeptics: Why some doubt biomedical models - and what it
+  takes to win them over.
+\newblock Biomedical Computation Review. 2012; p. 12 -- 18.
+
+\bibitem{Lee:2009fk}
+Lee MW, Vassiliadis VS, Park JM.
+\newblock Individual-based and stochastic modeling of cell population dynamics
+  considering substrate dependency.
+\newblock Biotechnol Bioeng. 2009;103(5):891--9.
+\newblock doi:{10.1002/bit.22327}.
+
+\bibitem{Swain:2002ve}
+Swain PS, Elowitz MB, Siggia ED.
+\newblock Intrinsic and extrinsic contributions to stochasticity in gene
+  expression.
+\newblock Proc Natl Acad Sci U S A. 2002;99(20):12795--800.
+\newblock doi:{10.1073/pnas.162041399}.
+
+\bibitem{Handl:2007fk}
+Handl J, Kell DB, Knowles J.
+\newblock Multiobjective optimization in bioinformatics and computational
+  biology.
+\newblock IEEE/ACM Trans Comput Biol Bioinform. 2007;4(2):279--92.
+\newblock doi:{10.1109/TCBB.2007.070203}.
+
+\bibitem{Vega:2004qf}
+Vega S, Morales AV, Oca{\~n}a OH, Vald{\'e}s F, Fabregat I, Nieto MA.
+\newblock Snail blocks the cell cycle and confers resistance to cell death.
+\newblock Genes Dev. 2004;18(10):1131--43.
+\newblock doi:{10.1101/gad.294104}.
+
+\bibitem{Nawshad:2003ff}
+Nawshad A, Hay ED.
+\newblock TGFbeta3 signaling activates transcription of the LEF1 gene to induce
+  epithelial mesenchymal transformation during mouse palate development.
+\newblock J Cell Biol. 2003;163(6):1291--301.
+\newblock doi:{10.1083/jcb.200306024}.
+
+\bibitem{Cano:2000kh}
+Cano A, P{\'e}rez-Moreno MA, Rodrigo I, Locascio A, Blanco MJ, del Barrio MG,
+  et~al.
+\newblock The transcription factor snail controls epithelial-mesenchymal
+  transitions by repressing E-cadherin expression.
+\newblock Nat Cell Biol. 2000;2(2):76--83.
+\newblock doi:{10.1038/35000025}.
+
+\bibitem{Arce:2009pr}
+Arce L, Pate KT, Waterman ML.
+\newblock Groucho binds two conserved regions of LEF-1 for HDAC-dependent
+  repression.
+\newblock BMC Cancer. 2009;9:159.
+\newblock doi:{10.1186/1471-2407-9-159}.
+
+\bibitem{Grumolato2013}
+Grumolato L, Liu G, Haremaki T, Mungamuri SK, Mong P, Akiri G, et~al.
+\newblock ?-Catenin-Independent Activation of TCF1/LEF1 in Human Hematopoietic
+  Tumor Cells through Interaction with ATF2 Transcription Factors.
+\newblock PLoS Genet. 2013;9(8):e1003603.
+\newblock doi:{10.1371/journal.pgen.1003603}.
+
+\bibitem{Wu:2009cr}
+Wu Y, Deng J, Rychahou PG, Qiu S, Evers BM, Zhou BP.
+\newblock Stabilization of snail by NF-kappaB is required for
+  inflammation-induced cell migration and invasion.
+\newblock Cancer Cell. 2009;15(5):416--28.
+\newblock doi:{10.1016/j.ccr.2009.03.016}.
+
+\bibitem{Sullivan:2009bh}
+Sullivan NJ, Sasser AK, Axel AE, Vesuna F, Raman V, Ramirez N, et~al.
+\newblock Interleukin-6 induces an epithelial-mesenchymal transition phenotype
+  in human breast cancer cells.
+\newblock Oncogene. 2009;28(33):2940--7.
+\newblock doi:{10.1038/onc.2009.180}.
+
+\bibitem{Niessen:2008nx}
+Niessen K, Fu Y, Chang L, Hoodless PA, McFadden D, Karsan A.
+\newblock Slug is a direct Notch target required for initiation of cardiac
+  cushion cellularization.
+\newblock J Cell Biol. 2008;182(2):315--25.
+\newblock doi:{10.1083/jcb.200710067}.
+
+\bibitem{Liu:2008aa}
+Liu Y, El-Naggar S, Darling DS, Higashi Y, Dean DC.
+\newblock Zeb1 links epithelial-mesenchymal transition and cellular senescence.
+\newblock Development. 2008;135(3):579--88.
+\newblock doi:{10.1242/dev.007047}.
+
+\bibitem{Hong:2015aa}
+Hong T, Watanabe K, Ta CH, Villarreal-Ponce A, Nie Q, Dai X.
+\newblock An Ovol2-Zeb1 Mutual Inhibitory Circuit Governs Bidirectional and
+  Multi-step Transition between Epithelial and Mesenchymal States.
+\newblock PLoS Comput Biol. 2015;11(11):e1004569.
+\newblock doi:{10.1371/journal.pcbi.1004569}.
+
+\bibitem{Bullock:2012aa}
+Bullock MD, Sayan AE, Packham GK, Mirnezami AH.
+\newblock MicroRNAs: critical regulators of epithelial to mesenchymal (EMT) and
+  mesenchymal to epithelial transition (MET) in cancer progression.
+\newblock Biol Cell. 2012;104(1):3--12.
+\newblock doi:{10.1111/boc.201100115}.
+
+\bibitem{Lu:2013aa}
+Lu M, Jolly MK, Levine H, Onuchic JN, Ben-Jacob E.
+\newblock MicroRNA-based regulation of epithelial-hybrid-mesenchymal fate
+  determination.
+\newblock Proc Natl Acad Sci U S A. 2013;110(45):18144--9.
+\newblock doi:{10.1073/pnas.1318192110}.
+
+\bibitem{Zhang:2014aa}
+Zhang J, Tian XJ, Zhang H, Teng Y, Li R, Bai F, et~al.
+\newblock TGF-beta-induced epithelial-to-mesenchymal transition proceeds
+  through stepwise activation of multiple feedback loops.
+\newblock Sci Signal. 2014;7(345):ra91.
+\newblock doi:{10.1126/scisignal.2005304}.
+
+\bibitem{Tian:2013aa}
+Tian XJ, Zhang H, Xing J.
+\newblock {Coupled reversible and irreversible bistable switches underlying
+  TGFb-induced epithelial to mesenchymal transition}.
+\newblock Biophys J. 2013;105(4):1079--89.
+\newblock doi:{10.1016/j.bpj.2013.07.011}.
+
+\bibitem{Linding:2007fk}
+Linding R, Jensen LJ, Ostheimer GJ, van Vugt MATM, J{\o}rgensen C, Miron IM,
+  et~al.
+\newblock Systematic discovery of in vivo phosphorylation networks.
+\newblock Cell. 2007;129(7):1415--26.
+\newblock doi:{10.1016/j.cell.2007.05.052}.
+
+\bibitem{Jensen:2009fk}
+Jensen LJ, Kuhn M, Stark M, Chaffron S, Creevey C, Muller J, et~al.
+\newblock STRING 8--a global view on proteins and their functional interactions
+  in 630 organisms.
+\newblock Nucleic Acids Res. 2009;37(Database issue):D412--6.
+\newblock doi:{10.1093/nar/gkn760}.
+
+\bibitem{Zhou:2004kx}
+Zhou BP, Deng J, Xia W, Xu J, Li YM, Gunduz M, et~al.
+\newblock Dual regulation of Snail by GSK-3beta-mediated phosphorylation in
+  control of epithelial-mesenchymal transition.
+\newblock Nat Cell Biol. 2004;6(10):931--40.
+\newblock doi:{10.1038/ncb1173}.
+
+\bibitem{Larue:2005vn}
+Larue L, Bellacosa A.
+\newblock Epithelial-mesenchymal transition in development and cancer: role of
+  phosphatidylinositol 3' kinase/AKT pathways.
+\newblock Oncogene. 2005;24(50):7443--54.
+\newblock doi:{10.1038/sj.onc.1209091}.
+
+\bibitem{Derynck:2003fc}
+Derynck R, Zhang YE.
+\newblock Smad-dependent and Smad-independent pathways in TGF-beta family
+  signalling.
+\newblock Nature. 2003;425(6958):577--84.
+\newblock doi:{10.1038/nature02006}.
+
+\bibitem{Massague:2005qc}
+Massagu{\'e} J, Seoane J, Wotton D.
+\newblock Smad transcription factors.
+\newblock Genes Dev. 2005;19(23):2783--810.
+\newblock doi:{10.1101/gad.1350705}.
+
+\bibitem{Nawshad:2007kl}
+Nawshad A, Medici D, Liu CC, Hay ED.
+\newblock TGFbeta3 inhibits E-cadherin gene expression in palate medial-edge
+  epithelial cells through a Smad2-Smad4-LEF1 transcription complex.
+\newblock J Cell Sci. 2007;120(Pt 9):1646--53.
+\newblock doi:{10.1242/jcs.003129}.
+
+\bibitem{Medici2011}
+Medici D, Potenta S, Kalluri R.
+\newblock Transforming growth factor-?2 promotes Snail-mediated
+  endothelial-mesenchymal transition through convergence of Smad-dependent and
+  Smad-independent signalling.
+\newblock Biochem J. 2011;437(3):515--520.
+\newblock doi:{10.1042/BJ20101500}.
+
+\bibitem{Suehiro:2014aa}
+Suehiro Ji, Kanki Y, Makihara C, Schadler K, Miura M, Manabe Y, et~al.
+\newblock Genome-wide approaches reveal functional vascular endothelial growth
+  factor (VEGF)-inducible nuclear factor of activated T cells (NFAT) c1 binding
+  to angiogenesis-related genes in the endothelium.
+\newblock J Biol Chem. 2014;289(42):29044--59.
+\newblock doi:{10.1074/jbc.M114.555235}.
+
+\bibitem{Hu:2015aa}
+Hu CY, Varner JD, Lucks JB.
+\newblock Generating Effective Models and Parameters for RNA Genetic Circuits.
+\newblock ACS Synth Biol. 2015;4(8):914--26.
+\newblock doi:{10.1021/acssynbio.5b00077}.
+
+\bibitem{pr3010138}
+Wayman JA, Sagar A, Varner JD.
+\newblock Dynamic Modeling of Cell-Free Biochemical Networks Using Effective
+  Kinetic Models.
+\newblock Processes. 2015;3(1):138.
+\newblock doi:{10.3390/pr3010138}.
+
+\bibitem{pr3010178}
+Sagar A, Varner JD.
+\newblock Dynamic Modeling of the Human Coagulation Cascade Using Reduced Order
+  Effective Kinetic Models.
+\newblock Processes. 2015;3(1):178.
+\newblock doi:{10.3390/pr3010178}.
+
+\bibitem{Julia}
+Bezanson J, Edelman A, Karpinski S, Shah VB.
+\newblock Julia: {A} Fresh Approach to Numerical Computing.
+\newblock CoRR. 2014;abs/1411.1607.
+
+\bibitem{Hindmarsh2005}
+Hindmarsh A, Brown P, Grant K, Lee S, Serban R, Shumaker D, et~al.
+\newblock SUNDIALS: Suite of nonlinear and differential/algebraic equation
+  solvers.
+\newblock ACM Transactions on Mathematical Software. 2005;31:363--396.
+
+\bibitem{JuPOETs-BioArXiv}
+Bassen D, Vilkhovoy M, Minot M, Butcher JT, Varner JD.
+\newblock {JuPOETs: A Constrained Multiobjective Optimization Approach to
+  Estimate Biochemical Model Ensembles in the Julia Programming Language}.
+\newblock bioRxiv. 2016;10.1101/056044.
+\newblock doi:{10.1101/056044}.
+
+\bibitem{kohavi1995study}
+Kohavi R.
+\newblock {A study of cross-validation and bootstrap for accuracy estimation
+  and model selection}.
+\newblock In: International joint Conference on artificial intelligence.
+  vol.~14. Citeseer; 1995. p. 1137--1145.
+
+\bibitem{ChangPP07}
+Chang H, Park C, Parvin B.
+\newblock Quantitative Representation of Three-dimensional Cell Culture Models.
+\newblock In: Proceedings of the 2007 {IEEE} International Symposium on
+  Biomedical Imaging: From Nano to Macro, Washington, DC, USA, April 12-16,
+  2007; 2007. p. 89--92.
+\newblock Available from: \url{http://dx.doi.org/10.1109/ISBI.2007.356795}.
+
+\bibitem{Polizzotti}
+Polizzotti L, Basak O, Bjornsson C, Shubert K, Yener B, Plopper G.
+\newblock Novel Image Analysis Approach Quantifies Morphological
+  Characteristics of 3D Breast Culture Acini with Varying Metastatic
+  Potentials.
+\newblock J Biomed Biotech. 2012;2012(doi:10.1155/2012/102036):1--16.
+
+\end{thebibliography}
 
@@
 \begin{figure}
-\includegraphics [width=1.0\linewidth] {./figs/Fig1_coarse_network.pdf}
+%\includegraphics [width=1.0\linewidth] {./figs/Fig1_coarse_network.pdf}
 \caption{Model connectivity recreates the core architecture during EMT. The EMT network contains 97 nodes (proteins, mRNA, and genes) interconnected by 169 interactions.
@@
 \begin{figure}
-\includegraphics [width=1.0\linewidth] {./figs/Fig2_training.pdf}
+%\includegraphics [width=1.0\linewidth] {./figs/Fig2_training.pdf}
 \caption{Training and validation simulations. The population of EMT models qualitatively captured TGF$-\beta$-induced EMT signaling.
@@
 \begin{figure}
-\includegraphics [width=1.0\linewidth] {./figs/Fig3_phenotype.pdf}
+%\includegraphics [width=1.0\linewidth] {./figs/Fig3_phenotype.pdf}
 \caption{Simulated VEGF-A and TGF$-\beta$1/2 exposure promoted phenotype heterogeneity.
@@
 \begin{figure}
-\includegraphics [width=1.0\linewidth] {./figs/Fig4_mechanism.pdf}
+%\includegraphics [width=1.0\linewidth] {./figs/Fig4_mechanism.pdf}
 \caption{Analysis of underlying signaling responses.
 (A) We examined the distribution of NFATc1 and AP1/SP1 in cells containing the hybrid phenotype (VEGF-A + TGF$-\beta$2 case), showing the potential for cells to express both SP1 and NFATc1 in a non exclusive manner.
-(B) We were able to show a fit to an additional objective demonstrating the actiation of GSK3 through PI3K.
-Our model captures this activation through TGF$-\beta$3 signaling.
-LY294002 is a PI3K inhibitor.
-(C) We identified a novel regulator of LEF1 called YREG1 that allows Snail/Slug to emulate an inducer by repressing YREG1, which was required to stabalize the untreated population.
+(B) We fit an additional objective demonstrating the activation of GSK3 through PI3K. The model captured this activation through TGF$-\beta$3 signaling. LY294002 is a PI3K inhibitor.
+(C) We identified a novel regulator of LEF1 called YREG1 that allows Snail/Slug to emulate an inducer by repressing YREG1, which was required to stabilize the untreated population.
 YREG1 overexpression revealed an enhanced epithelial phenotype, while some inherently transformed cells moved towards a hybrid phenotype.
@@
 \begin{figure}
-\includegraphics [width=1.0\linewidth] {./figs/Fig-5-MCF10A-Validation.pdf}
+%\includegraphics [width=1.0\linewidth] {./figs/Fig-5-MCF10A-Validation.pdf}
 \caption{Simultaneous TGF-$\beta$1/2 and VEGF-A treatment induced phenotype heterogeneity and is dependent upon NFAT activity \emph{in-vitro}.
@@
 \begin{figure}
-\includegraphics[width=0.5\linewidth] {./figs/Fig-6-Phenotyping.pdf}
+%\includegraphics[width=0.5\linewidth] {./figs/Fig-6-Phenotyping.pdf}
 \caption{Ductal branching is dependent upon phenotype heterogeneity within MCF10A in 3-D culture.
@@
 \clearpage
-\input{Supplemental-JB}
+
+\begin{figure}
+	%\includegraphics [width=1.0\linewidth] {./figs/Fig-Supplemental-Error-Table.pdf}
+	\caption{Training and prediction values as a function of condition for the 11 TGF-$\beta$ objective functions versus a random parameter control.}\label{fg:ObjTable}
+\end{figure}
+
+\begin{figure}
+%\includegraphics [width=1.0\linewidth] {./figs/Fig-3-Supplemental-DLD1.pdf}
+\caption{VEGF-A attenuates TGF-$\beta$1/2 to induce phenotype heterogeneity in DLD1.
+(A) In DLD1, we found that 5ng/ml of VEGFA increased NFATc1 and E-cadherin gene expression via qPCR and 50ng/ml potentiated this effect at 48 hrs.
+(B - C) These findings were confirmed at the protein level via immunofluorescence, as ecadherin levels and nuclear localization of NFATc1 increased.
+(D) Treatment with (10ng/ml) TGF$\beta$2 resulted in mesenchymal transformation as measured via qPCR against target genes Slug, ecadherin, vimentin, Sp1, and NFATc1.
+(E - F) Immunofluorescence and nuclear localization revealed a strong presence of phospho-Sp1.
+(G) Combination of VEGFA (50ng/ml) and TGF$\beta$2 (10ng/ml) treatment resulted in increased Slug, NFATc1, and vimentin expression, while also increasing ecadherin levels compared to control.
+(H) Immunofluorescence confirmed these results, as both ecadherin and vimentin levels were elevated.
+(I) A significant increase in nuclear localization of both NFATc1 and phospho-Sp1 were also found.
+Magnification, 40x. Scale bars: 50$\mu$m.  C=Control, T=TGF$\beta$2 , V=VEGFA, VI=NFAT inhibitor (VIVIT).
+Asterisks signify statistical differences from each other according to a one-way ANOVA with Tukey's post hoc (p$\prec$0.05).}\label{fg:S3}
+\end{figure}
+
+\begin{figure}
+%\includegraphics [width=1.0\linewidth] {./figs/Fig-4-Supplemental-DLD1-FlowCyto.pdf}
+\caption{E-cadherin expression is dependent upon NFAT activity in DLD1.
+(A) Treatment with VEGFA (50ng/ml) and NFAT inhibitory peptide VIVIT (10$\mu$M) resulted in significantly reduced ecadherin expression (qRT-PCR at 48hrs).
+Addition of TGF$\beta$2 (10ng/ml) and VIVIT resulted in increased Slug and vimentin expression, while inhibiting ecadherin levels.
+Combined TGF$\beta$2, VEGFA, and VIVIT treatment resulted in target genes Slug and vimentin expression increased, while inhibiting ecadherin levels.
+No change in Sp1 or NFATc1 expression was found.
+(B) These findings were confirmed via immunofluorescence as the VIVIT inhibitors was shown to inhibit ecadherin levels in all three cases.
+We also found no change in gene or nuclear localization of NFATc1 in all three cases, while phospho-Sp1 was found to increase in both TGF$\beta$ conditions.
+(C) Quantitative flow cytometry also confirmed this trend.
+(D,E)  TGF$\beta$2, VEGFA and VIVIT treatment in DLD1 and MCF10A resulted in no change of Sp1 expression or NFATc1 expression.
+(F)  Likewise, no change in nuclear localization of NFAT in all three cases, however phospho-Sp1 was found to increase in both TGF$\beta$ conditions.
+Magnification, 40x. Scale bars: 50$\mu$m.  C=Control, T=TGF$\beta$2 , V=VEGFA, VI= NFAT inhibitor (VIVIT).
+Asterisks signify statistical differences from each other according to a one-way ANOVA with Tukey's post hoc (p$\prec$0.05).}\label{fg:S4}
+\end{figure}
+
+\begin{figure}
+	%\includegraphics [width=0.75\linewidth] {./figs/Fig-Supplemental-Data-VEGFA.pdf}
+	\caption{VEGF-A qPCR data used to hand fit VEGF enhancement of E-cadherin expression. mRNA was harvested after 3hr and 48hr timepoint.}\label{fg:VEGFA-Data}
+\end{figure}
+
+\newpage
+%\include{S1_Text}
 

### 2016-11-28 S1_Text.tex
JV changes

@@
+% Reset the layout for the supplemental text -
+
 \nolinenumbers
+
 \section*{Supplemental Materials and Methods}
+
+\raggedright
+\setlength{\parindent}{0.5cm}
+\textwidth 6.25in
+\textheight 8.75in
 
@@
 In this study, the experimental data used for model training was typically the band intensity from Western or Northern blots.
-Band intensity was estimated using the ImageJ software package \cite{IMAGEJ}.
+Band intensity was estimated using the ImageJ software package.
 The scaled measurement for species $x$ at time $i=\{t_{1},t_{2},..,t_{n}\}$ in condition $j$ is given by:
@@
 
-\begin{figure}
-	\center
-	\includegraphics [width=1.0\linewidth] {./figs/Fig-Supplemental-Error-Table.pdf}
-	\caption{Training and prediction values as a function of condition for the 11 TGF-$\beta$ objective functions versus a random parameter control.}\label{fg:ObjTable}
-\end{figure}
+% \begin{figure}
+% 	\center
+% 	\includegraphics [width=1.0\linewidth] {./figs/Fig-Supplemental-Error-Table.pdf}
+% 	\caption{Training and prediction values as a function of condition for the 11 TGF-$\beta$ objective functions versus a random parameter control.}\label{fg:ObjTable}
+% \end{figure}
 
-\begin{figure}
-\includegraphics [width=1.0\linewidth] {./figs/Fig-3-Supplemental-DLD1.pdf}
-\caption{VEGF-A attenuates TGF-$\beta$1/2 to induce phenotype heterogeneity in DLD1.
-(A) In DLD1, we found that 5ng/ml of VEGFA increased NFATc1 and E-cadherin gene expression via qPCR and 50ng/ml potentiated this effect at 48 hrs.
-(B - C) These findings were confirmed at the protein level via immunofluorescence, as ecadherin levels and nuclear localization of NFATc1 increased.
-(D) Treatment with (10ng/ml) TGF$\beta$2 resulted in mesenchymal transformation as measured via qPCR against target genes Slug, ecadherin, vimentin, Sp1, and NFATc1.
-(E - F) Immunofluorescence and nuclear localization revealed a strong presence of phospho-Sp1.
-(G) Combination of VEGFA (50ng/ml) and TGF$\beta$2 (10ng/ml) treatment resulted in increased Slug, NFATc1, and vimentin expression, while also increasing ecadherin levels compared to control.
-(H) Immunofluorescence confirmed these results, as both ecadherin and vimentin levels were elevated.
-(I) A significant increase in nuclear localization of both NFATc1 and phospho-Sp1 were also found.
-Magnification, 40x. Scale bars: 50$\mu$m.  C=Control, T=TGF$\beta$2 , V=VEGFA, VI=NFAT inhibitor (VIVIT).
-Asterisks signify statistical differences from each other according to a one-way ANOVA with Tukey's post hoc (p$\prec$0.05).}\label{fg:S3}
-\end{figure}
+% \begin{figure}
+% \includegraphics [width=1.0\linewidth] {./figs/Fig-3-Supplemental-DLD1.pdf}
+% \caption{VEGF-A attenuates TGF-$\beta$1/2 to induce phenotype heterogeneity in DLD1.
+% (A) In DLD1, we found that 5ng/ml of VEGFA increased NFATc1 and E-cadherin gene expression via qPCR and 50ng/ml potentiated this effect at 48 hrs.
+% (B - C) These findings were confirmed at the protein level via immunofluorescence, as ecadherin levels and nuclear localization of NFATc1 increased.
+% (D) Treatment with (10ng/ml) TGF$\beta$2 resulted in mesenchymal transformation as measured via qPCR against target genes Slug, ecadherin, vimentin, Sp1, and NFATc1.
+% (E - F) Immunofluorescence and nuclear localization revealed a strong presence of phospho-Sp1.
+% (G) Combination of VEGFA (50ng/ml) and TGF$\beta$2 (10ng/ml) treatment resulted in increased Slug, NFATc1, and vimentin expression, while also increasing ecadherin levels compared to control.
+% (H) Immunofluorescence confirmed these results, as both ecadherin and vimentin levels were elevated.
+% (I) A significant increase in nuclear localization of both NFATc1 and phospho-Sp1 were also found.
+% Magnification, 40x. Scale bars: 50$\mu$m.  C=Control, T=TGF$\beta$2 , V=VEGFA, VI=NFAT inhibitor (VIVIT).
+% Asterisks signify statistical differences from each other according to a one-way ANOVA with Tukey's post hoc (p$\prec$0.05).}\label{fg:S3}
+% \end{figure}
 
-\begin{figure}
-\includegraphics [width=1.0\linewidth] {./figs/Fig-4-Supplemental-DLD1-FlowCyto.pdf}
-\caption{E-cadherin expression is dependent upon NFAT activity in DLD1.
-(A) Treatment with VEGFA (50ng/ml) and NFAT inhibitory peptide VIVIT (10$\mu$M) resulted in significantly reduced ecadherin expression (qRT-PCR at 48hrs).
-Addition of TGF$\beta$2 (10ng/ml) and VIVIT resulted in increased Slug and vimentin expression, while inhibiting ecadherin levels.
-Combined TGF$\beta$2, VEGFA, and VIVIT treatment resulted in target genes Slug and vimentin expression increased, while inhibiting ecadherin levels.
-No change in Sp1 or NFATc1 expression was found.
-(B) These findings were confirmed via immunofluorescence as the VIVIT inhibitors was shown to inhibit ecadherin levels in all three cases.
-We also found no change in gene or nuclear localization of NFATc1 in all three cases, while phospho-Sp1 was found to increase in both TGF$\beta$ conditions.
-(C) Quantitative flow cytometry also confirmed this trend.
-(D,E)  TGF$\beta$2, VEGFA and VIVIT treatment in DLD1 and MCF10A resulted in no change of Sp1 expression or NFATc1 expression.
-(F)  Likewise, no change in nuclear localization of NFAT in all three cases, however phospho-Sp1 was found to increase in both TGF$\beta$ conditions.
-Magnification, 40x. Scale bars: 50$\mu$m.  C=Control, T=TGF$\beta$2 , V=VEGFA, VI= NFAT inhibitor (VIVIT).
-Asterisks signify statistical differences from each other according to a one-way ANOVA with Tukey's post hoc (p$\prec$0.05).}\label{fg:S4}
-\end{figure}
+% \begin{figure}
+% \includegraphics [width=1.0\linewidth] {./figs/Fig-4-Supplemental-DLD1-FlowCyto.pdf}
+% \caption{E-cadherin expression is dependent upon NFAT activity in DLD1.
+% (A) Treatment with VEGFA (50ng/ml) and NFAT inhibitory peptide VIVIT (10$\mu$M) resulted in significantly reduced ecadherin expression (qRT-PCR at 48hrs).
+% Addition of TGF$\beta$2 (10ng/ml) and VIVIT resulted in increased Slug and vimentin expression, while inhibiting ecadherin levels.
+% Combined TGF$\beta$2, VEGFA, and VIVIT treatment resulted in target genes Slug and vimentin expression increased, while inhibiting ecadherin levels.
+% No change in Sp1 or NFATc1 expression was found.
+% (B) These findings were confirmed via immunofluorescence as the VIVIT inhibitors was shown to inhibit ecadherin levels in all three cases.
+% We also found no change in gene or nuclear localization of NFATc1 in all three cases, while phospho-Sp1 was found to increase in both TGF$\beta$ conditions.
+% (C) Quantitative flow cytometry also confirmed this trend.
+% (D,E)  TGF$\beta$2, VEGFA and VIVIT treatment in DLD1 and MCF10A resulted in no change of Sp1 expression or NFATc1 expression.
+% (F)  Likewise, no change in nuclear localization of NFAT in all three cases, however phospho-Sp1 was found to increase in both TGF$\beta$ conditions.
+% Magnification, 40x. Scale bars: 50$\mu$m.  C=Control, T=TGF$\beta$2 , V=VEGFA, VI= NFAT inhibitor (VIVIT).
+% Asterisks signify statistical differences from each other according to a one-way ANOVA with Tukey's post hoc (p$\prec$0.05).}\label{fg:S4}
+% \end{figure}
 
-\newpage
-
-\begin{figure}
-	\center
-	\includegraphics [width=0.75\linewidth] {./figs/Fig-Supplemental-Data-VEGFA.pdf}
-	\caption{VEGF-A qPCR data used to hand fit VEGF enhancement of E-cadherin expression. mRNA was harvested after 3hr and 48hr timepoint.}\label{fg:VEGFA-Data}
-\end{figure}
+% \begin{figure}
+% 	\center
+% 	\includegraphics [width=0.75\linewidth] {./figs/Fig-Supplemental-Data-VEGFA.pdf}
+% 	\caption{VEGF-A qPCR data used to hand fit VEGF enhancement of E-cadherin expression. mRNA was harvested after 3hr and 48hr timepoint.}\label{fg:VEGFA-Data}
+% \end{figure}
 
@@
 % in place ref %
-
-@article{IMAGEJ,
-	Author = {Abramoff, M.D. and Magelhaes, P.J. and Ram, S.J.},
-	Date-Added = {2011-11-07 13:49:32 -0500},
-	Date-Modified = {2011-11-07 13:49:32 -0500},
-	Journal = {Biophotonics International,},
-	Pages = {36-42},
-	Title = {Image Processing with ImageJ},
-	Volume = {11},
-	Year = {2004}}
+% @article{IMAGEJ,
+% 	Author = {Abramoff, M.D. and Magelhaes, P.J. and Ram, S.J.},
+% 	Date-Added = {2011-11-07 13:49:32 -0500},
+% 	Date-Modified = {2011-11-07 13:49:32 -0500},
+% 	Journal = {Biophotonics International,},
+% 	Pages = {36-42},
+% 	Title = {Image Processing with ImageJ},
+% 	Volume = {11},
+% 	Year = {2004}}

### 2016-12-06 Paper_v9.tex
fig order fix

@@
 TGF-$\beta$3 expression was also able to sustain $\beta$-catenin release by inhibiting its sequestration by the APC complex through PI3K mediated GSK3,
-which was captured by the model (Fig. \ref{fg:F4}B).
+which was captured by the model (Fig. \ref{fg:gsk3_alone}).
 Each phase involved the hierarchal expression and/or post-translational modification of several key transcription factors.
@@
 Low levels of YREG1 expression were present in all simulations to regulate the formation of the $\beta$-catenin-LEF-1 complex.
-To test the effect of YREG1 on the epithelial population, we conducted over-expression and knockdown simulations on untreated cells (Fig. \ref{fg:F4}C and \ref{fg:F4}D).
-In the absence of YREG1, the population of models failed to consistently to retain a stable epithelial state (Fig. \ref{fg:F4}D).
+To test the effect of YREG1 on the epithelial population, we conducted over-expression and knockdown simulations on untreated cells (Fig. \ref{fg:F4}B and \ref{fg:F4}C).
+In the absence of YREG1, the population of models failed to consistently to retain a stable epithelial state (Fig. \ref{fg:F4}B).
 Conversely, YREG1 amplification revealed an enhanced epithelial phenotype, while some inherently transformed cells moved towards a hybrid phenotype  (Fig. \ref{fg:F4}C).
@@
 (A) We examined the distribution of NFATc1 and AP1/SP1 in cells containing the hybrid phenotype (VEGF-A + TGF$-\beta$2 case), showing the potential for cells to express both SP1 and NFATc1 in a non exclusive manner.
-(B) We fit an additional objective demonstrating the activation of GSK3 through PI3K. The model captured this activation through TGF$-\beta$3 signaling. LY294002 is a PI3K inhibitor.
+(B) In the absence of YREG1, most of the population failed to consistently to retain a stable epithelial state.}\label{fg:F4}
 (C) We identified a novel regulator of LEF1 called YREG1 that allows Snail/Slug to emulate an inducer by repressing YREG1, which was required to stabilize the untreated population.
 YREG1 overexpression revealed an enhanced epithelial phenotype, while some inherently transformed cells moved towards a hybrid phenotype.
-(D) In the absence of YREG1, most of the population failed to consistently to retain a stable epithelial state.}\label{fg:F4}
 \end{figure}
@@
 \begin{figure}
+  %\includegraphics [width=0.75\linewidth] {./figs/Fig-Supplemental-Data-VEGFA.pdf}
+  \caption{VEGF-A qPCR data used to hand fit VEGF enhancement of E-cadherin expression. mRNA was harvested after 3hr and 48hr timepoint.}\label{fg:VEGFA-Data}
+\end{figure}
+
+\begin{figure}
+%\includegraphics [width=1.0\linewidth] {./figs/Fig_gsk3_sup.pdf}
+\caption{We fit an additional objective demonstrating the activation of GSK3 through PI3K. The model captured this activation through TGF$-\beta$3 signaling. LY294002 is a PI3K inhibitor.} 
+\label{fg:gsk3_alone}
+\end{figure}
+
+\begin{figure}
   %\includegraphics [width=1.0\linewidth] {./figs/Fig-Supplemental-Error-Table.pdf}
@@
 
-\begin{figure}
-	%\includegraphics [width=0.75\linewidth] {./figs/Fig-Supplemental-Data-VEGFA.pdf}
-	\caption{VEGF-A qPCR data used to hand fit VEGF enhancement of E-cadherin expression. mRNA was harvested after 3hr and 48hr timepoint.}\label{fg:VEGFA-Data}
-\end{figure}
-
 \newpage

### 2016-12-09 Paper_v9.tex
sup fig labels and missing betas in refs

@@
 Grumolato L, Liu G, Haremaki T, Mungamuri SK, Mong P, Akiri G, et~al.
-\newblock ?-Catenin-Independent Activation of TCF1/LEF1 in Human Hematopoietic
+\newblock $\beta$-Catenin-Independent Activation of TCF1/LEF1 in Human Hematopoietic
   Tumor Cells through Interaction with ATF2 Transcription Factors.
@@
 Medici D, Potenta S, Kalluri R.
-\newblock Transforming growth factor-?2 promotes Snail-mediated
+\newblock Transforming growth factor-$\beta$2 promotes Snail-mediated
   endothelial-mesenchymal transition through convergence of Smad-dependent and
@@
 Magnification, 40x. Scale bars: 50$\mu$m.  C=Control, T=TGF$\beta$2 , V=VEGFA, VI=NFAT inhibitor (VIVIT).
-Asterisks signify statistical differences from each other according to a one-way ANOVA with Tukey's post hoc (p$\prec$0.05).}\label{fg:S3}
+Asterisks signify statistical differences from each other according to a one-way ANOVA with Tukey's post hoc (p$\prec$0.05).}\label{fg:S4}
 \end{figure}
@@
 Magnification, 40x. Scale bars: 50$\mu$m.  C=Control, T=TGF$\beta$2 , V=VEGFA, VI= NFAT inhibitor (VIVIT).
-Asterisks signify statistical differences from each other according to a one-way ANOVA with Tukey's post hoc (p$\prec$0.05).}\label{fg:S4}
+Asterisks signify statistical differences from each other according to a one-way ANOVA with Tukey's post hoc (p$\prec$0.05).}\label{fg:S5}
 \end{figure}
